### 2015-09-14 report.tex
update images and add a .tex file for LANL report

\documentclass{report}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{parskip}
\usepackage[colorlinks,
            urlcolor=blue]{hyperref}
%\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{booktabs}
\usepackage{url}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pbox}
\usepackage[margin=1in]{geometry}

\author{Ashwin Srinath}
\title{An OpenCL-based tridiagonal solver for evaluation
    of compact finite differences on Graphics Processing Units}
\begin{document}
% \renewcommand{\bibname}{References}

\section{Introduction}

The motivation for this work comes from the need to evaluate
\emph{compact finite differences} arising in large simulations of
combustion.

\section{Compact finite differences}

The domain is shown in \ref{fig:domain}.

\begin{figure}[h]
\begin{center}
\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/domain.eps}
\end{center}
\caption{Domain}
\label{fig:domain}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/one-block.eps}
\end{center}
\caption{Single block}
\label{fig:one-block}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/one-thread.eps}
\end{center}
\caption{Single thread}
\label{fig:one-thread}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/domain.eps}
\end{center}
\caption{Domain}
\label{fig:domain}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/domain.eps}
\end{center}
\caption{Domain}
\label{fig:domain}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/domain.eps}
\end{center}
\caption{Domain}
\label{fig:domain}
\end{figure}

\section{Parallelism}

\section{}

\end{document}

### 2015-09-14 report.tex
update report

@@
-\documentclass{report}
+\documentclass{article}
 \usepackage{graphicx}
@@
 
-\section{Compact finite differences}
+We consider the following fourth-order
+compact finite difference scheme for evaluating the first derivative,
+with a 3-point stencil:
 
-The domain is shown in \ref{fig:domain}.
+\begin{equation}
+    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + f^{\prime}_i
+    =
+    a\frac{f_{i+1} - f_{i-1}}{dx}
+\end{equation}
 
+For the particular scheme used, $\alpha = 1/4$,
+and $ a = 3/4 $.
+It is easily noted that this leads to a tridiagonal system of equations,
+along with a 3-point stencil to evaluate the RHS.
+Further, the left boundary can be treated using the following implicit equation:
+
+\begin{equation}
+    f^{\prime}_1 + 2f^{\prime}_2 = \frac{-5f_1 + 4f_2 + f_3}{dx}
+\end{equation}
+
+The right boundary ($f^{\prime}_{n-1}$) is treated by utilizing
+the negative complex-conjugate of the Fourier image of the stencil
+at $f^{\prime}_1$:
+
+\begin{equation}
+    f^{\prime}_{n-1} + 2f^{\prime}_{n-2}
+    =
+    \frac{5f_{n-1} - 4f_{n-2} - f_{n-3}}{dx}
+\end{equation}
+
+The tridiagonal system that needs to be solved for the derivatives,
+then has an LHS of the form:
+
+\[ %\arraycolsep=4pt
+ \begin{bmatrix}
+     1&2\\
+     1/4&1&1/4\\
+     &1/4&1&1/4\\
+     &&1/4&1&1/4\\
+     &&&1/4&1&1/4\\
+     &&&&&\ddots\\
+     &&&&&&\ddots\\
+     &&&&&&&\ddots\\
+     &&&&&&&2&1
+  \end{bmatrix}
+\]
+
+And an RHS of the form:
+
+\[
+ \begin{bmatrix}
+     -5f_1 + 4f_2 + f_3\\
+     f_{3} - f_{1}\\
+     f_{4} - f_{2}\\
+     \vdots\\
+     \vdots\\
+     \vdots\\
+     \vdots\\
+     f_{n} - f_{n-2}\\
+     5f_{n} - 4f_{n-1} - f_{3}
+  \end{bmatrix}
+\]
+
+\section{Domain decomposition}
+
+The compact finite differences need to be evaluated
+at each grid point in the problem domain,
+shown in \ref{fig:domain}.
+Each grid point is represented as a cell in this representation.
+
+The domain has been shown as partitioned into blocks \ref{fig:one-block}.
+The blocks are three dimensional to facilitate
+the computation of derivatives in all three coordinate directions
+without incurring excessive communication costs.
+
+While evaluating the derivative, each block is assigned to a ``compute unit'',
+which may be a single core, a multi-core processor, or a GPU.
+Each block may be thought of as a collection of ``lines'' of grid points
+\ref{fig:one-line}.
+
+% left, lower, right, upper
 \begin{figure}[h]
 \begin{center}
-\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/domain.eps}
+\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=250pt]{img/domain.eps}
 \end{center}
@@
 
-\begin{figure}[h]
-\begin{center}
-\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/one-block.eps}
-\end{center}
+\begin{figure}[ht]
+\begin{minipage}[b]{0.45\linewidth}
+\centering
+\includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-block.eps}
 \caption{Single block}
 \label{fig:one-block}
+\end{minipage}
+\hspace{0.5cm}
+\begin{minipage}[b]{0.45\linewidth}
+\centering
+\includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-thread.eps}
+\caption{Single thread}
+\label{fig:one-line}
+\end{minipage}
 \end{figure}
@@
 \begin{center}
-\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/one-thread.eps}
+\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/parallel-x.eps}
 \end{center}
-\caption{Single thread}
-\label{fig:one-thread}
+\caption{Parallel x}
+\label{fig:parallel-x}
 \end{figure}
@@
 \begin{center}
-\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/domain.eps}
+\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/parallel-yz.eps}
 \end{center}
-\caption{Domain}
-\label{fig:domain}
+\caption{Parallel yz}
+\label{fig:parallel-yz}
 \end{figure}
@@
 \begin{center}
-\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/domain.eps}
+\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/parallel-both.eps}
 \end{center}
-\caption{Domain}
-\label{fig:domain}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/domain.eps}
-\end{center}
-\caption{Domain}
-\label{fig:domain}
+\caption{Parallel-both}
+\label{fig:parallel-both}
 \end{figure}

### 2015-09-18 report.tex
add to report

@@
 \title{An OpenCL-based tridiagonal solver for evaluation
-    of compact finite differences on Graphics Processing Units}
+    of compact finite differences on parallel architectures}
 \begin{document}
@@
     The motivation for this work comes from the need to evaluate
-\emph{compact finite differences} arising in large simulations of
-combustion.
+    \emph{compact finite differences} arising in large simulations of combustion.
+    Compact finite different schemes result in
+    high-order of accuracy for relatively small stencils.
+    Numerical methods with small stencil widths are especially benefitial
+    in managing communication costs in large-scale cluster computing.
+    This work is concerned with implementing algorithms for evaluating
+    compact finite differences to make full use of available parallel architecture.
+    The target architecture studied is the NVIDIA Tesla K20 GPU.
+    The algorithm design for this architecture is presented,
+    along with the associated issues.
+    Results are presented for domains sized up to $2048^3$ grid points,
+    using up to 64 GPUs.
+
+\section{Related work}
+
+\section{Numerical method}
 
@@
 
-\section{Domain decomposition}
+    A solution strategy is therefore concerned with the following:
+
+    \begin{enumerate}
+        \item{The evaluation of the right hand side.}
+        \item{The solution of the resulting matrix system.}
+    \end{enumerate}
+
+\section{Reference approach}
+    The current approach, developed for the CFDNS[ref] Compressible Navier-Stokes,
+    and adapted to the Roadrunner[ref] architecture,
+    is a parallel solver based on the standard LU algorithm,
+    but specialized for the solution of a constant matrix.
+    The problem is distributed among several ``processes'' and each
+    process uses a single CPU core to perform computations.
+
+    Assuming the coefficients of the LU algorithm are precomputed,
+    the following steps are required to solve the tridiagonal system
+    (only steps for the left-right sweep are listed, the right-left sweep
+    consists of similar steps):
+
+    \begin{enumerate}
+    \item Precompute the arrays of coefficients of the LU algorithm---
+    this is done once, in the beginning of the simulation process, and its cost
+    can be neglected.
+    \item Evaluate the RHS of the system.
+    \item Compute $\hat{\phi_i}$ and $\hat{\psi_i}$ for each grid point $i$,
+    with $\hat{\phi_i}$ and $\hat{\psi_i}$ both functions of the precomputed
+    coefficient arrays and the right hand side.
+    \item Every process posts the rightmost values of its $\hat{\phi_i}$ and $\hat{\psi_i}$,
+        $\widetilde{\phi_k}$ and $\widetilde{\psi_k}$.
+    \item Using the array of $\widetilde{\phi_k}$ and $\widetilde{\psi_k}$,
+        each process computes $\widetilde{x_k}$, and uses it to compute
+        $x_i,k = \phi_i + \psi_i\widetilde{x_k}$.
+    \end{enumerate}
+
+    The Message Passing Interface (MPI) is used for handling communication
+    between processes.
+
+\section{GPU architecture and parallelization strategies}
 
@@
     Each grid point is represented as a cell in this representation.
-
     The domain has been shown as partitioned into blocks \ref{fig:one-block}.
@@
     Each block may be thought of as a collection of ``lines'' of grid points
-\ref{fig:one-line}.
+    \ref{fig:one-line} in the x-direction,
+    and the entire domain may be thought of as a collection of ``lines'' of
+    blocks.
+    While computing derivatives in other directions,
+    (e.g., z-direction), we interpret each block as a collection of ``lines''
+    of grid points in the z-direction and the domain as a collection of
+    block in the z-direction.
+
+    A GPU affords massive parallelism by virtue of several (thousand)
+    lightweight compute threads (referred to as \emph{work items} in the OpenCL specification).
+    The work items on a GPU are scheduled as \emph{blocks},
+    which can be 1-D, 2-D, or 3-D.
+    Each block is assigned to a \emph{streaming microprocessor}
+    on the GPU---the Tesla K20 GPU has 13 streaming microprocessors---
+    and runs in parallel.
+    Threads within a block have common access to small, fast,
+    \emph{shared memory}, which can be explicitly controlled.
+
+    \subsection{Evaluation of the RHS}
+
+        The RHS is evaluated by a 3 point stencil computation of
+        the following form:
+
+        \begin{equation*}
+            r_i = \frac{3}{4dx} (f_{i+1} - f_{i-1})
+        \end{equation*}
+
+        This pointwise update can be performed easily in parallel
+        for the internal points (points not close to the boundary of the process).
+        Near the boundaries, the process will require information from its
+        neighbours.
+
+        Thus, each process will require information from its
+        neighbouring processes every time the derivative needs to be evaluated.
+
+        \begin{figure}[h]
+        \begin{center}
+        \includegraphics[trim={{100pt} {250pt} {100pt} {200pt}}, clip, height=200pt]{img/da-domain.eps}
+        \end{center}
+        \caption{Distribution of a 2-D domain among processes}
+        \label{fig:da-domain}
+        \end{figure}
+
+        \begin{figure}[h]
+        \begin{center}
+        \includegraphics[trim={{100pt} {250pt} {100pt} {200pt}}, clip, height=200pt]{img/da-local-portion.eps}
+        \end{center}
+        \caption{Elements required by a process to evaluate RHS}
+        \label{fig:da-local-portion}
+        \end{figure}
+
+        Thus, the RHS is evaluated in two steps:
+
+        \begin{enumerate}
+            \item Exchange boundary information with neighbouring processes
+            \item Evaluate the RHS at each grid point (in parallel)
+        \end{enumerate}
+
+        The GPU is specialized for performing such a large number of independent
+        computations as required in step 2.
+        The RHS at each grid point is computing by a single work item
+        (GPU thread).
+        Therefore the number of work items scheduled is
+        equal to the number of points in the local domain.
+
+        Another popular strategy is to take advantage of
+        shared memory, by launching 2-dimensional blocks of work-items,
+        loading each into shared memory, and streaming the third dimension
+        in registers.
+        This approach is discussed in [ref], and is much more effective
+        for larger stencils---as each thread in a block will access the same
+        value several times, it is effective to use shared memory as explicitly
+        controlled cache.
+
+    \subsection{Solving the tridiagonal system}
+
+        In the reference approach, a significantly larger portion of time is spent in solving
+        the tridiagonal system in parallel.
+        Because the problem does not exhibit the same degree of parallelism as
+        the pointwise stencil update,
+        the algorithms are necessarily more involved.
+
+\section{Code design}
+
+\section{Results and performance comparison}
+
+    \subsection{Results for a single node}
+
+    \begin{table}
+    \centering
+    \begin{tabular}{|l|r|r|r|r|}
+    \hline
+    Size    &      CPU (1 core) &  CPU (8 cores) & GPU (without memory transfers)  & GPU (with memory transfers) \\
+    \hline
+    32      &     0.0005        &   0.00029     &   0.00128 &   0.0014   \\
+    64      &     0.0038        &   0.0008      &   0.0018  &   0.00385  \\
+    128     &     0.042         &   0.0075      &   0.0073  &   0.02014  \\
+    256     &     0.34          &   0.062       &   0.057   &   0.148    \\
+    512     &     2.45          &   0.47        &   0.70    &   1.401    \\
+    \hline
+    \end{tabular}
+    \caption{Performance of solvers on a single GPU (Tesla k20m)}
+    \label{table:res1}
+    \end{table}
+
+    The results for a single node bring up some issues in performance evaluation.
+    First, it is clear that the major cost for the GPU is not in the computational
+    kernels, but rather in transferring data between host (CPU) and device (GPU) buffers.
+    In the context of a larger piece of software,
+    if only the computation of the compact finite difference was offloaded
+    to the GPU, then considering this cost is absolutely essential.
+    However, if the rest of the problem is also written to be solved on the GPU,
+    then there is no need for transfers to and from the host at each
+    time step.
+
+    It is instructive also to examine the the total time
+
+    We present the results for both
+
+    \subsection{Results for multi-node}
+
+    \subsection{Performance comparison}
+
+    \subsection{Major concerns}
+
+    \subsection{Performance improvement strategies}
+
+\section{Computational practices}
 
@@
 
-\section{Parallelism}
-
-\section{}
-
 \end{document}

### 2015-09-20 report.tex
add tesla block diagram; update report

@@
 
-\section{GPU architecture and parallelization strategies}
+\section{GPU architecture and OpenCL programming platform}
+
+    \begin{figure}[h]
+    \begin{center}
+    \includegraphics[height=200pt]{img/tesla-block-diagram.jpg}
+    \end{center}
+    \caption{Distribution of a 2-D domain among processes}
+    \label{fig:da-domain}
+    \end{figure}
+
+    A graphics processing unit (GPU) consists of several (thousand)
+    compute cores and is designed for high-throughput computations.
+
+    NVIDIA cards support extensions such as CUDA and OpenCL for
+    langauages like C, C++, Fortran and Python.
+    Applications written using these extensions are launched
+    with thousands of lightweight threads,
+    each operating on a small portion the problem domain (typically a single grid point).
+
+    Each GPU has several so-called streaming multiprocessors (SMs)---the Tesla K20M
+    has 13 SMs---each of which has its own control units, registers, caches, etc.
+    Instructions execute simultaneously on an SM.
+
+    All SMs (and therefore all threads) have access to the GPU's ``global'' memory.
+    Access to global memory is efficient when concurrent threads access neighbouring memory locations
+    (known as \emph{coalesced memory access}).
+
+    All threads in an SM also have access to a common fast, on-chip, \emph{shared memory}.
+    Access to shared memory, in general, is much faster than access to global memory,
+    but the amount of shared memory per SM is limited, typically a few dozen KB.
+    Certain memory access patterns by concurrent threads to shared memory may result in
+    serialization of the memory accesses, known as bank conflicts. These are detailed [here].
+
+    The OpenCL programming platform
+
+\section{Parallelization strategies}
 
@@
     \subsection{Results for a single node}
-
-    \begin{table}
-    \centering
-    \begin{tabular}{|l|r|r|r|r|}
-    \hline
-    Size    &      CPU (1 core) &  CPU (8 cores) & GPU (without memory transfers)  & GPU (with memory transfers) \\
-    \hline
-    32      &     0.0005        &   0.00029     &   0.00128 &   0.0014   \\
-    64      &     0.0038        &   0.0008      &   0.0018  &   0.00385  \\
-    128     &     0.042         &   0.0075      &   0.0073  &   0.02014  \\
-    256     &     0.34          &   0.062       &   0.057   &   0.148    \\
-    512     &     2.45          &   0.47        &   0.70    &   1.401    \\
-    \hline
-    \end{tabular}
-    \caption{Performance of solvers on a single GPU (Tesla k20m)}
-    \label{table:res1}
-    \end{table}
-
     The results for a single node bring up some issues in performance evaluation.

### 2015-09-21 report.tex
updates to report

@@
 \section{Related work}
+
+    [FastTridiagonal] discuss the GPU implementation of three algorithms
+    and their hybrid variants: Cyclic Reduction, Parallel Cyclic Reduction
+    and Recursive Doubling.
+    Their approach utilizes the GPUs \emph{shared memory},
+    and addresses the concerns of bank conflicts arising in these algorithms.
+    A global memory implementation of their solvers
+    runs about three times slower, but is able to accomodate larger systems.
+    The speedups they report are 12x over a 2.5 GHz quad-core CPU.
+
+    [ScalableTridiagonal] and [Autotuned] both propose hybrid
+    Parallel Cyclic Reduction and thread-level parallel Thomas algorithm (p-Thomas), and
+    methods to determine the switching point for the algorithms.
+    The speedups reported are 8.3x and 11x over a 3.33 GHz Intel i7 quad-core
+    processor and a 3.4 GHz Intel i5 dual-core respectively.
+
+    In [ScalableNumericallyStable],
+    the authors implement a heterogenous (GPU+MPI) solver using the
+    SPIKE algorithm for GPUs with a diagonal pivoting strategy for stability.
+    More recently, [ChunkedCyclic] propose a \emph{chunked} cyclic reduction
+    strategy for GPUss
+    In [GPUApplicationForHigherOrderCFD], the inverse of the coefficient matrix
+    is used to solve tridiagonal systems resulting from high-order compact
+    finite difference schemes. The work here is for a single GPU.
+    [An efficient GPU implementation of cyclic reduction solver] demonstrate
+    a method to improve memory accesses for Cyclic Reduction
+    using a reordering strategy, and apply it to compressible viscous flow
+    simulations.
+
+    --- Mention the speedups
+
+    In this work, we propose a variant of Cyclic Reduction for solving
+    tridiagonal systems arising in compact finite differences
+    in a cluster environment (several GPU nodes).
+    We introduce a strategy to
+    reduce global memory accesses and decrease memory requirements by
+    pre-computing coefficients appearing in Cyclic Reduction.
 
@@
 
-    A graphics processing unit (GPU) consists of several (thousand)
-    compute cores and is designed for high-throughput computations.
-
-    NVIDIA cards support extensions such as CUDA and OpenCL for
-    langauages like C, C++, Fortran and Python.
-    Applications written using these extensions are launched
-    with thousands of lightweight threads,
-    each operating on a small portion the problem domain (typically a single grid point).
-
-    Each GPU has several so-called streaming multiprocessors (SMs)---the Tesla K20M
-    has 13 SMs---each of which has its own control units, registers, caches, etc.
-    Instructions execute simultaneously on an SM.
-
+    GPUs afford massive parallelism by virtue of several (thousand)
+    lightweight compute threads, referred to as \emph{work items}.
+    The work items on a GPU are scheduled as \emph{blocks},
+    which can be 1-D, 2-D, or 3-D.
+    Each block is assigned to a \emph{streaming multiprocessor} (SM)
+    on the GPU---the Tesla K20 GPU has 13 streaming multiprocessor---and
+    all threads in a block are executed simultaneously by the SM.
     All SMs (and therefore all threads) have access to the GPU's ``global'' memory.
-    Access to global memory is efficient when concurrent threads access neighbouring memory locations
+    Access to global memory is efficient when concurrent threads
+    access neighbouring memory locations
     (known as \emph{coalesced memory access}).
-
     All threads in an SM also have access to a common fast, on-chip, \emph{shared memory}.
@@
 
-    The OpenCL programming platform
+\section{Parallelization strategies}
 
-\section{Parallelization strategies}
+    % left, lower, right, upper
+    \begin{figure}[h]
+    \begin{center}
+    \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=250pt]{img/domain.eps}
+    \end{center}
+    \caption{Problem domain}
+    \label{fig:domain}
+    \end{figure}
+
+    \begin{figure}[ht]
+    \begin{minipage}[b]{0.45\linewidth}
+    \centering
+    \includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-block.eps}
+    \caption{Single block assigned to a process}
+    \label{fig:block}
+    \end{minipage}
+
+    \hspace{0.5cm}
+    \begin{minipage}[b]{0.45\linewidth}
+    \centering
+    \includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-thread.eps}
+    \caption{Single grid line within a block}
+    \label{fig:grid-line}
+    \end{minipage}
+    \end{figure}
 
@@
 
-    While evaluating the derivative, each block is assigned to a ``compute unit'',
+    While evaluating the derivative in the x-direction (say),
+    each block is assigned to a \emph{compute unit}, or a \emph{process}
     which may be a single core, a multi-core processor, or a GPU.
     Each block may be thought of as a collection of ``lines'' of grid points
-    \ref{fig:one-line} in the x-direction,
-    and the entire domain may be thought of as a collection of ``lines'' of
-    blocks.
-    While computing derivatives in other directions,
-    (e.g., z-direction), we interpret each block as a collection of ``lines''
-    of grid points in the z-direction and the domain as a collection of
-    block in the z-direction.
-
-    A GPU affords massive parallelism by virtue of several (thousand)
-    lightweight compute threads (referred to as \emph{work items} in the OpenCL specification).
-    The work items on a GPU are scheduled as \emph{blocks},
-    which can be 1-D, 2-D, or 3-D.
-    Each block is assigned to a \emph{streaming microprocessor}
-    on the GPU---the Tesla K20 GPU has 13 streaming microprocessors---
-    and runs in parallel.
-    Threads within a block have common access to small, fast,
-    \emph{shared memory}, which can be explicitly controlled.
+    \ref{fig:one-line}---referred to as \emph{grid lines}---in the x-direction,
+    and the entire domain may be thought of as a collection of such blocks.
 
@@
 
+        \begin{figure}[h]
+        \begin{center}
+        \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=250pt]{img/block-dimensions.eps}
+        \end{center}
+        \caption{Domain}
+        \label{fig:block-dimensions}
+        \end{figure}
+
+        In the reference approach, the domain is decomposed into several blocks,
+        and each block is assigned to a CPU core.
+        Each core operates on a block with dimensions as shown
+        in the figure \ref{fig:block-dimensions}.
+        The computational pattern followed by each core for the steps listed in Section[SEC]
+        is a triple nested loop over the grid points in the block,
+        followed by a communication of the ``faces'' of the block with the other
+        cores:
+
+        \begin{listing}
+        \begin{minted}{fortran}
+           do j=1,n2
+              do i=1,n1
+                 x(1,i,j) = bet(1)*x(1,i,j)
+                 do k = 2, n3
+                    x(k,i,j) = (x(k,i,j) - a(k)*x(k-1,i,j))*bet(k)
+                 end do
+                 p_loc(i,j) = x(n3,i,j)
+              end do
+           end do
+
+           call MPI_Allgather(p_loc,n1*n2,double_real,p,n1*n2, &
+              double_real,comm,error)
+
+        \end{minted}
+        \caption{Computational pattern in reference approach}
+        \end{listing}
+
+        Conceptually, the innermost loop solves the tridiagonal system
+        for a single ``line'' of grid points.
+
+        In a multi-threaded environment,
+        rather than having a single thread perform the local part
+        of the computations for each line,
+        it is beneficial to have several threads ``sweep'' through the domain,
+        each performing the computations for a different tridiagonal system.
+        This is shown in Figure \ref{fig:block-multi-threaded} for a 4-core processor.
+        In each iteration of the inner-most loop,
+        4 lines of grid points are solved.
+
+        \begin{figure}[h]
+        \begin{center}
+        \includegraphics[trim={{50pt} {250pt} {50pt} {150pt}}, clip, height=300pt]{img/block-multi-threaded.eps}
+        \end{center}
+        \caption{Domain}
+        \label{fig:block-multi-threaded}
+        \end{figure}
+
+        Algorithms like LU and the Thomas Algorithm are inherently sequential.
+        While thread level parallelism can be used to solve several systems at once,
+        the number of steps to solve each system remains large.
+
+        Algorithms like Cyclic Reduction and Parallel Cyclic Reduction use
+        several parallel processors to solve a tridiagonal system.
+        Cyclic reduction
+
+\section{Description of the algorithm}
+
+    First, we discuss the cluster level parallelism.
+
+    Given N equations in N variables of the form:
+
+    \begin{align}
+    & A =
+     \begin{bmatrix}
+         b_1 & c_1 \\
+         a_2 & b_2 & c_2 \\
+             & a_3 & b_3 & c_3 \\
+             &     & a_4 & b_4 & c_4 \\
+             &     &     &     &  \ddots & c_{N-1}\\
+             &     &     &     &     a_N  & b_N
+      \end{bmatrix}&
+    \end{align}
+
+    each process $p$ is concerned with the solution of a \emph{subsystem}
+    of equations:
+
+    \begin{align}
+    &L_p =
+    \begin{bmatrix}
+        b_1^p & c_1^p \\
+        a_2^p & b_2^p & c_2^p \\
+              & a_3^p & b_3^p & c_3^p \\
+              &       & a_4^p & b_4^p & c_4^p \\
+              &       &       &       &  \ddots & c_{M-1}^p\\
+              &       &       &       &     a_{M}^p  & b_{M}^p
+     \end{bmatrix}&
+    \end{align}
+
+    For each subsystem, the following equations are solved:
+
+    \begin{align}
+        & L_p\boldsymbol{x}_p^R = \boldsymbol{r_p} &  \label{eqn:local-eqn-1} \\
+        & L_p\boldsymbol{x}_p^{UH} = (-a_1^p, 0, 0 \hdots 0)^T &  \label{eqn:local-eqn-2} \\
+        & L_p\boldsymbol{x}_p^{LH} = (0, 0, 0, \hdots -c_M^p)^T \label{eqn:local-eqn-3} &
+    \end{align}
+
+    The general solution is obtained as a linear combination of the above solutions:
+
+    \begin{equation} \label{eqn:linear-combo}
+        \boldsymbol{x}_p = {\xi}_p^{UH} \boldsymbol{x}_p^R + {\xi}_p^{LH} \boldsymbol{x}_p^{LH}
+    \end{equation}
+
+    Where the parameters ${\xi}_p^{UH}$ and ${\xi}_p^{LH}$ are obtained
+    by solving the following ``reduced'' system:
+
+    \begin{equation} \label{eqn:reduced-system}
+     \begin{bmatrix}
+         x_{1,M}^{LH}   &   -1                                                          \\
+         -1             &   x_{2,1}^{UH} &   x_{2,1}^{LH}                               \\
+                        &   x_{2,M}^{LH} &   x_{2,M}^{LH}  &  -1                        \\
+         &              &   -1           &   x_{3,M}^{UH}  &  x_{1,M}^{LH}              \\
+         &              &   &                x_{3,M}^{UH}  &  x_{1,M}^{LH} &    -1      \\
+         &              &   &            &   &             \ddots   \\
+         &              &   &            &   &             &        \ddots \\
+         &  &   &   &   &   &   & -1 x_{P,1}^{UH}
+      \end{bmatrix}
+    \begin{bmatrix}
+        \xi_1^{LH} \\
+        \xi_2^{UH} \\
+        \xi_2^{LH} \\
+        \xi_3^{UH} \\
+        \xi_3^{LH} \\
+        \vdots \\
+        \vdots \\
+        \xi_P^{UH} \\
+     \end{bmatrix}
+     =
+     \begin{bmatrix}
+         x_{1,M}^R \\
+         x_{2,1}^R \\
+         x_{2,M}^R \\
+         x_{3,1}^R \\
+         x_{3,M}^R \\
+         \vdots \\
+         \vdots \\
+         x_P^R \\
+      \end{bmatrix}
+  \end{equation}
+
+    The solution procedure is then:
+
+    \begin{enumerate}
+        \item Each process solves the three equations
+        \ref{eqn:local-eqn-1}-\ref{eqn:local-eqn-3}.
+        \item Each process communicates the parameters required to
+        assemble the ``reduced'' system eqn:reduced-system
+        \item The reduced system is solved for the coupling coefficients
+        $\xi_i^{UH}$ and $_i^{LH}$, which are sent to their respective processes
+        (if required)
+        \item Each process solves \ref{eqn:linear-combo} for the local part
+        of the solution to the global system.
+    \end{enumerate}
+
+    In the context of solving tridiagonal systems arising in compact finite differeces,
+    we note the following:
+
+    \begin{enumerate}
+        \item Each process is involved in solving \emph{several} equations,
+            one for each grid line.
+        \item
+        \item
+    \end{enumerate}
+
+    * Each process is involved in solving a number of equations, and not a single one
+    * The left hand side for all the equations is the same,
+        - LHS of \ref{eqn:local-eqn-1}-\ref{eqn:local-eqn-3}
+        - RHS of \ref{eqn:local-eqn-2} and \ref{eqn:local-eqn-3} are the same
+        - The above two equations need only be solved for a single grid line
+    * Only equation \ref{eqn:local-eqn-3} is solved for every grid line,
+    which amounts to solving the same equation for several different RHS
+
+    Cyclic Reduction
+    Precomputing coefficients in cyclic reduction
+    OpenCL implementations
+    Shared memory and associated issues
+    Repeated kernel invocations
+    Final kernel listings
+
+    Overall solution strategy
+
 \section{Code design}
@@
 \section{Results and performance comparison}
+    Notes on performance comparison
+        - Fair comparisons
+        - Memory transfers
+        - Time spent on writing code
+        - Compiler optimizations
+    Problem size v/s solution time
+    Breakup of time taken for various problem sizes and nprocs
+    Results for CPU
 
@@
 
-    It is instructive also to examine the the total time
-
-    We present the results for both
-
     \subsection{Results for multi-node}
@@
 \section{Computational practices}
-
-% left, lower, right, upper
-\begin{figure}[h]
-\begin{center}
-\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=250pt]{img/domain.eps}
-\end{center}
-\caption{Domain}
-\label{fig:domain}
-\end{figure}
-
-\begin{figure}[ht]
-\begin{minipage}[b]{0.45\linewidth}
-\centering
-\includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-block.eps}
-\caption{Single block}
-\label{fig:one-block}
-\end{minipage}
-\hspace{0.5cm}
-\begin{minipage}[b]{0.45\linewidth}
-\centering
-\includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-thread.eps}
-\caption{Single thread}
-\label{fig:one-line}
-\end{minipage}
-\end{figure}
 

### 2015-09-21 report.tex
updates to report and forward-reduction diagram

@@
 
-    The right boundary ($f^{\prime}_{n-1}$) is treated by utilizing
+    The right boundary ($f^{\prime}_{n}$) is treated by utilizing
     the negative complex-conjugate of the Fourier image of the stencil
@@
     \begin{equation}
-        f^{\prime}_{n-1} + 2f^{\prime}_{n-2}
+        f^{\prime}_{n} + 2f^{\prime}_{n-1s}
         =
-        \frac{5f_{n-1} - 4f_{n-2} - f_{n-3}}{dx}
+        \frac{5f_{n} - 4f_{n-2} 1 f_{n-2}}{dx}
     \end{equation}
@@
     The tridiagonal system that needs to be solved for the derivatives,
-    then has an LHS of the form:
+    then has the form:
 
-    \[ %\arraycolsep=4pt
+    \begin{equation} \label{eqn:compact-tridiagonal-system}
      \begin{bmatrix}
@@
       \end{bmatrix}
-    \]
-
-    And an RHS of the form:
-
-    \[
+      \begin{bmatrix}
+          f^{\prime}_1 \\
+          f^{\prime}_2 \\
+          f^{\prime}_3 \\
+          \vdots \\
+          \vdots \\
+          \vdots \\
+          \vdots \\
+          f^{\prime}_{n-1} \\
+          f^{\prime}_n
+       \end{bmatrix}
+     =
      \begin{bmatrix}
@@
          f_{n} - f_{n-2}\\
-         5f_{n} - 4f_{n-1} - f_{3}
+         5f_{n} - 4f_{n-1} - f_{n-2}
       \end{bmatrix}
-    \]
+    \end{equation}
 
@@
     GPUs afford massive parallelism by virtue of several (thousand)
-    lightweight compute threads, referred to as \emph{work items}.
-    The work items on a GPU are scheduled as \emph{blocks},
+    lightweight compute \emph{threads} or \emph{work-items}.
+    The threads on a GPU are scheduled as \emph{blocks},
     which can be 1-D, 2-D, or 3-D.
@@
     \begin{center}
-    \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=250pt]{img/domain.eps}
+    \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=200pt]{img/domain.eps}
     \end{center}
@@
     \end{figure}
-
+    %
     \begin{figure}[ht]
-    \begin{minipage}[b]{0.45\linewidth}
-    \centering
-    \includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-block.eps}
-    \caption{Single block assigned to a process}
+    \begin{minipage}[b]{0.40\linewidth}
+    \begin{center}
+    \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=200pt]{img/one-block.eps}
+    \end{center}
+    \caption{A subdomain}
     \label{fig:block}
     \end{minipage}
-
+    %
     \hspace{0.5cm}
-    \begin{minipage}[b]{0.45\linewidth}
-    \centering
-    \includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-thread.eps}
-    \caption{Single grid line within a block}
+    \begin{minipage}[b]{0.40\linewidth}
+    \begin{center}
+    \includegraphics[trim={{50pt} {150pt} {100pt} {150pt}}, clip, height=200pt]{img/one-thread.eps}
+    \end{center}
+    \caption{Single grid line within a subdomain}
     \label{fig:grid-line}
@@
     \end{figure}
+    %
+    \begin{figure}[h]
+    \begin{center}
+    \includegraphics[trim={{100pt} {250pt} {100pt} {150pt}}, clip, height=200pt]{img/block-dimensions.eps}
+    \end{center}
+    \caption{Domain}
+    \label{fig:block-dimensions}
+    \end{figure}
+
+    \pagebreak
 
@@
     Each grid point is represented as a cell in this representation.
-    The domain has been shown as partitioned into blocks \ref{fig:one-block}.
-    The blocks are three dimensional to facilitate
+    The domain has been shown as partitioned into subdomains \ref{fig:block}.
+    The subdomains are three dimensional to facilitate
     the computation of derivatives in all three coordinate directions
@@
     While evaluating the derivative in the x-direction (say),
-    each block is assigned to a \emph{compute unit}, or a \emph{process}
+    each subdomain is assigned to a \emph{compute unit}, or a \emph{process}
     which may be a single core, a multi-core processor, or a GPU.
-    Each block may be thought of as a collection of ``lines'' of grid points
-    \ref{fig:one-line}---referred to as \emph{grid lines}---in the x-direction,
-    and the entire domain may be thought of as a collection of such blocks.
+    Each subdomain may be thought of as a collection of ``lines'' of grid points
+    \ref{fig:grid-line}---referred to as \emph{grid lines}---in the x-direction,
+    and the entire domain may be thought of as a collection of such subdomains.
 
-    \subsection{Evaluation of the RHS}
-
-        The RHS is evaluated by a 3 point stencil computation of
-        the following form:
-
-        \begin{equation*}
-            r_i = \frac{3}{4dx} (f_{i+1} - f_{i-1})
-        \end{equation*}
-
-        This pointwise update can be performed easily in parallel
-        for the internal points (points not close to the boundary of the process).
-        Near the boundaries, the process will require information from its
-        neighbours.
-
-        Thus, each process will require information from its
-        neighbouring processes every time the derivative needs to be evaluated.
-
-        \begin{figure}[h]
-        \begin{center}
-        \includegraphics[trim={{100pt} {250pt} {100pt} {200pt}}, clip, height=200pt]{img/da-domain.eps}
-        \end{center}
-        \caption{Distribution of a 2-D domain among processes}
-        \label{fig:da-domain}
-        \end{figure}
-
-        \begin{figure}[h]
-        \begin{center}
-        \includegraphics[trim={{100pt} {250pt} {100pt} {200pt}}, clip, height=200pt]{img/da-local-portion.eps}
-        \end{center}
-        \caption{Elements required by a process to evaluate RHS}
-        \label{fig:da-local-portion}
-        \end{figure}
-
-        Thus, the RHS is evaluated in two steps:
-
-        \begin{enumerate}
-            \item Exchange boundary information with neighbouring processes
-            \item Evaluate the RHS at each grid point (in parallel)
-        \end{enumerate}
-
-        The GPU is specialized for performing such a large number of independent
-        computations as required in step 2.
-        The RHS at each grid point is computing by a single work item
-        (GPU thread).
-        Therefore the number of work items scheduled is
-        equal to the number of points in the local domain.
-
-        Another popular strategy is to take advantage of
-        shared memory, by launching 2-dimensional blocks of work-items,
-        loading each into shared memory, and streaming the third dimension
-        in registers.
-        This approach is discussed in [ref], and is much more effective
-        for larger stencils---as each thread in a block will access the same
-        value several times, it is effective to use shared memory as explicitly
-        controlled cache.
-
-    \subsection{Solving the tridiagonal system}
-
-        In the reference approach, a significantly larger portion of time is spent in solving
-        the tridiagonal system in parallel.
-        Because the problem does not exhibit the same degree of parallelism as
-        the pointwise stencil update,
-        the algorithms are necessarily more involved.
-
-        \begin{figure}[h]
-        \begin{center}
-        \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=250pt]{img/block-dimensions.eps}
-        \end{center}
-        \caption{Domain}
-        \label{fig:block-dimensions}
-        \end{figure}
-
-        In the reference approach, the domain is decomposed into several blocks,
-        and each block is assigned to a CPU core.
-        Each core operates on a block with dimensions as shown
+    In the reference approach, each subdomain is assigned to a CPU core.
+    Each core operates on a subdomain with dimensions as shown
     in the figure \ref{fig:block-dimensions}.
     The computational pattern followed by each core for the steps listed in Section[SEC]
-        is a triple nested loop over the grid points in the block,
-        followed by a communication of the ``faces'' of the block with the other
+    is a triple nested loop over the grid points in the subdomain,
+    followed by a communication of the ``faces'' of the subdomain with the other
     cores:
@@
     \begin{minted}{fortran}
+
        do j=1,n2
@@
     \caption{Computational pattern in reference approach}
+    \label{listing:triple-loop}
     \end{listing}
@@
     Conceptually, the innermost loop solves the tridiagonal system
-        for a single ``line'' of grid points.
+    for a single grid line,
+    while the outer loops navigate the different grid lines.
+    The evalution of such triple loops are found to be the most time-consuming steps
+    in the solution process,
+    and it is desirable to introduce a second level of parallelism to replace these loops.
+    Broadly speaking, there are two forms of parallelism that can be exploited:
 
+    \textbf{Thread-level parallelism} \\
         In a multi-threaded environment,
@@
         of the computations for each line,
-        it is beneficial to have several threads ``sweep'' through the domain,
+        it is beneficial to have several threads ``sweep'' through the subdomain,
         each performing the computations for a different tridiagonal system.
-        This is shown in Figure \ref{fig:block-multi-threaded} for a 4-core processor.
-        In each iteration of the inner-most loop,
-        4 lines of grid points are solved.
+        This is a sort of \emph{thread parallel} operations (Figure \ref{fig:thread-parallel}).
+        Algorithms that are inherently sequential,
+        such as the Thomas algorithm lend themselves easily to thread-level parallelism
+        With reference to the triple-loop in Listing 1,
+        algorithms employing thread-level parallelism reduce the
+        size of the outer two loops.
 
@@
         \begin{center}
-        \includegraphics[trim={{50pt} {250pt} {50pt} {150pt}}, clip, height=300pt]{img/block-multi-threaded.eps}
+        \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/thread-parallel.eps}
         \end{center}
-        \caption{Domain}
-        \label{fig:block-multi-threaded}
+        \caption{Parallel x}
+        \label{fig:thread-parallel}
         \end{figure}
 
-        Algorithms like LU and the Thomas Algorithm are inherently sequential.
-        While thread level parallelism can be used to solve several systems at once,
-        the number of steps to solve each system remains large.
+    \textbf{Data parallelism} \\
+        Algorithms like Cyclic Reduction and Parallel Cyclic Reduction
+        use parallel processors to perform
+        perform the same operation at different points in the tridiagonal system.
+        They are said to exhibit \emph{data parallelism} (Figure \ref{fig:data-parallel}).
+        Although these algorithms perform a larger number of computations
+        per step compared to the Thomas algorithm,
+        they can take far fewer steps to run,
+        depending on the number of available parallel processors.
+        With reference to the triple-loop in Listing 1,
+        algorithms employing data parallelism reduce the size of the innermost loop.
 
-        Algorithms like Cyclic Reduction and Parallel Cyclic Reduction use
-        several parallel processors to solve a tridiagonal system.
-        Cyclic reduction
+        \begin{figure}[h]
+        \begin{center}
+        \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/data-parallel.eps}
+        \end{center}
+        \caption{Parallel yz}
+        \label{fig:data-parallel}
+        \end{figure}
+
+    In this work, we evaluate the performance of a variant of Cyclic Reduction
+    on the GPU that is both data-parallel and thread-parallel,
+    and a thread-parallel implementation of the Thomas Algorithm on
+    multi-core CPUs.
+
+    Both implementations are compared against the reference implementation
+    described in Section [SEC].
+    The thread-parallel implementation of Thomas Algorithm is
+    straightforward.
+    However, implementation Cyclic Reduction on GPUs involves several issues,
+    and our algorithm is described in detail in Section [SEC].
+    %
+    % \subsection{Evaluation of the RHS}
+    %
+    %     The RHS is evaluated by a 3 point stencil computation of
+    %     the following form:
+    %
+    %     \begin{equation*}
+    %         r_i = \frac{3}{4dx} (f_{i+1} - f_{i-1})
+    %     \end{equation*}
+    %
+    %     This pointwise update can be performed easily in parallel
+    %     for the internal points (points not close to the boundary of the process).
+    %     Near the boundaries, the process will require information from its
+    %     neighbours.
+    %
+    %     Thus, each process will require information from its
+    %     neighbouring processes every time the derivative needs to be evaluated.
+    %
+    %     \begin{figure}[h]
+    %     \begin{center}
+    %     \includegraphics[trim={{100pt} {250pt} {100pt} {200pt}}, clip, height=200pt]{img/da-domain.eps}
+    %     \end{center}
+    %     \caption{Distribution of a 2-D domain among processes}
+    %     \label{fig:da-domain}
+    %     \end{figure}
+    %
+    %     \begin{figure}[h]
+    %     \begin{center}
+    %     \includegraphics[trim={{100pt} {250pt} {100pt} {200pt}}, clip, height=200pt]{img/da-local-portion.eps}
+    %     \end{center}
+    %     \caption{Elements required by a process to evaluate RHS}
+    %     \label{fig:da-local-portion}
+    %     \end{figure}
+    %
+    %     Thus, the RHS is evaluated in two steps:
+    %
+    %     \begin{enumerate}
+    %         \item Exchange boundary information with neighbouring processes
+    %         \item Evaluate the RHS at each grid point (in parallel)
+    %     \end{enumerate}
+    %
+    %     The GPU is specialized for performing such a large number of independent
+    %     computations as required in step 2.
+    %     The RHS at each grid point is computing by a single work item
+    %     (GPU thread).
+    %     Therefore the number of work items scheduled is
+    %     equal to the number of points in the local domain.
+    %
+    %     Another popular strategy is to take advantage of
+    %     shared memory, by launching 2-dimensional blocks of work-items,
+    %     loading each into shared memory, and streaming the third dimension
+    %     in registers.
+    %     This approach is discussed in [ref], and is much more effective
+    %     for larger stencils---as each thread in a block will access the same
+    %     value several times, it is effective to use shared memory as explicitly
+    %     controlled cache.
 
@@
 
-    First, we discuss the cluster level parallelism.
+    The size of the problems solved using the high-order compact finite
+    difference schemes discussed is usually much larger than can be accomodated
+    on a single compute node.
+    The problem domain is thus divided into ``subdomains'' and distributed among
+    several processes.
+    First, we discuss this cluster level parallelism.
 
@@
      \end{bmatrix}
-     =
+     = -
      \begin{bmatrix}
@@
         \item The reduced system is solved for the coupling coefficients
-        $\xi_i^{UH}$ and $_i^{LH}$, which are sent to their respective processes
+        $\xi_i^{UH}$ and $xi_i^{LH}$, which are sent to their respective processes
         (if required)
@@
     \begin{enumerate}
-        \item Each process is involved in solving \emph{several} equations,
-            one for each grid line.
-        \item
-        \item
+        \item Each process solves equations \ref{eqn:local-eqn-1}-\ref{eqn:local-eqn-3}
+            for \emph{each grid line} in its subdomain.
+        \item The left-hand side of these equations is the same
+            for each grid line, coming from equation \ref{eqn:compact-tridiagonal-system}).
+        \item The right-hand side for equations \ref{eqn:local-eqn-1} and
+            \ref{eqn:local-eqn-2} are the same for each grid line.
+            Thus, the equations need only be solved for a single grid line.
+        \item Only \ref{eqn:local-eqn-3} is evaluated for each grid line
+            in the subdomain. This is expected to be the most computationally expensive step.
     \end{enumerate}
 
-    * Each process is involved in solving a number of equations, and not a single one
-    * The left hand side for all the equations is the same,
-        - LHS of \ref{eqn:local-eqn-1}-\ref{eqn:local-eqn-3}
-        - RHS of \ref{eqn:local-eqn-2} and \ref{eqn:local-eqn-3} are the same
-        - The above two equations need only be solved for a single grid line
-    * Only equation \ref{eqn:local-eqn-3} is solved for every grid line,
-    which amounts to solving the same equation for several different RHS
+    \subsection{Cyclic reduction}
 
-    Cyclic Reduction
-    Precomputing coefficients in cyclic reduction
-    OpenCL implementations
-    Shared memory and associated issues
-    Repeated kernel invocations
-    Final kernel listings
+        The classic Cyclic Reduction algorithm consists of two phases:
+        forward reduction and backward substitution.
+        The forward substitution phase begins with a system of size $N$,
+        and by expressing every even-indexed equation $i$ as a linear
+        combination of equations $i$, $i-1$ and $i+1$, reduces it to a
+        system of size $N/2$.
+        The process is repeated until a system of 2 equations in 2 unknowns
+        is left.
 
-    Overall solution strategy
+        For each even-indexed $i$, we define:
 
-\section{Code design}
+        \begin{equation*}
+        k_1 = \frac{a_i}{b_{i-1}},
+        k_2 = \frac{c_i}{b_{i+1}}
+        \end{equation*}
+
+        Then, we update the values of $a$, $b$, $c$ and $d$ as:
+
+        \begin{align} \label{eqn:forward-reduction}
+        & a^{\prime}_i = -a_{i-1}k_1 & \\
+        & b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 & \\
+        & c^{\prime}_i = -c_{i+1}k_2 & \\
+        & d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 &
+        \end{align}
+
+        The two equations are solved, yielding $x_N$ and $x_{N/2}$.
+        In the backward substitution phase, every odd-indexed unknown $x_i$
+        is solved for by substituting the known values of $x_{i-1}$ and $x_{i+1}$:
+
+        \begin{align} \label{eqn:back-substitution}
+        x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - c^{\prime}_ix_{i+1}}{b^{\prime}_i}
+        \end{align}
+
+        At each step, for the last index $i$, the forward reduction step is
+        instead:
+        \begin{align} \label{eqn:forward-reduction-right}
+            & a^{\prime}_i = -a_{i-1}k_1 & \\
+            & b^{\prime}_i = b_i - c_{i-1}k_1 & \\
+            & d^{\prime}_i = d_i - d_{i-1}k_1 &
+        \end{align}
+
+        And the backward reduction step for $i=1$ is instead:
+        \begin{align} \label{eqn:back-substitution-right}
+        x_i = \frac{d^{\prime}_i - c^{\prime}_ix_{i+1}}{b^{\prime}_i}
+        \end{align}
+
+        \begin{figure}[h]
+        \begin{center}
+        \includegraphics[trim={{50pt} {300pt} {100pt} {50pt}}, clip, height=200pt]{img/forward-reduction.eps}
+        \end{center}
+        \caption{Cyclic reduction}
+        \label{fig:forward-reduction}
+        \end{figure}
+
+        The cyclic reduction algorithm is well suited to the massively
+        parallel architecture of the GPU.
+        In the most straightforward implementation,
+        we may launch a number of threads equal to
+        half the number of equations,
+        and assign each thread the task of
+        solving the equations \ref{eqn:forward-reduction} for each even-indexed
+        $i$.
+        We can then perform the successive steps of forward reduction
+        by halving the number of threads at each step.
+        The backward substitution phase is completed in a similar way,
+        with the number of threads doubling at each step.
+
+        This naive approach leads to the following issues:
+
+        \begin{enumerate}
+            \item Most of the threads are inactive,
+            except at the very beginning of forward reduction
+            and the very end of backward substitution.
+
+            \item The later steps of forward reduction and the first few
+            steps of backward substitution involve memory accesses with
+            large strides (see Figure \ref{fig:forward-reduction})
+            If the threads are accessing global memory,
+            this can lead to poor performance.
+        \end{enumerate}
+
+        The first issue is fairly straightforward to tackle.
+        The OpenCL API allows \emph{scheduling} of a large number of
+        threads---far more than the number of threads actually available
+        on the computing device.
+        Rather than scheduling threads for a single system (grid line),
+        we may schedule threads to perform
+        forward reduction steps for every grid line.
+
+        To avoid non-coalesced accesses to global memory,
+        we can instead have threads working with the fast, on-chip,
+        \emph{shared memory}. In this case, the threads in each
+        \emph{block} read data from global memory on to the block's
+        shared memory.
+        The threads then perform cyclic reduction on shared memory
+        and write the data back to global memory.
+        Conceptually, this approach has the advantage of avoiding
+        non-coalesced global memory accesses, and facilitates data reuse.
+        However, as described in [FastTridiagonalSolvers], the method
+        suffers from so-called \emph{bank conflicts} in which,
+        for certain memory access patters,
+        data access by concurrent threads from shared memory may be serialized.
+        This can seriously affect performance.
+        Several approaches for avoiding bank conflicts have been proposed,
+        this is at the cost of additional code complexity.
+        Besides, an approach using shared memory suffers from the penalty
+        of \emph{synchronization} between blocks and inactive threads.
+
+    \subsection{Precomputing scheme}
+
+        \begin{figure}[h]
+        \begin{center}
+        \includegraphics[trim={{50pt} {300pt} {100pt} {50pt}}, clip, height=200pt]{img/forward-reduction-coefficients.eps}
+        \end{center}
+        \caption{Forward reduction: indices with identical coefficients are shown in color}
+        \label{fig:forward-reduction-coefficients}
+        \end{figure}
+
+        We introduce a variant of cyclic reduction,
+        specialized for constant tridiagonal systems, that:
+
+        \begin{enumerate}
+        \item Reduces the number of accesses to global memory
+        \item Reduces the memory requirement
+        \item Reduces the number of computations in the forward reduction step
+        \end{enumerate}
+
+        We recall the form of the tridiagonal matrix:
+
+        \[
+        \begin{bmatrix}
+            1&2\\
+            1/4&1&1/4\\
+            &1/4&1&1/4\\
+            &&1/4&1&1/4\\
+            &&&1/4&1&1/4\\
+            &&&&&\ddots\\
+            &&&&&&\ddots\\
+            &&&&&&&\ddots\\
+            &&&&&&&2&1
+         \end{bmatrix}
+         \]
+
+        The matrix is the same for each grid line in the subdomain.
+        Thus,
+        the first three equations in \ref{eqn:forward-reduction}
+        are the same for every grid line.
+        For each forward reduction step,
+        the parameters $a^{\prime}_i$, $b^{\prime}_i$ and $c^{\prime}_i$
+        can be pre-computed.
+        For a general tridiagonal system, this would require storage for
+        $N/2 + N/4 + ... = 2N-2$ coefficients.
+        However, it is readily apparent that,
+        with the exception of the boundaries (near the first and last equations),
+        the coefficients $k_1$, $k_2$,
+        $a_i$, $b_i$ and $c_i$ are the same for each $i$.
+
+        The size of the precomputed coefficient arrays is then $log_2(N)-1$
+        (with an additional element for $a_N$ and $b_N$ as described below).
+        For event modestly sized $N$ such as 512, this represents large savings
+        in the memory requirements for the coefficient arrays.
+
+        More importantly, at each forward reduction step,
+        the number of non-coalesced memory accesses is greatly reduced.
+        With the above precomputed coefficients,
+        the forward reduction step now involves solving only a single equation
+        for $d_i$ at each $i$.
+
+        \begin{align} \label{eqn:forward-reduction-right}
+            & a^{\prime}_i = -a_{i-1}k_1 & \\
+            & b^{\prime}_i = b_i - c_{i-1}k_1 & \\
+            & d^{\prime}_i = d_i - d_{i-1}k_1 &
+        \end{align}
+
+        The backward reduction step benefits from improved access to
+        global memory.
+
+        \subsubsection*{Boundary conditions}
+
+        The tridiagonal systems for all grid lines are the same for all subdomains,
+        except near the boundaries.
+        From \ref{eqn:forward-reduction}, it is clear that
+        to handle the boundaries, three additional sets of coefficients
+        need to be stored:
+
+        \begin{enumerate}
+         \item The values for $b_1$ at each step
+         \item The values for $k_1$ evaluated at $i=1$ for each step
+         \item The values for $k_1$ evaluated at $i=n$ for each step,
+            where $n$ is the last index of the reduction step.
+        \end{enumerate}
+
+        The values of $a_N$ and $b_N$ are required only for the
+        final step (2-by-2 solve) of forward reduction.
+        These two scalar values may be stored separately.
 
 \section{Results and performance comparison}
-    Notes on performance comparison
-        - Fair comparisons
-        - Memory transfers
-        - Time spent on writing code
-        - Compiler optimizations
-    Problem size v/s solution time
-    Breakup of time taken for various problem sizes and nprocs
-    Results for CPU
-
-    \subsection{Results for a single node}
-    The results for a single node bring up some issues in performance evaluation.
-    First, it is clear that the major cost for the GPU is not in the computational
-    kernels, but rather in transferring data between host (CPU) and device (GPU) buffers.
-    In the context of a larger piece of software,
-    if only the computation of the compact finite difference was offloaded
-    to the GPU, then considering this cost is absolutely essential.
-    However, if the rest of the problem is also written to be solved on the GPU,
-    then there is no need for transfers to and from the host at each
-    time step.
-
-    \subsection{Results for multi-node}
 
@@
 \begin{center}
-\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/parallel-x.eps}
-\end{center}
-\caption{Parallel x}
-\label{fig:parallel-x}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/parallel-yz.eps}
-\end{center}
-\caption{Parallel yz}
-\label{fig:parallel-yz}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/parallel-both.eps}
+\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/data-and-thread-parallel.eps}
 \end{center}

### 2015-09-21 results/graphs/plot_graphs.py
add bibliography; graphs

@@ -1,0 +1,59 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib
+
+sizes = np.array([64, 128, 256, 512, 1024, 2048])
+
+# results on 4 nodes:
+CPU_4 = np.array([0.000654, 0.00144, 0.020, 0.147, 1.09, None])
+GPU_4 = np.array([0.0043, 0.0052, 0.0094, 0.034, 0.207, None])
+MCPU_4 = np.array([None, 0.01, 0.034, 0.124, 0.75, None])
+
+fig, ax = plt.subplots()
+ax.plot(sizes[:-1], CPU_4[:-1], '-o', linewidth=2, label='Reference approach: 64 CPU cores')
+ax.plot(sizes[:-1], GPU_4[:-1], '-o', linewidth=2, label='Cyclic reduction: 8 GPUs')
+ax.plot(sizes[1:-1], MCPU_4[1:-1], '-o', linewidth=2, label='pThomas algorithm: 8 CPUs')
+ax.set_xlabel('Problem size in one direction')
+ax.set_ylabel('Time to solve tridiagonal system')
+
+ax.set_yscale('log', basey=2)
+ax.set_xscale('log', basex=2)
+ax.get_xaxis().set_major_formatter(matplotlib.ticker.ScalarFormatter())
+
+ax.tick_params(labelright=True)
+
+ax.legend(loc='best')
+ax.grid()
+fig.savefig('4-nodes.png')
+plt.close(fig)
+
+# results on 32 nodes:
+CPU_32 = np.array([None, 0.0017, 0.010, 0.028, 0.226, 1.75])
+GPU_32 = np.array([None, 0.017, 0.007, 0.026, 0.084, 0.41])
+MCPU_32 = np.array([None, None, 0.011, 0.046, 0.19, 1.09])
+
+fig, ax = plt.subplots()
+ax.loglog(sizes[1:], CPU_32[1:], '-o', basex=2, basey=2, linewidth=2, label='Reference approach: 512 CPU cores')
+ax.loglog(sizes[1:], GPU_32[1:], '-o', basex=2, basey=2, linewidth=2, label='Cyclic reduction: 64 GPUs')
+ax.loglog(sizes[2:], MCPU_32[2:], '-o', basex=2, basey=2, linewidth=2, label='pThomas algorithm: 64 CPUs')
+ax.set_xlabel('Problem size in one direction')
[…]

### 2015-09-21 results/graphs/plot_timings.py

@@ -1,0 +1,27 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib
+
+def plot_stacked_bar(offs, width, values, colors):
+    y = 0
+    for color, value in zip(colors, values):
+        p = plt.bar(offs, value, width, bottom=y, color=color)
+        y += value
+
+sizes = np.array([512, 1024, 2048])
+
+CPU_512_64 = np.array([0.0096, 0.1375])
+CPU_1024_64 =
+CPU_2048_64 = 
+
+CPU_512_512 = np.array([0.0044, 0.02465])
+GPU_512_8 = np.array([0.02157, 0.01, 0.0035])
+GPU_512_64 = np.array([0.0064, 0.018, 0.0015])
+MCPU_512_8 = np.array([0.0844, 0.0205, 0.0225])
+MCPU_512_64 = np.array([0.03, 0.0068, 0.00437])
+
+plot_stacked_bar(1, 0.2, CPU_512_64, ['g', 'r'])
+plot_stacked_bar(2, 0.2, CPU_512_512, ['g', 'r'])
+
+plt.savefig('example.png')
+plt.close()

### 2015-09-27 2015-09-27-report/report.tex
move all materials relating to september report to new directory

--- 2015-09-21 report.tex
@@
 \author{Ashwin Srinath}
-\title{An OpenCL-based tridiagonal solver for evaluation
+\title{OpenCL-based tridiagonal solvers for evaluation
     of compact finite differences on parallel architectures}
+
 \begin{document}
-% \renewcommand{\bibname}{References}
+\renewcommand{\refname}{References}
+
+\maketitle
+
+\begin{abstract}
+    We present methods to evaluate compact finite differences in a
+    heterogeneous cluster environment
+    on two different architectures: NVIDIA GPUs and Intel multicore CPUs.
+    Our methods are compared against the CFDNS implementation,
+    which uses a parallelized version of the LU algorithm.
+    The implementations are tested on the Clemson University Palmetto supercomputer---each
+    node is equipped with 2 NVIDIA Tesla K20 GPUs and a
+    2.4 GHz 16 core Intel Xeon Processor.
+    The OpenCL framework is used to develop platform-independent code
+    so that other parallel architectures can be studied.
+    For a problem sized $2048^{3}$,
+    our GPU implementation on 64 GPUs provides a
+    5x speedup against the CFDNS implementation on the equivalent 512 CPU cores,
+    and our CPU implementation provides a 2x speedup against
+    the CFDNS implementation on 512 CPU cores.
+    The speedup is shown to increase with problem size.
+    Compiler optimizations are enabled for all three implementations.
+    Our methods achieve speedup by exploiting the parallel architecture in two ways:
+    (1) employing cyclic reduction (a parallel algorithm) for each grid line (in the case of GPU), and
+    (2) solving multiple grid lines concurrently (both CPU and GPU).
+    Our GPU implementation uses a novel strategy
+    based on precomputing coefficients,
+    to improve the memory access pattern,
+    and reduce the number of computations involved.
+    The results are of direct interest for implementing compact finite difference
+    solvers for other parallel architectures, such as MICs.
+
+\end{abstract}
 
@@
     The motivation for this work comes from the need to evaluate
-    \emph{compact finite differences} arising in large simulations of combustion.
+    \emph{compact finite differences} arising in large CFD simulations.
     Compact finite different schemes result in
-    high-order of accuracy for relatively small stencils.
+    high order of accuracy for relatively small stencils.
     Numerical methods with small stencil widths are especially benefitial
     in managing communication costs in large-scale cluster computing.
-    This work is concerned with implementing algorithms for evaluating
-    compact finite differences to make full use of available parallel architecture.
-    The target architecture studied is the NVIDIA Tesla K20 GPU.
-    The algorithm design for this architecture is presented,
-    along with the associated issues.
-    Results are presented for domains sized up to $2048^3$ grid points,
-    using up to 64 GPUs.
+    Compact finite difference schemes lead to
+    several banded tridiagonal systems,
+    the solution of which is generally the most time-consuming step in the simulation process.
+    It is thus desirable to exploit available parallel architectures to speed up
+    the solution of the tridiagonal systems.
+    The target architectures studied are the NVIDIA Tesla K20 GPU,
+    and and a 2.4 GHz Intel Xeon processor with 16 cores.
+    Other parallel architectures are of interest---specifically the
+    Intel Xeon Phi many-integrated core.
 
-\section{Related work}
-
-    [FastTridiagonal] discuss the GPU implementation of three algorithms
+    Several recent efforts have been made to exploit massively parallel architectures
+    in the solution of tridiagonal systems.
+    \cite{Zhang:2010:FTS:1837853.1693472} discuss the GPU implementation of three algorithms
     and their hybrid variants: Cyclic Reduction, Parallel Cyclic Reduction
@@
     The speedups they report are 12x over a 2.5 GHz quad-core CPU.
-
-    [ScalableTridiagonal] and [Autotuned] both propose hybrid
+    \cite{kim2011scalable} and \cite{davidson2011auto} both propose hybrid
     Parallel Cyclic Reduction and thread-level parallel Thomas algorithm (p-Thomas), and
@@
     processor and a 3.4 GHz Intel i5 dual-core respectively.
-
-    In [ScalableNumericallyStable],
+    In \cite{chang2012scalable},
     the authors implement a heterogenous (GPU+MPI) solver using the
     SPIKE algorithm for GPUs with a diagonal pivoting strategy for stability.
-    More recently, [ChunkedCyclic] propose a \emph{chunked} cyclic reduction
-    strategy for GPUss
-    In [GPUApplicationForHigherOrderCFD], the inverse of the coefficient matrix
+    More recently, \cite{zhao2015efficiently} propose a \emph{chunked} cyclic reduction
+    strategy for GPUs.
+    In \cite{tutkun2012gpu}, the inverse of the coefficient matrix
     is used to solve tridiagonal systems resulting from high-order compact
     finite difference schemes. The work here is for a single GPU.
-    [An efficient GPU implementation of cyclic reduction solver] demonstrate
+    \cite{esfahanian2014efficient} demonstrate
     a method to improve memory accesses for Cyclic Reduction
@@
 
-    --- Mention the speedups
-
-    In this work, we propose a variant of Cyclic Reduction for solving
-    tridiagonal systems arising in compact finite differences
-    in a cluster environment (several GPU nodes).
-    We introduce a strategy to
-    reduce global memory accesses and decrease memory requirements by
-    pre-computing coefficients appearing in Cyclic Reduction.
+    In this work,
+    we demonstrate the performance of a thread-parallel Thomas algorithm
+    on multi-core CPUs.
+    For GPUs,
+    we propose novel variant of Cyclic Reduction algorithm that
+    reduces global memory accesses and decreases memory requirements by
+    pre-computing coefficients.
 
@@
 
-\section{Reference approach}
-    The current approach, developed for the CFDNS[ref] Compressible Navier-Stokes,
-    and adapted to the Roadrunner[ref] architecture,
-    is a parallel solver based on the standard LU algorithm,
+\section{Reference approach} \label{sec:ref-approach}
+    The current approach, developed for the CFDNS \cite{livescu2009cfdns}
+    Direct Numerical Simulation solver,
+    uses a parallel tridiagonal solver based on the standard LU algorithm,
     but specialized for the solution of a constant matrix.
@@
     process uses a single CPU core to perform computations.
-
     Assuming the coefficients of the LU algorithm are precomputed,
@@
 
-\section{GPU architecture and OpenCL programming platform}
-
-    \begin{figure}[h]
-    \begin{center}
-    \includegraphics[height=200pt]{img/tesla-block-diagram.jpg}
-    \end{center}
-    \caption{Distribution of a 2-D domain among processes}
-    \label{fig:da-domain}
-    \end{figure}
-
-    GPUs afford massive parallelism by virtue of several (thousand)
-    lightweight compute \emph{threads} or \emph{work-items}.
-    The threads on a GPU are scheduled as \emph{blocks},
-    which can be 1-D, 2-D, or 3-D.
-    Each block is assigned to a \emph{streaming multiprocessor} (SM)
-    on the GPU---the Tesla K20 GPU has 13 streaming multiprocessor---and
-    all threads in a block are executed simultaneously by the SM.
-    All SMs (and therefore all threads) have access to the GPU's ``global'' memory.
-    Access to global memory is efficient when concurrent threads
-    access neighbouring memory locations
-    (known as \emph{coalesced memory access}).
-    All threads in an SM also have access to a common fast, on-chip, \emph{shared memory}.
-    Access to shared memory, in general, is much faster than access to global memory,
-    but the amount of shared memory per SM is limited, typically a few dozen KB.
-    Certain memory access patterns by concurrent threads to shared memory may result in
-    serialization of the memory accesses, known as bank conflicts. These are detailed [here].
-
 \section{Parallelization strategies}
@@
     \begin{center}
-    \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=200pt]{img/one-block.eps}
+    \includegraphics[trim={{50pt} {50pt} {50pt} {50pt}}, clip, height=200pt]{img/one-block.eps}
     \end{center}
-    \caption{A subdomain}
+    \caption{A subdomain (assigned to a process)}
     \label{fig:block}
@@
     \begin{center}
-    \includegraphics[trim={{50pt} {150pt} {100pt} {150pt}}, clip, height=200pt]{img/one-thread.eps}
+    \includegraphics[trim={{50pt} {50pt} {50pt} {50pt}}, clip, height=200pt]{img/one-thread.eps}
     \end{center}
@@
     \end{center}
-    \caption{Domain}
+    \caption{Dimensions of subdomain}
     \label{fig:block-dimensions}
@@
     Each grid point is represented as a cell in this representation.
-    The domain has been shown as partitioned into subdomains \ref{fig:block}.
-    The subdomains are three dimensional to facilitate
-    the computation of derivatives in all three coordinate directions
-    without incurring excessive communication costs.
+    The domain has been shown as partitioned into subdomains (Figure \ref{fig:block}).
 
@@
     Each subdomain may be thought of as a collection of ``lines'' of grid points
-    \ref{fig:grid-line}---referred to as \emph{grid lines}---in the x-direction,
+    ---referred to as \emph{grid lines}---in the x-direction (Figure \ref{fig:grid-line}),
     and the entire domain may be thought of as a collection of such subdomains.
@@
     in the figure \ref{fig:block-dimensions}.
-    The computational pattern followed by each core for the steps listed in Section[SEC]
+    The computational pattern followed by each core for the steps listed in Section
+    \ref{sec:ref-approach}
     is a triple nested loop over the grid points in the subdomain,
@@
 
+    \pagebreak
+
     \textbf{Thread-level parallelism} \\
@@
         each performing the computations for a different tridiagonal system.
-        This is a sort of \emph{thread parallel} operations (Figure \ref{fig:thread-parallel}).
+        This is a sort of \emph{thread parallel} operation (Figure \ref{fig:thread-parallel}).
         Algorithms that are inherently sequential,
@@
         \begin{center}
-        \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/thread-parallel.eps}
+        \includegraphics[trim={{100pt} {250pt} {100pt} {150pt}}, clip, height=200pt]{img/thread-parallel.eps}
         \end{center}
-        \caption{Parallel x}
+        \caption{Thread-level parallelism}
         \label{fig:thread-parallel}
@@
         \begin{center}
-        \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/data-parallel.eps}
+        \includegraphics[trim={{100pt} {200pt} {100pt} {150pt}}, clip, height=200pt]{img/data-parallel.eps}
         \end{center}
-        \caption{Parallel yz}
+        \caption{Data parallelism}
         \label{fig:data-parallel}
@@
 
-    In this work, we evaluate the performance of a variant of Cyclic Reduction
+    In this work, we evaluate the performance of a variant of cyclic reduction
     on the GPU that is both data-parallel and thread-parallel,
@@
     Both implementations are compared against the reference implementation
-    described in Section [SEC].
+    described in Section \ref{sec:ref-approach}.
     The thread-parallel implementation of Thomas Algorithm is
     straightforward.
-    However, implementation Cyclic Reduction on GPUs involves several issues,
-    and our algorithm is described in detail in Section [SEC].
+    However, implementation of cyclic reduction on GPUs involves several issues,
+    and our algorithm is described in detail in Section \ref{sec:gpu-algorithm}.
     %
@@
 
-\section{Description of the algorithm}
+\section{Description of the GPU algorithm} \label{sec:gpu-algorithm}
+
+    \subsection{Cluster-level parallelization strategy}
 
@@
     First, we discuss this cluster level parallelism.
+    This strategy is common to both our CPU and GPU solvers,
+    and was first proposed in \cite{mattor1995algorithm}
 
@@
         \item Each process communicates the parameters required to
-        assemble the ``reduced'' system eqn:reduced-system
+        assemble the ``reduced'' system \ref{eqn:reduced-system}.
         \item The reduced system is solved for the coupling coefficients
-        $\xi_i^{UH}$ and $xi_i^{LH}$, which are sent to their respective processes
+        $\xi_i^{UH}$ and $\xi_i^{LH}$, which are sent to their respective processes
         (if required)
@@
     \end{enumerate}
+
+    \subsection{GPU architecture}
+
+        \begin{figure}[h]
+        \begin{center}
+        \includegraphics[height=200pt]{img/tesla-block-diagram.jpg}
+        \end{center}
+        \caption{Block diagram of the K20 GPU}
+        \label{fig:k20}
+        \end{figure}
+
+        GPUs afford massive parallelism by virtue of several (thousand)
+        lightweight compute \emph{threads} or \emph{work-items}.
+        The threads on a GPU are scheduled as \emph{blocks},
+        which can be 1-D, 2-D, or 3-D.
+        Each block is assigned to a \emph{streaming multiprocessor} (SM)
+        on the GPU---the Tesla K20 GPU (Figure \ref{fig:k20})
+        has 13 streaming multiprocessor---and
+        all threads in a block are executed simultaneously by the SM.
+        All SMs (and therefore all threads) have access to the GPU's ``global'' memory.
+        Access to global memory is efficient when concurrent threads
+        access neighbouring memory locations
+        (known as \emph{coalesced memory access}).
+        All threads in an SM also have access to a common fast, on-chip, \emph{shared memory}.
+        Access to shared memory, in general, is much faster than access to global memory,
+        but the amount of shared memory per SM is limited, typically a few dozen KB.
+        Certain memory access patterns by concurrent threads to shared memory may result in
+        serialization of the memory accesses, known as bank conflicts.
 
@@
 
-        And the backward reduction step for $i=1$ is instead:
+        And the backward substitution step for $i=1$ is instead:
         \begin{align} \label{eqn:back-substitution-right}
@@
         \end{center}
-        \caption{Cyclic reduction}
+        \caption{Cyclic reduction: memory access pattern for forward reduction}
         \label{fig:forward-reduction}
@@
         forward reduction steps for every grid line.
+        This ensures that threads that would otherwise be inactive
+        now perform computations for other grid lines.
 
@@
         non-coalesced global memory accesses, and facilitates data reuse.
-        However, as described in [FastTridiagonalSolvers], the method
+        However, as described in \cite{Zhang:2010:FTS:1837853.1693472}, the method
         suffers from so-called \emph{bank conflicts} in which,
@@
 
-        The backward reduction step benefits from improved access to
+        The backward substitution step benefits from improved access to
         global memory.
@@
 
-\section{Results and performance comparison}
+\section{Performance comparison}
 
-    \subsection{Performance comparison}
+    The reference implementation is compared against our implementations
+    of the following algorithms:
 
-    \subsection{Major concerns}
+    \begin{enumerate}
+        \item Cyclic reduction with precomputed coefficients (described above)
+            for GPUs
+        \item p-Thomas or thread-parallel Thomas algorithm
+            for multi-core CPUs
+    \end{enumerate}
 
-    \subsection{Performance improvement strategies}
+    In most of the literature, GPU performance is measured
+    by comparing the performance of the GPU with
+    a single CPU core, or with 2-4 CPU cores.
+    We consider a ``node-for-node'' comparison.
+    Each node on the Palmetto cluster has 16 CPU cores and 2 GPUs,
+    so we maintain a 1:8 ratio of GPUs to CPU cores in our performance evaluation.
+    \texttt{O2} level compiler optimizations are applied for the reference implementation,
+    the default OpenCL optimizations are applied for our implementations.
+    We assume that data loaded on to the GPU is used in other computations
+    (evaluation of the RHS, computation of the derivative in other directions),
+    and that the computations take long enough that
+    the cost of CPU-GPU transfers may be ignored.
 
-\section{Computational practices}
-
-\begin{figure}[h]
+    \begin{figure}[h!]
     \begin{center}
-\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/data-and-thread-parallel.eps}
+    \includegraphics[height=200pt]{results/graphs/4-nodes.png}
     \end{center}
-\caption{Parallel-both}
-\label{fig:parallel-both}
+    \caption{Performance of solvers on 4 compute nodes (64 CPU cores, 8 GPUs))}
+    \label{fig:4-nodes}
     \end{figure}
 
+    \begin{figure}[h!]
+    \begin{center}
+    \includegraphics[height=200pt]{results/graphs/32-nodes.png}
+    \end{center}
+    \caption{Performance of solvers on 32 compute nodes (512 CPU cores, 64 GPUs))}
+    \label{fig:32-nodes}
+    \end{figure}
+
+    From Figures \ref{fig:4-nodes} and \ref{fig:32-nodes},
+    it is clear that the OpenCL implementations run faster than the reference
+    implementation for larger problem sizes.
+    We also note an increase in the speedup with the problem size.
+    For smaller problem sizes,
+    the data---specifically for the communication phases of the algorithms---was
+    noted to be somewhat noisy.
+
+\section{Performance improvement and future work}
+
+    There is much scope for even better parallel performance and future study.
+    Initial profiling results suggest that most of the time is spent solving
+    the primary tridiagonal systems,
+    but a significant portion of time is spent in assembling and solving
+    the secondary ``reduced'' system of equations.
+    Our implementations need to be carefully profiled
+    to understand where further opimization efforts are required.
+    Secondly, finite differences in the other directions can be implemented similarly
+    in a straightforward way,
+    but it is possible to use a \emph{one-shot}
+    strategy, as shown in \cite{esfahanian2014efficient}
+    to evaulate all three derivatives concurrently.
+    We would also like to develop and evaluate the performance
+    a thread-parallel version of the current CFDNS solver,
+    which will solve multiple grid lines in a single sweep.
+    Finally, we propose to evaluate the performance of our solvers
+    on the Intel Many Integrated Core architecture, where we expect to see
+    similar results.
+
+    \bibliography{references}
+    \bibliographystyle{abbrv}
+
 \end{document}

### 2015-10-14 paper-draft/paper.tex
add paper draft

\documentclass{elsarticle}

\usepackage{amsmath}

\begin{document}

\begin{frontmatter}
\author{Ashwin Srinath}
\title{A}
\maketitle

\begin{abstract}
    We present a computation strategy for evaluating
    compact finite differences efficiently on GPU clusters.
    We describe a novel algorithm for solving the
    near-Toeplitz tridiagonal systems associated with
    the compact finite difference schemes.
    Our specialized tridiagonal solver
    takes advantage of the near-Toeplitz nature of the
    tridiagonal system to precompute the coefficients
    appearing in the Cyclic Reduction algorithm.
    This allows us to solve the tridiagonal systems
    up to 2 times faster than with the NVIDIA CUSPARSE
    gtsvStridedBatch routine.
    Additionally, we present a methodology to solve for
    compact finite differences fully on multiple GPUs on a cluster
    without intermediate host-device transfers.
\end{abstract}

\end{frontmatter}

\section{Introduction}

\section{Compact Finite Difference Schemes}

For example,
in a uniformly spaced one-dimensional grid with spacing $dx$,
if $f_i$ represents the value of
the function evaluated at the $i$th node,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:

\begin{equation}
\begin{split}
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
        f^{\prime}_i
    = 
    c\frac{f_{i+3} - f_{i-3}}{dx} + \
    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
    a\frac{f_{i+1} - f_{i-1}}{dx} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}

The derivatives near the boundaries are approximated using
one-sided stencil operators of the form:

\begin{equation}
f^{\prime}_1 + \alpha_1 f^{\prime}_2 = \
    \frac{1}{dx} (a_1 f_1 + b_1 f_2 + c_1 f_3 + d_1 f_4) 
\end{equation}

and

\begin{equation}
    f^{\prime}_n + \alpha_n f^{\prime}_{n-1} = \
    \frac{1}{dx} (a_n f_n + b_n f_{n-1} + c_n f_{n-2} + d_n f_{n-3}) 
\end{equation}

As described by Lele \cite{lele1992compact},
setting $\beta = 0$ in Equation \ref{eqn:general-compact} leads to
a family of tridiagonal schemes with a single parameter ($\alpha$).
The coefficient matrix associated with such tridiagonal schemes
has the general form:

\begin{equation} \label{eqn:compact-tridiagonal-system}
\begin{bmatrix}
     1 &  \alpha_1  \\
     \alpha   &  1   &  \alpha \\
         &  \alpha   &  1  &  \alpha  \\
         &      &  \alpha  &  1  &  \alpha  \\
         &      &     &     &  \ddots \\
         &      &     &     &     &  \ddots  \\
         &      &     &     &     &  \alpha_n &  1
\end{bmatrix}
\begin{bmatrix}
    f^{\prime}_1 \\
    f^{\prime}_2 \\
    f^{\prime}_3 \\
    \vdots \\
    \vdots \\
    f^{\prime}_{n-1} \\
    f^{\prime}_n
 \end{bmatrix}
=
\begin{bmatrix}
   d_1 \\
   d_2 \\
   d_3 \\
   \vdots \\
   \vdots \\
   d_{n-1} \\
   d_{n}
\end{bmatrix}
\end{equation}

We note that the system is \emph{near Toeplitz tridiagonal},
i.e., the left-hand-side matrix is tridiagonal with
constant coefficients along each diagonal,
except for the first and last equations.

\section{Distributed compact-finite difference solution}

We use the algorithm discussed by Mattor et al.
\cite{mattor1995algorithm}
to distribute and solve the problem among multiple processes. 
We present the algorithm to solve a general tridiagonal system,
and in subsequent sections discuss its adaptation to solving
near-Teoplitz tridiagonal systems on GPUs.

Given a tridiagonal system with  $n$ equations,
to be solved by $P$ processes:

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
x_{r,1}^p \\
x_{r,2}^p \\
x_{r,3}^p \\
x_{r,4}^p \\
\vdots \\
x_{r,m}^p
\end{bmatrix}
=
\begin{bmatrix}
r_1^p \\
r_2^p \\
r_3^p \\
r_4^p \\
\vdots \\
r_m^p
\end{bmatrix} & \label{eqn:global-system} 
\end{align}

each process $p$ proceeds by solving the following three ``local''
tridiagonal systems:

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
x_{r,1}^p \\
x_{r,2}^p \\
x_{r,3}^p \\
x_{r,4}^p \\
\vdots \\
x_{r,m}^p
\end{bmatrix}
=
\begin{bmatrix}
r_1^p \\
r_2^p \\
r_3^p \\
r_4^p \\
\vdots \\
r_m^p
\end{bmatrix} & \label{eqn:primary-system} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
u_1^p \\
u_2^p \\
u_3^p \\
u_4^p \\
\vdots \\
u_m^p
\end{bmatrix}
=
\begin{bmatrix}
-a_1^p \\
0 \\
0 \\
0 \\
\vdots \\
0
\end{bmatrix} & \label{eqn:secondary-system-1} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
l_1^p \\
l_2^p \\
l_3^p \\
l_4^p \\
\vdots \\
l_m^p
\end{bmatrix}
=
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
\vdots \\
-c_m^p
\end{bmatrix} & \label{eqn:secondary-system-2}
\end{align}

We call the system in Equation \ref{eqn:primary-system}
the ``primary'' system, and the systems in
Equations
\ref{eqn:secondary-system-1} and \ref{eqn:secondary-system-2}
the ``secondary'' systems.
The local part of the solution to the ``global'' tridiagonal system
(Equation \ref{eqn:global-system})
is obtained as a linear combination of
the solutions to the primary and secondary sytems:

\begin{equation}
    \boldsymbol{x}^p = \boldsymbol{x}_r^p + \
        \alpha^p \boldsymbol{u}^p + \beta^p \boldsymbol{l}^p
    \label{eqn:sum-of-systems}
\end{equation}

where the  parameters $\alpha^p$ and $\beta^p$ are obtained by
solving the following ``reduced'' system of equations:

\begin{equation} \label{eqn:reduced-system}
\begin{bmatrix}
l^1_m & -1 \\
-1    & u^2_1 & l^2_1 \\
      & u^2_m & l^2_m & -1 \\
      &       & -1    & u^3_1 & l^3_1 \\
      &       &       & u^3_m & l^3_n  & -1 \\
      &       &       &       & \ddots & \ddots & \ddots \\
      &       &       &       &        & -1     & u^P_1
\end{bmatrix}
\begin{bmatrix}
\beta^1 \\
\alpha^2 \\
\beta^2 \\
\alpha^3 \\
\beta^3 \\
\vdots \\
\alpha^P
\end{bmatrix}
=
\begin{bmatrix}
x_{r,m}^1 \\
x_{r,1}^2 \\
x_{r,m}^2 \\
x_{r,1}^3 \\
x_{r,m}^3 \\
\vdots \\
x_{r,1}^P \\
\end{bmatrix}
\end{equation}

$P = n/m$ is the number of processes.
We note that the reduced system is sized $2P-2$,
and that its assembly requires communication between the processes.
The general solution procedure is then:

\begin{enumerate}
    \item Each process $p$ solves the primary and secondary systems
        to obtain $\boldsymbol{x}_r^p$, $u^p$ and $l^p$
    \item The ``reduced'' system is assembled and solved for
        the coefficients $\alpha^p$ and $\beta^p$
    \item The local solution is computed from Equation \ref{eqn:sum-of-systems}
\end{enumerate}

\subsection{Evaluation of the right-hand side}

\subsection{Evaluation of the right-hand side}

\subsection{Solution of primary Near-Toeplitz tridiagonal systems}

\subsection{Solution of reduced system}

\subsection{Solution of secondary systems}

\section{Results}

\section*{References}

\bibliography{references}
\bibliographystyle{elsarticle-num}
\end{document}

### 2015-10-16 paper-draft/paper.tex
add images, updates to paper draft

@@
 \section{Introduction}
+
+\section{Related work}
 
@@
 
+\section{Evaluation of compact finite differences}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{img/computational-domain.eps}
+\end{center}
+\caption{Computational domain in 3-D}
+\label{fig:computational-domain}
+\end{figure}
+
+The computational domain (Figure \ref{fig:computational-domain})
+is a structured grid comprised of $nx \times ny \times nz$ grid points.
+While evaluating the derivative in, say, the x-direction,
+the computational domain is thought of as
+a collection of \emph{grid lines}---each consisting of
+$nx$ grid points---oriented in the x-direction.
+For each grid line, we may assemble the system in
+Equation \ref{eqn:compact-tridiagonal-system}.
+Solving for the derivatives in the $x$ direction
+at each grid point in the domain then entails
+solving a tridiagonal system for each of the $ny \times nz$ grid lines.
+We note immediately that the left-hand-side (LHS) matrix
+is the same for each of the tridiagonal systems,
+and only the right-hand-side (RHS) vector is different.
+
+\section{Cyclic reduction}
+
+The main step in evaluating compact-finite differences is the
+solution of tridiagonal systems
+for each grid line in the computational domain.
+The standard approach for solving tridiagonal systems is
+the Thomas algorithm.
+However, on parallel architectures such as GPUs,
+it is advantageous to use other algorithms like
+cyclic reduction \cite{Zhang2010FTS}.
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{img/cyclic-reduction.eps}
+\end{center}
+\caption{Cyclic reduction -
+the values that are solved for at each step of backward reduction
+are highlighted}
+\label{fig:cyclic-reduction}
+\end{figure}
+
+The Cyclic Reduction algorithm consists of two phases:
+\emph{forward reduction} and \emph{backward substitution}
+(see Figure \ref{fig:cyclic-reduction}).
+The forward substitution phase begins with a system of size $n$,
+and by expressing every even-indexed equation $i$ as a linear
+combination of equations $i$, $i-1$ and $i+1$, reduces it to a
+system of size $n/2$.
+The process is repeated until a system of 2 equations in 2 unknowns
+is left.
+
+For each even-indexed $i$, we define:
+
+\begin{equation*}
+k_1 = \frac{a_i}{b_{i-1}},
+k_2 = \frac{c_i}{b_{i+1}}
+\end{equation*}
+
+Then, we update the values of $a$, $b$, $c$ and $d$ as:
+
+\begin{align} \label{eqn:forward-reduction}
+& a^{\prime}_i = -a_{i-1}k_1 & \\
+& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 & \\
+& c^{\prime}_i = -c_{i+1}k_2 & \\
+& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 &
+\end{align}
+
+The 2-by-2 system of equations is solved trivially,
+yielding $x_n$ and $x_{n/2}$.
+In the backward substitution phase,
+every odd-indexed unknown $x_i$ is solved for by
+substituting the known values of $x_{i-1}$ and $x_{i+1}$:
+
+\begin{align} \label{eqn:back-substitution}
+x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - c^{\prime}_ix_{i+1}}{b^{\prime}_i}
+\end{align}
+
+At each step, for the last index $i=n$,
+the forward reduction step is instead:
+\begin{align} \label{eqn:forward-reduction-last}
+    & a^{\prime}_n = -a_{n-1}k_1 & \\
+    & b^{\prime}_n = b_n - c_{n-1}k_1 & \\
+    & d^{\prime}_n = d_n - d_{n-1}k_1 &
+\end{align}
+
+And the backward substitution step for $i=1$ is instead:
+\begin{align} \label{eqn:back-substitution-first}
+x_1 = \frac{d^{\prime}_1 - c^{\prime}_1x_{2}}{b^{\prime}_1}
+\end{align}
+
+While cyclic reduction performs far more work per step
+compared to the Thomas algorithm,
+it is able to solve the system in far fewer steps
+on massively parallel architectures.
+As pointed out by \cite{Zhang2010FTS},
+$n$ parallel processors can solve a tridiagonal system
+of size $n$ in $2log(n) - 1$ steps.
+The Thomas algorithm, on the other hand,
+requires $2n$ steps.
+
+\section{Cyclic reduction for near-Toeplitz systems}
+
+We note from
+Equations \ref{eqn:forward-reduction}-\ref{eqn:back-substitution-last}
+that each forward reduction step involves 12 operations.
+As the LHS matrix is constant,
+it is advantageous to pre-compute the coefficients appearing
+in the forward reduction step.
+For a general tridiagonal matrix,
+this requires a substantial amount of storage
+(2$n$ for each of the arrays $a$, $b$ and $c$):
+$\frac{n}{2} + \frac{n}{4} + \frac{n}{8} + ... = n$
+for the reduction steps,
+and an additional $n$ for the original coefficients.
 \section{Distributed compact-finite difference solution}

### 2015-10-18 paper-draft/paper.tex
add more images and update paper

@@
 
+%----------------------------------------------------------------------%
+
 \section{Introduction}
 
-\section{Related work}
+Compact finite difference schemes, developed by
+Lele\cite{lele1992compact} 
+are extremely popular in the computational fluid dynamics community
+for their high order of accuracy at relatively small stencil widths.
+These schemes require the solution of
+linear systems that are
+in general, \emph{banded}, and specifically,
+for several popular schemes, \emph{tridiagonal} in nature.
+Tridiagonal systems also appear in other numerical schemes
+such as
+Alternating Direction Implicit (ADI) methods
+\cite{1955ADI}
+and numerical solutions to one-dimensional differential equations.
+In such CFD codes,
+the evaluation of these tridiagonal systems constitute
+a significant portion of the runtime.
+Recently, there has been considerable interest in
+the CFD community
+in developing tridiagonal solvers
+that exploit the massive parallelism afforded by
+Graphics Processing Units (GPUs)
+to speed up the computations
+\cite{tutkun2012gpu}
+\cite{esfahanian2014efficient}
+\cite{GoSt11CR}.
+The performance of tridiagonal solvers on the GPU is
+dependent on several factors,
+such as
+choice of algorithm and the associated complexity,
+thread utilization,
+use of the available memory heirarchy,
+and synchronization and control costs
+\cite{Zhang2010FTS}.
+The different approaches described in the literature
+seek to make tradeoffs between these factors
+to arrive at a solver that best fits the requirements.
+But to our knowledge, the algorithms proposed so far
+are applicable to general tridiagonal systems,
+and make no attempt to take advantage of the known structure
+of the tridiagonal matrix.
+The tridiagonal systems that appear in several numerical schemes
+are known to have a \emph{near-Toeplitz} structure.
+We describe the implementation of a GPU solver that
+makes use of this fact to
+reduce the number of computations and
+improve memory access.
+The resulting solver is thus able to perform twice as fast as
+the NVIDIA CUSPARSE equivalent
+and $<$add a comparison to the CPU Thomas algorithm$>$.
 
-\section{Compact Finite Difference Schemes}
+This paper is organized as follows:
+
+Section \ref{sec:preliminaries}
+lists related works,
+describes the cyclic reduction algorithm
+and the specific issues in its implementation on GPUs.
+Section \ref{sec:proposed-algorithm}
+describes our proposed solver for near-Toeplitz tridiagonal systems.
+Section \ref{sec:results-single-gpu}
+provides an overview of the performance of our solver,
+and compares it with other implementations.
+
+%----------------------------------------------------------------------%
+\section{Preliminaries} \label{sec:preliminaries}
+
+The classical algorithm for solving tridiagonal systems is
+the Thomas algorithm.
+The algorithm itself exhibits no parallelism,
+and while the GPU can still be exploited to solve
+several independent tridiagonal systems,
+the performance remains poor.
+Zhang et al. \cite{Zhange2010FTS}
+describe the implementation of three different algorithms:
+cyclic reduction, parallel cyclic reduction, and
+recursive doubling,
+along with hybrid variants,
+on GPUs.
+These algorithms perform more work per step than
+the Thomas algorithm,
+but exhibit more fine-grained parallelism
+that allows them to finish in
+far fewer than the $2n$ steps required by the Thomas algorithm.
+In the best case ($n$ parallel processors),
+cyclic reduction requires 
+$2log(n) - 1$ steps,
+and parallel cyclic reduction requires
+$log(n)$ steps
+(while doing more work per step than cyclic reduction).
+This feature makes them a natural choice for
+implementation on GPUs.
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{img/cyclic-reduction.eps}
+\end{center}
+\caption{Cyclic reduction - the highlighted elements
+show the order in which values are solved for}
+\label{fig:cyclic-reduction}
+\end{figure}
+
+The Cyclic Reduction algorithm consists of two phases:
+\emph{forward reduction} and \emph{backward substitution}
+(see Figure \ref{fig:cyclic-reduction}).
+The forward substitution phase begins with a system of size $n$,
+and by expressing every even-indexed equation $i$ as a linear
+combination of equations $i$, $i-1$ and $i+1$, reduces it to a
+system of size $n/2$.
+The process is repeated until a system of 2 equations in 2 unknowns
+is left.
+
+For each even-indexed $i$, we define:
+
+\begin{equation*}
+k_1 = \frac{a_i}{b_{i-1}},
+k_2 = \frac{c_i}{b_{i+1}}
+\end{equation*}
+
+Then, we update the values of $a$, $b$, $c$ and $d$ as:
+
+\begin{align} \label{eqn:forward-reduction}
+& a^{\prime}_i = -a_{i-1}k_1 & \\
+& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 & \\
+& c^{\prime}_i = -c_{i+1}k_2 & \\
+& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 \
+    \label{eqn:forward-reduction-rhs}&
+\end{align}
+
+The 2-by-2 system of equations is solved trivially,
+yielding $x_n$ and $x_{n/2}$.
+In the backward substitution phase,
+every odd-indexed unknown $x_i$ is solved for by
+substituting the known values of $x_{i-1}$ and $x_{i+1}$:
+
+\begin{align} \label{eqn:back-substitution}
+x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - \
+    c^{\prime}_ix_{i+1}}{b^{\prime}_i}
+\end{align}
+
+At each step, for the last index $i=n$,
+the forward reduction step is instead:
+\begin{align} \label{eqn:forward-reduction-last}
+    & a^{\prime}_n = -a_{n-1}k_1 & \\
+    & b^{\prime}_n = b_n - c_{n-1}k_1 & \\
+    & d^{\prime}_n = d_n - d_{n-1}k_1&
+\end{align}
+
+And the backward substitution step for $i=1$ is instead:
+\begin{align} \label{eqn:back-substitution-first}
+x_1 = \frac{d^{\prime}_1 - c^{\prime}_1x_{2}}{b^{\prime}_1}
+\end{align}
+
+%----------------------------------------------------------------------%
+\section{Proposed algorithm} \label{sec:proposed-algorithm}
+
+%----------------------------------------------------------------------%
+\section{Results: single GPU} \label{sec:results-single-gpu}
+
+%----------------------------------------------------------------------%
 
@@
     f^{\prime}_2 \\
-    f^{\prime}_3 \\
+    f^{\prime}_3 
     \vdots \\
@@
 
-\section{Evaluation of compact finite differences}
+%----------------------------------------------------------------------%
 
@@
 
-\section{Cyclic reduction}
+%----------------------------------------------------------------------%
 
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[height=200pt]{img/cyclic-reduction.eps}
-\end{center}
-\caption{Cyclic reduction -
-the values that are solved for at each step of backward reduction
-are highlighted}
-\label{fig:cyclic-reduction}
-\end{figure}
+%----------------------------------------------------------------------%
 
-The Cyclic Reduction algorithm consists of two phases:
-\emph{forward reduction} and \emph{backward substitution}
-(see Figure \ref{fig:cyclic-reduction}).
-The forward substitution phase begins with a system of size $n$,
-and by expressing every even-indexed equation $i$ as a linear
-combination of equations $i$, $i-1$ and $i+1$, reduces it to a
-system of size $n/2$.
-The process is repeated until a system of 2 equations in 2 unknowns
-is left.
-
-For each even-indexed $i$, we define:
-
-\begin{equation*}
-k_1 = \frac{a_i}{b_{i-1}},
-k_2 = \frac{c_i}{b_{i+1}}
-\end{equation*}
-
-Then, we update the values of $a$, $b$, $c$ and $d$ as:
-
-\begin{align} \label{eqn:forward-reduction}
-& a^{\prime}_i = -a_{i-1}k_1 & \\
-& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 & \\
-& c^{\prime}_i = -c_{i+1}k_2 & \\
-& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 &
-\end{align}
-
-The 2-by-2 system of equations is solved trivially,
-yielding $x_n$ and $x_{n/2}$.
-In the backward substitution phase,
-every odd-indexed unknown $x_i$ is solved for by
-substituting the known values of $x_{i-1}$ and $x_{i+1}$:
-
-\begin{align} \label{eqn:back-substitution}
-x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - c^{\prime}_ix_{i+1}}{b^{\prime}_i}
-\end{align}
-
-At each step, for the last index $i=n$,
-the forward reduction step is instead:
-\begin{align} \label{eqn:forward-reduction-last}
-    & a^{\prime}_n = -a_{n-1}k_1 & \\
-    & b^{\prime}_n = b_n - c_{n-1}k_1 & \\
-    & d^{\prime}_n = d_n - d_{n-1}k_1 &
-\end{align}
-
-And the backward substitution step for $i=1$ is instead:
-\begin{align} \label{eqn:back-substitution-first}
-x_1 = \frac{d^{\prime}_1 - c^{\prime}_1x_{2}}{b^{\prime}_1}
-\end{align}
-
-While cyclic reduction performs far more work per step
-compared to the Thomas algorithm,
-it is able to solve the system in far fewer steps
-on massively parallel architectures.
-As pointed out by \cite{Zhang2010FTS},
-$n$ parallel processors can solve a tridiagonal system
-of size $n$ in $2log(n) - 1$ steps.
-The Thomas algorithm, on the other hand,
-requires $2n$ steps.
-
-\section{Cyclic reduction for near-Toeplitz systems}
+%----------------------------------------------------------------------%
 
 We note from
-Equations \ref{eqn:forward-reduction}-\ref{eqn:back-substitution-last}
+Equations \ref{eqn:forward-reduction}-\ref{eqn:forward-reduction-rhs}
 that each forward reduction step involves 12 operations.
@@
 For a general tridiagonal matrix,
-this requires a substantial amount of storage
-(2$n$ for each of the arrays $a$, $b$ and $c$):
-$\frac{n}{2} + \frac{n}{4} + \frac{n}{8} + ... = n$
-for the reduction steps,
-and an additional $n$ for the original coefficients.
-\section{Distributed compact-finite difference solution}
+in addition to the $3n$ original coefficient arrays
+$a$, $b$ and $c$,
+this requires extra storage of 
+$3(\frac{n}{2}+\frac{n}{4}+\frac{n}{8}+...)=3n$
+coefficients.
+This storage cost is greatly amortized
+when solving multiple right-hand sides.
 
@@
 
-\subsection{Evaluation of the right-hand side}
-
-\subsection{Evaluation of the right-hand side}
-
-\subsection{Solution of primary Near-Toeplitz tridiagonal systems}
-
-\subsection{Solution of reduced system}
-
-\subsection{Solution of secondary systems}
-
-\section{Results}
-
 \section*{References}

### 2015-10-20 paper-draft/paper.tex
updates to thesis

@@
 Section \ref{sec:preliminaries}
-lists related works,
-describes the cyclic reduction algorithm
-and the specific issues in its implementation on GPUs.
+describes algorithms for solving tridiagonal systems on GPUs.
+The cyclic reduction algorithm in particular, is discussed in detail,
+and issues with its implementation on GPUs is discussed.
 Section \ref{sec:proposed-algorithm}
@@
 \section{Preliminaries} \label{sec:preliminaries}
+
+\subsection{Solving tridiagonal systems on the GPU}
 
@@
 
-\begin{align} \label{eqn:forward-reduction}
-& a^{\prime}_i = -a_{i-1}k_1 & \\
-& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 & \\
-& c^{\prime}_i = -c_{i+1}k_2 & \\
+\begin{align}
+& a^{\prime}_i = -a_{i-1}k_1 \
+    \label{eqn:forward-reduction-1}& \\
+& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 \
+    \label{eqn:forward-reduction-2}& \\
+& c^{\prime}_i = -c_{i+1}k_2 \
+    \label{eqn:forward-reduction-3}& \\
 & d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 \
-    \label{eqn:forward-reduction-rhs}&
+    \label{eqn:forward-reduction-4}&
 \end{align}
@@
 The 2-by-2 system of equations is solved trivially,
-yielding $x_n$ and $x_{n/2}$.
+yielding $x_n$ and $\frac{x_n}{2}$.
 In the backward substitution phase,
@@
 
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{img/parallel-cyclic-reduction-single-path.eps}
+\end{center}
+\caption{Parallel cyclic reduction - one reduction path highlighted}
+\label{fig:parallel-cyclic-reduction-single-path}
+\end{figure}
+
+Parallel cyclic reduction begins by
+performing the forward reduction step on
+both the odd and even indexed equations.
+This results in two systems of size $\frac{n}{2}$ each.
+Forward reduction is applied again to each of these systems
+resulting in four systems of size $\frac{n}{4}$ each.
+The process is repeated till $\frac{n}{2}$ 2-by-2 systems are produced,
+which can each be solved trivially.
+Figure \ref{fig:parallel-cyclic-reduction-single-path}
+shows PCR with one of these ``reduction paths'' highlighted.
+
+Our proposed solver is a specialization of cyclic reduction,
+so we discuss the pertinent implementation issues on GPUs.
+
+\subsection{GPU architecture}
+
+Writing applications that effectively exploit the GPU
+requires a thorough understanding of the underlying architecure.
+We provide the pertinent details here,
+and refer to \cite{GPUcomputingera} for a more complete picture.
+Our tests are performed on the NVIDIA Tesla K20 and K40 GPUs,
+built on the ``Kepler'' architecture,
+an overview of which is available at \cite{Keplerwhitepaper}.
+
+Applications that use the GPU include special
+pieces of code that are executed on the GPU,
+referred to as \emph{kernels}.
+In C, for example, kernels are written as functions,
+and are called with (almost) the same conventions.
+A kernel is executed in parallel by several threads,
+which are organized into a \emph{grid} of \emph{thread blocks}
+(or just \emph{blocks}).
+As a requirement, thread blocks must be able to run
+independent of each other,
+in series or in parallel.
+
+From the hardware perspective, an NVIDIA GPU may be viewed primarily as
+a collection of so-called Streaming Microprocessors (SMs).
+When a kernel is launched (with a specified number of thread blocks),
+each block is assigned to an SM.
+Threads within a thread block can execute concurrently on the SM,
+and an SM can execute several thread blocks concurrently.
+The number of thread blocks that an SM can execute concurrently
+is limited by a number of factors,
+and maximizing this number is often key to obtaining good performance.
+
+Every thread in every block has access to so-called \emph{global memory},
+which can also be accessed by the CPU (or \emph{host}) via the PCI-e bus.
+Further,
+threads within a thread block have access
+to a common, fast, limited \emph{shared memory}.
+The contents of shared memory are managed by the kernel code,
+and shared memory is often viewed as an explicitly controlled cache.
+Each thread also has private \emph{local memory},
+and access to extremely fast registers.
+Unlike CPUs, the GPU has a large number of registers---a
+thread executing a kernel will
+typically attempt to store the variables defined in registers,
+before spilling over to local memory, which is much slower.
+
+Thread access to global memory is slow,
+and is especially inefficient when
+successive threads in a block
+access locations that are far apart in memory---known
+as \emph{uncoalesced} memory access.
+Shared memory access can be expected to be
+much faster than global memory---however,
+the amount of shared memory per SM is limited
+(48 KiB for current GPUs).
+The amount of shared memory actually used by each block,
+can thus limit the number of blocks that concurrently run on each SM.
+The number of registers per SM is also limited,
+and the use of registers can similarly limit
+the number of concucrrently running blocks.
+Finally, memory transfers between the CPU (host) and GPU (device)
+are extremely expensive, limited by the bandwidth of the PCI-e bus.
+
+\subsection{Cyclic reduction on GPU}
+
 %----------------------------------------------------------------------%
 \section{Proposed algorithm} \label{sec:proposed-algorithm}
+
+We note that solving the equations
+\ref{eqn:forward-reduction-1}-\ref{eqn:forward-reduction-3}
+for each grid line and at each time step of the simulation
+is wasteful.
+It may be benefitial to
+\emph{precompute} and store the coefficients appearing at
+each step of the forward reduction,
+and to load these values from memory when required.
+
+\subsection{General tridiagonal matrix}
+
+For a general tridiagonal matrix,
+in addition to the $3n$ original coefficient arrays
+$a$, $b$ and $c$,
+this requires extra storage of 
+$\frac{n}{2}+\frac{n}{4}+\frac{n}{8}+...=n$
+elements each
+for $a_i$, $b_i$, $c_i$, $k1$ and $k2$
+appearing at the forward reduction steps.
+This additional storage cost is greatly amortized
+when solving the same system for multiple right-hand sides.
+But the additional arrays
+still need to be loaded into shared memory for each thread block.
+This raises the shared memory storage from $4n$ to $9n$,
+which significantly impacts the number of
+concurrently running blocks.
+
+\subsection{Near-Toeplitz matrix}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{img/cyclic-reduction-boundaries.eps}
+\end{center}
+\caption{Indices affected by boundaries in cyclic reduction}
+\label{fig:cyclic-reduction-boundaries}
+\end{figure}
+
+Precomputing the forward reduction coefficients
+becomes especially effective for near-Toeplitz matrices.
+At each reduction step,
+the coefficients $a_i$, $b_i$, $c_i$, $k1_i$ and $k2_i$ 
+are the same for all indices $i$,
+except for indices ``polluted'' by the boundary conditions
+(Figure \ref{fig:cyclic-reduction-boundaries}).
+Ignoring the boundaries, then, we find that the precomputed coefficients
+require $5(log_2(n) - 1)$ storage---far lower than the $5n$ required for the
+general case.
+Further, a careful analysis of the equations
+\ref{eqn:forward-reduction-1}-\ref{eqn:forward-reduction-3}
+reveals that the following extra arrays are required
+to store the boundary information:
+
+\begin{enumerate}
+    \item The values of $b_1$ at each reduction step
+    \item The values of $k_{1,1}$ at each reduction step
+    \item The values of $k_{1,n}$ at each reduction step
+    \item The values of $a_n$ and $b_n$ at the final reduction step
+        (required for the 2-by-2 solve)
+\end{enumerate}
+
+This puts the total storage for precomputed coefficients
+at $8(log_2(n) - 1)+2$.
 
@@
 We note from
-Equations \ref{eqn:forward-reduction}-\ref{eqn:forward-reduction-rhs}
+Equations \ref{eqn:forward-reduction-1}-\ref{eqn:forward-reduction-3}
 that each forward reduction step involves 12 operations.

### 2015-10-21 paper-draft/results/2015-10-15/scripts/cusparse-vs-ours-K20.py
add scripts that generates plots

@@ -1,0 +1,15 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import os
+import sys
+
+prob_sizes = [64, 128, 256, 512]
+cusparse = [0.000520, 0.00335, 0.0286, None]
+ours_global = [0.0008, 0.00376, 0.0213, 0.1456]
+ours_shared = [0.000476, 0.00237, 0.0136, 0.1325]
+
+pt.loglog(prob_sizes, cusparse, '-o', basey=2, basex=2, label='CUSPARSE dgtsvStridedBatch')
+plt.loglog(prob_sizes, ours_global, '-o', basey=2, basex=2, label='Our solver, global memory')
+plt.loglog(prob_sizes, ours_shared, '-o', basey=2, basex=2, label='Our solver, shared memory')
+plt.legend(loc='best')
+plt.savefig('cusparse-vs-ours-K20.png')

### 2015-10-21 paper-draft/results/2015-10-15/scripts/cusparse-vs-ours-K40.py

@@ -1,0 +1,15 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import os
+import sys
+
+prob_sizes = [64, 128, 256, 512]
+cusparse = [0.000446, 0.002864, 0.0244, 0.2178]
+ours_global = [0.00065, 0.0030, 0.0172, 0.1127]
+ours_shared = [0.00039, 0.00193, 0.0111, 0.104]
+
+plt.loglog(prob_sizes, cusparse, '-o', basey=2, basex=2, label='CUSPARSE dgtsvStridedBatch')
+plt.loglog(prob_sizes, ours_global, '-o', basey=2, basex=2, label='Our solver, global memory')
+plt.loglog(prob_sizes, ours_shared, '-o', basey=2, basex=2, label='Our solver, shared memory')
+plt.legend(loc='best')
+plt.savefig('cusparse-vs-ours-K40.png')

### 2015-10-23 paper-draft/paper.tex
reorganize results and make additions to paper draft

@@
 \usepackage{amsmath}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{minted}
+\usepackage{xcolor}
 
@@
 \begin{abstract}
-We present a computation strategy for evaluating
-compact finite differences efficiently on GPU clusters.
-We describe a novel algorithm for solving the
-near-Toeplitz tridiagonal systems associated with
-the compact finite difference schemes.
-Our specialized tridiagonal solver
-takes advantage of the near-Toeplitz nature of the
-tridiagonal system to precompute the coefficients
-appearing in the Cyclic Reduction algorithm.
-This allows us to solve the tridiagonal systems
-up to 2 times faster than with the NVIDIA CUSPARSE
-gtsvStridedBatch routine.
-Additionally, we present a methodology to solve for
-compact finite differences fully on multiple GPUs on a cluster
-without intermediate host-device transfers.
+
 \end{abstract}
@@
 
-Compact finite difference schemes, developed by
-Lele\cite{lele1992compact} 
-are extremely popular in the computational fluid dynamics community
-for their high order of accuracy at relatively small stencil widths.
-These schemes require the solution of
-linear systems that are
-in general, \emph{banded}, and specifically,
-for several popular schemes, \emph{tridiagonal} in nature.
-Tridiagonal systems also appear in other numerical schemes
+The solution of a tridiagonal system
+for several different right hand sides
+appears in several numerical schemes
+in Computational Fluid Dynamics (CFD)
 such as
-Alternating Direction Implicit (ADI) methods
-\cite{1955ADI}
-and numerical solutions to one-dimensional differential equations.
-In such CFD codes,
-the evaluation of these tridiagonal systems constitute
+compact finite differences \cite{lele1992compact},
+Alternating Direction Implicit (ADI) methods \cite{1955ADI}, and
+numerical solutions to one-dimensional differential equations.
+In such applications,
+the evaluation of the tridiagonal systems constitutes
 a significant portion of the runtime.
-Recently, there has been considerable interest in
-the CFD community
-in developing tridiagonal solvers
-that exploit the massive parallelism afforded by
+Recently, there has been considerable interest in the CFD community
+in developing tridiagonal solvers that
+exploit the massive parallelism afforded by
 Graphics Processing Units (GPUs)
@@
 \cite{GoSt11CR}.
-The performance of tridiagonal solvers on the GPU is
-dependent on several factors,
+The performance of tridiagonal solvers on the GPU
+has been shown to depend on several factors,
 such as
@@
 \cite{Zhang2010FTS}.
-The different approaches described in the literature
+Many approaches described in the literature
 seek to make tradeoffs between these factors
 to arrive at a solver that best fits the requirements.
-But to our knowledge, the algorithms proposed so far
+But to our knowledge,
+the algorithms proposed so far
 are applicable to general tridiagonal systems,
-and make no attempt to take advantage of the known structure
-of the tridiagonal matrix.
-The tridiagonal systems that appear in several numerical schemes
-are known to have a \emph{near-Toeplitz} structure.
+and make no attempt to take advantage of
+the known structure of the tridiagonal matrix.
+
+For the relatively common case of structured grids
+with constant spacing in each coordinate direction,
+the above schemes lead to tridiagonal systems with
+\emph{near-Toeplitz} structure
 We describe the implementation of a GPU solver that
-makes use of this fact to
-reduce the number of computations and
-improve memory access.
+uses this information to
+reduce the number of computations
+and access memory more efficiently.
 The resulting solver is thus able to perform twice as fast as
-the NVIDIA CUSPARSE equivalent
-and $<$add a comparison to the CPU Thomas algorithm$>$.
+the equivalent NVIDIA CUSPARSE solver,
+and an Intel MKL solver running with 16 independent cores.
 
 This paper is organized as follows:
-
 Section \ref{sec:preliminaries}
@@
 
-\subsection{Solving tridiagonal systems on the GPU}
+\subsection{Cyclic reduction}
 
-The classical algorithm for solving tridiagonal systems is
-the Thomas algorithm.
-The algorithm itself exhibits no parallelism,
-and while the GPU can still be exploited to solve
-several independent tridiagonal systems,
-the performance remains poor.
-Zhang et al. \cite{Zhange2010FTS}
-describe the implementation of three different algorithms:
-cyclic reduction, parallel cyclic reduction, and
-recursive doubling,
-along with hybrid variants,
-on GPUs.
-These algorithms perform more work per step than
-the Thomas algorithm,
-but exhibit more fine-grained parallelism
-that allows them to finish in
-far fewer than the $2n$ steps required by the Thomas algorithm.
-In the best case ($n$ parallel processors),
-cyclic reduction requires 
-$2log(n) - 1$ steps,
-and parallel cyclic reduction requires
-$log(n)$ steps
-(while doing more work per step than cyclic reduction).
-This feature makes them a natural choice for
-implementation on GPUs.
+\begin{equation} \label{eqn:general-tridiagonal-system}
+\begin{pmatrix}
+     b_1 & c_1  \\
+     a_2 & b_2  &  c_2  \\
+         & a_3  &  b_3 &  c_3  \\
+         &      &  a_4 &  b_4 &  c_4  \\
+         &      &      &      &  \ddots \\
+         &      &      &      &     &  \ddots  \\
+         &      &      &      &     &  a_n  &  b_n
+\end{pmatrix}
+\begin{bmatrix}
+    x_1 \\
+    x_2 \\
+    x_3 \\
+    x_4 \\
+    \vdots \\
+    \vdots \\
+    x_n
+ \end{bmatrix}
+=
+\begin{bmatrix}
+   d_1 \\
+   d_2 \\
+   d_3 \\
+   d_4 \\
+   \vdots \\
+   \vdots \\
+   d_{n}
+\end{bmatrix}
+\end{equation}
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[height=200pt]{img/cyclic-reduction.eps}
-\end{center}
-\caption{Cyclic reduction - the highlighted elements
-show the order in which values are solved for}
-\label{fig:cyclic-reduction}
-\end{figure}
-
-The Cyclic Reduction algorithm consists of two phases:
-\emph{forward reduction} and \emph{backward substitution}
-(see Figure \ref{fig:cyclic-reduction}).
-The forward substitution phase begins with a system of size $n$,
-and by expressing every even-indexed equation $i$ as a linear
-combination of equations $i$, $i-1$ and $i+1$, reduces it to a
-system of size $n/2$.
-The process is repeated until a system of 2 equations in 2 unknowns
-is left.
-
-For each even-indexed $i$, we define:
-
-\begin{equation*}
-k_1 = \frac{a_i}{b_{i-1}},
-k_2 = \frac{c_i}{b_{i+1}}
-\end{equation*}
-
-Then, we update the values of $a$, $b$, $c$ and $d$ as:
+Cyclic reduction is applicable to the solution of
+general tridiagonal systems of the form
+(\ref{eqn:general-tridiagonal-system}).
+The algorithm consists of two phases:
+\emph{forward reduction} and \emph{backward substitution}.
+In the forward reduction phase,
+every even-indexed equation $i$
+is expressed as a
+linear combination of equations $i$, $i-1$ and $i+1$,
+yielding a new tridiagonal system of
+$n/2$ equations in $n/2$ unknowns
+(Equations \ref{eqn:forward-reduction-1} - \ref{eqn:forward-reduction-4}).
+The process is repeated until a system of
+2 equations in 2 unknowns is left.
 
@@
 
+where,
+
+\begin{equation*}
+k_1 = \frac{a_i}{b_{i-1}},
+k_2 = \frac{c_i}{b_{i+1}}
+\end{equation*}
+
 The 2-by-2 system of equations is solved trivially,
-yielding $x_n$ and $\frac{x_n}{2}$.
+yielding $x_n$ and $x_{n/2}$.
 In the backward substitution phase,
 every odd-indexed unknown $x_i$ is solved for by
-substituting the known values of $x_{i-1}$ and $x_{i+1}$:
+substituting the known values of $x_{i-1}$ and $x_{i+1}$
+(Equation \ref{eqn:back-substitution}) .
 
@@
 
-At each step, for the last index $i=n$,
+For the last index $i=n$,
 the forward reduction step is instead:
@@
 
-And the backward substitution step for $i=1$ is instead:
+And for $i=1$, the backward substitution step is instead:
 \begin{align} \label{eqn:back-substitution-first}
@@
 
+Thus, in the best case ($n$ parallel processors),
+cyclic reduction requires 
+$2log(n) - 1$ steps.
+For even moderately large $n$,
+this is significantly smaller than
+the $2n$ steps required by the classical Thomas algorithm.
+This makes cyclic reduction a good fit
+for massively parallel architectures like GPUs.
+
 \begin{figure}[h!]
 \begin{center}
-\includegraphics[height=200pt]{img/parallel-cyclic-reduction-single-path.eps}
+\includegraphics[height=200pt]{img/cyclic-reduction.eps}
 \end{center}
-\caption{Parallel cyclic reduction - one reduction path highlighted}
-\label{fig:parallel-cyclic-reduction-single-path}
+\caption{Cyclic reduction - the highlighted elements
+show the order in which values are solved for}
+\label{fig:cyclic-reduction}
 \end{figure}
 
-Parallel cyclic reduction begins by
-performing the forward reduction step on
-both the odd and even indexed equations.
-This results in two systems of size $\frac{n}{2}$ each.
-Forward reduction is applied again to each of these systems
-resulting in four systems of size $\frac{n}{4}$ each.
-The process is repeated till $\frac{n}{2}$ 2-by-2 systems are produced,
-which can each be solved trivially.
-Figure \ref{fig:parallel-cyclic-reduction-single-path}
-shows PCR with one of these ``reduction paths'' highlighted.
-
-Our proposed solver is a specialization of cyclic reduction,
-so we discuss the pertinent implementation issues on GPUs.
-
-\subsection{GPU architecture}
+\subsection{GPU architecture} \label{sec:gpu-architecture}
 
@@
 The amount of shared memory actually used by each block,
-can thus limit the number of blocks that concurrently run on each SM.
+can thus limit the number of blocks that
+concurrently run on each SM---known as the \emph{occupancy}.
 The number of registers per SM is also limited,
-and the use of registers can similarly limit
-the number of concucrrently running blocks.
+and the use of registers can similarly affect occupancy.
 Finally, memory transfers between the CPU (host) and GPU (device)
@@
 
-\subsection{Cyclic reduction on GPU}
+\subsection{Cyclic reduction implementation on GPU}
+
+As shown in Figure \ref{fig:cyclic-reduction},
+in the forward reduction phase,
+elements are accessed in a \emph{strided} fashion.
+At every subsequent step,
+this stride is doubled, while the number of active threads is halved.
+In the backward substitution phase,
+the strides are halved at each step,
+while the number of active threads is doubled.
+
+Before any computations begin,
+The coefficient arrays $a$, $b$, $c$
+and right-hand sides $d_k$
+for all tridiagonal systems $k$
+reside in global memory---the
+traditional approach is store these right hand sides in
+a contiguous buffer.
+
+\subsubsection{Global memory}
+
+The advantage of a global memory implementation is that
+threads are constantly busy
+
+\subsubsection{Shared memory}
+
+are stored in global memory,
+the large strides in the later stages of forward reduction
+and the earlier stages of backward substitution
+(Figure \ref{fig:cyclic-reduction})
+introduce the problem of uncoalesced memory access.
+
+To overcome this limitation,
+each block loads its system to shared memory,
+and writes the result back to the appropriate location in
+global memory.
+The limited amount of shared memory places restrictions
+on the size of tridiagonal systems that can be solved.
+Shared memory usage is also associated with
+lowered GPU occupancy (Section \ref{sec:gpu-architecture}.
+Further, as described by Zhang et al.,
+the memory access pattern in
+cyclic reduction involves \emph{bank conflicts},
+leading to performance penalties.
+Zhang et al. \cite{Zhang2010FTS} propose a
+\emph{hybrid} solver
+that uses both cyclic reduction and parallal cyclic reduction
+to reduce bank conflicts.
+G{\"o}dekke et al. \cite{GoSt11CR}
+present an approach that is free of bank conflicts,
+but uses additional shared memory.
+Davidson et al. describe the method of \emph{register packing}
+as a means to reduce shared memory usage in
+downsweep-upsweep patterns.
+This approach launches fewer threads,
+with each thread computing several elements,
+and working with registers rather than shared memory.
+The above approaches are applicable
 
@@
 elements each
-for $a_i$, $b_i$, $c_i$, $k1$ and $k2$
+for $a_i$, $b_i$, $c_i$, $k_{1,i}$ and $k_{2,i}$
 appearing at the forward reduction steps.
@@
 At each reduction step,
-the coefficients $a_i$, $b_i$, $c_i$, $k1_i$ and $k2_i$ 
+the coefficients $a_i$, $b_i$, $c_i$, $k_{1,i}$ and $k_{2,i}$ 
 are the same for all indices $i$,
@@
 (Figure \ref{fig:cyclic-reduction-boundaries}).
-Ignoring the boundaries, then, we find that the precomputed coefficients
-require $5(log_2(n) - 1)$ storage---far lower than the $5n$ required for the
-general case.
 Further, a careful analysis of the equations
@@
 reveals that the following extra arrays are required
-to store the boundary information:
+to store the coefficients at the boundaries:
 
@@
 at $8(log_2(n) - 1)+2$.
-
-%----------------------------------------------------------------------%
-\section{Results: single GPU} \label{sec:results-single-gpu}
+For symmetric near-Toeplitz matrices,
+$c_i = a_i$, and $k_{1,i} = k_{2,i}$ away from the boundaries,
+reducing storage to $6(log_2(n) - 1)+2$.
 
@@
 
-For example,
-in a uniformly spaced one-dimensional grid with spacing $dx$,
-if $f_i$ represents the value of
-the function evaluated at the $i$th node,
-the first derivative $f^{\prime}_i$ can be approximated from
-a relation of the form:
-
-\begin{equation}
-\begin{split}
-    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
-    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
-        f^{\prime}_i
-    = 
-    c\frac{f_{i+3} - f_{i-3}}{dx} + \
-    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
-    a\frac{f_{i+1} - f_{i-1}}{dx} + \
-    \hdots
-\end{split}
-\label{eqn:general-compact}
-\end{equation}
-
-The derivatives near the boundaries are approximated using
-one-sided stencil operators of the form:
-
-\begin{equation}
-f^{\prime}_1 + \alpha_1 f^{\prime}_2 = \
-    \frac{1}{dx} (a_1 f_1 + b_1 f_2 + c_1 f_3 + d_1 f_4) 
-\end{equation}
-
-and
-
-\begin{equation}
-    f^{\prime}_n + \alpha_n f^{\prime}_{n-1} = \
-    \frac{1}{dx} (a_n f_n + b_n f_{n-1} + c_n f_{n-2} + d_n f_{n-3}) 
-\end{equation}
-
-As described by Lele \cite{lele1992compact},
-setting $\beta = 0$ in Equation \ref{eqn:general-compact} leads to
-a family of tridiagonal schemes with a single parameter ($\alpha$).
-The coefficient matrix associated with such tridiagonal schemes
-has the general form:
-
-\begin{equation} \label{eqn:compact-tridiagonal-system}
-\begin{bmatrix}
-     1 &  \alpha_1  \\
-     \alpha   &  1   &  \alpha \\
-         &  \alpha   &  1  &  \alpha  \\
-         &      &  \alpha  &  1  &  \alpha  \\
-         &      &     &     &  \ddots \\
-         &      &     &     &     &  \ddots  \\
-         &      &     &     &     &  \alpha_n &  1
-\end{bmatrix}
-\begin{bmatrix}
-    f^{\prime}_1 \\
-    f^{\prime}_2 \\
-    f^{\prime}_3 
-    \vdots \\
-    \vdots \\
-    f^{\prime}_{n-1} \\
-    f^{\prime}_n
- \end{bmatrix}
-=
-\begin{bmatrix}
-   d_1 \\
-   d_2 \\
-   d_3 \\
-   \vdots \\
-   \vdots \\
-   d_{n-1} \\
-   d_{n}
-\end{bmatrix}
-\end{equation}
-
-We note that the system is \emph{near Toeplitz tridiagonal},
-i.e., the left-hand-side matrix is tridiagonal with
-constant coefficients along each diagonal,
-except for the first and last equations.
-
-%----------------------------------------------------------------------%
+\section{Results: single GPU} \label{sec:results-single-gpu}
 
@@
 \begin{center}
-\includegraphics[height=200pt]{img/computational-domain.eps}
+\includegraphics[height=200pt]{fig/compare-solvers-K20.eps}
 \end{center}
-\caption{Computational domain in 3-D}
-\label{fig:computational-domain}
+\label{fig:compare-solvers-K20}
 \end{figure}
 
-The computational domain (Figure \ref{fig:computational-domain})
-is a structured grid comprised of $nx \times ny \times nz$ grid points.
-While evaluating the derivative in, say, the x-direction,
-the computational domain is thought of as
-a collection of \emph{grid lines}---each consisting of
-$nx$ grid points---oriented in the x-direction.
-For each grid line, we may assemble the system in
-Equation \ref{eqn:compact-tridiagonal-system}.
-Solving for the derivatives in the $x$ direction
-at each grid point in the domain then entails
-solving a tridiagonal system for each of the $ny \times nz$ grid lines.
-We note immediately that the left-hand-side (LHS) matrix
-is the same for each of the tridiagonal systems,
-and only the right-hand-side (RHS) vector is different.
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{fig/compare-solvers-K40.eps}
+\end{center}
+\label{fig:compare-solvers-K40}
+\end{figure}
 
-%----------------------------------------------------------------------%
+\definecolor{light-gray}{gray}{0.95}
+\begin{minted}[bgcolor=light-gray,linenos,fontsize=\footnotesize]{C++}
+__global__ void forwardReduction( double *a_d, double *b_d,
+        double *c_d, double *d_d, int n, int stride) {
+    int tix = threadIdx.x;
+    int i = (stride-1) + tix*stride;
+    int offset = blockIdx.x*n;
+    int stride;
+    int m, n;
+    double k1, k2;
 
-The main step in evaluating compact-finite differences is the
-solution of tridiagonal systems
-for each grid line in the computational domain.
-The standard approach for solving tridiagonal systems is
-the Thomas algorithm.
-However, on parallel architectures such as GPUs,
-it is advantageous to use other algorithms like
-cyclic reduction \cite{Zhang2010FTS}.
+    k1 = a_d[i]/b_d[i-stride/2];
+    k2 = c_d[i]/b_d[i+stride/2];
+    a_d[i] = -a_d[i-stride/2]*k1;
+    b_d[i] = b_d[i] - c_d[i-stride/2]*k1 - a_d[i+stride/2]*k2;
+    c_d[i] = -c_d[i+stride/2]*k2;
+    d_d[offset+i] = (d_d[offset+i] - d_d[offset+i-stride/2]*k1 -
+                        d_d[offset+i+stride/2]*k2);
+}
 
-%----------------------------------------------------------------------%
+__global__ void backwardSubstitution( double *a_d, double *b_d,
+        double *c_d, double *d_d, int n, int stride) {
+    int tix = threadIdx.x;
+    int i = (stride/2-1) + tix*stride;
+    int offset = blockIdx.x*n;
+    d_d[offset+i] = (d_d[goffset+i] - a_d[i]*d_d[offset+i-stride/2] -
+                        c_d[i]*d_d[offset+i+stride/2])/b_d[i];
+}
 
-%----------------------------------------------------------------------%
+\end{minted}
 
-We note from
-Equations \ref{eqn:forward-reduction-1}-\ref{eqn:forward-reduction-3}
-that each forward reduction step involves 12 operations.
-As the LHS matrix is constant,
-it is advantageous to pre-compute the coefficients appearing
-in the forward reduction step.
-For a general tridiagonal matrix,
-in addition to the $3n$ original coefficient arrays
-$a$, $b$ and $c$,
-this requires extra storage of 
-$3(\frac{n}{2}+\frac{n}{4}+\frac{n}{8}+...)=3n$
-coefficients.
-This storage cost is greatly amortized
-when solving multiple right-hand sides.
+\begin{minted}[linenos,bgcolor=light-gray,fontsize=\footnotesize]{c++}
 
-We use the algorithm discussed by Mattor et al.
-\cite{mattor1995algorithm}
-to distribute and solve the problem among multiple processes. 
-We present the algorithm to solve a general tridiagonal system,
-and in subsequent sections discuss its adaptation to solving
-near-Teoplitz tridiagonal systems on GPUs.
+__global__ cyclicReduction(double *a_d, double *b_d, double *c_d,
+    double *d_d, int n) {
 
-Given a tridiagonal system with  $n$ equations,
-to be solved by $P$ processes:
+    __shared__ double a_l[n];
+    __shared__ double b_l[n];
+    __shared__ double c_l[n];
+    __shared__ double d_l[n];
+    int i = threadIdx.x;
+    int offset = blockIdx.x*n;
+    int stride;
 
-\begin{align}
-& \begin{bmatrix}
-b_1^p & c_1^p \\
-a_2^p & b_2^p & c_2^p \\
-      & a_3^p & b_3^p & c_3^p \\
-      &       & a_4^p & b_4^p & c_4^p \\
-      &       &       &       &  \ddots & c_{m-1}^p\\
-      &       &       &       &     a_{m}^p  & b_{m}^p
-\end{bmatrix}
-\begin{bmatrix}
-x_{r,1}^p \\
-x_{r,2}^p \\
-x_{r,3}^p \\
-x_{r,4}^p \\
-\vdots \\
-x_{r,m}^p
-\end{bmatrix}
-=
-\begin{bmatrix}
-r_1^p \\
-r_2^p \\
-r_3^p \\
-r_4^p \\
-\vdots \\
-r_m^p
-\end{bmatrix} & \label{eqn:global-system} 
-\end{align}
+    /* Load coefficients and RHS into shared memory */
+    a_l[i] = a_d[offset+i]; b_l[i] = b_d[offset+i];
+    c_l[i] = c_d[offset+i]; d_l[i] = d_d[offset+i];
 
-each process $p$ proceeds by solving the following three ``local''
-tridiagonal systems:
+    /* Forward reduction */
+    stride = 1;
+    for (int k=0; k<log2(nx)-1; k++) {
+        if (i < nx/stride) {
+            k1 = a_l[i]/b_l[i-stride/2];
+            k2 = c_l[i]/b_l[i+stride/2];
+            a_l[i] = -a_l[i-stride/2]*k1;
+            b_l[i] = b_l[i] - c_l[i-stride/2]*k1 - a_l[i+stride/2]*k2;
+            c_l[i] = -c_l[i+stride/2]*k2;
+            d_l[i] = d_l[i] - d_l[i-stride/2]*k1 - d_l[i+stride/2]*k2;
+        }
+        __syncthreads();
+    }
 
-\begin{align}
-& \begin{bmatrix}
-b_1^p & c_1^p \\
-a_2^p & b_2^p & c_2^p \\
-      & a_3^p & b_3^p & c_3^p \\
-      &       & a_4^p & b_4^p & c_4^p \\
-      &       &       &       &  \ddots & c_{m-1}^p\\
-      &       &       &       &     a_{m}^p  & b_{m}^p
-\end{bmatrix}
-\begin{bmatrix}
-x_{r,1}^p \\
-x_{r,2}^p \\
-x_{r,3}^p \\
-x_{r,4}^p \\
-\vdots \\
-x_{r,m}^p
-\end{bmatrix}
-=
-\begin{bmatrix}
-r_1^p \\
-r_2^p \\
-r_3^p \\
-r_4^p \\
-\vdots \\
-r_m^p
-\end{bmatrix} & \label{eqn:primary-system} \\
-%
-%
-%
-& \begin{bmatrix}
-b_1^p & c_1^p \\
-a_2^p & b_2^p & c_2^p \\
-      & a_3^p & b_3^p & c_3^p \\
-      &       & a_4^p & b_4^p & c_4^p \\
-      &       &       &       &  \ddots & c_{m-1}^p\\
-      &       &       &       &     a_{m}^p  & b_{m}^p
-\end{bmatrix}
-\begin{bmatrix}
-u_1^p \\
-u_2^p \\
-u_3^p \\
-u_4^p \\
-\vdots \\
-u_m^p
-\end{bmatrix}
-=
-\begin{bmatrix}
--a_1^p \\
-0 \\
-0 \\
-0 \\
-\vdots \\
-0
-\end{bmatrix} & \label{eqn:secondary-system-1} \\
-%
-%
-%
-& \begin{bmatrix}
-b_1^p & c_1^p \\
-a_2^p & b_2^p & c_2^p \\
-      & a_3^p & b_3^p & c_3^p \\
-      &       & a_4^p & b_4^p & c_4^p \\
-      &       &       &       &  \ddots & c_{m-1}^p\\
-      &       &       &       &     a_{m}^p  & b_{m}^p
-\end{bmatrix}
-\begin{bmatrix}
-l_1^p \\
-l_2^p \\
-l_3^p \\
-l_4^p \\
-\vdots \\
-l_m^p
-\end{bmatrix}
-=
-\begin{bmatrix}
-0 \\
-0 \\
-0 \\
-0 \\
-\vdots \\
--c_m^p
-\end{bmatrix} & \label{eqn:secondary-system-2}
-\end{align}
+    if (i == 0) {
+        /* Solve the 2-by-2 system for d_l[n-1] & d_l[n/2-1] */
+    }
 
-We call the system in Equation \ref{eqn:primary-system}
-the ``primary'' system, and the systems in
-Equations
-\ref{eqn:secondary-system-1} and \ref{eqn:secondary-system-2}
-the ``secondary'' systems.
-The local part of the solution to the ``global'' tridiagonal system
-(Equation \ref{eqn:global-system})
-is obtained as a linear combination of
-the solutions to the primary and secondary sytems:
+    /* Backward substitution */
+    stride = n;
+    for (int k=0; k<log2(nx)-1; k++) {
+        stride = stride/2;
+        if (lix < nx/stride){
+            d_l[i] = (d_l[i] - a_l[i]*d_l[i-stride/2] -
+                        c_l[i]*d_l[i+stride/2])/b_l[i];
+        }
+        __syncthreads();
+    }
 
-\begin{equation}
-    \boldsymbol{x}^p = \boldsymbol{x}_r^p + \
-        \alpha^p \boldsymbol{u}^p + \beta^p \boldsymbol{l}^p
-    \label{eqn:sum-of-systems}
-\end{equation}
+    /* Load the solution from shared memory to global memory */
+    d_d[offset+i] = d_l[i];
+}
 
-where the  parameters $\alpha^p$ and $\beta^p$ are obtained by
-solving the following ``reduced'' system of equations:
-
-\begin{equation} \label{eqn:reduced-system}
-\begin{bmatrix}
-l^1_m & -1 \\
--1    & u^2_1 & l^2_1 \\
-      & u^2_m & l^2_m & -1 \\
-      &       & -1    & u^3_1 & l^3_1 \\
-      &       &       & u^3_m & l^3_n  & -1 \\
-      &       &       &       & \ddots & \ddots & \ddots \\
-      &       &       &       &        & -1     & u^P_1
-\end{bmatrix}
-\begin{bmatrix}
-\beta^1 \\
-\alpha^2 \\
-\beta^2 \\
-\alpha^3 \\
-\beta^3 \\
-\vdots \\
-\alpha^P
-\end{bmatrix}
-=
-\begin{bmatrix}
-x_{r,m}^1 \\
-x_{r,1}^2 \\
-x_{r,m}^2 \\
-x_{r,1}^3 \\
-x_{r,m}^3 \\
-\vdots \\
-x_{r,1}^P \\
-\end{bmatrix}
-\end{equation}
-
-$P = n/m$ is the number of processes.
-We note that the reduced system is sized $2P-2$,
-and that its assembly requires communication between the processes.
-The general solution procedure is then:
-
-\begin{enumerate}
-    \item Each process $p$ solves the primary and secondary systems
-        to obtain $\boldsymbol{x}_r^p$, $u^p$ and $l^p$
-    \item The ``reduced'' system is assembled and solved for
-        the coefficients $\alpha^p$ and $\beta^p$
-    \item The local solution is computed from Equation \ref{eqn:sum-of-systems}
-\end{enumerate}
+\end{minted}
 

### 2015-10-23 paper-draft/sandbox.tex
reorganize results and make additions to paper draft

%----------------------------------------------------------------------%

For example,
in a uniformly spaced one-dimensional grid with spacing $dx$,
if $f_i$ represents the value of
the function evaluated at the $i$th node,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:

\begin{equation}
\begin{split}
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
        f^{\prime}_i
    = 
    c\frac{f_{i+3} - f_{i-3}}{dx} + \
    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
    a\frac{f_{i+1} - f_{i-1}}{dx} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}

The derivatives near the boundaries are approximated using
one-sided stencil operators of the form:

\begin{equation}
f^{\prime}_1 + \alpha_1 f^{\prime}_2 = \
    \frac{1}{dx} (a_1 f_1 + b_1 f_2 + c_1 f_3 + d_1 f_4) 
\end{equation}

and

\begin{equation}
    f^{\prime}_n + \alpha_n f^{\prime}_{n-1} = \
    \frac{1}{dx} (a_n f_n + b_n f_{n-1} + c_n f_{n-2} + d_n f_{n-3}) 
\end{equation}

As described by Lele \cite{lele1992compact},
setting $\beta = 0$ in Equation \ref{eqn:general-compact} leads to
a family of tridiagonal schemes with a single parameter ($\alpha$).
The coefficient matrix associated with such tridiagonal schemes
has the general form:

\begin{equation} \label{eqn:compact-tridiagonal-system}
\begin{bmatrix}
     1 &  \alpha_1  \\
     \alpha   &  1   &  \alpha \\
         &  \alpha   &  1  &  \alpha  \\
         &      &  \alpha  &  1  &  \alpha  \\
         &      &     &     &  \ddots \\
         &      &     &     &     &  \ddots  \\
         &      &     &     &     &  \alpha_n &  1
\end{bmatrix}
\begin{bmatrix}
    f^{\prime}_1 \\
    f^{\prime}_2 \\
    f^{\prime}_3 
    \vdots \\
    \vdots \\
    f^{\prime}_{n-1} \\
    f^{\prime}_n
 \end{bmatrix}
=
\begin{bmatrix}
   d_1 \\
   d_2 \\
   d_3 \\
   \vdots \\
   \vdots \\
   d_{n-1} \\
   d_{n}
\end{bmatrix}
\end{equation}

We note that the system is \emph{near Toeplitz tridiagonal},
i.e., the left-hand-side matrix is tridiagonal with
constant coefficients along each diagonal,
except for the first and last equations.

%----------------------------------------------------------------------%

\begin{figure}[h!]
\begin{center}
\includegraphics[height=200pt]{img/computational-domain.eps}
\end{center}
\caption{Computational domain in 3-D}
\label{fig:computational-domain}
\end{figure}

The computational domain (Figure \ref{fig:computational-domain})
is a structured grid comprised of $nx \times ny \times nz$ grid points.
While evaluating the derivative in, say, the x-direction,
the computational domain is thought of as
a collection of \emph{grid lines}---each consisting of
$nx$ grid points---oriented in the x-direction.
For each grid line, we may assemble the system in
Equation \ref{eqn:compact-tridiagonal-system}.
Solving for the derivatives in the $x$ direction
at each grid point in the domain then entails
solving a tridiagonal system for each of the $ny \times nz$ grid lines.
We note immediately that the left-hand-side (LHS) matrix
is the same for each of the tridiagonal systems,
and only the right-hand-side (RHS) vector is different.

%----------------------------------------------------------------------%

The main step in evaluating compact-finite differences is the
solution of tridiagonal systems
for each grid line in the computational domain.
The standard approach for solving tridiagonal systems is
the Thomas algorithm.
However, on parallel architectures such as GPUs,
it is advantageous to use other algorithms like
cyclic reduction \cite{Zhang2010FTS}.

%----------------------------------------------------------------------%

%----------------------------------------------------------------------%

We note from
Equations \ref{eqn:forward-reduction-1}-\ref{eqn:forward-reduction-3}
that each forward reduction step involves 12 operations.
As the LHS matrix is constant,
it is advantageous to pre-compute the coefficients appearing
in the forward reduction step.
For a general tridiagonal matrix,
in addition to the $3n$ original coefficient arrays
$a$, $b$ and $c$,
this requires extra storage of 
$3(\frac{n}{2}+\frac{n}{4}+\frac{n}{8}+...)=3n$
coefficients.
This storage cost is greatly amortized
when solving multiple right-hand sides.

We use the algorithm discussed by Mattor et al.
\cite{mattor1995algorithm}
to distribute and solve the problem among multiple processes. 
We present the algorithm to solve a general tridiagonal system,
and in subsequent sections discuss its adaptation to solving
near-Teoplitz tridiagonal systems on GPUs.

Given a tridiagonal system with  $n$ equations,
to be solved by $P$ processes:

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
x_{r,1}^p \\
x_{r,2}^p \\
x_{r,3}^p \\
x_{r,4}^p \\
\vdots \\
x_{r,m}^p
\end{bmatrix}
=
\begin{bmatrix}
r_1^p \\
r_2^p \\
r_3^p \\
r_4^p \\
\vdots \\
r_m^p
\end{bmatrix} & \label{eqn:global-system} 
\end{align}

each process $p$ proceeds by solving the following three ``local''
tridiagonal systems:

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
x_{r,1}^p \\
x_{r,2}^p \\
x_{r,3}^p \\
x_{r,4}^p \\
\vdots \\
x_{r,m}^p
\end{bmatrix}
=
\begin{bmatrix}
r_1^p \\
r_2^p \\
r_3^p \\
r_4^p \\
\vdots \\
r_m^p
\end{bmatrix} & \label{eqn:primary-system} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
u_1^p \\
u_2^p \\
u_3^p \\
u_4^p \\
\vdots \\
u_m^p
\end{bmatrix}
=
\begin{bmatrix}
-a_1^p \\
0 \\
0 \\
0 \\
\vdots \\
0
\end{bmatrix} & \label{eqn:secondary-system-1} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
l_1^p \\
l_2^p \\
l_3^p \\
l_4^p \\
\vdots \\
l_m^p
\end{bmatrix}
=
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
\vdots \\
-c_m^p
\end{bmatrix} & \label{eqn:secondary-system-2}
\end{align}

We call the system in Equation \ref{eqn:primary-system}
the ``primary'' system, and the systems in
Equations
\ref{eqn:secondary-system-1} and \ref{eqn:secondary-system-2}
the ``secondary'' systems.
The local part of the solution to the ``global'' tridiagonal system
(Equation \ref{eqn:global-system})
is obtained as a linear combination of
the solutions to the primary and secondary sytems:

\begin{equation}
    \boldsymbol{x}^p = \boldsymbol{x}_r^p + \
        \alpha^p \boldsymbol{u}^p + \beta^p \boldsymbol{l}^p
    \label{eqn:sum-of-systems}
\end{equation}

where the  parameters $\alpha^p$ and $\beta^p$ are obtained by
solving the following ``reduced'' system of equations:

\begin{equation} \label{eqn:reduced-system}
\begin{bmatrix}
l^1_m & -1 \\
-1    & u^2_1 & l^2_1 \\
      & u^2_m & l^2_m & -1 \\
      &       & -1    & u^3_1 & l^3_1 \\
      &       &       & u^3_m & l^3_n  & -1 \\
      &       &       &       & \ddots & \ddots & \ddots \\
      &       &       &       &        & -1     & u^P_1
\end{bmatrix}
\begin{bmatrix}
\beta^1 \\
\alpha^2 \\
\beta^2 \\
\alpha^3 \\
\beta^3 \\
\vdots \\
\alpha^P
\end{bmatrix}
=
\begin{bmatrix}
x_{r,m}^1 \\
x_{r,1}^2 \\
x_{r,m}^2 \\
x_{r,1}^3 \\
x_{r,m}^3 \\
\vdots \\
x_{r,1}^P \\
\end{bmatrix}
\end{equation}

$P = n/m$ is the number of processes.
We note that the reduced system is sized $2P-2$,
and that its assembly requires communication between the processes.
The general solution procedure is then:

\begin{enumerate}
    \item Each process $p$ solves the primary and secondary systems
        to obtain $\boldsymbol{x}_r^p$, $u^p$ and $l^p$
    \item The ``reduced'' system is assembled and solved for
        the coefficients $\alpha^p$ and $\beta^p$
    \item The local solution is computed from Equation \ref{eqn:sum-of-systems}
\end{enumerate}

### 2015-10-23 paper-draft/results/2015-10-15/scripts/cusparse-vs-ours-K20.py

@@ -1,15 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-import os
-import sys
-
-prob_sizes = [64, 128, 256, 512]
-cusparse = [0.000520, 0.00335, 0.0286, None]
-ours_global = [0.0008, 0.00376, 0.0213, 0.1456]
-ours_shared = [0.000476, 0.00237, 0.0136, 0.1325]
-
-pt.loglog(prob_sizes, cusparse, '-o', basey=2, basex=2, label='CUSPARSE dgtsvStridedBatch')
-plt.loglog(prob_sizes, ours_global, '-o', basey=2, basex=2, label='Our solver, global memory')
-plt.loglog(prob_sizes, ours_shared, '-o', basey=2, basex=2, label='Our solver, shared memory')
-plt.legend(loc='best')
-plt.savefig('cusparse-vs-ours-K20.png')

### 2015-10-23 paper-draft/results/2015-10-15/scripts/cusparse-vs-ours-K40.py

@@ -1,15 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-import os
-import sys
-
-prob_sizes = [64, 128, 256, 512]
-cusparse = [0.000446, 0.002864, 0.0244, 0.2178]
-ours_global = [0.00065, 0.0030, 0.0172, 0.1127]
-ours_shared = [0.00039, 0.00193, 0.0111, 0.104]
-
-plt.loglog(prob_sizes, cusparse, '-o', basey=2, basex=2, label='CUSPARSE dgtsvStridedBatch')
-plt.loglog(prob_sizes, ours_global, '-o', basey=2, basex=2, label='Our solver, global memory')
-plt.loglog(prob_sizes, ours_shared, '-o', basey=2, basex=2, label='Our solver, shared memory')
-plt.legend(loc='best')
-plt.savefig('cusparse-vs-ours-K40.png')

### 2015-10-23 paper-draft/results/scripts/compare-solvers-K20.py

@@ -1,0 +1,53 @@
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import os
+import sys
+
+plt.style.use('bmh')
+prob_sizes = [64, 128, 256, 512]
+intel_mkl_2 = np.array([0.97, 5.884, 75.16, 703.38])
+intel_mkl_4 = np.array([0.836, 7.10, 50.24, 392.7])
+intel_mkl_16 = np.array([1.522, 2.84, 19.64, 221.0])
+cusparse = np.array([0.000520, 0.00335, 0.0286, np.nan])*1000
+ours_global = np.array([0.0008, 0.00376, 0.0213, 0.1456])*1000
+ours_shared = np.array([0.000476, 0.00237, 0.0136, 0.1325])*1000
+
+ind = np.arange(4)*2
+fig, ax = plt.subplots()
+width = 0.2
+mkl_2_bars = ax.bar(ind, intel_mkl_2, width, color='y')
+mkl_4_bars = ax.bar(ind+width, intel_mkl_4, width, color='b')
+mkl_16_bars = ax.bar(ind+2*width, intel_mkl_16, width, color='c')
+cusparse_bars = ax.bar(ind+3*width, cusparse, width, color='m')
+global_bars = ax.bar(ind+4*width, ours_global, width, color='r')
+shared_bars = ax.bar(ind+5*width, ours_shared, width, color='g')
+
+ax.set_yscale('log', basey=2)
+ax.get_yaxis().set_major_formatter(
+        matplotlib.ticker.ScalarFormatter())
+
+ax.yaxis.set_major_formatter(matplotlib.ticker.FormatStrFormatter('%.1f'))
+ax.set_xlabel('Problem size')
+ax.set_ylabel('Time (ms)')
+ax.set_xticks(ind+width)
+ax.set_xticklabels( ('$64^3$', '$128^3$', '$256^3$', '$512^3$') )
+ax.tick_params(labelright=True)
+ax.set_title('Performance of solvers on Tesla K20')
+ax.grid(True)
+leg = ax.legend((mkl_2_bars[0],
+            mkl_4_bars[0],
+            mkl_16_bars[0],
+            cusparse_bars[0],
+            global_bars[0],
+            shared_bars[0]),
+            ('Intel MKL 2 cores',
+            'Intel MKL 4 cores',
+            'Intel MKL 16 cores',
+            'CUSPARSE',
+            'Our solver (global memory)',
[…]

### 2015-10-23 paper-draft/results/scripts/compare-solvers-K40.py

@@ -1,0 +1,41 @@
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import os
+import sys
+
+
+plt.style.use('bmh')
+prob_sizes = [64, 128, 256, 512]
+intel_mkl = np.array([1.13, 8.90, 77.67, 654])
+cusparse = np.array([0.000446, 0.002864, 0.0244, 0.2178])*1000
+ours_global = np.array([0.00065, 0.0030, 0.0172, 0.1127])*1000
+ours_shared = np.array([0.00039, 0.00193, 0.0111, 0.104])*1000
+
+ind = np.arange(4)*2
+fig, ax = plt.subplots()
+width = 0.25
+mkl_bars = ax.bar(ind, intel_mkl, width, color='r')
+cusparse_bars = ax.bar(ind+width, cusparse, width, color='b')
+global_bars = ax.bar(ind+2*width, ours_global, width, color='y')
+shared_bars = ax.bar(ind+3*width, ours_shared, width, color='g')
+
+ax.set_yscale('log')
+ax.get_yaxis().set_major_formatter(
+        matplotlib.ticker.ScalarFormatter())
+ax.set_xlabel('Problem size')
+ax.set_ylabel('Time (ms)')
+ax.set_xticks(ind+width)
+ax.set_xticklabels( ('$64^3$', '$128^3$', '$256^3$', '$512^3$') )
+ax.tick_params(labelright=True)
+ax.set_title('Performance of solvers on Tesla K40')
+ax.grid(True)
+leg = ax.legend( (mkl_bars[0],
+            cusparse_bars[0],
+            global_bars[0],
+            shared_bars[0]),
+            ('Intel MKL', 'CUSPARSE',
+             'Our solver (global memory)',
+             'Our solver (shared memory)'), loc='best')
+leg.draw_frame(False)
+plt.savefig('compare-solvers-K40.eps')

### 2015-10-26 paper-draft/paper.tex
add more images; update paper draft

@@
 
-\usepackage{amsmath}
+\usepackage{amsmath,bm}
 \usepackage{algorithm}
@@
 \author{Ashwin Srinath}
-\title{A}
+\title{A fast GPU solver for near-Toeplitz tridiagonal systems}
 \maketitle
@@
 
-The solution of a tridiagonal system
-for several different right hand sides
+In this paper, we consider the solution of
+linear systems of the form $A\bm{x} = \bm{d}$,
+where $A$ has the following structure
+
+\begin{equation} \label{eqn:toeplitz-matrix}
+A = 
+\begin{pmatrix}
+     b_1 & c_1  \\
+     a_0 & b_0  &  c_0  \\
+         & a_0  &  b_0 &  c_0  \\
+         &      &  a_0 &  b_0 &  c    \\
+         &      &      &      &  \ddots \\
+         &      &      &      &     &  \ddots  \\
+         &      &      &      &     &  a_n  &  b_n
+\end{pmatrix}
+\end{equation}
+
+Where, in general
+\begin{align*}
+    & a_0 \neq a_n & \\
+    & b_1 \neq b_0 \neq b_n &\\
+    & c_1 \neq c_0 &
+\end{align*}
+
+We refer to matrices with the specific structure of $A$
+as \emph{near-Toeplitz tridiagonal matrices},
+and the linear system $A\bm{x}=\bm{d}$ as \emph{near-Toeplitz tridiagonal systems}.
+
+The solution of such tridiagonal systems
 appears in several numerical schemes
@@
 Recently, there has been considerable interest in the CFD community
-in developing tridiagonal solvers that
+in developing general tridiagonal solvers that
 exploit the massive parallelism afforded by
@@
 thread utilization,
-use of the available memory heirarchy,
+use of the available memory hierarchy,
 and synchronization and control costs
@@
 Many approaches described in the literature
-seek to make tradeoffs between these factors
+seek to make trade-offs between these factors
 to arrive at a solver that best fits the requirements.
@@
 the known structure of the tridiagonal matrix.
-
-For the relatively common case of structured grids
-with constant spacing in each coordinate direction,
-the above schemes lead to tridiagonal systems with
-\emph{near-Toeplitz} structure
 We describe the implementation of a GPU solver that
@@
 reduce the number of computations
-and access memory more efficiently.
+and memory accesses.
 The resulting solver is thus able to perform twice as fast as
@@
 
-\subsection{Cyclic reduction}
+\subsection{Cyclic reduction} \label{subsec:cyclic-reduction}
 
@@
 Cyclic reduction is applicable to the solution of
-general tridiagonal systems of the form
-(\ref{eqn:general-tridiagonal-system}).
+general linear systems of the form $Ax = d$,
+where $A$ is a tridiagonal matrix with diagonals
+$a$, $b$ and $c$
+(Equation \ref{eqn:general-tridiagonal-system}).
 The algorithm consists of two phases:
@@
 
-\begin{equation*}
-k_1 = \frac{a_i}{b_{i-1}},
-k_2 = \frac{c_i}{b_{i+1}}
-\end{equation*}
+\begin{align}
+& k_1 = \frac{a_i}{b_{i-1}} \label{eqn:k1-update}& \\
+& k_2 = \frac{c_i}{b_{i+1}} \label{eqn:k2-update}&
+\end{align}
 
@@
 substituting the known values of $x_{i-1}$ and $x_{i+1}$
-(Equation \ref{eqn:back-substitution}) .
+(Equation \ref{eqn:backward-substitution}) .
 
-\begin{align} \label{eqn:back-substitution}
+\begin{align} \label{eqn:backward-substitution}
 x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - \
@@
 And for $i=1$, the backward substitution step is instead:
-\begin{align} \label{eqn:back-substitution-first}
+\begin{align} \label{eqn:backward-substitution-first}
 x_1 = \frac{d^{\prime}_1 - c^{\prime}_1x_{2}}{b^{\prime}_1}
@@
 \begin{center}
-\includegraphics[height=200pt]{img/cyclic-reduction.eps}
+\includegraphics[height=250pt]{img/cyclic-reduction.eps}
 \end{center}
-\caption{Cyclic reduction - the highlighted elements
-show the order in which values are solved for}
+\caption{Cyclic reduction}
 \label{fig:cyclic-reduction}
@@
 
-\subsection{GPU architecture} \label{sec:gpu-architecture}
+\subsection{GPU architecture} \label{subsec:gpu-architecture}
 
 Writing applications that effectively exploit the GPU
-requires a thorough understanding of the underlying architecure.
+requires a thorough understanding of the underlying architecture.
 We provide the pertinent details here,
@@
 and are called with (almost) the same conventions.
-A kernel is executed in parallel by several threads,
+A kernel is executed in parallel by several lightweight threads,
 which are organized into a \emph{grid} of \emph{thread blocks}
@@
 
-As shown in Figure \ref{fig:cyclic-reduction},
-in the forward reduction phase,
-elements are accessed in a \emph{strided} fashion.
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{img/forward-reduction-step.eps}
+\end{center}
+\caption{Updating $\bm{b}$ in the first forward reduction step}
+\label{fig:forward-reduction-step}
+\end{figure}
+
+As described in Section \ref{subsec:gpu-architecture},
+GPU instructions are executed concurrently
+by several threads and
+threads are organized into thread blocks.
+In most GPU implementations of cyclic reduction,
+blocks are assigned to tridiagonal systems
+(when solving multiple systems),
+and threads within a block 
+are assigned to equations, or \emph{indices}.
+(Figure \ref{fig:gpu-mapping})
+During the forward reduction phase,
+the threads assigned to each even index $i$
+compute the coefficients
+$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$.
+In practice, the coefficients are updated \emph{in-place}
+(Figure \ref{fig:forward-reduction-step} shows the updates
+to the coefficient array $\bm{b}$ in the first forward reduction step).
+
+In each step of forward reduction,
+a thread accesses values from the coefficient arrays
+$\bm{a}$, $\bm{b}$ and $\bm{c}$ in a \emph{strided} fashion.
 At every subsequent step,
@@
 while the number of active threads is doubled.
-
-Before any computations begin,
-The coefficient arrays $a$, $b$, $c$
-and right-hand sides $d_k$
-for all tridiagonal systems $k$
-reside in global memory---the
-traditional approach is store these right hand sides in
-a contiguous buffer.
-
-\subsubsection{Global memory}
-
-The advantage of a global memory implementation is that
-threads are constantly busy
-
-\subsubsection{Shared memory}
-
-are stored in global memory,
-the large strides in the later stages of forward reduction
-and the earlier stages of backward substitution
-(Figure \ref{fig:cyclic-reduction})
-introduce the problem of uncoalesced memory access.
-
-To overcome this limitation,
-each block loads its system to shared memory,
-and writes the result back to the appropriate location in
-global memory.
-The limited amount of shared memory places restrictions
+The long-strided memory accesses towards the end
+of the forward reduction
+and the beginning of backward substitution
+motivates the use of shared memory---strided global memory access is,
+in general, associated with performance penalties.
+But the limited amount of shared memory places restrictions
 on the size of tridiagonal systems that can be solved.
-Shared memory usage is also associated with
-lowered GPU occupancy (Section \ref{sec:gpu-architecture}.
-Further, as described by Zhang et al.,
+Further,
 the memory access pattern in
-cyclic reduction involves \emph{bank conflicts},
-leading to performance penalties.
+cyclic reduction leads to \emph{bank conflicts}
+\cite{Zhang2010FTS}.
 Zhang et al. \cite{Zhang2010FTS} propose a
 \emph{hybrid} solver
-that uses both cyclic reduction and parallal cyclic reduction
+that uses both cyclic reduction and parallel cyclic reduction
 to reduce bank conflicts.
 G{\"o}dekke et al. \cite{GoSt11CR}
-present an approach that is free of bank conflicts,
-but uses additional shared memory.
-Davidson et al. describe the method of \emph{register packing}
-as a means to reduce shared memory usage in
-downsweep-upsweep patterns.
-This approach launches fewer threads,
-with each thread computing several elements,
-and working with registers rather than shared memory.
-The above approaches are applicable
+use a method of separately storing
+the even and odd indexed equations
+to arrive at a bank-conflict free solver
+at the cost of additional shared memory usage.
+Shared memory usage is also associated with
+lowered GPU occupancy (Section \ref{subsec:gpu-architecture}).
+Davidson et al. \cite{davidson2011register}
+describe the method of
+\emph{register packing}---performing more computations
+on \emph{registers}, rather than shared memory---as
+a means to reduce shared memory usage in cyclic reduction.
+More recently, Esfahanian et al. \cite{esfahanian2014efficient}
+describe an implementation of cyclic reduction
+that works entirely in global memory,
+using additional kernels to reorder data so as to
+coalesce memory access.
 
@@
 
-We note that solving the equations
-\ref{eqn:forward-reduction-1}-\ref{eqn:forward-reduction-3}
-for each grid line and at each time step of the simulation
-is wasteful.
-It may be benefitial to
-\emph{precompute} and store the coefficients appearing at
-each step of the forward reduction,
-and to load these values from memory when required.
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{img/gpu-mapping.eps}
+\end{center}
+\label{fig:gpu-mapping}
+\caption{Mapping work to blocks and threads:
+systems are mapped to blocks and
+indices are mapped to individual threads}
+\end{figure}
 
-\subsection{General tridiagonal matrix}
+The method is based on precomputing elements 
+appearing in the forward reduction phase
+so that they can be reused for multiple right hand sides.
+While this can be done for general tridiagonal systems,
+the applicability to near-Toeplitz systems is
+of particular interest due to the lower storage costs.
+In Figure \ref{fig:forward-reduction-step},
+we list the updates to
+$b_i$, for $i=2, 4, 6 ... n$
+during the first forward reduction step
+for a general tridiagonal matrix.
+Similar equations can be written describing
+the updates to $a_i$ and  $c_i$.
 
-For a general tridiagonal matrix,
-in addition to the $3n$ original coefficient arrays
-$a$, $b$ and $c$,
-this requires extra storage of 
-$\frac{n}{2}+\frac{n}{4}+\frac{n}{8}+...=n$
-elements each
-for $a_i$, $b_i$, $c_i$, $k_{1,i}$ and $k_{2,i}$
-appearing at the forward reduction steps.
-This additional storage cost is greatly amortized
-when solving the same system for multiple right-hand sides.
-But the additional arrays
-still need to be loaded into shared memory for each thread block.
-This raises the shared memory storage from $4n$ to $9n$,
-which significantly impacts the number of
-concurrently running blocks.
+When the matrix $A$ is near-Toeplitz tridiagonal,
+if we substitute in these equations:
 
-\subsection{Near-Toeplitz matrix}
+\begin{itemize}
+    \item [] $a_2 = a_3 = a_4 = \hdots \equiv a_0$
+    \item [] $b_2 = b_3 = b_4 = \hdots \equiv b_0$
+    \item [] $c_2 = c_3 = c_4 = \hdots \equiv c_0$
+\end{itemize}
+
+we observe that the resulting coefficients
+$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
+correspond to the coefficients of a tridiagonal matrix with
+exactly the near-Toeplitz structure of $A$.
+This form-preserving property of cyclic reduction
+has been reported for block Toeplitz tridiagonal systems
+by Bini et al.
+
+At each reduction step,
+for every index $i$,
+the coefficients $a_i^\prime$, $b_i^\prime$, $c_i^\prime$
+and auxiliary variables $k_1$ and $k_2$
+(hereby denoted as $k_{1,i}$ and $k_{2,i}$) are identical,
+except for those influenced by the
+boundary coefficients of the previous step.
+It is possible to \emph{precompute} these values
+and store them in separate arrays.
+In Figure \ref{fig:cyclic-reduction-precomputing},
+it is seen that the space required for storing each of the
+forward reduction coefficients
+$a_i^\prime$, $b_i^\prime$, $c_i^\prime$,
+$k_{1,i}$ and $k_{2,i}$
+for a near-Toeplitz matrix
+is, at most, $log_2n-2 + 2(log_2n-1)$.
 
@@
 \begin{center}
-\includegraphics[height=200pt]{img/cyclic-reduction-boundaries.eps}
+\includegraphics[height=300pt]{img/cyclic-reduction-precomputing.eps}
 \end{center}
-\caption{Indices affected by boundaries in cyclic reduction}
-\label{fig:cyclic-reduction-boundaries}
+\caption{Storing forward reduction coefficients
+    from each reduction step $m$}
+\label{fig:cyclic-reduction-precomputing}
 \end{figure}
 
-Precomputing the forward reduction coefficients
-becomes especially effective for near-Toeplitz matrices.
-At each reduction step,
-the coefficients $a_i$, $b_i$, $c_i$, $k_{1,i}$ and $k_{2,i}$ 
-are the same for all indices $i$,
-except for indices ``polluted'' by the boundary conditions
-(Figure \ref{fig:cyclic-reduction-boundaries}).
-Further, a careful analysis of the equations
-\ref{eqn:forward-reduction-1}-\ref{eqn:forward-reduction-3}
-reveals that the following extra arrays are required
-to store the coefficients at the boundaries:
+By precomputing and storing the forward reduction
+coefficients,
+the $m^{th}$ forward reduction step is reduced to:
 
-\begin{enumerate}
-    \item The values of $b_1$ at each reduction step
-    \item The values of $k_{1,1}$ at each reduction step
-    \item The values of $k_{1,n}$ at each reduction step
-    \item The values of $a_n$ and $b_n$ at the final reduction step
-        (required for the 2-by-2 solve)
-\end{enumerate}
+\begin{equation}
+d^{\prime}_i = d_i - d_{i-1}k_1^{m}  - d_{i+1}k_2^{m}
+\label{eqn:precomputed-forward-reduction-step}
+\end{equation}
 
-This puts the total storage for precomputed coefficients
-at $8(log_2(n) - 1)+2$.
-For symmetric near-Toeplitz matrices,
-$c_i = a_i$, and $k_{1,i} = k_{2,i}$ away from the boundaries,
-reducing storage to $6(log_2(n) - 1)+2$.
+And the $m^{th}$ backward substitution step is:
+
+\begin{equation}
+x_i = \frac{d^{\prime}_i - a^mx_{i-1} - \
+    c^{m}x_{i+1}}{b^m}
+\label{eqn:precomputed-backward-substitution-step}
+\end{equation}
+
+Where $a^m$, $b^m$, $c^m$, $k_1^m$ and $k_2^m$
+are the appropriate values
+from the precomputed coefficient arrays.
+
+\section{GPU implementation} \label{sec:gpu-implementation}
+
+In this section, we describe the implementation of a GPU solver
+for solving a given near-Toeplitz tridiagonal system
+for $N_{rhs}$ right hand sides.
+
+We describe two approaches based on the common idea
+of precomputed forward reduction coefficients---one that leverages
+the GPU's \emph{shared memory},
+and the other working directly with global memory.
+In both approaches,
+we map each of the $N_{rhs}$ tridiagonal systems
+to blocks,
+and equations or \emph{indices} of a system
+to individual threads within a block.
+
+Following the algorithm described in \ref{sec:proposed-algorithm},
+the forward reduction coefficients
+are first computed and loaded into global memory.
+The right hand sides are stored in a single
+contiguous region global memory.
+
+\subsubsection*{Global memory}
+
+In the global memory approach,
+we perform several kernel launches
+one for each single forward reduction step,
+and another for each backward substitution step.
+The precomputed coefficient arrays and right hand
+are accessed by the kernels from global memory directly.
+Because the coefficients
+$k_{1,i}$ and $k_{2,i}$
+required in the forward reduction step
+(Equation \ref{eqn:precomputed-forward-reduction-step})
+are precomputed and available in global memory,
+we do not compute
+$a_i^\prime$, $b_i^\prime$, $c_i^\prime$,
+$k_{1,i}$ and $k_{2,i}$.
+This reduces the number of uncoalesced accesses to global memory.
+In the backward substitution steps
+(Equation \ref{eqn:precomputed-backward-substitution-step}),
+the coefficients $a^m$, $b^m$ and $c^m$
+are accessed from global memory without coalescing problems.
+
+The advantage of the global memory approach is that
+very large tridiagonal systems can be solved without
+being limited by the available shared memory per block.
+Further,
+because we launch a kernel for each
+forward reduction and back substitution step,
+we can specify the number of threads per block at each launch so that
+there are no idle threads throughout the solution process.
+($n/2$ threads per block for the first forward reduction step,
+$n/4$ for the second, and so on).
+The number of blocks launched is equal to the number of right
+hand sides.
+Exit from a kernel marks a global synchronization point,
+so no explicit synchronization is required within the kernel.
+
+\subsubsection*{Shared memory}
+
+In the shared memory approach,
+we launch a single kernel with $n/2$ threads per block,
+and as many blocks as right hand sides.
+Each block allocates shared memory with size $n/2$.
+We note that the arrays containing the precomputed coefficients,
+are kept in global memory.
+Each thread performs the first reduction step
+(Equation \ref{eqn:precomputed-forward-reduction-step})
+by accessing the required values
+$d_i$, $d_{i-1}$ and $d_{i+1}$ from global memory,
+storing the result in shared memory.
+In subsequent reduction steps,
+$d_i$, $d_{i-1}$ and $d_{i+1}$
+are accessed from shared memory,
+avoiding uncoalesced global memory accesses.
+In each back substitution step,
+threads overwrite the existing values in shared memory
+with the values of the solution $x$.
+In the final step,
+shared memory is filled completeley
+with the even-indexed solution values.
+Each thread then computes an odd-indexed solution value,
+storing it directly in global memory
+and copies the even-indexed solution value
+from shared memory to global memory.
+Explicit synchronization is required at the end of each step.
 
@@
 
-\definecolor{light-gray}{gray}{0.95}
-\begin{minted}[bgcolor=light-gray,linenos,fontsize=\footnotesize]{C++}
-__global__ void forwardReduction( double *a_d, double *b_d,
-        double *c_d, double *d_d, int n, int stride) {
-    int tix = threadIdx.x;
-    int i = (stride-1) + tix*stride;
-    int offset = blockIdx.x*n;
-    int stride;
-    int m, n;
-    double k1, k2;
-
-    k1 = a_d[i]/b_d[i-stride/2];
-    k2 = c_d[i]/b_d[i+stride/2];
-    a_d[i] = -a_d[i-stride/2]*k1;
-    b_d[i] = b_d[i] - c_d[i-stride/2]*k1 - a_d[i+stride/2]*k2;
-    c_d[i] = -c_d[i+stride/2]*k2;
-    d_d[offset+i] = (d_d[offset+i] - d_d[offset+i-stride/2]*k1 -
-                        d_d[offset+i+stride/2]*k2);
-}
-
-__global__ void backwardSubstitution( double *a_d, double *b_d,
-        double *c_d, double *d_d, int n, int stride) {
-    int tix = threadIdx.x;
-    int i = (stride/2-1) + tix*stride;
-    int offset = blockIdx.x*n;
-    d_d[offset+i] = (d_d[goffset+i] - a_d[i]*d_d[offset+i-stride/2] -
-                        c_d[i]*d_d[offset+i+stride/2])/b_d[i];
-}
-
-\end{minted}
-
-\begin{minted}[linenos,bgcolor=light-gray,fontsize=\footnotesize]{c++}
-
-__global__ cyclicReduction(double *a_d, double *b_d, double *c_d,
-    double *d_d, int n) {
-
-    __shared__ double a_l[n];
-    __shared__ double b_l[n];
-    __shared__ double c_l[n];
-    __shared__ double d_l[n];
-    int i = threadIdx.x;
-    int offset = blockIdx.x*n;
-    int stride;
-
-    /* Load coefficients and RHS into shared memory */
-    a_l[i] = a_d[offset+i]; b_l[i] = b_d[offset+i];
-    c_l[i] = c_d[offset+i]; d_l[i] = d_d[offset+i];
-
-    /* Forward reduction */
-    stride = 1;
-    for (int k=0; k<log2(nx)-1; k++) {
-        if (i < nx/stride) {
-            k1 = a_l[i]/b_l[i-stride/2];
-            k2 = c_l[i]/b_l[i+stride/2];
-            a_l[i] = -a_l[i-stride/2]*k1;
-            b_l[i] = b_l[i] - c_l[i-stride/2]*k1 - a_l[i+stride/2]*k2;
-            c_l[i] = -c_l[i+stride/2]*k2;
-            d_l[i] = d_l[i] - d_l[i-stride/2]*k1 - d_l[i+stride/2]*k2;
-        }
-        __syncthreads();
-    }
-
-    if (i == 0) {
-        /* Solve the 2-by-2 system for d_l[n-1] & d_l[n/2-1] */
-    }
-
-    /* Backward substitution */
-    stride = n;
-    for (int k=0; k<log2(nx)-1; k++) {
-        stride = stride/2;
-        if (lix < nx/stride){
-            d_l[i] = (d_l[i] - a_l[i]*d_l[i-stride/2] -
-                        c_l[i]*d_l[i+stride/2])/b_l[i];
-        }
-        __syncthreads();
-    }
-
-    /* Load the solution from shared memory to global memory */
-    d_d[offset+i] = d_l[i];
-}
-
-\end{minted}
-
+\pagebreak
 \section*{References}

### 2015-10-26 paper-draft/results/scripts/compare-solvers-K20.py

@@ -49,5 +49,5 @@ leg = ax.legend((mkl_2_bars[0],
             'Our solver (shared memory)'),
             loc='best',
             fontsize='medium')
-leg.draw_frame(False)
+leg.draw_frame(True)
 plt.savefig('compare-solvers-K20.eps')

### 2015-10-26 paper-draft/results/scripts/compare-solvers-K40.py

@@ -7,22 +7,28 @@ import sys
 
 plt.style.use('bmh')
 prob_sizes = [64, 128, 256, 512]
-intel_mkl = np.array([1.13, 8.90, 77.67, 654])
+intel_mkl_2 = np.array([0.97, 5.884, 75.16, 703.38])
+intel_mkl_4 = np.array([0.836, 7.10, 50.24, 392.7])
+intel_mkl_16 = np.array([1.522, 2.84, 19.64, 221.0])
 cusparse = np.array([0.000446, 0.002864, 0.0244, 0.2178])*1000
 ours_global = np.array([0.00065, 0.0030, 0.0172, 0.1127])*1000
 ours_shared = np.array([0.00039, 0.00193, 0.0111, 0.104])*1000
 
 ind = np.arange(4)*2
 fig, ax = plt.subplots()
-width = 0.25
-mkl_bars = ax.bar(ind, intel_mkl, width, color='r')
-cusparse_bars = ax.bar(ind+width, cusparse, width, color='b')
-global_bars = ax.bar(ind+2*width, ours_global, width, color='y')
-shared_bars = ax.bar(ind+3*width, ours_shared, width, color='g')
+width = 0.2
+mkl_2_bars = ax.bar(ind, intel_mkl_2, width, color='y')
+mkl_4_bars = ax.bar(ind+width, intel_mkl_4, width, color='b')
+mkl_16_bars = ax.bar(ind+2*width, intel_mkl_16, width, color='c')
+cusparse_bars = ax.bar(ind+3*width, cusparse, width, color='m')
+global_bars = ax.bar(ind+4*width, ours_global, width, color='r')
+shared_bars = ax.bar(ind+5*width, ours_shared, width, color='g')
 
-ax.set_yscale('log')
+ax.set_yscale('log', basey=2)
 ax.get_yaxis().set_major_formatter(
         matplotlib.ticker.ScalarFormatter())
+
+ax.yaxis.set_major_formatter(matplotlib.ticker.FormatStrFormatter('%.1f'))
 ax.set_xlabel('Problem size')
 ax.set_ylabel('Time (ms)')
 ax.set_xticks(ind+width)
@@ -30,12 +36,19 @@ ax.set_xticklabels( ('$64^3$', '$128^3$', '$256^3$', '$512^3$') )
 ax.tick_params(labelright=True)
 ax.set_title('Performance of solvers on Tesla K40')
 ax.grid(True)
[…]

### 2015-11-02 paper-draft/paper.tex
update benchmarks

@@
 \begin{abstract}
+    In this paper, we consider the solution of linear
+    systems of the form $A\bm{x} = \bm{d}$,
+    where $A$ is \emph{near-Toeplitz}.
+    These systems appear in numerical schemes
+    such as
+    Alternating Direction Implicit (ADI) methods,
+    compact finite differences,
+    and one-dimensional ordinary
+    and partial differential equations.
+    We develop a GPU solver that takes advantage
+    of the known structure of the matrix
+    to reduce the number of computations
+    and improve memory access.
+    We also describe the implementation of a multi-GPU solver
+    using MPI and CUDA.
+    We describe two applications:
+    2-D ADI applied to the heat equation on a single GPU,
+    and computation of
+    3-D compact finite differences on multiple GPUs.
+    Our solver is able to achieve 6x speedup
+    over an Intel MKL solver running on a single core,
+    and 2x speedup
+    over the NVIDIA CUSPARSE solver.
+    Additionally, we demonstrate scalability up to 64 GPUs.
 
@@
 
-%----------------------------------------------------------------------%
+%---------------------------------------------------------------------%
+\section{Introduction}
+%---------------------------------------------------------------------%
 
-\section{Introduction}
-
-In this paper, we consider the solution of
+The focus of this paper is the solution of
 linear systems of the form $A\bm{x} = \bm{d}$,
@@
 
-The solution of such tridiagonal systems
-appears in several numerical schemes
+Such tridiagonal systems
+appear in several numerical schemes
 in Computational Fluid Dynamics (CFD)
@@
 Alternating Direction Implicit (ADI) methods \cite{1955ADI}, and
-numerical solutions to one-dimensional differential equations.
+numerical solutions of one-dimensional differential equations.
 In such applications,
@@
 
-%----------------------------------------------------------------------%
+%---------------------------------------------------------------------%
 \section{Preliminaries} \label{sec:preliminaries}
+%---------------------------------------------------------------------%
 
@@
 \section{Proposed algorithm} \label{sec:proposed-algorithm}
+%----------------------------------------------------------------------%
 
@@
 
+%----------------------------------------------------------------------%
 \section{GPU implementation} \label{sec:gpu-implementation}
+%----------------------------------------------------------------------%
 
@@
 In the global memory approach,
-we perform several kernel launches
-one for each single forward reduction step,
-and another for each backward substitution step.
+we launch several kernels---one
+for each single forward reduction step,
+an each backward substitution step.
 The precomputed coefficient arrays and right hand
@@
 threads overwrite the existing values in shared memory
-with the values of the solution $x$.
+with the values of the solution.
 In the final step,
@@
 
+\section{3-D compact finite difference on multi-GPU}
+
 %----------------------------------------------------------------------%
-
 \section{Results: single GPU} \label{sec:results-single-gpu}
+%----------------------------------------------------------------------%
 

### 2015-11-02 paper-draft/results/scripts/compare-solvers-K20.py
add benchmark script

@@ -1,53 +1,0 @@
-import numpy as np
-import matplotlib
-import matplotlib.pyplot as plt
-import os
-import sys
-
-plt.style.use('bmh')
-prob_sizes = [64, 128, 256, 512]
-intel_mkl_2 = np.array([0.97, 5.884, 75.16, 703.38])
-intel_mkl_4 = np.array([0.836, 7.10, 50.24, 392.7])
-intel_mkl_16 = np.array([1.522, 2.84, 19.64, 221.0])
-cusparse = np.array([0.000520, 0.00335, 0.0286, np.nan])*1000
-ours_global = np.array([0.0008, 0.00376, 0.0213, 0.1456])*1000
-ours_shared = np.array([0.000476, 0.00237, 0.0136, 0.1325])*1000
-
-ind = np.arange(4)*2
-fig, ax = plt.subplots()
-width = 0.2
-mkl_2_bars = ax.bar(ind, intel_mkl_2, width, color='y')
-mkl_4_bars = ax.bar(ind+width, intel_mkl_4, width, color='b')
-mkl_16_bars = ax.bar(ind+2*width, intel_mkl_16, width, color='c')
-cusparse_bars = ax.bar(ind+3*width, cusparse, width, color='m')
-global_bars = ax.bar(ind+4*width, ours_global, width, color='r')
-shared_bars = ax.bar(ind+5*width, ours_shared, width, color='g')
-
-ax.set_yscale('log', basey=2)
-ax.get_yaxis().set_major_formatter(
-        matplotlib.ticker.ScalarFormatter())
-
-ax.yaxis.set_major_formatter(matplotlib.ticker.FormatStrFormatter('%.1f'))
-ax.set_xlabel('Problem size')
-ax.set_ylabel('Time (ms)')
-ax.set_xticks(ind+width)
-ax.set_xticklabels( ('$64^3$', '$128^3$', '$256^3$', '$512^3$') )
-ax.tick_params(labelright=True)
-ax.set_title('Performance of solvers on Tesla K20')
-ax.grid(True)
-leg = ax.legend((mkl_2_bars[0],
-            mkl_4_bars[0],
-            mkl_16_bars[0],
-            cusparse_bars[0],
-            global_bars[0],
-            shared_bars[0]),
-            ('Intel MKL 2 cores',
-            'Intel MKL 4 cores',
-            'Intel MKL 16 cores',
-            'CUSPARSE',
-            'Our solver (global memory)',
[…]

### 2015-11-02 paper-draft/results/scripts/k20-bench.py

@@ -1,0 +1,135 @@
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import os
+import sys
+
+#plt.style.use('bmh')
+# 2-D plots
+prob_sizes = [32, 64, 128, 256, 512., 1024, 2048]
+intel_mkl_1 = np.array(
+        [0.045, 0.048, 0.089, 0.263, 0.959, 3.775, 16.272])
+intel_mkl_2 = np.array(
+        [0.043, 0.052, 0.086, 0.217, 0.688, 2.612, 10.953])
+intel_mkl_4 = np.array(
+        [0.045, 0.051, 0.068, 0.151, 0.412, 1.577, 6.815])
+intel_mkl_8 = np.array(
+        [0.042, 0.048, 0.084, 0.107, 0.299, 1.023, 4.823])
+intel_mkl_16 = np.array(
+        [0.050, 0.249, 0.145, 0.202, 0.299, 0.864, 4.111])
+cusparse = np.array(
+        [0.273, 0.271, 0.271, 0.305, 0.629, 1.939, 7.607])
+neato_global = np.array(
+        [0.235, 0.279, 0.345, 0.350, 0.402, 1.378, 5.823])
+
+ind = np.arange(7)*5
+fig, ax = plt.subplots()
+width = 0.5
+mkl_1_bars = ax.bar(ind+0*width, intel_mkl_1, width, color='r')
+mkl_2_bars = ax.bar(ind+1*width, intel_mkl_2, width, color='g')
+mkl_4_bars = ax.bar(ind+2*width, intel_mkl_4, width, color='b')
+mkl_8_bars = ax.bar(ind+3*width, intel_mkl_8, width, color='orange')
+mkl_16_bars = ax.bar(ind+4*width, intel_mkl_16, width, color='m')
+cusparse_bars = ax.bar(ind+5*width, cusparse, width, color='y')
+neato_bars = ax.bar(ind+6*width, neato_global, width, color='c')
+
+ax.set_yscale('log', basey=2)
+ax.get_yaxis().set_major_formatter(
+        matplotlib.ticker.ScalarFormatter())
+ax.yaxis.set_major_formatter(
+        matplotlib.ticker.FormatStrFormatter('%.1f'))
+ax.set_xlabel('Problem size')
+
+ax.set_xlabel('Number of systems = system size')
+ax.set_ylabel('Time to solve systems (ms)')
[…]

### 2015-11-02 paper-draft/results/scripts/bench/k20-bench.py
add new results

@@ -19,8 +19,8 @@ intel_mkl_16 = np.array(
         [0.050, 0.249, 0.145, 0.202, 0.299, 0.864, 4.111])
 cusparse = np.array(
         [0.273, 0.271, 0.271, 0.305, 0.629, 1.939, 7.607])
-neato_global = np.array(
-        [0.235, 0.279, 0.345, 0.350, 0.402, 1.378, 5.823])
+neato_shared = np.array(
+        [0.024, 0.025, 0.032, 0.066, 0.272, 1.252, 10.407])
 
 ind = np.arange(7)*5
 fig, ax = plt.subplots()
@@ -31,7 +31,7 @@ mkl_4_bars = ax.bar(ind+2*width, intel_mkl_4, width, color='b')
 mkl_8_bars = ax.bar(ind+3*width, intel_mkl_8, width, color='orange')
 mkl_16_bars = ax.bar(ind+4*width, intel_mkl_16, width, color='m')
 cusparse_bars = ax.bar(ind+5*width, cusparse, width, color='y')
-neato_bars = ax.bar(ind+6*width, neato_global, width, color='c')
+neato_bars = ax.bar(ind+6*width, neato_shared, width, color='c')
 
 ax.set_yscale('log', basey=2)
 ax.get_yaxis().set_major_formatter(
@@ -40,7 +40,7 @@ ax.yaxis.set_major_formatter(
         matplotlib.ticker.FormatStrFormatter('%.1f'))
 ax.set_xlabel('Problem size')
 
-ax.set_xlabel('Number of systems = system size')
+ax.set_xlabel('System size (Number of systems = system size)')
 ax.set_ylabel('Time to solve systems (ms)')
 ax.set_xticks(ind+width)
 ax.set_xticklabels( 
@@ -83,7 +83,7 @@ intel_mkl_16 = np.array(
         [0.257, 0.403, 1.931, 16.610, 159.394])
 cusparse = np.array(
         [0.310, 0.556, 3.128, 28.495, np.nan])
-neato_global = np.array(
+neato_shared = np.array(
         [0.093, 0.409, 2.224, 12.568, 125.326])
 
 ind = np.arange(5)*5
@@ -95,7 +95,7 @@ mkl_4_bars = ax.bar(ind+2*width, intel_mkl_4, width, color='b')
 mkl_8_bars = ax.bar(ind+3*width, intel_mkl_8, width, color='orange')
 mkl_16_bars = ax.bar(ind+4*width, intel_mkl_16, width, color='m')
[…]

### 2015-11-02 paper-draft/results/scripts/compare-solvers-K40.py

@@ -1,54 +1,0 @@
-import numpy as np
-import matplotlib
-import matplotlib.pyplot as plt
-import os
-import sys
-
-
-plt.style.use('bmh')
-prob_sizes = [64, 128, 256, 512]
-intel_mkl_2 = np.array([0.97, 5.884, 75.16, 703.38])
-intel_mkl_4 = np.array([0.836, 7.10, 50.24, 392.7])
-intel_mkl_16 = np.array([1.522, 2.84, 19.64, 221.0])
-cusparse = np.array([0.000446, 0.002864, 0.0244, 0.2178])*1000
-ours_global = np.array([0.00065, 0.0030, 0.0172, 0.1127])*1000
-ours_shared = np.array([0.00039, 0.00193, 0.0111, 0.104])*1000
-
-ind = np.arange(4)*2
-fig, ax = plt.subplots()
-width = 0.2
-mkl_2_bars = ax.bar(ind, intel_mkl_2, width, color='y')
-mkl_4_bars = ax.bar(ind+width, intel_mkl_4, width, color='b')
-mkl_16_bars = ax.bar(ind+2*width, intel_mkl_16, width, color='c')
-cusparse_bars = ax.bar(ind+3*width, cusparse, width, color='m')
-global_bars = ax.bar(ind+4*width, ours_global, width, color='r')
-shared_bars = ax.bar(ind+5*width, ours_shared, width, color='g')
-
-ax.set_yscale('log', basey=2)
-ax.get_yaxis().set_major_formatter(
-        matplotlib.ticker.ScalarFormatter())
-
-ax.yaxis.set_major_formatter(matplotlib.ticker.FormatStrFormatter('%.1f'))
-ax.set_xlabel('Problem size')
-ax.set_ylabel('Time (ms)')
-ax.set_xticks(ind+width)
-ax.set_xticklabels( ('$64^3$', '$128^3$', '$256^3$', '$512^3$') )
-ax.tick_params(labelright=True)
-ax.set_title('Performance of solvers on Tesla K40')
-ax.grid(True)
-leg = ax.legend((mkl_2_bars[0],
-            mkl_4_bars[0],
-            mkl_16_bars[0],
-            cusparse_bars[0],
-            global_bars[0],
-            shared_bars[0]),
-            ('Intel MKL 2 cores',
-            'Intel MKL 4 cores',
-            'Intel MKL 16 cores',
-            'CUSPARSE',
-            'Our solver (global memory)',
[…]

### 2015-11-02 paper-draft/results/scripts/global-vs-shared/k20-global-vs-shared.py

@@ -1,0 +1,83 @@
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import os
+import sys
+
+#plt.style.use('bmh')
+# 2-D plots
+prob_sizes = [32, 64, 128, 256, 512., 1024, 2048]
+globalmem = np.array(
+        [0.201, 0.247, 0.284, 0.326, 0.403, 1.375, 5.811])
+sharedmem = np.array(
+        [0.024, 0.025, 0.032, 0.066, 0.273, 1.252, 10.407])
+
+ind = np.arange(7)*2
+fig, ax = plt.subplots()
+width = 0.5
+globalmem_bars = ax.bar(ind+0*width, globalmem, width, color='r')
+sharedmem_bars = ax.bar(ind+1*width, sharedmem, width, color='g')
+
+ax.set_yscale('log', basey=2)
+ax.get_yaxis().set_major_formatter(
+        matplotlib.ticker.ScalarFormatter())
+ax.yaxis.set_major_formatter(
+        matplotlib.ticker.FormatStrFormatter('%.1f'))
+ax.set_xlabel('Problem size')
+
+ax.set_xlabel('System size (Number of systems = system size)')
+ax.set_ylabel('Time to solve systems (ms)')
+ax.set_xticks(ind+width)
+ax.set_xticklabels( 
+    ('$32$', '$64$', '$128$', '$256$', '$512$', '$1024$', '$2048$') )
+ax.tick_params(labelright=True)
+ax.set_title('Global v/s shared memory solver performance for 2-D problems')
+ax.grid(True)
+leg = ax.legend((
+            globalmem_bars[0],
+            sharedmem_bars[0]),
+            ('Global memory',
+            'Shared memory'),
+            loc='best',
+            fontsize='medium')
+leg.draw_frame(True)
+plt.savefig('k20-global-vs-shared-2d.eps')
+
+#plt.style.use('bmh')
+# 3-D plots
+prob_sizes = [32, 64, 128, 256, 512]
+globalmem = np.array(
+        [0.207, 0.751, 3.669, 21.148, 145.34])
+sharedmem = np.array(
+        [0.092, 0.409, 2.225, 12.565, 125.311])
+
+ind = np.arange(5)*2
+fig, ax = plt.subplots()
+width = 0.5
+globalmem_bars = ax.bar(ind+0*width, globalmem, width, color='r')
+sharedmem_bars = ax.bar(ind+1*width, sharedmem, width, color='g')
+
[…]

### 2015-11-02 paper-draft/paper.tex
add benchmark script
add new results
add more results and update paper draft

@@
 \usepackage{xcolor}
+\usepackage{multirow}
 
@@
 
-\section{3-D compact finite difference on multi-GPU}
-
+\pagebreak
 %----------------------------------------------------------------------%
@@
 \begin{center}
-\includegraphics[height=200pt]{fig/compare-solvers-K20.eps}
+\includegraphics[height=200pt]{fig/k20-bench-2d.eps}
 \end{center}
-\label{fig:compare-solvers-K20}
+\label{fig:k20-bench-2d}
 \end{figure}
@@
 \begin{center}
-\includegraphics[height=200pt]{fig/compare-solvers-K40.eps}
+\includegraphics[height=200pt]{fig/k20-bench-3d.eps}
 \end{center}
-\label{fig:compare-solvers-K40}
+\label{fig:k20-bench-3d}
 \end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{fig/k20-global-vs-shared-2d.eps}
+\end{center}
+\label{fig:k20-global-vs-shared-2d}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{fig/k20-global-vs-shared-3d.eps}
+\end{center}
+\label{fig:k20-global-vs-shared-3d}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=250pt]{fig/k20-forward-vs-backward.eps}
+\end{center}
+\label{fig:k20-forward-vs-backward}
+\end{figure}
+
+\input{table/global-vs-shared.tex}
 

### 2015-11-02 paper-draft/results/scripts/phases-analysis/k20-phases-analysis.py

@@ -1,0 +1,85 @@
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import os
+import sys
+
+plt.style.use('bmh')
+# 3-D plots
+prob_sizes = [32, 64, 128, 256, 512]
+
+forward_globalmem = np.array(
+        [0.105, 0.317, 1.585, 9.645, 68.635])
+solve_globalmem = np.array(
+        [0.026, 0.057, 0.226, 0.883, 3.526])
+backward_globalmem = np.array(
+        [0.103, 0.373, 1.850, 10.588, 73.086])
+forward_sharedmem = np.array(
+        [0.031, 0.136, 0.758, 3.523, 35.192])
+solve_sharedmem = np.array(
+        [0.024, 0.069, 0.256, 1.084, 8.520])
+backward_sharedmem = np.array(
+        [0.065, 0.313, 1.484, 7.382, 70.943])
+
+forward_globalmem_percent = forward_globalmem/(
+        forward_globalmem+backward_globalmem+solve_globalmem)*100
+solve_globalmem_percent = solve_globalmem/(
+        forward_globalmem+backward_globalmem+solve_globalmem)*100
+backward_globalmem_percent = backward_globalmem/(
+        forward_globalmem+backward_globalmem+solve_globalmem)*100
+forward_sharedmem_percent = forward_sharedmem/(
+        forward_sharedmem+backward_sharedmem+solve_sharedmem)*100
+solve_sharedmem_percent = solve_sharedmem/(
+        forward_sharedmem+backward_sharedmem+solve_sharedmem)*100
+backward_sharedmem_percent = backward_sharedmem/(
+        forward_sharedmem+backward_sharedmem+solve_sharedmem)*100
+
+
+ind = np.arange(5)*2
+fig, ax = plt.subplots(figsize=(8, 8))
+width = 0.5
+forward_globalmem_bars = ax.barh(
+        ind+0*width, forward_globalmem_percent, width, color='b')
+solve_globalmem_bars = ax.barh(
+        ind+0*width, solve_globalmem_percent, width,
+        left=forward_globalmem_percent, color='r')
+backward_globalmem_bars = ax.barh(
+        ind+0*width, backward_globalmem_percent, width,
+        left=solve_globalmem_percent+forward_globalmem_percent, color='g')
+forward_sharedmem_bars = ax.barh(
+        ind+1*width, forward_sharedmem_percent, width, color='b')
+solve_sharedmem_bars = ax.barh(
[…]

### 2015-11-02 paper-draft/paper.tex
move stuff around

@@
 %----------------------------------------------------------------------%
-\section{Results: single GPU} \label{sec:results-single-gpu}
+\section{Results} \label{sec:results-single-gpu}
 %----------------------------------------------------------------------%
+
+\subsection{Global memory v/s shared memory solver performance}
 
@@
 \begin{center}
-\includegraphics[height=200pt]{fig/k20-bench-2d.eps}
+\includegraphics[height=200pt]{fig/global-vs-shared-2d.eps}
+\end{center}
+\label{fig:global-vs-shared-2d}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{fig/global-vs-shared-3d.eps}
+\end{center}
+\label{fig:global-vs-shared-3d}
+\end{figure}
+
+In Figures \ref{fig:global-vs-shared-2d} and \ref{fig:global-vs-shared-3d},
+we report the performance of the two solvers for the case
+$N_{rhs} = n$ and $N_{rhs} = n^2$.
+These cases correspond to tridiagonal systems
+arising in 2-D and 3-D problems respectively.
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=250pt]{fig/bench-2d.eps}
 \end{center}
@@
 \begin{center}
-\includegraphics[height=200pt]{fig/k20-bench-3d.eps}
+\includegraphics[width=250pt]{fig/bench-3d.eps}
 \end{center}
@@
 \begin{center}
-\includegraphics[height=200pt]{fig/k20-global-vs-shared-2d.eps}
-\end{center}
-\label{fig:k20-global-vs-shared-2d}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[height=200pt]{fig/k20-global-vs-shared-3d.eps}
-\end{center}
-\label{fig:k20-global-vs-shared-3d}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[height=250pt]{fig/k20-forward-vs-backward.eps}
+\includegraphics[width=250pt]{fig/phases-analysis.eps}
 \end{center}

### 2015-11-02 paper-draft/results/scripts/bench/bench.py

@@ -66,7 +66,7 @@ leg = ax.legend((
             loc='best',
             fontsize='medium')
 leg.draw_frame(True)
-plt.savefig('k20-bench-2d.eps')
+plt.savefig('bench-2d.eps')
 
 #plt.style.use('bmh')
 # 3-D plots
@@ -130,6 +130,6 @@ leg = ax.legend((
             loc='best',
             fontsize='medium')
 leg.draw_frame(True)
-plt.savefig('k20-bench-3d.eps')
+plt.savefig('bench-3d.eps')
 
 

### 2015-11-02 paper-draft/results/scripts/global-vs-shared/global-vs-shared.py

@@ -41,7 +41,7 @@ leg = ax.legend((
             loc='best',
             fontsize='medium')
 leg.draw_frame(True)
-plt.savefig('k20-global-vs-shared-2d.eps')
+plt.savefig('global-vs-shared-2d.eps')
 
 #plt.style.use('bmh')
 # 3-D plots
@@ -80,4 +80,4 @@ leg = ax.legend((
             loc='best',
             fontsize='medium')
 leg.draw_frame(True)
-plt.savefig('k20-global-vs-shared-3d.eps')
+plt.savefig('global-vs-shared-3d.eps')

### 2015-11-02 paper-draft/results/scripts/phases-analysis/phases-analysis.py

@@ -82,4 +82,4 @@ leg = ax.legend((
             fontsize='medium',
             bbox_to_anchor=(0.5, 1.1))
 leg.draw_frame(True)
-plt.savefig('k20-phases-analysis.eps', bbox_extra_artists=(leg,), bbox_inches='tight')
+plt.savefig('phases-analysis.eps', bbox_extra_artists=(leg,), bbox_inches='tight')

### 2015-11-02 paper-draft/results/scripts/bench/bench.py
better graph

@@ -37,7 +37,7 @@ ax.set_yscale('log', basey=2)
 ax.get_yaxis().set_major_formatter(
         matplotlib.ticker.ScalarFormatter())
 ax.yaxis.set_major_formatter(
-        matplotlib.ticker.FormatStrFormatter('%.1f'))
+        matplotlib.ticker.FormatStrFormatter('%.2f'))
 ax.set_xlabel('Problem size')
 
 ax.set_xlabel('System size (Number of systems = system size)')
@@ -101,7 +101,7 @@ ax.set_yscale('log', basey=2)
 ax.get_yaxis().set_major_formatter(
         matplotlib.ticker.ScalarFormatter())
 ax.yaxis.set_major_formatter(
-        matplotlib.ticker.FormatStrFormatter('%.1f'))
+        matplotlib.ticker.FormatStrFormatter('%.2f'))
 ax.set_xlabel('Problem size')
 
 ax.set_xlabel('System size (Number of systems = system size$^2$)')

### 2015-11-03 paper-draft/paper.tex
update

@@
 
+In this section we present a performance overview
+of our Near-Toeplitz tridiagonal solver (NEATO)
+against a multi-threaded Intel MKL solver and
+the CUSPARSE GPU solver
+(\texttt{dgtsv} and \texttt{dgtsvStridedBatch} respectively).
+
+The CPU code is compiled with the Intel C++ compiler
+and run on up to 16 cores of the
+Intel Xeon Processor E5-2670 v2 (25M Cache, 2.50 GHz).
+GPU code is compiled with the CUDA toolkit (version 6.5.14),
+and run on the
+NVIDIA Tesla K20 Accelerator.
+\texttt{-O2} level optimizations are turned on for both
+CPU and GPU code.
+
 \subsection{Global memory v/s shared memory solver performance}
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[height=200pt]{fig/global-vs-shared-2d.eps}
-\end{center}
+\begin{figure}[!htb]
+    \centering
+    \begin{minipage}{0.75\textwidth}
+        \centering
+        \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
+        \caption{Comparison of global memory and shared
+            memory implementations of NEATO (2D problems)}
         \label{fig:global-vs-shared-2d}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[height=200pt]{fig/global-vs-shared-3d.eps}
-\end{center}
+    \end{minipage}
+    \\
+    \begin{minipage}{0.75\textwidth}
+        \centering
+        \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
+        \caption{Comparison of global memory and shared memory
+            implementations of NEATO (2D problems)}
         \label{fig:global-vs-shared-3d}
+    \end{minipage}
 \end{figure}
@@
 arising in 2-D and 3-D problems respectively.
+We note that the shared memory implementation
+offers better performance in nearly all cases.
+However, the relative speedup from using
+shared memory diminishes with increasing problem size.
+For larger problem sizes,
+the reduced occupancy and increase in number of inactive threads
+leads to poor shared memory performance.
+
+\subsection{Comparison with Intel MKL and CUSPARSE solvers}
+
+In Figures \ref{fig:bench-2d} and \ref{fig:bench-3d},
+we provide the relative performance of
+Intel MKL and CUSPARSE solvers and compare against
+the NEATO shared memory implementation.
+The Intel MKL solver can use multiple threads
+to solve several systems simaltaneously.
+The CUSPARSE solver uses a hybrid algorithm composed
+of Cyclic Reduction and Parallel Cyclic Reduction (CR+PCR).
+
+The relative performance for
+each problem size is obtained by
+normalizing the solver timings
+by the timing for the fastest solver for that problem size.
+
+Table \ref{table:bench} shows
+timings for different solvers.
 
@@
 \begin{center}
-\includegraphics[width=250pt]{fig/bench-2d.eps}
+\includegraphics[width=300pt]{fig/bench-2d.eps}
 \end{center}
-\label{fig:k20-bench-2d}
+\caption{Relative solver performance for 2-D problems}
+\label{fig:bench-2d}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=250pt]{fig/bench-3d.eps}
+\includegraphics[width=300pt]{fig/bench-3d.eps}
 \end{center}
-\label{fig:k20-bench-3d}
+\caption{Relative solver performance for 3-D problems}
+\label{fig:bench-3d}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=250pt]{fig/phases-analysis.eps}
+\includegraphics[width=300pt]{fig/phases-analysis.eps}
 \end{center}
@@
 
-\input{table/global-vs-shared.tex}
+\begin{table}[]
+\scriptsize
+\centering
+\caption{Performance of Intel MKL, CUSPARSE and NEATO solvers}
+\label{table:bench}
+\input{table/bench.tex}
+\end{table}
 
+\section{Application: compact finite difference on multiple GPUs}
+
+\pagebreak
 \pagebreak

### 2015-11-03 paper-draft/results/scripts/bench/bench.py

@@ -4,7 +4,7 @@ import matplotlib.pyplot as plt
 import os
 import sys
 
-#plt.style.use('bmh')
+plt.style.use('bmh')
 # 2-D plots
 prob_sizes = [32, 64, 128, 256, 512., 1024, 2048]
 intel_mkl_1 = np.array(
@@ -21,6 +21,23 @@ cusparse = np.array(
         [0.273, 0.271, 0.271, 0.305, 0.629, 1.939, 7.607])
 neato_shared = np.array(
         [0.024, 0.025, 0.032, 0.066, 0.272, 1.252, 10.407])
+matrix = np.vstack([intel_mkl_1,
+                    intel_mkl_2,
+                    intel_mkl_4,
+                    intel_mkl_8,
+                    intel_mkl_16,
+                    cusparse,
+                    neato_shared])
+mins = np.min(matrix, axis=0)
+
+# normalize timings:
+intel_mkl_1 = intel_mkl_1/mins
+intel_mkl_2 = intel_mkl_2/mins
+intel_mkl_4 = intel_mkl_4/mins
+intel_mkl_8 = intel_mkl_8/mins
+intel_mkl_16 = intel_mkl_16/mins
+cusparse = cusparse/mins
+neato_shared = neato_shared/mins
 
 ind = np.arange(7)*5
 fig, ax = plt.subplots()
@@ -33,20 +50,18 @@ mkl_16_bars = ax.bar(ind+4*width, intel_mkl_16, width, color='m')
 cusparse_bars = ax.bar(ind+5*width, cusparse, width, color='y')
 neato_bars = ax.bar(ind+6*width, neato_shared, width, color='c')
 
-ax.set_yscale('log', basey=2)
 ax.get_yaxis().set_major_formatter(
         matplotlib.ticker.ScalarFormatter())
 ax.yaxis.set_major_formatter(
-        matplotlib.ticker.FormatStrFormatter('%.2f'))
+        matplotlib.ticker.FormatStrFormatter('%d'))
+
+ax.set_xticks(ind+3*width)
+ax.set_xticklabels(['32', '64', '128', '256', '512', '1024', '2048'])
 ax.set_xlabel('Problem size')
 
 ax.set_xlabel('System size (Number of systems = system size)')
-ax.set_ylabel('Time to solve systems (ms)')
-ax.set_xticks(ind+width)
-ax.set_xticklabels( 
-    ('$32$', '$64$', '$128$', '$256$', '$512$', '$1024$', '$2048$') )
[…]

### 2015-11-03 paper-draft/results/scripts/global-vs-shared/global-vs-shared.py

@@ -4,7 +4,7 @@ import matplotlib.pyplot as plt
 import os
 import sys
 
-#plt.style.use('bmh')
+plt.style.use('bmh')
 # 2-D plots
 prob_sizes = [32, 64, 128, 256, 512., 1024, 2048]
 globalmem = np.array(
@@ -15,29 +15,28 @@ sharedmem = np.array(
 ind = np.arange(7)*2
 fig, ax = plt.subplots()
 width = 0.5
-globalmem_bars = ax.bar(ind+0*width, globalmem, width, color='r')
-sharedmem_bars = ax.bar(ind+1*width, sharedmem, width, color='g')
+globalmem_bars = ax.bar(ind+0*width, globalmem, width, color='c')
+sharedmem_bars = ax.bar(ind+1*width, sharedmem, width, color='y')
 
 ax.set_yscale('log', basey=2)
 ax.get_yaxis().set_major_formatter(
         matplotlib.ticker.ScalarFormatter())
 ax.yaxis.set_major_formatter(
-        matplotlib.ticker.FormatStrFormatter('%.1f'))
+        matplotlib.ticker.FormatStrFormatter('%.2f'))
 ax.set_xlabel('Problem size')
 
 ax.set_xlabel('System size (Number of systems = system size)')
 ax.set_ylabel('Time to solve systems (ms)')
 ax.set_xticks(ind+width)
 ax.set_xticklabels( 
-    ('$32$', '$64$', '$128$', '$256$', '$512$', '$1024$', '$2048$') )
+    ('32', '64', '128', '256', '512', '1024', '2048') )
 ax.tick_params(labelright=True)
-ax.set_title('Global v/s shared memory solver performance for 2-D problems')
 ax.grid(True)
 leg = ax.legend((
             globalmem_bars[0],
             sharedmem_bars[0]),
-            ('Global memory',
-            'Shared memory'),
+            ('NEATO (global memory)',
+            'NEATO (shared memory)'),
             loc='best',
             fontsize='medium')
 leg.draw_frame(True)
@@ -54,29 +53,28 @@ sharedmem = np.array(
 ind = np.arange(5)*2
 fig, ax = plt.subplots()
 width = 0.5
-globalmem_bars = ax.bar(ind+0*width, globalmem, width, color='r')
-sharedmem_bars = ax.bar(ind+1*width, sharedmem, width, color='g')
+globalmem_bars = ax.bar(ind+0*width, globalmem, width, color='c')
[…]

### 2015-11-04 paper-draft/paper.tex
update paper

@@
 \usepackage{multirow}
+\usepackage{rotating}
 
@@
 \begin{frontmatter}
-\author{Ashwin Srinath}
+\author{Author names}
 \title{A fast GPU solver for near-Toeplitz tridiagonal systems}
@@
     and improve memory access.
+    Our solver is able to achieve up to 5x speedup
+    over an Intel MKL solver running on a single core,
+    and 1.5x-11x speedup
+    over the NVIDIA CUSPARSE solver.
+    For larger problem sizes,
+    our solver is able to outperform
+    CUSPARSE as well as Intel MKL (on up to 16 cores).
     We also describe the implementation of a multi-GPU solver
@@
     3-D compact finite differences on multiple GPUs.
-    Our solver is able to achieve 6x speedup
-    over an Intel MKL solver running on a single core,
-    and 2x speedup
-    over the NVIDIA CUSPARSE solver.
-    Additionally, we demonstrate scalability up to 64 GPUs.
 
@@
 describes our proposed solver for near-Toeplitz tridiagonal systems.
+Section \ref{sec:gpu-implementation}
+discusses the implementation of the algorithm on GPUs.
 Section \ref{sec:results-single-gpu}
@@
 \begin{center}
-\includegraphics[height=300pt]{img/cyclic-reduction-precomputing.eps}
+\includegraphics[height=250pt]{img/cyclic-reduction-precomputing.eps}
 \end{center}
@@
 for $N_{rhs}$ right hand sides.
-
-We describe two approaches based on the common idea
+We develop two approaches based on the common idea
 of precomputed forward reduction coefficients---one that leverages
@@
 and the other working directly with global memory.
-In both approaches,
-we map each of the $N_{rhs}$ tridiagonal systems
-to blocks,
-and equations or \emph{indices} of a system
-to individual threads within a block.
 
@@
 contiguous region global memory.
+In both approaches,
+we map each of the $N_{rhs}$ tridiagonal systems
+to a thread block,
+and equations or \emph{indices} of a system
+to individual threads within a block.
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=400pt]{img/global-and-shared.eps}
+\end{center}
+\caption{Shared memory (top) and Global memory (bottom) implementations---
+    multiple kernel launches allow us to dedicate appropriate number
+    of threads per block in the global memory approach}
+\label{fig:global-and-shared}
+\end{figure}
 
@@
 
-In the global memory approach,
-we launch several kernels---one
-for each single forward reduction step,
-an each backward substitution step.
+This implementation works entirely on the GPU's global memory.
+Each right-hand side is mapped to a single thread block.
+Each block executes two kernels,
+\texttt{forwardReduction} and \texttt{backSubstitution},
+which are called each called $log_2(n)-1$ times.
+An extra call to the
+forward reduction kernel additionally performs
+the two-by-two solve.
+Exit from a kernel marks a global synchronization point,
+so no explicit synchronization is performed in the kernels.
+Because we launch a kernel for each
+forward reduction and back substitution step,
+we can specify the appropriate number of threads per block---$n/2$
+threads per block for the first forward reduction step,
+$n/4$ for the second, and so on (Figure \ref{fig:global-and-shared}).
+This guarantees that
+all threads within a block are always active.
 The precomputed coefficient arrays and right hand
@@
 
-The advantage of the global memory approach is that
-very large tridiagonal systems can be solved without
-being limited by the available shared memory per block.
-Further,
-because we launch a kernel for each
-forward reduction and back substitution step,
-we can specify the number of threads per block at each launch so that
-there are no idle threads throughout the solution process.
-($n/2$ threads per block for the first forward reduction step,
-$n/4$ for the second, and so on).
-The number of blocks launched is equal to the number of right
-hand sides.
-Exit from a kernel marks a global synchronization point,
-so no explicit synchronization is required within the kernel.
-
 \subsubsection*{Shared memory}
 
-In the shared memory approach,
-we launch a single kernel with $n/2$ threads per block,
-and as many blocks as right hand sides.
-Each block allocates shared memory with size $n/2$.
-We note that the arrays containing the precomputed coefficients,
-are kept in global memory.
+In the shared memory approach (Figure \ref{fig:global-and-shared}),
+we again map every right-hand side to a thread block.
+Each thread block executes a single kernel \emph{once},
+which performs forward reduction,
+backward substitution, and the two-by-two solve.
+Shared memory sized $n/2$ is allocated by each block.
 Each thread performs the first reduction step
@@
 from shared memory to global memory.
-Explicit synchronization is required at the end of each step.
+The entire solution is done in a single kernel launch
+to hide the latency of
+memory transfers between global and shared memory.
+Explicit synchronization between the threads of a block
+is required within the kernel at the end of each step.
 
-\pagebreak
 %----------------------------------------------------------------------%
@@
 (\texttt{dgtsv} and \texttt{dgtsvStridedBatch} respectively).
-
-The CPU code is compiled with the Intel C++ compiler
+The CPU code is compiled with the Intel C compiler
 and run on up to 16 cores of the
@@
 \begin{figure}[!htb]
-    \centering
-    \begin{minipage}{0.75\textwidth}
-        \centering
         \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
@@
         \label{fig:global-vs-shared-2d}
-    \end{minipage}
-    \\
-    \begin{minipage}{0.75\textwidth}
-        \centering
         \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
@@
         \label{fig:global-vs-shared-3d}
-    \end{minipage}
 \end{figure}
@@
 of Cyclic Reduction and Parallel Cyclic Reduction (CR+PCR).
-
 The relative performance for
@@
 by the timing for the fastest solver for that problem size.
-
-Table \ref{table:bench} shows
-timings for different solvers.
+Table \ref{fig:bench} shows the timings of the various
+solvers to solve different problem sizes.
 
@@
 \includegraphics[width=300pt]{fig/bench-2d.eps}
-\end{center}
 \caption{Relative solver performance for 2-D problems}
 \label{fig:bench-2d}
+\end{center}
 \end{figure}
@@
 \includegraphics[width=300pt]{fig/bench-3d.eps}
-\end{center}
 \caption{Relative solver performance for 3-D problems}
 \label{fig:bench-3d}
+\end{center}
 \end{figure}
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[width=300pt]{fig/phases-analysis.eps}
-\end{center}
-\label{fig:k20-forward-vs-backward}
-\end{figure}
-
-\begin{table}[]
+\begin{sidewaystable}[]
 \scriptsize
@@
 \input{table/bench.tex}
-\end{table}
+\end{sidewaystable}
 
@@
 
-\pagebreak
-\pagebreak
+\subsection{Numerical algorithm}
+
+In a uniformly spaced one-dimensional grid with spacing $dx$,
+if $f_i$ represents the value of
+the function evaluated at the $i$th node,
+the first derivative $f^{\prime}_i$ can be approximated from
+a relation of the form:
+
+\begin{equation}
+\begin{split}
+    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
+    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
+        f^{\prime}_i
+    = 
+    c\frac{f_{i+3} - f_{i-3}}{dx} + \
+    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
+    a\frac{f_{i+1} - f_{i-1}}{dx} + \
+    \hdots
+\end{split}
+\label{eqn:general-compact}
+\end{equation}
+
+For the particular scheme used, $\alpha = 1/4$,
+and $ a = 3/4 $.
+It is easily noted that this leads to a tridiagonal system of equations,
+along with a 3-point stencil to evaluate the RHS.
+Further, the left boundary can be treated using the following implicit equation:
+
+\begin{equation}
+    f^{\prime}_1 + 2f^{\prime}_2 = \frac{-5f_1 + 4f_2 + f_3}{dx}
+\end{equation}
+
+The right boundary ($f^{\prime}_{n}$) is treated by utilizing
+the negative complex-conjugate of the Fourier image of the stencil
+at $f^{\prime}_1$:
+
+\begin{equation}
+    f^{\prime}_{n} + 2f^{\prime}_{n-1s}
+    =
+    \frac{5f_{n} - 4f_{n-2} 1 f_{n-2}}{dx}
+\end{equation}
+
+This boundary treatment maintains the tridiagonal identity
+of the syste to be solved, which then has the form:
+
+\begin{equation} \label{eqn:compact-tridiagonal-system}
+ \begin{bmatrix}
+     1&2\\
+     1/4&1&1/4\\
+     &1/4&1&1/4\\
+     &&1/4&1&1/4\\
+     &&&1/4&1&1/4\\
+     &&&&&\ddots\\
+     &&&&&&\ddots\\
+     &&&&&&&\ddots\\
+     &&&&&&&2&1
+  \end{bmatrix}
+  \begin{bmatrix}
+      f^{\prime}_1 \\
+      f^{\prime}_2 \\
+      f^{\prime}_3 \\
+      \vdots \\
+      \vdots \\
+      \vdots \\
+      \vdots \\
+      f^{\prime}_{n-1} \\
+      f^{\prime}_n
+   \end{bmatrix}
+ =
+ \begin{bmatrix}
+     -5f_1 + 4f_2 + f_3\\
+     f_{3} - f_{1}\\
+     f_{4} - f_{2}\\
+     \vdots\\
+     \vdots\\
+     \vdots\\
+     \vdots\\
+     f_{n} - f_{n-2}\\
+     5f_{n} - 4f_{n-1} - f_{n-2}
+  \end{bmatrix}
+\end{equation}
+
+\subsection{Computational domain}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{img/computational-domain.eps}
+\end{center}
+\caption{Computational domain in 3-D}
+\label{fig:computational-domain}
+\end{figure}
+
+The computational domain (Figure \ref{fig:computational-domain})
+is a structured grid comprised of
+$nx \times ny \times nz$ grid points.
+While evaluating the derivative in, say, the x-direction,
+the computational domain is thought of as
+a collection of \emph{grid lines}---each consisting of
+$nx$ grid points---oriented in the x-direction.
+For each grid line,
+we may assemble a tridiagonal system of size $nx$, as in
+Equation \ref{eqn:compact-tridiagonal-system}.
+Solving for the all derivatives in the $x$ direction then involves
+solving this tridiagonal system for $ny \times nz$ right hand sides.
+
+\subsection{Multi-GPU solver}
+
+For multiple GPUs, we use the algorithm
+proposed by Mattor et al. \cite{mattor1995algorithm}.
+Here, we distribute the tridiagonal system among
+$p$ procesors.
+We solve for the ``local''  
+portion of the global tridiagonal system
+by first solving the following three ``local''
+tridiagonal systems:
+
+\begin{align}
+& \begin{bmatrix}
+b_1^p & c_1^p \\
+a_2^p & b_2^p & c_2^p \\
+      & a_3^p & b_3^p & c_3^p \\
+      &       & a_4^p & b_4^p & c_4^p \\
+      &       &       &       &  \ddots & c_{m-1}^p\\
+      &       &       &       &     a_{m}^p  & b_{m}^p
+\end{bmatrix}
+\begin{bmatrix}
+x_{r,1}^p \\
+x_{r,2}^p \\
+x_{r,3}^p \\
+x_{r,4}^p \\
+\vdots \\
+x_{r,m}^p
+\end{bmatrix}
+=
+\begin{bmatrix}
+r_1^p \\
+r_2^p \\
+r_3^p \\
+r_4^p \\
+\vdots \\
+r_m^p
+\end{bmatrix} & \label{eqn:primary-system} \\
+%
+%
+%
+& \begin{bmatrix}
+b_1^p & c_1^p \\
+a_2^p & b_2^p & c_2^p \\
+      & a_3^p & b_3^p & c_3^p \\
+      &       & a_4^p & b_4^p & c_4^p \\
+      &       &       &       &  \ddots & c_{m-1}^p\\
+      &       &       &       &     a_{m}^p  & b_{m}^p
+\end{bmatrix}
+\begin{bmatrix}
+u_1^p \\
+u_2^p \\
+u_3^p \\
+u_4^p \\
+\vdots \\
+u_m^p
+\end{bmatrix}
+=
+\begin{bmatrix}
+-a_1^p \\
+0 \\
+0 \\
+0 \\
+\vdots \\
+0
+\end{bmatrix} & \label{eqn:secondary-system-1} \\
+%
+%
+%
+& \begin{bmatrix}
+b_1^p & c_1^p \\
+a_2^p & b_2^p & c_2^p \\
+      & a_3^p & b_3^p & c_3^p \\
+      &       & a_4^p & b_4^p & c_4^p \\
+      &       &       &       &  \ddots & c_{m-1}^p\\
+      &       &       &       &     a_{m}^p  & b_{m}^p
+\end{bmatrix}
+\begin{bmatrix}
+l_1^p \\
+l_2^p \\
+l_3^p \\
+l_4^p \\
+\vdots \\
+l_m^p
+\end{bmatrix}
+=
+\begin{bmatrix}
+0 \\
+0 \\
+0 \\
+0 \\
+\vdots \\
+-c_m^p
+\end{bmatrix} & \label{eqn:secondary-system-2}
+\end{align}
+
+and additionally, the following ``reduced'' system:
+
+\begin{align} \label{eqn:reduced-system}
+&
+\begin{bmatrix}
+l^1_m & -1 \\
+-1    & u^2_1 & l^2_1 \\
+      & u^2_m & l^2_m & -1 \\
+      &       & -1    & u^3_1 & l^3_1 \\
+      &       &       & u^3_m & l^3_n  & -1 \\
+      &       &       &       & \ddots & \ddots & \ddots \\
+      &       &       &       &        & -1     & u^P_1
+\end{bmatrix}
+\begin{bmatrix}
+\beta^1 \\
+\alpha^2 \\
+\beta^2 \\
+\alpha^3 \\
+\beta^3 \\
+\vdots \\
+\alpha^P
+\end{bmatrix}
+=
+\begin{bmatrix}
+x_{r,m}^1 \\
+x_{r,1}^2 \\
+x_{r,m}^2 \\
+x_{r,1}^3 \\
+x_{r,m}^3 \\
+\vdots \\
+x_{r,1}^P \\
+\end{bmatrix}
+&
+\end{align}
+
+We call the system in Equation \ref{eqn:primary-system}
+the ``primary'' system, and the systems in
+Equations
+\ref{eqn:secondary-system-1} and \ref{eqn:secondary-system-2}
+the ``secondary'' systems.
+The local part of the solution to the ``global'' tridiagonal system
+is obtained as a linear combination of
+the solutions to the primary and secondary sytems:
+
+\begin{equation}
+    \boldsymbol{x}^p = \boldsymbol{x}_r^p + \
+        \alpha^p \boldsymbol{u}^p + \beta^p \boldsymbol{l}^p
+    \label{eqn:sum-of-systems}
+\end{equation}
+
+Where $\alpha^p$ and $\beta^p$ are obtained
+from the solution of the reduced system.
+
+\subsection{GPU implementation}
+
+\begin{enumerate}
+    \item Computation of the RHS (requires communication)
+    \item Solving the primary and secondary systems
+    \item Constructing and solving the reduced systems,
+        and scattering the paramters $\alpha^p$ and $\beta^p$ (requires communication)
+    \item Summing the solutions (Equation \ref{eqn:sum-of-systems})
+\end{enumerate}
+
+Each of these is performed on the GPU,
+one of the primary concerns is to eliminate
+CPU-GPU transfers as much as possible in the solution process.
+
+The right-hand side computation involves stencil computation,
+which requires communication at the boundaries.
+Communication of non-contiguous MPI
+data types is expensive on the GPU,
+so we have dedicated
+contiguous halo arrays for performing swaps in each direction
+
+\subsection{Results}
+
+% Time for dfdx, dfdy, dfdz at various problem sizes
+% and nprocs
+
+% Breakdown - time allocated for each step
+
+% Weak scaling
+
+% Strong scaling
+
+\pagebreak[4]
 \section*{References}

### 2015-11-04 paper-draft/results/scripts/bench/bench.py
update compact results

@@ -28,16 +28,15 @@ matrix = np.vstack([intel_mkl_1,
                     intel_mkl_16,
                     cusparse,
                     neato_shared])
-mins = np.min(matrix, axis=0)
 
 # normalize timings:
-intel_mkl_1 = intel_mkl_1/mins
-intel_mkl_2 = intel_mkl_2/mins
-intel_mkl_4 = intel_mkl_4/mins
-intel_mkl_8 = intel_mkl_8/mins
-intel_mkl_16 = intel_mkl_16/mins
-cusparse = cusparse/mins
-neato_shared = neato_shared/mins
+intel_mkl_1 = intel_mkl_1/neato_shared
+intel_mkl_2 = intel_mkl_2/neato_shared
+intel_mkl_4 = intel_mkl_4/neato_shared
+intel_mkl_8 = intel_mkl_8/neato_shared
+intel_mkl_16 = intel_mkl_16/neato_shared
+cusparse = cusparse/neato_shared
+neato_shared = neato_shared/neato_shared
 
 ind = np.arange(7)*5
 fig, ax = plt.subplots()
@@ -108,16 +107,15 @@ matrix = np.vstack([intel_mkl_1,
                     intel_mkl_16,
                     cusparse,
                     neato_shared])
-mins = np.min(matrix, axis=0)
 
 # normalize timings:
-intel_mkl_1 = intel_mkl_1/mins
-intel_mkl_2 = intel_mkl_2/mins
-intel_mkl_4 = intel_mkl_4/mins
-intel_mkl_8 = intel_mkl_8/mins
-intel_mkl_16 = intel_mkl_16/mins
-cusparse = cusparse/mins
-neato_shared = neato_shared/mins
+intel_mkl_1 = intel_mkl_1/neato_shared
+intel_mkl_2 = intel_mkl_2/neato_shared
+intel_mkl_4 = intel_mkl_4/neato_shared
+intel_mkl_8 = intel_mkl_8/neato_shared
+intel_mkl_16 = intel_mkl_16/neato_shared
+cusparse = cusparse/neato_shared
+neato_shared = neato_shared/neato_shared
 
 ind = np.arange(5)*5
 fig, ax = plt.subplots()

### 2015-11-04 paper-draft/results/scripts/compact/strong-scaling.py

@@ -1,0 +1,52 @@
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import os
+import sys
+
+plt.style.use('bmh')
+
+ngpus = np.array([1, 2, 3])
+
+
+# prob size:  256
+dfdx = np.array([
+    12.5, 5.13, 2.7])
+dfdy = np.array([
+    17.2, 5.17, 2.87])
+dfdz = np.array([
+    19.2, 5.93, 2.803])
+
+fig, ax = plt.subplots()
+ax.set_yscale("log", nonposy='clip')
+ax.set_xscale("linear")
+ax.plot(dfdx, label='dfdx')
+ax.plot(dfdy, label='dfdy')
+ax.plot(dfdz, label='dfdz')
+ax.set_xticks(range(len(ngpus)))
+ax.set_xticklabels(['1', '8', '64'])
+ax.legend()
+ax.set_xlabel('Number of GPUs')
+ax.set_ylabel('Time to solve in ms')
+plt.savefig('strong-scaling-256.eps')
+
+# prob size: 512
+dfdx = np.array([
+    114.0, 22.8, 6.2])
+dfdy = np.array([
+    155.0, 26.3, 5.7])
+dfdz = np.array([
+    166.0, 29.1, 6.3])
+
+fig, ax = plt.subplots()
+ax.set_yscale("log", nonposy='clip')
+ax.set_xscale("linear")
+ax.plot(dfdx, label='dfdx')
+ax.plot(dfdy, label='dfdy')
+ax.plot(dfdz, label='dfdz')
+ax.set_xticks(range(len(ngpus)))
+ax.set_xticklabels(['1', '8', '64'])
+ax.legend()
+ax.set_xlabel('Number of GPUs')
+ax.set_ylabel('Time to solve in ms')
+plt.savefig('strong-scaling-512.eps')

### 2015-11-04 paper-draft/results/scripts/compact/weak-scaling.py

@@ -1,0 +1,51 @@
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import os
+import sys
+
+plt.style.use('bmh')
+
+ngpus = np.array([1, 4, 27, 64])
+
+# size per proc = 128
+dfdx = np.array([
+    2.117, 5.12, 5.49, 6.20])
+dfdy = np.array([
+    2.66, 5.15, 5.53, 5.73])
+dfdz = np.array([
+    2.96, 5.53, 5.93, 6.32])
+
+fig, ax = plt.subplots()
+ax.set_yscale("log", nonposy='clip')
+ax.set_xscale("linear")
+ax.plot(dfdx, label='dfdx')
+ax.plot(dfdy, label='dfdy')
+ax.plot(dfdz, label='dfdz')
+ax.set_xticks(range(len(ngpus)))
+ax.set_xticklabels(['1', '8', '64'])
+ax.legend()
+ax.set_xlabel('Number of GPUs')
+ax.set_ylabel('Time to solve in ms')
+plt.savefig('weak-scaling-256.eps')
+
+# prob size: 256
+dfdx = np.array([
+    12.41, 22.9, 24.5, 24.4])
+dfdy = np.array([
+    17.30, 26.7, 38.5, 30.0])
+dfdz = np.array([
+    19.30, 28.8, 40.0, 40.0])
+
+fig, ax = plt.subplots()
+ax.set_yscale("log", nonposy='clip')
+ax.set_xscale("linear")
+ax.plot(dfdx, label='dfdx')
+ax.plot(dfdy, label='dfdy')
+ax.plot(dfdz, label='dfdz')
+ax.set_xticks(range(len(ngpus)))
+ax.set_xticklabels(['1', '8', '64'])
+ax.legend()
+ax.set_xlabel('Number of GPUs')
+ax.set_ylabel('Time to solve in ms')
+plt.savefig('weak-scaling-512.eps')

### 2015-11-07 paper-draft/paper.tex
fix formatting issues

@@
     systems of the form $A\bm{x} = \bm{d}$,
-    where $A$ is \emph{near-Toeplitz}.
+    where $A$ is \emph{near-Toeplitz} tridiagonal.
     These systems appear in numerical schemes
@@
     and improve memory access.
-    Our solver is able to achieve up to 5x speedup
-    over an Intel MKL solver running on a single core,
-    and 1.5x-11x speedup
-    over the NVIDIA CUSPARSE solver.
-    For larger problem sizes,
-    our solver is able to outperform
-    CUSPARSE as well as Intel MKL (on up to 16 cores).
+    Our solver is able to outperform
+    the NVIDIA CUSPARSE and
+    multi-core Intel MKL tridiagonal solvers
+    (on up to 16 cores).
     We also describe the implementation of a multi-GPU solver
-    using MPI and CUDA.
-    We describe two applications:
-    2-D ADI applied to the heat equation on a single GPU,
-    and computation of
-    3-D compact finite differences on multiple GPUs.
-
+    using MPI and CUDA,
+    and show scalability for up to 64 GPUs
+    for a 3-D compact finite difference application.
 \end{abstract}
@@
 \end{equation}
-
-Where, in general
+%
+where, in general
 \begin{align*}
@@
 the known structure of the tridiagonal matrix.
-We describe the implementation of a GPU solver that
-uses this information to
-reduce the number of computations
-and memory accesses.
-The resulting solver is thus able to perform twice as fast as
-the equivalent NVIDIA CUSPARSE solver,
-and an Intel MKL solver running with 16 independent cores.
+The contribution of this paper is twofold:
+First, we describe the implementation of a GPU solver
+for the specific case of near-Toeplitz systems
+that uses the known structure to reduce
+computations and memory accesses.
+Secondly, we describe the implementation for
+for multi-GPU clusters using MPI and CUDA,
+and apply it to the evaluation of compact finite differences.
 
@@
 and compares it with other implementations.
+Section \ref{sec-compact-finite-differences}
+describes the implementation on multiple GPUs,
+and application to the solution of compact finite differences.
 
@@
 Cyclic reduction is applicable to the solution of
-general linear systems of the form $Ax = d$,
+general linear systems of the form $A\bm{x} = \bm{d}$,
 where $A$ is a tridiagonal matrix with diagonals
-$a$, $b$ and $c$
+$\bm{a}$, $\bm{b}$ and $\bm{c}$
 (Equation \ref{eqn:general-tridiagonal-system}).
 The algorithm consists of two phases:
-\emph{forward reduction} and \emph{backward substitution}.
+\emph{forward reduction} and \emph{backward substitution}
+\ref{fig:cyclic-reduction}.
 In the forward reduction phase,
@@
 \end{align}
-
+%
 where,
@@
 \end{align}
-
+%
 For the last index $i=n$,
@@
 \end{align}
-
+%
 And for $i=1$, the backward substitution step is instead:
@@
 Writing applications that effectively exploit the GPU
-requires a thorough understanding of the underlying architecture.
+requires an understanding of the underlying architecture.
 We provide the pertinent details here,
@@
 \end{itemize}
-
+%
 we observe that the resulting coefficients
@@
 \end{equation}
-
+%
 And the $m^{th}$ backward substitution step is:
@@
 \end{equation}
-
-Where $a^m$, $b^m$, $c^m$, $k_1^m$ and $k_2^m$
+%
+where $a^m$, $b^m$, $c^m$, $k_1^m$ and $k_2^m$
 are the appropriate values
 from the precomputed coefficient arrays.
+In practice,
+the backward substitution step
+can overwrite the right-hand side vector $\bm{d}$
+with the solution values.
 
@@
 and equations or \emph{indices} of a system
-to individual threads within a block.
+to individual threads within a block
+(Figure \ref{fig:gpu-mapping}).
 
@@
 \end{center}
-\caption{Shared memory (top) and Global memory (bottom) implementations---
-    multiple kernel launches allow us to dedicate appropriate number
-    of threads per block in the global memory approach}
+\caption{Shared memory (top) and Global memory (bottom) implementations}
 \label{fig:global-and-shared}
@@
 
-This implementation works entirely on the GPU's global memory.
-Each right-hand side is mapped to a single thread block.
-Each block executes two kernels,
-\texttt{forwardReduction} and \texttt{backSubstitution},
-which are called each called $log_2(n)-1$ times.
+This implementation works entirely on the GPU's global memory,
+Here, we define two kernels - one for the forward reduction step,
+and the other for the backward substitution step.
+Each kernel is called $log_2(n)-1$ times.
 An extra call to the
-forward reduction kernel additionally performs
+forward reduction kernel performs
 the two-by-two solve.
-Exit from a kernel marks a global synchronization point,
-so no explicit synchronization is performed in the kernels.
-Because we launch a kernel for each
-forward reduction and back substitution step,
-we can specify the appropriate number of threads per block---$n/2$
-threads per block for the first forward reduction step,
-$n/4$ for the second, and so on (Figure \ref{fig:global-and-shared}).
-This guarantees that
-all threads within a block are always active.
+At each step,
+the size of the grid block is determined by the \emph{stride}
+between elements accessed at that step.
+For the forward reduction phase,
+we use $n/2$ threads per block for the first step,
+$n/4$ threads for the second step, and so on.
+The pattern is reversed for the backward substitution phase,
+beginning with 2 threads per block for the first step.
+Although this ensures that there are no inactive threads
+at any stage,
+the occupancy of the GPU is still very low
+during the end of forward reduction
+and the end of backward substitution.
+
 The precomputed coefficient arrays and right hand
-are accessed by the kernels from global memory directly.
-Because the coefficients
-$k_{1,i}$ and $k_{2,i}$
-required in the forward reduction step
-(Equation \ref{eqn:precomputed-forward-reduction-step})
-are precomputed and available in global memory,
-we do not compute
-$a_i^\prime$, $b_i^\prime$, $c_i^\prime$,
-$k_{1,i}$ and $k_{2,i}$.
-This reduces the number of uncoalesced accesses to global memory.
-In the backward substitution steps
-(Equation \ref{eqn:precomputed-backward-substitution-step}),
-the coefficients $a^m$, $b^m$ and $c^m$
-are accessed from global memory without coalescing problems.
+are accessed by the kernels from global memory.
+The kernel suffers from
+strided memory access for the right-hand side,
+but the precomputed coefficient values are accessed
+without major coalescing problems.
+Further, by precomputing the forward reduction coefficients,
+we greatly reduce the number of computations
+(and thus the number of uncoalesced memory accesses).
 
@@
 In the shared memory approach (Figure \ref{fig:global-and-shared}),
-we again map every right-hand side to a thread block.
-Each thread block executes a single kernel \emph{once},
-which performs forward reduction,
-backward substitution, and the two-by-two solve.
-Shared memory sized $n/2$ is allocated by each block.
+we launch a single kernel to perform the entire
+cyclic reduction solve.
+The kernel is launched with $n/2$ threads per block.
+Each thread block allocated a block of shared memory of size $n/2$.
 Each thread performs the first reduction step
@@
 are accessed from shared memory,
-avoiding uncoalesced global memory accesses.
+avoiding the uncoalesced global memory accesses
+seen in the global memory implementation.
 In each back substitution step,
@@
 
+The shared memory implementation suffers from two major issues:
+first, the number of active threads is halved at each forward reduction step,
+(and subsequently doubled at each backward reduction step).
+Synchronization between threads of a block is necessary at each step.
+Thus, a significant portion of the kernel execution time is
+spent by idle threads waiting for active threads to complete execution.
+Secondly, the strided access to the
+right-hand side values leads to \emph{bank-conflicts}.
+However, the number of bank conflicts is significantly smaller
+than in cyclic reduction implementations for general tridiagonal systems,
+due the reduced number of computations performed.
+
 %----------------------------------------------------------------------%
@@
 %----------------------------------------------------------------------%
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=300pt]{fig/bench-2d.eps}
+\caption{Relative solver performance for 2-D problems}
+\label{fig:bench-2d}
+\end{center}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=300pt]{fig/bench-3d.eps}
+\caption{Relative solver performance for 3-D problems}
+\label{fig:bench-3d}
+\end{center}
+\end{figure}
 
@@
 
-\begin{figure}[!htb]
+\begin{figure}[h!]
+\begin{center}
 \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
@@
 \label{fig:global-vs-shared-2d}
-        \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
+\end{center}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=1.0\linewidth]{fig/global-vs-shared-3d.eps}
 \caption{Comparison of global memory and shared memory
@@
 \label{fig:global-vs-shared-3d}
+\end{center}
 \end{figure}
@@
 The Intel MKL solver can use multiple threads
-to solve several systems simaltaneously.
+to solve several systems simultaneously.
 The CUSPARSE solver uses a hybrid algorithm composed
@@
 by the timing for the fastest solver for that problem size.
-Table \ref{fig:bench} shows the timings of the various
+Table \ref{table:bench} shows the timings of the various
 solvers to solve different problem sizes.
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[width=300pt]{fig/bench-2d.eps}
-\caption{Relative solver performance for 2-D problems}
-\label{fig:bench-2d}
-\end{center}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[width=300pt]{fig/bench-3d.eps}
-\caption{Relative solver performance for 3-D problems}
-\label{fig:bench-3d}
-\end{center}
-\end{figure}
 
@@
 
+%----------------------------------------------------------------------
 \section{Application: compact finite difference on multiple GPUs}
+\label{sec-compact-finite-differences}
+%----------------------------------------------------------------------
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{img/computational-domain.eps}
+\caption{Computational domain in 3-D}
+\label{fig:computational-domain}
+\end{center}
+\end{figure}
 
@@
 \end{equation}
-
+%
 The right boundary ($f^{\prime}_{n}$) is treated by utilizing
@@
 \end{equation}
-
+%
 This boundary treatment maintains the tridiagonal identity
-of the syste to be solved, which then has the form:
+of the system to be solved, which then has the form:
 
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[height=200pt]{img/computational-domain.eps}
-\end{center}
-\caption{Computational domain in 3-D}
-\label{fig:computational-domain}
-\end{figure}
-
 The computational domain (Figure \ref{fig:computational-domain})
@@
 Here, we distribute the tridiagonal system among
-$p$ procesors.
+$p$ processors.
 We solve for the ``local''  
@@
 \end{align}
-
+%
 and additionally, the following ``reduced'' system:
@@
 is obtained as a linear combination of
-the solutions to the primary and secondary sytems:
+the solutions to the primary and secondary systems:
 
@@
 
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=200pt]{img/constructing-reduced-system.eps}
+\caption{Construction of the reduced system and scattering of parameters}
+\label{fig:reduced-system}
+\end{center}
+\end{figure}
+
+In this section we describe the implementation of
+a CUDA+MPI solver for compact finite differences.
+The algorithm 
+consists of the following sequence of steps:
+
 \begin{enumerate}
@@
 \item Constructing and solving the reduced systems,
-        and scattering the paramters $\alpha^p$ and $\beta^p$ (requires communication)
-    \item Summing the solutions (Equation \ref{eqn:sum-of-systems})
+    and scattering the parameters $\alpha^p$ and $\beta^p$ (requires communication)
+\item Summing the above solutions (Equation \ref{eqn:sum-of-systems})
 \end{enumerate}
@@
 Each of these is performed on the GPU,
-one of the primary concerns is to eliminate
+and one of the primary concerns is to eliminate
 CPU-GPU transfers as much as possible in the solution process.
 
-The right-hand side computation involves stencil computation,
-which requires communication at the boundaries.
-Communication of non-contiguous MPI
-data types is expensive on the GPU,
-so we have dedicated
-contiguous halo arrays for performing swaps in each direction
+The right-hand side evaluation is a pointwise stencil computation,
+which requires communication of the function values at the boundaries.
+This is achieved using a halo-swapping technique
+with dedicated contiguous halo arrays for each subdomain,
+as communication of non-contiguous MPI
+data types is expensive on the GPU.
+After the halo information is exchanged,
+a pointwise stencil kernel computes the right-hand side values.
+
+The secondary systems are set up and solved only once
+as it is dependent only on the
+coefficients of the tridiagonal matrix.
+As the computation is offline,
+any convenient method can be used.
+The primary system is solved at every time step
+for different right hand sides,
+and is evaluated using the NEATO solver.
+
+The set up and solution of the reduced system
+requires global communication of the boundary information,
+which may be associated with a high overhead.
+While several techniques have been proposed to implement
+this type of communication pattern efficiently,
+we find it sufficient for moderately large problems
+to communicate all information to rank 0,
+using \texttt{MPI\_Gather}, and distribute the results
+back using \texttt{MPI\_Scatter} (Figure \ref{fig:reduced-system}).
+
+Finally, the summing of the solutions
+is a pointwise operation that is easily implemented
+on the GPU.
+
+For evaluation of compact finite differences in different
+coordinate directions,
+a local permutation of the data is performed.
 
@@
 
-% Time for dfdx, dfdy, dfdz at various problem sizes
-% and nprocs
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=300pt]{fig/strong-scaling-256.eps}
+\caption{Strong scaling for multi-GPU compact finite difference, problem size: $256^3$}
+\label{fig:strong-scaling-256}
+\end{center}
+\end{figure}
 
-% Breakdown - time allocated for each step
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=300pt]{fig/strong-scaling-512.eps}
+\caption{Strong scaling for multi-GPU compact finite difference, problem size: $512^3$}
+\label{fig:strong-scaling-512}
+\end{center}
+\end{figure}
 
-% Weak scaling
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=300pt]{fig/weak-scaling-128.eps}
+\caption{Weak scaling for multi-GPU compact finite difference, problem size: $128^3$}
+\label{fig:weak-scaling-128}
+\end{center}
+\end{figure}
 
-% Strong scaling
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=300pt]{fig/weak-scaling-256.eps}
+\caption{Weak scaling for multi-GPU compact finite difference, problem size: $256^3$}
+\label{fig:weak-scaling-256}
+\end{center}
+\end{figure}
 

### 2015-11-07 paper-draft/results/scripts/compact/weak-scaling.py

@@ -23,19 +23,22 @@ ax.plot(dfdx, label='dfdx')
 ax.plot(dfdy, label='dfdy')
 ax.plot(dfdz, label='dfdz')
 ax.set_xticks(range(len(ngpus)))
-ax.set_xticklabels(['1', '8', '64'])
+ax.set_xticklabels(['1', '8', '27', '64'])
 ax.legend()
 ax.set_xlabel('Number of GPUs')
 ax.set_ylabel('Time to solve in ms')
-plt.savefig('weak-scaling-256.eps')
+plt.savefig('weak-scaling-128.eps')
+
+
+ngpus = np.array([1, 4, 64])
 
 # prob size: 256
 dfdx = np.array([
-    12.41, 22.9, 24.5, 24.4])
+    12.41, 22.9, 24.4])
 dfdy = np.array([
-    17.30, 26.7, 38.5, 30.0])
+    17.30, 26.7, 30.0])
 dfdz = np.array([
-    19.30, 28.8, 40.0, 40.0])
+    19.30, 28.8, 40.0])
 
 fig, ax = plt.subplots()
 ax.set_yscale("log", nonposy='clip')
@@ -48,4 +51,4 @@ ax.set_xticklabels(['1', '8', '64'])
 ax.legend()
 ax.set_xlabel('Number of GPUs')
 ax.set_ylabel('Time to solve in ms')
-plt.savefig('weak-scaling-512.eps')
+plt.savefig('weak-scaling-256.eps')

### 2015-11-08 paper-draft/paper.tex
change Figure to Fig.

@@
 \emph{forward reduction} and \emph{backward substitution}
-\ref{fig:cyclic-reduction}.
+(Fig. \ref{fig:cyclic-reduction}).
 In the forward reduction phase,
@@
 and threads within a block 
-are assigned to equations, or \emph{indices}.
-(Figure \ref{fig:gpu-mapping})
+are assigned to equations, or \emph{indices}
+(Fig. \ref{fig:gpu-mapping}).
 During the forward reduction phase,
@@
 $a_i^\prime$, $b_i^\prime$ and $c_i^\prime$.
-In practice, the coefficients are updated \emph{in-place}
-(Figure \ref{fig:forward-reduction-step} shows the updates
+In practice, the coefficients are updated \emph{in-place}.
+Figure \ref{fig:forward-reduction-step} shows the updates
 to the coefficient array $\bm{b}$ in the first forward reduction step).
@@
 of particular interest due to the lower storage costs.
-In Figure \ref{fig:forward-reduction-step},
+In Fig. \ref{fig:forward-reduction-step},
 we list the updates to
@@
 and store them in separate arrays.
-In Figure \ref{fig:cyclic-reduction-precomputing},
+In Fig. \ref{fig:cyclic-reduction-precomputing},
 it is seen that the space required for storing each of the
@@
 to individual threads within a block
-(Figure \ref{fig:gpu-mapping}).
+(Fig. \ref{fig:gpu-mapping}).
 
@@
 
-In the shared memory approach (Figure \ref{fig:global-and-shared}),
+In the shared memory approach (Fig. \ref{fig:global-and-shared}),
 we launch a single kernel to perform the entire
@@
 
-In Figures \ref{fig:global-vs-shared-2d} and \ref{fig:global-vs-shared-3d},
+In Figs. \ref{fig:global-vs-shared-2d} and \ref{fig:global-vs-shared-3d},
 we report the performance of the two solvers for the case
@@
 
-In Figures \ref{fig:bench-2d} and \ref{fig:bench-3d},
+In Fig. \ref{fig:bench-2d} and \ref{fig:bench-3d},
 we provide the relative performance of
@@
 
-The computational domain (Figure \ref{fig:computational-domain})
+The computational domain (Fig. \ref{fig:computational-domain})
 is a structured grid comprised of
@@
 using \texttt{MPI\_Gather}, and distribute the results
-back using \texttt{MPI\_Scatter} (Figure \ref{fig:reduced-system}).
+back using \texttt{MPI\_Scatter} (Fig. \ref{fig:reduced-system}).
 

### 2015-11-09 paper-draft/paper.tex
add full stop to captions

@@
 \end{center}
-\caption{Cyclic reduction}
+\caption{Cyclic reduction.}
 \label{fig:cyclic-reduction}
@@
 \end{center}
-\caption{Updating $\bm{b}$ in the first forward reduction step}
+\caption{Updating $\bm{b}$ in the first forward reduction step.}
 \label{fig:forward-reduction-step}
@@
 systems are mapped to blocks and
-indices are mapped to individual threads}
+indices are mapped to individual threads.}
 \end{figure}
@@
 \caption{Storing forward reduction coefficients
-    from each reduction step $m$}
+    from each reduction step $m$.}
 \label{fig:cyclic-reduction-precomputing}
@@
 \end{center}
-\caption{Shared memory (top) and Global memory (bottom) implementations}
+\caption{Thread activity in shared memory (top) and global memory (bottom) implementations.}
 \label{fig:global-and-shared}
@@
 \includegraphics[width=300pt]{fig/bench-2d.eps}
-\caption{Relative solver performance for 2-D problems}
+\caption{Relative solver performance for 2-D problems.}
 \label{fig:bench-2d}
@@
 \includegraphics[width=300pt]{fig/bench-3d.eps}
-\caption{Relative solver performance for 3-D problems}
+\caption{Relative solver performance for 3-D problems.}
 \label{fig:bench-3d}
@@
 \caption{Comparison of global memory and shared
-    memory implementations of NEATO (2D problems)}
+    memory implementations of NEATO (2D problems).}
 \label{fig:global-vs-shared-2d}
@@
 \caption{Comparison of global memory and shared memory
-    implementations of NEATO (2D problems)}
+    implementations of NEATO (3D problems).}
 \label{fig:global-vs-shared-3d}
@@
 \centering
-\caption{Performance of Intel MKL, CUSPARSE and NEATO solvers}
+\caption{Performance of Intel MKL, CUSPARSE and NEATO solvers.}
 \label{table:bench}
@@
 \includegraphics[height=200pt]{img/computational-domain.eps}
-\caption{Computational domain in 3-D}
+\caption{Computational domain in 3-D.}
 \label{fig:computational-domain}
@@
 \includegraphics[width=200pt]{img/constructing-reduced-system.eps}
-\caption{Construction of the reduced system and scattering of parameters}
+\caption{Construction of the reduced system and scattering of parameters.}
 \label{fig:reduced-system}
@@
 \includegraphics[width=300pt]{fig/strong-scaling-256.eps}
-\caption{Strong scaling for multi-GPU compact finite difference, problem size: $256^3$}
+\caption{Strong scaling for multi-GPU compact finite difference, problem size: $256^3$.}
 \label{fig:strong-scaling-256}
@@
 \includegraphics[width=300pt]{fig/strong-scaling-512.eps}
-\caption{Strong scaling for multi-GPU compact finite difference, problem size: $512^3$}
+\caption{Strong scaling for multi-GPU compact finite difference, problem size: $512^3$.}
 \label{fig:strong-scaling-512}
@@
 \includegraphics[width=300pt]{fig/weak-scaling-128.eps}
-\caption{Weak scaling for multi-GPU compact finite difference, problem size: $128^3$}
+\caption{Weak scaling for multi-GPU compact finite difference, problem size: $128^3$ per process.}
 \label{fig:weak-scaling-128}
@@
 \includegraphics[width=300pt]{fig/weak-scaling-256.eps}
-\caption{Weak scaling for multi-GPU compact finite difference, problem size: $256^3$}
+\caption{Weak scaling for multi-GPU compact finite difference, problem size: $256^3$ per process.}
 \label{fig:weak-scaling-256}

### 2015-11-09 paper-draft/paper.tex
replace Equation with Eq.

@@
 $\bm{a}$, $\bm{b}$ and $\bm{c}$
-(Equation \ref{eqn:general-tridiagonal-system}).
+[Eq. (\ref{eqn:general-tridiagonal-system})].
 The algorithm consists of two phases:
@@
 $n/2$ equations in $n/2$ unknowns
-(Equations \ref{eqn:forward-reduction-1} - \ref{eqn:forward-reduction-4}).
+[Eqs. (\ref{eqn:forward-reduction-1}) - (\ref{eqn:forward-reduction-4})].
 The process is repeated until a system of
@@
 substituting the known values of $x_{i-1}$ and $x_{i+1}$
-(Equation \ref{eqn:backward-substitution}) .
+[Eq. (\ref{eqn:backward-substitution})] .
 
@@
 Each thread performs the first reduction step
-(Equation \ref{eqn:precomputed-forward-reduction-step})
+[Eq. (\ref{eqn:precomputed-forward-reduction-step})]
 by accessing the required values
@@
 we may assemble a tridiagonal system of size $nx$, as in
-Equation \ref{eqn:compact-tridiagonal-system}.
+[Eq. (\ref{eqn:compact-tridiagonal-system})].
 Solving for the all derivatives in the $x$ direction then involves
@@
 
-We call the system in Equation \ref{eqn:primary-system}
+We call the system in [Eq. (\ref{eqn:primary-system})]
 the ``primary'' system, and the systems in
-Equations
-\ref{eqn:secondary-system-1} and \ref{eqn:secondary-system-2}
+[Eqs.
+(\ref{eqn:secondary-system-1}) and (\ref{eqn:secondary-system-2})]
 the ``secondary'' systems.
@@
     and scattering the parameters $\alpha^p$ and $\beta^p$ (requires communication)
-\item Summing the above solutions (Equation \ref{eqn:sum-of-systems})
+\item Summing the above solutions [Eq. (\ref{eqn:sum-of-systems})]
 \end{enumerate}

### 2015-11-09 paper-draft/paper.tex
update some images

@@
 Secondly, we describe the implementation for
-for multi-GPU clusters using MPI and CUDA,
+multi-GPU clusters using MPI and CUDA,
 and apply it to the evaluation of compact finite differences.
@@
 Section \ref{sec:preliminaries}
-describes algorithms for solving tridiagonal systems on GPUs.
-The cyclic reduction algorithm in particular, is discussed in detail,
-and issues with its implementation on GPUs is discussed.
+reviews the cyclic reduction algorithm,
+and its implementation on GPUs.
+We also provide some relevant details about GPU architecture here.
 Section \ref{sec:proposed-algorithm}
@@
 substituting the known values of $x_{i-1}$ and $x_{i+1}$
-[Eq. (\ref{eqn:backward-substitution})] .
+[Eq. (\ref{eqn:backward-substitution})].
 
@@
 \begin{center}
-\includegraphics[height=200pt]{img/forward-reduction-step.eps}
-\end{center}
-\caption{Updating $\bm{b}$ in the first forward reduction step.}
-\label{fig:forward-reduction-step}
+\includegraphics[height=200pt]{img/gpu-mapping.eps}
+\end{center}
+\caption{Mapping work to blocks and threads:
+systems are mapped to blocks and
+indices are mapped to individual threads.}
+\label{fig:gpu-mapping}
 \end{figure}
@@
 
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[height=200pt]{img/forward-reduction-step.eps}
+\end{center}
+\caption{Updating $\bm{b}$ in the first forward reduction step.}
+\label{fig:forward-reduction-step}
+\end{figure}
+
 %----------------------------------------------------------------------%
@@
 %----------------------------------------------------------------------%
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[height=200pt]{img/gpu-mapping.eps}
-\end{center}
-\label{fig:gpu-mapping}
-\caption{Mapping work to blocks and threads:
-systems are mapped to blocks and
-indices are mapped to individual threads.}
-\end{figure}
 

### 2015-11-11 paper-draft/paper.tex
updates

@@
 for $N_{rhs}$ right hand sides.
+Our implementation uses the NVIDIA CUDA platform,
 We develop two approaches based on the common idea
@@
 and the other working directly with global memory.
-
-Following the algorithm described in \ref{sec:proposed-algorithm},
+The GPU kernels for both implementations are relatively
+straightforward and compact,
+spanning no more than 100 lines.
+
+Following the algorithm described in Section \ref{sec:proposed-algorithm},
 the forward reduction coefficients
@@
 The right hand sides are stored in a single
-contiguous region global memory.
+contiguous region in global memory.
 In both approaches,
@@
 In the final step,
-shared memory is filled completeley
+shared memory is filled completely
 with the even-indexed solution values.
@@
 \includegraphics[width=300pt]{fig/bench-2d.eps}
-\caption{Relative solver performance for 2-D problems.}
+\caption{Relative solver performance for 2-D problems. Relative time defined as:}
+$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
 \label{fig:bench-2d}
@@
 \includegraphics[width=300pt]{fig/bench-3d.eps}
-\caption{Relative solver performance for 3-D problems.}
+\caption{Relative solver performance for 3-D problems. Relative time defined as:}
+$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
 \label{fig:bench-3d}
@@
 (\texttt{dgtsv} and \texttt{dgtsvStridedBatch} respectively).
-The CPU code is compiled with the Intel C compiler
-and run on up to 16 cores of the
-Intel Xeon Processor E5-2670 v2 (25M Cache, 2.50 GHz).
+The MKL solver is based on Gaussian Elimination with partial pivoting,
+Boundary conditions may prevent
+the matrix from being symmetric and/or diagonally dominant,
+precluding the use of more specialized tridiagonal solvers.
+and the CUSPARSE solver uses a combination of
+Cyclic Reduction and Parallel Cyclic Reduction.
+Both solvers are optimized for the underlying architectures.
+
+The CPU code is compiled with the Intel C compiler (version 15.0),
+and run (with OpenMP support) on up to
+16 independent cores of the the same shared-memory node
+(one thread per core).
+The CPU is an
+Intel Xeon Processor E5-2670 v2 (2.50 GHz, 25 MB Smart Cache).
 GPU code is compiled with the CUDA toolkit (version 6.5.14),
@@
 NVIDIA Tesla K20 Accelerator.
-\texttt{-O2} level optimizations are turned on for both
-CPU and GPU code.
-
-\subsection{Global memory v/s shared memory solver performance}
+When measuring GPU performance (both CUSPARSE and NEATO),
+we do \emph{not} include the cost
+of data transfer between the CPU and GPU.
+\texttt{-O2} level compiler optimizations are turned on for both
+CPU and GPU code;
+no further optimization options are enabled in either case.
+Of course, we use double precision for all solvers.
+All timings reported are for the tridiagonal solution
+and are averaged over several (100) solves.
+
+\subsection{NEATO: global memory v/s shared memory performance}
 
@@
 For larger problem sizes,
-the reduced occupancy and increase in number of inactive threads
+the synchronization costs associated with inactive threads
 leads to poor shared memory performance.
 
-\subsection{Comparison with Intel MKL and CUSPARSE solvers}
+\subsection{Comparison of NEATO with Intel MKL and CUSPARSE solvers}
 
@@
 the NEATO shared memory implementation.
-The Intel MKL solver can use multiple threads
-to solve several systems simultaneously.
-The CUSPARSE solver uses a hybrid algorithm composed
-of Cyclic Reduction and Parallel Cyclic Reduction (CR+PCR).
 The relative performance for
@@
     = 
+    a\frac{f_{i+1} - f_{i-1}}{dx} + \
+    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
     c\frac{f_{i+3} - f_{i-3}}{dx} + \
-    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
-    a\frac{f_{i+1} - f_{i-1}}{dx} + \
     \hdots
@@
 
-For the particular scheme used, $\alpha = 1/4$,
-and $ a = 3/4 $.
+For the particular scheme used, $\beta=0$, $\alpha = 1/4$,
+$a = 3/4$, and $b = c = \hdots = 0$.
 It is easily noted that this leads to a tridiagonal system of equations,
@@
 This boundary treatment maintains the tridiagonal identity
-of the system to be solved, which then has the form:
+of the system to be solved, which then has the near-Toeplitz form:
 
@@
 
-\subsection{Computational domain}
-
-The computational domain (Fig. \ref{fig:computational-domain})
+\subsection{Computational domain and problem distribution}
+
+The computational domain we consider
+(Fig. \ref{fig:computational-domain})
 is a structured grid comprised of
@@
 we may assemble a tridiagonal system of size $nx$, as in
-[Eq. (\ref{eqn:compact-tridiagonal-system})].
+Eq. (\ref{eqn:compact-tridiagonal-system}).
 Solving for the all derivatives in the $x$ direction then involves
 solving this tridiagonal system for $ny \times nz$ right hand sides.
-
-\subsection{Multi-GPU solver}
-
-For multiple GPUs, we use the algorithm
-proposed by Mattor et al. \cite{mattor1995algorithm}.
-Here, we distribute the tridiagonal system among
-$p$ processors.
-We solve for the ``local''  
-portion of the global tridiagonal system
-by first solving the following three ``local''
-tridiagonal systems:
+On a single GPU,
+this is done easily using the NEATO solver.
+
+When the problem is too large to fit in a single GPU,
+it is necessary to \emph{partition} the domain
+among multiple \emph{processes} (GPUs).
+In this case, each GPU works on a \emph{chunk}, or \emph{subdomain}.
+For evaluating compact finite differences along a given direction,
+we need only consider the solution strategy for
+one line of subdomains.
+The other lines may be solved independently and concurrently
+in exactly the same manner.
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=350pt]{img/subdomains-grid-lines.eps}
+\caption{A line of subdomains and a single grid line composed of ``local'' grid lines.
+    Each subdomain is mapped to a single GPU, or \emph{process},
+    and each process is identified by its ``rank''.}
+\label{fig:subdomains-grid-lines}
+\end{center}
+\end{figure}
+
+Figure \ref{fig:subdomains-grid-lines}
+shows a line of subdomains
+(each of which is mapped to a single process),
+and a single grid line
+broken into ``local'' grid lines.
+For each line of subdomains,
+the evaluation of the derivative requires the solution
+of Eq. (\ref{eqn:compact-tridiagonal-system})
+for each grid line.
+The primary difference from the single GPU case is that
+the tridiagonal systems are now \emph{distributed}
+among several processes.
+
+\subsection{Solving distributed tridiagonal systems}
+
+We use the algorithm
+proposed by Mattor et al. \cite{mattor1995algorithm}
+for solving the distributed tridiagonal systems.
+For a general tridiagonal system $A\bm{x}=\bm{r}$,
+the algorithm begins by partitioning
+the matrix and right-hand side among the $P$ processes.
+Each process $p$ is then associated with the
+following $\emph{subsystems}$
+
+\begin{align}
+    & A^p\bm{u}^p = \bm{r_u}^p & \label{eqn:secondary-system-1} \\ 
+    & A^p\bm{l}^p = \bm{r_l}^p & \label{eqn:secondary-system-2} \\
+    & A^p\bm{x_r}^p = \bm{r}^p & \label{eqn:primary-system} 
+\end{align}
+%
+expanded as:
 
@@
 r_m^p
-\end{bmatrix} & \label{eqn:primary-system} \\
+\end{bmatrix} & \label{eqn:primary-system-expanded} \\
 %
@@
 0
-\end{bmatrix} & \label{eqn:secondary-system-1} \\
+\end{bmatrix} & \label{eqn:secondary-system-1-expanded} \\
 %
@@
 -c_m^p
-\end{bmatrix} & \label{eqn:secondary-system-2}
+\end{bmatrix} & \label{eqn:secondary-system-2-expanded}
 \end{align}
 %
-and additionally, the following ``reduced'' system:
+We refer to the subsystem in Eq. (\ref{eqn:primary-system-expanded})
+the ``primary'' system, and the subsystems in
+Eqs. (\ref{eqn:secondary-system-1-expanded}) and
+(\ref{eqn:secondary-system-2-expanded})
+the ``secondary'' systems.
+Additionally, the following ``reduced'' system is solved:
 
@@
 \end{align}
-
-We call the system in [Eq. (\ref{eqn:primary-system})]
-the ``primary'' system, and the systems in
-[Eqs.
-(\ref{eqn:secondary-system-1}) and (\ref{eqn:secondary-system-2})]
-the ``secondary'' systems.
-The local part of the solution to the ``global'' tridiagonal system
+The local part of the solution
 is obtained as a linear combination of
@@
 \begin{equation}
-    \boldsymbol{x}^p = \boldsymbol{x}_r^p + \
-        \alpha^p \boldsymbol{u}^p + \beta^p \boldsymbol{l}^p
+    \bm{x}^p = \bm{x}_r^p + \
+        \alpha^p \bm{u}^p + \beta^p \bm{l}^p
     \label{eqn:sum-of-systems}
 \end{equation}
-
-Where $\alpha^p$ and $\beta^p$ are obtained
+%
+where $\alpha^p$ and $\beta^p$ are obtained
 from the solution of the reduced system.
@@
 \subsection{GPU implementation}
+
+We present a specialization of the above algorithm
+applied to the distributed near-Toeplitz systems that
+appear in compact finite difference evaluations,
+and adapted for GPUs.
+Our algorithm is outlined in \ref{fig:compact-algorithm},
+and is described for the case of evaluating derivatives
+in the direction along which successive
+function values in a subdomain are stored contigiously in memory,
+i.e., the ``fastest'' coordinate direction.
+Each step of the algorithm must be implemented on the GPU
+to avoid data transfer to and from the CPU,
+which is prohibitively expensive for large problems.
+Thus, we have several kernels to implement the algorithm.
+For communication of data between processes,
+we use the Message Passing Interface (MPI),
+and leverage the NVIDIA GPUDirect Technology
+for GPU-GPU communication.
+The purpose of each CUDA kernel and MPI call used
+is described in Table \ref{table:compact-algorithm}.
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[trim={0 7cm 0 7cm},clip,width=500pt]{img/compact-algorithm.pdf}
+\centering
+\caption{Algorithm for evaluating compact finite differences on multiple GPUs,
+(right: CUDA kernels and MPI calls used)}
+\label{fig:compact-algorithm}
+\end{center}
+\end{figure}
+
+\begin{table}[]
+\centering
+\caption{Purpose of kernels and MPI calls}
+\label{table:compact-algorithm}
+\begin{tabular}{|l|p{6cm}|}
+\hline
+Memcpy3d            & Copy noncontiguous boundary information
+                      of the function values
+                      to and from contiguous halo arrays \\ \hline
+ISend, IRecv        & Perform halo swaps with i-1 and i+1 processes \\ \hline
+computeRHSKernel    & Apply pointwise stencil operator to compute
+                      RHS at each grid point in the subdomain,
+                      using the halo values near the boundaries\\ \hline
+NEATO               & Solve the primary near-Toeplitz tridiagonal systems
+                      for the computed right hand sides, giving $x_r$ \\ \hline
+copyFacesKernel     & Copy the left and right faces of $x_r$
+                      into a single contiguous array \\ \hline
+MPI\_Gather         & Gather the data required to assemble
+                      the reduced system at rank 0 \\ \hline
+reducedSolverKernel & Solve the reduced systems for parameters
+                      $\alpha^p$ and $\beta^p$
+                      for each grid line \\ \hline
+MPI\_Scatter        & Scatter the parameters 
+                      $\alpha^p$ and $\beta^p$
+                      from rank 0 to all the processes \\ \hline
+sumSolutionsKernel  & Sum the primary and secondary solutions
+                      to compute the local part of the solution \\ \hline
+\end{tabular}
+\end{table}
 
@@
 \caption{Construction of the reduced system and scattering of parameters.}
-\label{fig:reduced-system}
-\end{center}
-\end{figure}
-
-In this section we describe the implementation of
-a CUDA+MPI solver for compact finite differences.
-The algorithm 
-consists of the following sequence of steps:
-
-\begin{enumerate}
-\item Computation of the RHS (requires communication)
-\item Solving the primary and secondary systems
-\item Constructing and solving the reduced systems,
-    and scattering the parameters $\alpha^p$ and $\beta^p$ (requires communication)
-\item Summing the above solutions [Eq. (\ref{eqn:sum-of-systems})]
-\end{enumerate}
-
-Each of these is performed on the GPU,
-and one of the primary concerns is to eliminate
-CPU-GPU transfers as much as possible in the solution process.
-
-The right-hand side evaluation is a pointwise stencil computation,
-which requires communication of the function values at the boundaries.
+\label{fig:constructing-reduced-system}
+\end{center}
+\end{figure}
+
+First, we note that the secondary systems
+[Eq. (\ref{eqn:secondary-system-1-expanded}) and
+ Eq. (\ref{eqn:secondary-system-2-expanded})]
+are dependent only on $\bm{a}$, $\bm{b}$ and $\bm{c}$,
+the tridiagonal coefficients of the global tridiagonal system.
+These are the same for all local grid lines,
+so they are solved for a single local grid line in each subdomain.
+
+The primary system [Eq. (\ref{eqn:primary-system-expanded})] must be solved
+for each of the local grid lines in a subdomain,
+as the right hand sides are different at each of the local grid lines. 
+The evaluation of the right hand sides are
+pointwise stencil computations,
+which require communication of the
+function values at the boundaries of the subdomains.
 This is achieved using a halo-swapping technique
@@
 data types is expensive on the GPU.
-After the halo information is exchanged,
-a pointwise stencil kernel computes the right-hand side values.
-
-The secondary systems are set up and solved only once
-as it is dependent only on the
-coefficients of the tridiagonal matrix.
-As the computation is offline,
-any convenient method can be used.
-The primary system is solved at every time step
-for different right hand sides,
-and is evaluated using the NEATO solver.
+The primary system is solved for all the local grid lines
+using the NEATO solver.
 
 The set up and solution of the reduced system
-requires global communication of the boundary information,
-which may be associated with a high overhead.
-While several techniques have been proposed to implement
-this type of communication pattern efficiently,
-we find it sufficient for moderately large problems
-to communicate all information to rank 0,
-using \texttt{MPI\_Gather}, and distribute the results
-back using \texttt{MPI\_Scatter} (Fig. \ref{fig:reduced-system}).
+requires global communication of the boundary information
+from the solutions $\bm{u}^p$, $\bm{l}^p$ and $\bm{x_r}^p$.
+The tridiagonal coefficients of the reduced system
+are set up easily,
+by communicating the boundary elements from
+$\bm{u}$ and $\bm{l}$,
+which are the same for every local grid line in each subdomain.
+
+The right hand sides require significantly more communication,
+as they are assembled from the boundary elements of $\bm{x_r}$,
+which are different for each local grid line in each subdomain.
+Thus, the boundary ``faces'' of each subdomain need to be communicated.
+These faces are first copied into a contiguous array,
+and these arrays are gathered at rank 0
+to assemble the right-hand sides of the reduced system
+(Fig. \ref{fig:constructing-reduced-system}).
+This communication strategy has the effect of
+producing right hand sides aligned along the
+``slowest'' co-ordinate direction, i.e.,
+the right hand side values for neighbouring grid points
+are located far apart in memory.
+As the reduced systems are quite small relative to the primary systems,,
+we use a convenient, but rather inefficient
+\emph{p-Thomas} algorithm to solve the systems,
+wherein each GPU thread solves the reduced system
+for a single grid line using the Thomas algorithm.
+
+The solution of the reduced systems produces the parameters
+$\alpha^p$ and $\beta^p$,
+which are scattered back to the respective processes, $p$.
 
@@
 
-For evaluation of compact finite differences in different
+For evaluation of compact finite differences in other
 coordinate directions,
-a local permutation of the data is performed.
+a local permutation of the data is performed on the input data
+(function values)
+before applying the above algorithm.
+and again on the output data
+(derivative values).
 
 \subsection{Results}
+
+Talk about which GPUs, cluster information
+
+Figures \ref{fig:strong-scaling-256}-\ref{weak-scaling-256}
+show the strong and weak scaling of the compact finite difference solver
+for evaluating derivatives in all three coordinate directions.
 

### 2015-11-11 paper-draft/scripts/compact-profiling/compact-profiling.py

@@ -1,0 +1,42 @@
+import numpy as np
+import pandas
+import matplotlib
+import matplotlib.pyplot as plt
+import os
+import sys
+import collections
+
+
+
+plt.style.use('bmh')
+fig, ax = plt.subplots()
+
+# horizontal bar with timing breakdown
+
+# size 256, 64 GPUs
+results_lists = [
+ [6.640, 2.700, 12.864, 5.336, 2.869, 0],
+ [6.322, 2.771, 12.729, 5.427, 2.708, 6.226],
+ [6.515, 2.721, 12.892, 5.465, 2.536, 8.679]]
+
+df = pandas.DataFrame(results_lists, index=['dfdx', 'dfdy', 'dfdz'],
+    columns=['Halo swaps', 'RHS', 'Primary systems', 'Reduced systems', 'Summing solutions', 'Permutations'])
+df.plot(kind='bar', stacked=True, ax=ax)
+leg=plt.legend(loc='center left', bbox_to_anchor=(1.0, 0.5))
+ax.set_ylabel('Time taken (ms)')
+ax.set_xticklabels(df.index, rotation=0)
+plt.savefig('profiling-256.eps', extra_artists=(leg,), bbox_inches='tight')
+
+# size 512, 64 GPUs
+results_lists = [
+[35.43, 16.88, 98.97, 10.63, 13.68, 0],
+[35.69, 16.84, 98.43, 6.83, 13.66, 41.53],
+[34.47, 16.83, 99.00, 6.74, 13.67, 52.08]]
+df = pandas.DataFrame(results_lists, index=['dfdx', 'dfdy', 'dfdz'],
+    columns=['Halo swaps', 'RHS', 'Primary systems', 'Reduced systems', 'Summing solutions', 'Permutations'])
+df.plot(kind='bar', stacked=True, ax=ax)
+leg=plt.legend(loc='center left', bbox_to_anchor=(1.0, 0.5))
+ax.set_ylabel('Time taken (ms)')
+ax.set_xticklabels(df.index, rotation=0)
+plt.savefig('profiling-512.eps', bbox_extra_artists=(leg,), bbox_inches='tight')
+

### 2015-11-11 paper-draft/scripts/compact-scaling/strong-scaling.py

@@ -19,6 +19,11 @@ dfdz = np.array([
 
 fig, ax = plt.subplots()
 ax.set_yscale("log", nonposy='clip')
+ax.get_yaxis().set_major_formatter(
+        matplotlib.ticker.ScalarFormatter())
+ax.yaxis.set_major_formatter(
+        matplotlib.ticker.FormatStrFormatter('%d'))
+
 ax.set_xscale("linear")
 ax.plot(dfdx, label='dfdx')
 ax.plot(dfdy, label='dfdy')
@@ -40,10 +45,14 @@ dfdz = np.array([
 
 fig, ax = plt.subplots()
 ax.set_yscale("log", nonposy='clip')
+ax.get_yaxis().set_major_formatter(
+        matplotlib.ticker.ScalarFormatter())
+ax.yaxis.set_major_formatter(
+        matplotlib.ticker.FormatStrFormatter('%d'))
 ax.set_xscale("linear")
-ax.plot(dfdx, label='dfdx')
-ax.plot(dfdy, label='dfdy')
-ax.plot(dfdz, label='dfdz')
+ax.plot(dfdx, 'o-', label='dfdx')
+ax.plot(dfdy, 'o-', label='dfdy')
+ax.plot(dfdz, 'o-', label='dfdz')
 ax.set_xticks(range(len(ngpus)))
 ax.set_xticklabels(['1', '8', '64'])
 ax.legend()

### 2015-11-11 paper-draft/scripts/compact-scaling/weak-scaling.py

@@ -18,6 +18,10 @@ dfdz = np.array([
 
 fig, ax = plt.subplots()
 ax.set_yscale("log", nonposy='clip')
+ax.get_yaxis().set_major_formatter(
+        matplotlib.ticker.ScalarFormatter())
+ax.yaxis.set_major_formatter(
+        matplotlib.ticker.FormatStrFormatter('%d'))
 ax.set_xscale("linear")
 ax.plot(dfdx, label='dfdx')
 ax.plot(dfdy, label='dfdy')
@@ -42,6 +46,10 @@ dfdz = np.array([
 
 fig, ax = plt.subplots()
 ax.set_yscale("log", nonposy='clip')
+ax.get_yaxis().set_major_formatter(
+        matplotlib.ticker.ScalarFormatter())
+ax.yaxis.set_major_formatter(
+        matplotlib.ticker.FormatStrFormatter('%d'))
 ax.set_xscale("linear")
 ax.plot(dfdx, label='dfdx')
 ax.plot(dfdy, label='dfdy')

### 2015-11-12 paper-draft/paper.tex
updates

@@
 such as
-compact finite differences \cite{lele1992compact},
+compact finite differences
+\cite{lele1992compact,kennedy1994several},
 Alternating Direction Implicit (ADI) methods \cite{1955ADI}, and
@@
 to speed up the computations
-\cite{tutkun2012gpu}
-\cite{esfahanian2014efficient}
-\cite{GoSt11CR}.
+\cite{tutkun2012gpu,esfahanian2014efficient,GoSt11CR}.
 The performance of tridiagonal solvers on the GPU
@@
 
-\begin{figure}[h!]
+\begin{figure}
 \begin{center}
@@
 requires an understanding of the underlying architecture.
-We provide the pertinent details here,
+We provide the pertinent details here
 and refer to \cite{GPUcomputingera} for a more complete picture.
@@
 (48 KiB for current GPUs).
-The amount of shared memory actually used by each block,
+The amount of shared memory actually used by each block
 can thus limit the number of blocks that
@@
 \cite{Zhang2010FTS}.
+
+Much work has been done on optimizing cyclic reduction performance
+ong GPUs.
 Zhang et al. \cite{Zhang2010FTS} propose a
@@
 a means to reduce shared memory usage in cyclic reduction.
-More recently, Esfahanian et al. \cite{esfahanian2014efficient}
-describe an implementation of cyclic reduction
-that works entirely in global memory,
-using additional kernels to reorder data so as to
-coalesce memory access.
-
-\begin{figure}[h!]
+Esfahanian et. al, \cite{esfahanian2014efficient}
+avoid shared memory (and associated bank conflicts entirely)
+using a data rearrangement scheme to improve global memory access.
+Our approach takes a different route,
+and is focussed on near-Toeplitz systems---exploiting
+the relatively simple matrix structure
+to reduce the number of computations and memory accesses
+performed by each thread
+at every cyclic reduction step.
+
+\begin{figure}
 \begin{center}
@@
 has been reported for block Toeplitz tridiagonal systems
-by Bini et al.
+by Bini et al \cite{bini}.
 
@@
 for $N_{rhs}$ right hand sides.
-Our implementation uses the NVIDIA CUDA platform,
+Our implementation uses the NVIDIA CUDA platform for writing GPU kernels,
+interfaced via the Python library PyCUDA \cite{kloeckner_pycuda_2012}.
 We develop two approaches based on the common idea
@@
 straightforward and compact,
-spanning no more than 100 lines.
+spanning no more than 100 lines of code.
 
@@
 At each step,
-the size of the grid block is determined by the \emph{stride}
+the size of the thread blocks is determined by the \emph{stride}
 between elements accessed at that step.
@@
 The kernel is launched with $n/2$ threads per block.
-Each thread block allocated a block of shared memory of size $n/2$.
-Each thread performs the first reduction step
+Each thread block is allocated a block of shared memory of size $n/2$.
+Each thread of a block performs the first reduction step
 [Eq. (\ref{eqn:precomputed-forward-reduction-step})]
@@
 Of course, we use double precision for all solvers.
-All timings reported are for the tridiagonal solution
-and are averaged over several (100) solves.
+The timings reported are
+kernel \emph{execution} times, i.e.,
+the time for all kernel(s) to execute completely
+before returning the program control to the CPU.
+All timings are averaged over 100 tridiagonal solves.
 
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[height=200pt]{img/computational-domain.eps}
+\begin{figure}
+\begin{center}
+\includegraphics[height=150pt]{img/computational-domain.eps}
 \caption{Computational domain in 3-D.}
@@
 \begin{equation}
-    f^{\prime}_{n} + 2f^{\prime}_{n-1s}
+    f^{\prime}_{n} + 2f^{\prime}_{n-1}
     =
-    \frac{5f_{n} - 4f_{n-2} 1 f_{n-2}}{dx}
+    \frac{5f_{n} - 4f_{n-2} -  f_{n-1}}{dx}
 \end{equation}
@@
  \begin{bmatrix}
-     -5f_1 + 4f_2 + f_3\\
-     f_{3} - f_{1}\\
-     f_{4} - f_{2}\\
+     \frac{-5f_1 + 4f_2 + f_3}{2}\\
+     \frac{3(f_{3} - f_{1})}{4}\\
+     \frac{3(f_{4} - f_{2})}{4}\\
      \vdots\\
@@
      \vdots\\
-     f_{n} - f_{n-2}\\
-     5f_{n} - 4f_{n-1} - f_{n-2}
+     \frac{3(f_{n} - f_{n-2})}{4}\\
+     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2}
   \end{bmatrix}
@@
 solving this tridiagonal system for $ny \times nz$ right hand sides.
-On a single GPU,
-this is done easily using the NEATO solver.
+Some work has been done on the solution of these
+tridiagonal systems using GPUs---Tutkun et. al
+\cite{tutkun2012gpu} compute the inverse of the coefficient matrix,
+and use the GPU to multiply the inverse with the right-hand sides.
+Wei et al. \cite{wei2013parallelizing}
+use the parallel cyclic reduction (PCR) algorithm
+to solve similar systems appearing in
+the 2-D ADI problem.
+More recently, Esfahanian et. al \cite{esfahanian2014efficient}
+use a variant of cyclic reduction with
+data rearrangements kernels for better global memory access patterns.
+Our approach uses the NEATO solver described above
+to solve for the right hand sides at each grid line,
+which takes advantage of the known near-Toepltiz structure
+of the coefficient matrix arising
+in the compact finite difference scheme.
 
@@
 \end{bmatrix} & \label{eqn:primary-system-expanded} \\
-%
-%
-%
+\end{align}
+
+\begin{align}
 & \begin{bmatrix}
@@
 \end{bmatrix} & \label{eqn:secondary-system-1-expanded} \\
+\end{align}
 %
-%
-%
+\begin{align}
 & \begin{bmatrix}
@@
 for GPU-GPU communication.
+MPI is interfaced via the mpi4py \cite{dalcin2005mpi}
+Python library.
 The purpose of each CUDA kernel and MPI call used
@@
 \centering
-\caption{Purpose of kernels and MPI calls}
+\caption{Purpose of kernels and MPI calls in compact finite difference application}
 \label{table:compact-algorithm}
@@
 by communicating the boundary elements from
-$\bm{u}$ and $\bm{l}$,
+$\bm{u}^p$ and $\bm{l}^p$,
 which are the same for every local grid line in each subdomain.
@@
 The right hand sides require significantly more communication,
-as they are assembled from the boundary elements of $\bm{x_r}$,
+as they are assembled from the boundary elements of $\bm{x_r}^p$,
 which are different for each local grid line in each subdomain.
@@
 for a single grid line using the Thomas algorithm.
-
 The solution of the reduced systems produces the parameters
@@
 on the GPU.
-
 For evaluation of compact finite differences in other
@@
 
-Talk about which GPUs, cluster information
-
-Figures \ref{fig:strong-scaling-256}-\ref{weak-scaling-256}
+The experiments were run on the
+Clemson University Palmetto Cluster,
+using NVIDIA Tesla K20 and K40 GPUs.
+The K40 GPUs were used for the largest problem sizes.
+Each compute node on the cluster is equipped with up to 2 GPUs,
+and nodes are connected by 56 Gbps Infiniband interconnect.
+We use Open MPI 1.8.1 configured with OpenFabrics support.
+
+The timings reported are \emph{wall clock} times
+with global synchronization between processes performed 
+before and after evaluation of the derivatives.
+All timings are averaged over 100 evaluations of the function derivatives
+in each coordinate direction.
+We make it clear that our reported
+problem sizes represent the \emph{actual size of problem data}.
+Although it may be considered sufficient to
+run tests on a single line of processes
+for measuring the compact finite difference solver performance,
+we set up and solve the problem for the entire computational domain.
+In the context of a larger simulation,
+global synchronization between the processes
+is typically performed before and after the evalaution of derivatives,
+and it is important to consider the related overhead.
+
+Figures \ref{fig:compact-profiling-1024-64}-\ref{fig:compact-profiling-2048-64}
+show the time taken by the different steps of
+the compact finite difference solver.
+We note that for the larger problem size,
+the evaluation of the primary systems (using the NEATO solver)
+constitutes a larger majority of the total runtime,
+which justifies our efforts in optimizing the tridiagonal solver.
+For evaluation of the derivatives in the y- and z- directions,
+we note that a significant portion of the runtime is dedicated
+to performing permutations of the input and output data.
+We attribute this to our naive implementation of the
+permutation kernels (no shared memory).
+Figures \ref{fig:strong-scaling-256}-\ref{fig:weak-scaling-256}
 show the strong and weak scaling of the compact finite difference solver
 for evaluating derivatives in all three coordinate directions.
-
-\begin{figure}[h!]
+We note good scaling for both 2-D and 3-D cases.
+The strong scaling for 3-D problems is somewhat better
+as the GPU is kept busy for all problem sizes.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/profiling-1024-64.eps}
+\caption{Solving problem sized $1024^3$ on 64 GPUs}
+\label{fig:compact-profiling-1024-64}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/profiling-2048-64.eps}
+\caption{Solving problem sized $2048^3$ on 64 GPUs}
+\label{fig:compact-profiling-2048-64}
+\end{center}
+\end{figure}
+
+\begin{figure}
 \begin{center}
@@
 
-\pagebreak[4]
+\section{Conclusions}
+
+We present a GPU solver for near-Toeplitz tridiagonal systems
+based on the cyclic reduction algorithm,
+that is applicable to several numerical schemes
+in computational fluid dynamics.
+Our solver gives better performance than
+the current multicore Intel MKL and CUSPARSE solvers
+Additionally, we describe an implementation of our solver
+for GPU clusters,
+and provide performance results for a
+compact finite difference application,
+run on up to 64 GPUs.
+
+\pagebreak
 \section*{References}
-
 \bibliography{references}

### 2015-11-12 paper-draft/scripts/compact-profiling/compact-profiling.py

@@ -25,7 +25,10 @@ df.plot(kind='bar', stacked=True, ax=ax)
 leg=plt.legend(loc='center left', bbox_to_anchor=(1.0, 0.5))
 ax.set_ylabel('Time taken (ms)')
 ax.set_xticklabels(df.index, rotation=0)
-plt.savefig('profiling-256.eps', extra_artists=(leg,), bbox_inches='tight')
+plt.savefig('profiling-1024-64.eps', extra_artists=(leg,), bbox_inches='tight')
+plt.close()
+
+fig, ax = plt.subplots()
 
 # size 512, 64 GPUs
 results_lists = [
@@ -38,5 +41,5 @@ df.plot(kind='bar', stacked=True, ax=ax)
 leg=plt.legend(loc='center left', bbox_to_anchor=(1.0, 0.5))
 ax.set_ylabel('Time taken (ms)')
 ax.set_xticklabels(df.index, rotation=0)
-plt.savefig('profiling-512.eps', bbox_extra_artists=(leg,), bbox_inches='tight')
+plt.savefig('profiling-2048-64.eps', bbox_extra_artists=(leg,), bbox_inches='tight')
 

### 2015-11-12 paper-draft/scripts/compact-scaling/strong-scaling.py

@@ -25,14 +25,15 @@ ax.yaxis.set_major_formatter(
         matplotlib.ticker.FormatStrFormatter('%d'))
 
 ax.set_xscale("linear")
-ax.plot(dfdx, label='dfdx')
-ax.plot(dfdy, label='dfdy')
-ax.plot(dfdz, label='dfdz')
+ax.plot(dfdx, 'o-', label='dfdx')
+ax.plot(dfdy, 'o-', label='dfdy')
+ax.plot(dfdz, 'o-', label='dfdz')
 ax.set_xticks(range(len(ngpus)))
 ax.set_xticklabels(['1', '8', '64'])
 ax.legend()
 ax.set_xlabel('Number of GPUs')
 ax.set_ylabel('Time to solve in ms')
+ax.set_xlim((-0.1, 2.2))
 plt.savefig('strong-scaling-256.eps')
 
 # prob size: 512
@@ -58,4 +59,5 @@ ax.set_xticklabels(['1', '8', '64'])
 ax.legend()
 ax.set_xlabel('Number of GPUs')
 ax.set_ylabel('Time to solve in ms')
+ax.set_xlim((-0.1, 2.2))
 plt.savefig('strong-scaling-512.eps')

### 2015-11-12 paper-draft/scripts/compact-scaling/weak-scaling.py

@@ -23,16 +23,16 @@ ax.get_yaxis().set_major_formatter(
 ax.yaxis.set_major_formatter(
         matplotlib.ticker.FormatStrFormatter('%d'))
 ax.set_xscale("linear")
-ax.plot(dfdx, label='dfdx')
-ax.plot(dfdy, label='dfdy')
-ax.plot(dfdz, label='dfdz')
+ax.plot(dfdx, 'o-', label='dfdx')
+ax.plot(dfdy, 'o-', label='dfdy')
+ax.plot(dfdz, 'o-', label='dfdz')
 ax.set_xticks(range(len(ngpus)))
 ax.set_xticklabels(['1', '8', '27', '64'])
-ax.legend()
+ax.legend(loc='best')
 ax.set_xlabel('Number of GPUs')
 ax.set_ylabel('Time to solve in ms')
+ax.set_xlim((-0.1, 3.2))
 plt.savefig('weak-scaling-128.eps')
-
 
 ngpus = np.array([1, 4, 64])
 
@@ -51,12 +51,13 @@ ax.get_yaxis().set_major_formatter(
 ax.yaxis.set_major_formatter(
         matplotlib.ticker.FormatStrFormatter('%d'))
 ax.set_xscale("linear")
-ax.plot(dfdx, label='dfdx')
-ax.plot(dfdy, label='dfdy')
-ax.plot(dfdz, label='dfdz')
+ax.plot(dfdx, 'o-', label='dfdx')
+ax.plot(dfdy, 'o-', label='dfdy')
+ax.plot(dfdz, 'o-', label='dfdz')
 ax.set_xticks(range(len(ngpus)))
 ax.set_xticklabels(['1', '8', '64'])
-ax.legend()
+ax.legend(loc='best')
 ax.set_xlabel('Number of GPUs')
 ax.set_ylabel('Time to solve in ms')
+ax.set_xlim((-0.1, 2.2))
 plt.savefig('weak-scaling-256.eps')

### 2015-11-14 thesis/acknowledgments.tex
delete junk files and reorganize

\chapter*{Acknowledgments}

I acknowledge everyone.

### 2015-11-14 thesis/introduction-and-background.tex
delete junk files and reorganize

\chapter{Preliminaries}

\section{Graphics Processing Units (GPUs)}

\subsection{Motivation}

Most numerical approaches for modeling systems and phenomena
approximate continuous quantities as discrete quantities.
The larger the number of discrete entities used,
the better the approximation.
In CFD, the physical system (fluid)
is represented as a set of discrete points (finite difference methods),
or discrete cells (finite volume method),
and this discrete representation is termed as a \emph{mesh}.
In CFD applications
a mesh containing several points is used
to represent the underlying system,
and numerical simulations are performed
to study how physical quantities
change with time at these mesh points (or cells).
Typically, a \emph{time stepping} scheme is used to study this evolution,
wherein the simulation proceeds as a series of \emph{time steps}.
For each time step, the quantities at each mesh point at
\emph{updated} from quantities computed at the previous time step.

In the direct numerical simulation (DNS) of fluid systems,
the aim is to resolve the smallest features of the flow.
This requires the use of extremely fine meshes.
At each of the mesh points,
data about the flow must be stored
(for instance, the pressure, temperature and velocities at that point).
The large amount of data storage required places
restrictions on the problem sizes that can be accommodated
on a single workstation/PC.
Further,
the numerical simulation requires a large amount of
computations with this data.
In general, the computations are of a repetitive nature:
the same computations are performed repeatedly
for the different mesh points,
and for successive time steps.

Thus, CFD is typically a computationally expensive endeavour,
and it has long been of interest to use
\emph{parallel processing} to
perform numerical simulations of fluid flows.
This involves distributing computing tasks
among multiple independent processors,
which can run concurrently.
Of course, this requires that the computations be \emph{independent}
of each other.
Fortunately, the computational patterns that arise in CFD
typically satisfy this requirement:
the computations at different mesh points are largely independent
of each other and can be done in parallel.

The traditional approach to parallel computing has been
to distribute parallel tasks among CPU cores.
The cores may all have access to a common memory space
(shared memory systems),
or may have distinct memory spaces
(distributed memory systems).
Much more common than purely distributed systems are
\emph{hybrid} systems wherein the tasks may be distributed
among so-called \emph{nodes},
which are themselves shared-memory systems.
Again, the computational workhorse in all these systems has traditionally
been the CPU core.

A more recent parallel system that has seen
wide adoption
is the Graphics Processing Unit (GPU).
GPUs were developed primarily for performing
graphics computations, i.e.,
calculations with pixel data.
This is a \emph{highly} parallelizable task;
pixels can be processed independently and in parallel.
Thus, the GPU evolved as a system
to perform hundreds of computations concurrently,
and it should be no surprise that it soon
gained attention from the scientific computation community.
Now, several scientific software packages have been
written for GPUs,
and the worlds largest supercomputers employ GPUs
as accelerators.

\section{Compact finite differences}

Numerical evaluation of derivatives is a central component
in scientific computing.
The simplest and most widely-used approach for numerical differentiation
is the \emph{finite difference} approximation,
wherein the numerical approximation of the derivative
is expressed as a \emph{difference equation}.
A key application of the finite difference approximation
is in \emph{finite difference methods},
a family of numerical methods for solving differential equations
in which the derivatives are approximated using
finite difference approximations.
%
For example, we may consider a uniformly sampled
function $f(x)$,
sampled at points $x_1, x_2, x_3, \hdots x_n$.
At each sample point $i$, we may approximate the derivative as
a combination of the function values at
$i$, and its neighbouring points:

\begin{enumerate}
    \item When the derivative at $i$ is expressed
        as some combination of the function values
        at $i$, $i+1$, $i+2$, $\hdots$,
        we refer to the approximation as a
        \emph{forward} difference.

    \item When the derivative is expressed as
        some combination of the function values
        at $i$, $i-1$, $i-2$, $\hdots$,
        we refer to the approximation as a
        \emph{backward} difference.

    \item When the derivative is expressed as
        some combination of the function values
        on \emph{both} sides of $i$, i.e,
        at $\hdots$, $i-2$, $i-1$, $i$, $i-1$, $i-2$, $\hdots$,
        we refer to the approximation as a
        \emph{central} difference.
\end{enumerate}

The approximation of higher order derivatives generally requires
inclusion of a larger number of points in the
finite difference approximation
(referred to as the finite difference \emph{stencil}).
For a given order of derivative,
finite difference approximation of arbitrary stencil widths
may be derived,
with larger stencils associated with higher accuracy~\cite{fornberg1988generation}.
In most applications,
large stencil widths are undesirable.
This is because:

\begin{enumerate}
    \item The arithmetic intensity increases, i.e.,
        a larger number of computations must be performed per grid point.
    \item When the solution is done in parallel,
        the amount of boundary information that must be exchanged
        between processors increases.
\end{enumerate}

To overcome the problems associated with larger stencil widths,
another class of finite difference schemes,
known as \emph{compact} schemes may be used.
First proposed by Sanjeeva Lele ~\cite{lele1992compact},
compact schemes express the derivative at a point $i$
in terms of the derivatives at neighbouring points,
For example,
if $f_i$ represents the value of
a uniformly sampled function evaluated at the $i$th sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:

\begin{equation}
\begin{split}
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
        f^{\prime}_i
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{dx} + \
    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
    c\frac{f_{i+3} - f_{i-3}}{dx} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%
where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
are parameters that must satisfy certain constraints.
We note that for equations of the form in Eq. \ref{eqn:general-compact},
the derivative at any points $i$ cannot be computed explicitly.
Instead, we must write similar equations for \emph{all} points $i$
in the range.
This results in a system of linear equations
with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
which may be represented by the matrix system $A\bm{x}=\bm{d}$,
where $\bm{x}$ is the vector
\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
$\bm{d}$ is a vector of right hand sides [Eq.\ref{eqn:general-compact}],
and $A$ is, in general, a \emph{banded} matrix.

The simplest classes of banded matrix include
diagonal, tridiagonal and pentadiagonal matrices.
When $A$ is diagonal ($\alpha$ = $\beta$ = $\hdots$ = 0),
the scheme is explicit and straightforward to evaluate:
the derivatives are simply computed pointwise
(in series or in parallel),
by application of the right-hand side stencil at each point.
When $A$ is tridiagonal ($\alpha \neq 0$, $\beta$ = $\hdots$ = 0),
the evaluation of the derivatives requires the solution
of the tridiagonal system, which yields the derivatives
at all points simultaneously.
The classical Thomas algorithm
also referred to as the Tridiagonal Matrix Algorithm,
is widely employed for this purpose.
The algorithm is straightforward to implement,
and stable for diagonally dominant matrices ~\cite{numericalrecipes}.

### 2015-11-15 thesis/compact-finite-difference.tex
update sections and add content to thesis

\chapter{Application to compact finite difference evaluation}

\section{Compact finite difference evaluation strategy}
    \subsection{Shared memory systems}
    \subsection{Distributed memory systems}
    \subsection{Heterogeneous systems}

\section{Distributed tridiagonal solver}
    \subsection{General algorithm}
    \subsection{Specialization for compact finite difference
        evaluations}

\section{GPU application}
    \subsection{General strategy}
    \subsection{GPU-GPU communication}
        \subsubsection{Naive strategy}
        \subsubsection{NVIDIA GPUDirect Technology}
    \subsection{Implementation}
        \subsubsection{Precomputing coefficients}
        \subsubsection{Solving secondary systems}
        \subsubsection{Computing right hand sides}
        \subsubsection{Solving primary systems}
        \subsubsection{Constructing and solving reduced systems}
        \subsubsection{Summing primary and secondary solutions}
        \subsubsection{Evaluating derivatives in other
            coordinate directions}

### 2015-11-15 thesis/introduction-and-background.tex
update sections and add content to thesis

@@
-\chapter{Preliminaries}
+\chapter{Introduction and Background}
 
-\section{Graphics Processing Units (GPUs)}
-
-\subsection{Motivation}
-
-Most numerical approaches for modeling systems and phenomena
-approximate continuous quantities as discrete quantities.
-The larger the number of discrete entities used,
-the better the approximation.
-In CFD, the physical system (fluid)
-is represented as a set of discrete points (finite difference methods),
-or discrete cells (finite volume method),
-and this discrete representation is termed as a \emph{mesh}.
-In CFD applications
-a mesh containing several points is used
-to represent the underlying system,
-and numerical simulations are performed
-to study how physical quantities
-change with time at these mesh points (or cells).
-Typically, a \emph{time stepping} scheme is used to study this evolution,
-wherein the simulation proceeds as a series of \emph{time steps}.
-For each time step, the quantities at each mesh point at
-\emph{updated} from quantities computed at the previous time step.
+\section{Motivation}
 
 In the direct numerical simulation (DNS) of fluid systems,
-the aim is to resolve the smallest features of the flow.
-This requires the use of extremely fine meshes.
-At each of the mesh points,
-data about the flow must be stored
-(for instance, the pressure, temperature and velocities at that point).
-The large amount of data storage required places
-restrictions on the problem sizes that can be accommodated
-on a single workstation/PC.
-Further,
-the numerical simulation requires a large amount of
-computations with this data.
-In general, the computations are of a repetitive nature:
-the same computations are performed repeatedly
-for the different mesh points,
-and for successive time steps.
+the aim is to resolve the smallest features of the flow
+without resorting to any modeling of the turbulence.
+This involves the use of
+extremely fine computational grids to
+discretize the flow domain,
+and some numerical method for solving the
+flow equations on this grid.
+By far the most popular numerical approach for
+solving the flow equations
+is the \emph{finite difference method}.
+The finite difference method approximates the
+spatial and temporal derivatives appearing in the
+partial differential equations that describe the flow
+using finite difference schemes.
+\emph{Compact finite difference schemes} are a class
+of finite difference schemes that have found widespread
+adoption in DNS codes
+for their high order of accuracy.
+The evaluation of compact finite differences
+requires the repeated solution of \emph{banded} linear systems,
+making them fairly complex and expensive computationally.
 
-Thus, CFD is typically a computationally expensive endeavour,
-and it has long been of interest to use
-\emph{parallel processing} to
-perform numerical simulations of fluid flows.
-This involves distributing computing tasks
-among multiple independent processors,
-which can run concurrently.
-Of course, this requires that the computations be \emph{independent}
-of each other.
-Fortunately, the computational patterns that arise in CFD
-typically satisfy this requirement:
-the computations at different mesh points are largely independent
-of each other and can be done in parallel.
+As mentioned, DNS involves the use of extremely fine grids.
+At each of the grid points,
+various data about the flow must be stored---for
+example, the geometric coordinates (x, y and z),
+pressure, temperature and velocities at that point.
+The amount of memory required to store this data
+for even small problem sizes
+can quickly exceed the capacity of modern workstations/PCs.
+Thus, a \emph{distributed memory} parallel system
+is generally required for performing DNS.
+Here, the traditional approach has been
+to distribute parallel tasks among individual CPU cores,
+or groups of CPU cores
+that share common memory spaces.
+In the latter case,
+each group of cores constitutesa \emph{shared memory} system,
+and the overall system is referred to as a \emph{hybrid system}.
 
-The traditional approach to parallel computing has been
-to distribute parallel tasks among CPU cores.
-The cores may all have access to a common memory space
-(shared memory systems),
-or may have distinct memory spaces
-(distributed memory systems).
-Much more common than purely distributed systems are
-\emph{hybrid} systems wherein the tasks may be distributed
-among so-called \emph{nodes},
-which are themselves shared-memory systems.
-Again, the computational workhorse in all these systems has traditionally
-been the CPU core.
+The workhorse for computation in the above described
+parallel systems is the CPU core,
+however, more recently,
+Graphics Processing Units (GPUs)
+are being used for performing intensive calculations.
+GPUs, while themselves being highly parallel processors,
+can also function as accelerators in distributed memory systems
+(GPU clusters).
+However,
+applications that exploit such systems
+need careful redesign of algorithms---and
+sometimes, substantial changes to code---to
+see significant performance improvements.
+This is because the CPU and GPU have very different architectures,
+and any na\"{\i}ve ``parallelization'' of algorithms
+designed for the CPU
+is likely not take full advantage of the GPU's
+memory heirarchy and compute ability.
 
-A more recent parallel system that has seen
-wide adoption
-is the Graphics Processing Unit (GPU).
-GPUs were developed primarily for performing
-graphics computations, i.e.,
-calculations with pixel data.
-This is a \emph{highly} parallelizable task;
-pixels can be processed independently and in parallel.
-Thus, the GPU evolved as a system
-to perform hundreds of computations concurrently,
-and it should be no surprise that it soon
-gained attention from the scientific computation community.
-Now, several scientific software packages have been
-written for GPUs,
-and the worlds largest supercomputers employ GPUs
-as accelerators.
+The objective of this work is to develop an
+approach for evaluating
+compact finite differences
+on multiple GPUs in a distributed system.
+This is of interest because
+the evaluation of spatial derivatives using
+compact finite difference schemes is one of the most
+expensive tasks in DNS.
+But perhaps more significantly,
+it encompasses several computational patterns
+such as pointwise updates, stencil evaluations,
+and solutions of distributed tridiagonal systems.
+Efficient implementation of these computational patterns
+on GPUs is of interest in other areas of CFD,
+and scientific computation in general.
 
@@
 with larger stencils associated with higher accuracy~\cite{fornberg1988generation}.
+
+\subsection{Wavenumber analysis of finite difference methods}
+
+For DNS applications, the relevant measure of accuracy of a
+finite difference scheme is obtained from the so-called
+\emph{modified wavenumber} approach.
+Here, we test the scheme's ability to
+accurately estimate the derivative of sinusoidal functions
+with increasing wavenumbers (frequencies),
+given a specified grid size. 
+We expect that the approximation of the derivative
+becomes more difficult with increasing wavenumbers,
+as the function value varies more rapidly.
+The wavenumber of the approximate derivative
+as given by the finite difference scheme is compared with
+the wavenumber of the exact derivative.
+The result for different schemes is shown
+in Fig. \ref{fig:modified-wavenumbers}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=250pt]{fig/modified-wavenumbers.eps}
+\end{center}
+\caption{Modified wavenumbers for three different schemes
+    with the same stencil width.
+    The compact schemes better estimate the derivative
+    for higher wavenumbers.}
+\label{fig:modified-wavenumbers}
+\end{figure}
+
+The ability of a finite difference scheme
+to accomodate large wavenumbers
+is extremely relevant in
+computational fluid dynamics applications
+~\cite{KravchenkoEffNumErr}.
+In DNS, the scheme
+\emph{must} capture the
+rapidly varying characteristics of the flow
+associated with the turbulence.
+For this purpose,
+higher order explicit schemes
+may be considered.
+While they are traightforward to compute,
+they are associated with large stencil widths.
 In most applications,
 large stencil widths are undesirable.
-This is because:
+This is because
+the arithmetic intensity increases with stencil size, i.e.,
+a larger number of computations must be performed per grid point.
+Further,
+the amount of boundary information that must be exchanged
+between parallel processes increases,
+which can seriously affect overall performance.
 
-\begin{enumerate}
-    \item The arithmetic intensity increases, i.e.,
-        a larger number of computations must be performed per grid point.
-    \item When the solution is done in parallel,
-        the amount of boundary information that must be exchanged
-        between processors increases.
-\end{enumerate}
+From \ref{fig:modified-wavenumbers},
+it is clear that the compact schemes are better
+able to compute derivatives for larger wavenumbers
+for a given stencil width.
+Thus, they are widely used in
+CFD for evaluation of spatial derivatives.
 
-To overcome the problems associated with larger stencil widths,
-another class of finite difference schemes,
-known as \emph{compact} schemes may be used.
-First proposed by Sanjeeva Lele ~\cite{lele1992compact},
-compact schemes express the derivative at a point $i$
-in terms of the derivatives at neighbouring points,
+\subsection{General form of compact finite difference schemes}
+\label{subsec:gen-form-compact}
+
+Compact schemes express the derivative at a point $i$
+in terms of
+function values \emph{and derivatives}
+at neighbouring points,
 For example,
@@
 a relation of the form:
-
+%
 \begin{equation}
@@
 where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
-are parameters that must satisfy certain constraints.
+are parameters that must satisfy certain constraints ~\cite{lele1992compact}.
 We note that for equations of the form in Eq. \ref{eqn:general-compact},
@@
 
+\subsection{Boundary conditions}
+
+One of the advantages of the compact finite difference approach
+is that it accommodates \emph{non-periodic} boundary conditions.
+This is in contrast to other methods used in DNS,
+such as the \emph{spectral} methods.
+We note that the Eq. (\ref{eqn:general-compact})
+cannot be applied near the boundary points.
+At the boundaries,
+\emph{non-centered} or \emph{one-sided}
+finite difference approximations are required.
+Some considerations are made in choosing
+these approximations:
+firstly, the bandwidth of the resulting banded matrix
+must be preserved.
+Secondly,
+the width of the boundary stencils
+must be lower than the interior stencils,
+as higher order boundary stencils are unstable
+~\cite{kennedy1994several}.
+In general, boundary equations for the first derivative
+are of the following form ~\cite{lele1992compact}:
+
+\begin{equation}
+    f_1^{\prime} + \alpha f_2^{\prime} = \
+        \frac{1}{h}(af_1 + bf_2 + cf_3 + df_4)
+\label{eqn:boundary-compact}
+\end{equation}
+
+\subsection{Tridiagonal compact schemes}
+
+As mentioned in Sec. \ref{subsec:gen-form-compact},
+compact finite difference schemes lead to
+\emph{banded} linear systems.
 The simplest classes of banded matrix include
 diagonal, tridiagonal and pentadiagonal matrices.
-When $A$ is diagonal ($\alpha$ = $\beta$ = $\hdots$ = 0),
-the scheme is explicit and straightforward to evaluate:
-the derivatives are simply computed pointwise
-(in series or in parallel),
-by application of the right-hand side stencil at each point.
-When $A$ is tridiagonal ($\alpha \neq 0$, $\beta$ = $\hdots$ = 0),
-the evaluation of the derivatives requires the solution
-of the tridiagonal system, which yields the derivatives
-at all points simultaneously.
-The classical Thomas algorithm
-also referred to as the Tridiagonal Matrix Algorithm,
-is widely employed for this purpose.
+When $\alpha$ = $\beta$ = $\hdots$ = 0,
+$A$ is a $\emph{diagonal}$ matrix.
+In this case, the
+scheme is \emph{explicit}
+and the derivatives are straightforward to evaluate:
+this simply involves the
+application of the right-hand side stencil at each point.
+When $\alpha \neq 0, \beta = \hdots$ = 0,
+$A$ is \emph{tridiagonal}.
+The evaluation of the derivatives requires
+the solution of the resulting tridiagonal system.
+The classical Thomas algorithm---also
+referred to as the Tridiagonal Matrix Algorithm---is
+widely employed for this purpose.
 The algorithm is straightforward to implement,
 and stable for diagonally dominant matrices ~\cite{numericalrecipes}.
+When $\alpha \neq 0$ and $\beta \neq  0$,
+pentadiagonal and higher-bandwidth systems arise.
+These can be more expensive and difficult to evaluate
+than tridiagonal systems
+using a direct method.
+
+This work will focus on
+compact finite difference schemes that lead
+to tridiagonal systems.
+An example of such a scheme is obtained
+by substituting 
+$\beta = 0$, $\alpha = \frac{1}{4}$ and $a = \frac{3}{4}$
+in Eq. \ref{eqn:general-compact}
+(all other coefficients are set to zero).
+This leads to a fourth-order accurate
+tridiagonal compact finite difference scheme,
+known also as the Pad\'{e} scheme.
+At the boundaries,
+the following \emph{third-order} accurate
+approximations are used:
+
+\begin{align}
+    f^{\prime}_1 + 2f^{\prime}_2 = \frac{-5f_1 + 4f_2 + f_3}{dx} \\
+    f^{\prime}_{n} + 2f^{\prime}_{n-1}
+    =
+    \frac{5f_{n} - 4f_{n-2} -  f_{n-1}}{dx}
+\end{align}
+%
+The resulting tridiagonal system is then:
+
+\begin{equation}
+ \label{eqn:compact-tridiagonal-system}
+ \begin{bmatrix}
+     1&2\\
+     1/4&1&1/4\\
+     &1/4&1&1/4\\
+     &&1/4&1&1/4\\
+     &&&1/4&1&1/4\\
+     &&&&&\ddots\\
+     &&&&&&\ddots\\
+     &&&&&&&\ddots\\
+     &&&&&&&2&1
+  \end{bmatrix}
+  \begin{bmatrix}
+      f^{\prime}_1 \\
+
+      f^{\prime}_2 \\
+      f^{\prime}_3 \\
+      \vdots \\
+      \vdots \\
+      \vdots \\
+      \vdots \\
+      f^{\prime}_{n-1} \\
+      f^{\prime}_n
+   \end{bmatrix}
+ =
+ \begin{bmatrix}
+     \frac{-5f_1 + 4f_2 + f_3}{2}\\
+     \frac{3(f_{3} - f_{1})}{4}\\
+     \frac{3(f_{4} - f_{2})}{4}\\
+     \vdots\\
+     \vdots\\
+     \vdots\\
+     \vdots\\
+     \frac{3(f_{n} - f_{n-2})}{4}\\
+     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2}
+  \end{bmatrix}
+\end{equation}
+
+Similar tridiagonal schemes are available
+to evaluate higher order derivatives.
+Compact schemes for
+second, third and fourth derivatives
+are derived in ~\cite{lele1992compact}.
+
+\section{Graphics processing units}
+
+\subsection{Introduction}
+\subsection{Memory model}
+\subsubsection{Global memory}
+\subsubsection{Shared memory}
+\subsubsection{Registers and local memory}
+\subsection{CUDA Programming model}
+\subsubsection{Kernels}
+\subsubsection{Thread organization}
+
+\section{Tridiagonal solvers}
+\subsection{Thomas algorithm}
+\subsection{Cyclic reduction}
+\subsection{Other algorithms}

### 2015-11-15 thesis/proposed-tridiagonal-algorithm.tex
update sections and add content to thesis

\chapter{Proposed tridiagonal algorithm}

\section{Modified cyclic reduction}
\subsection{Analysis of forward reduction coefficiets}
\subsection{Precomputing coefficients}
\subsection{Forward reduction}
\subsection{Back substitution}

\section{Implementation}
\subsection{General strategy}
\subsection{Global memory implementation}
\subsubsection{Algorithm}
\subsubsection{Kernels}
\subsubsection{Grid and thread block configuration}

\subsection{Shared memory implementation}
\subsubsection{Algorithm}
\subsubsection{Kernel}
\subsubsection{Grid and thread block configuration}

### 2015-11-15 thesis/results.tex
update sections and add content to thesis

\chapter{Results}

\section{Validation}
    \subsection{Validation of tridiagonal solver}
    \subsection{Validation of compact finite difference application}
        \subsubsection{Tests for individual kernels}
        \subsubsection{Tests for communication routines}
        \subsubsection{Test functions for derivatives}
        \subsubsection{Order of convergence tests}
    \subsubsection{Validation results}

\section{Performance of GPU tridiagonal solver}
\section{Performance of compact finite difference application}

### 2015-11-15 thesis/fig/plot-wavenumbers.py
add script to generate wavenumbers plot

@@ -1,0 +1,25 @@
+import numpy as np
+from numpy import sin, cos
+import matplotlib.pyplot as plt
+
+plt.style.use('bmh')
+
+def get_wprime(w, a, b, c, alpha, beta):
+    return (a*sin(w) + (b/2)*sin(2*w) + (c/3)*sin(3*2))/(1 + 2*alpha*cos(w) + 2*beta*cos(2*w))
+
+w = np.linspace(0, np.pi, 100)
+w_prime_second_central = get_wprime(w, 1.0, 0.0, 0.0, 0.0, 0.0)
+w_prime_fourth_central = get_wprime(w, 4./3, -1./3, 0.0, 0.0, 0.0)
+#w_prime_sixth_central = get_wprime(w, 1.0, 0.0, 0.0, 0.0, 0.0)
+w_prime_pade = get_wprime(w, 3./2, 0.0, 0.0, 1./4, 0.0)
+w_prime_sixth_tridiagonal = get_wprime(w, 14./9, 1./9, 0, 1./3, 0.0)
+plt.plot(w, w_prime_second_central, 'k', linewidth=2, label='2nd order central')
+plt.plot(w, w_prime_fourth_central, 'k:', linewidth=2, label='4th order central')
+plt.plot(w, w_prime_pade, 'k--', linewidth=2, label='4th order, compact tridiagonal')
+plt.plot(w, w_prime_sixth_tridiagonal, 'k-.', linewidth=2, label='6th order, compact tridiagonal')
+plt.plot(w, w, 'k-', linewidth=3, label='exact')
+
+plt.legend(loc='best')
+plt.ylabel('$hk^{\prime}$')
+plt.xlabel('$hk$')
+plt.savefig('modified-wavenumbers.eps')

### 2015-11-16 thesis/introduction-and-background.tex
updates to thesis

@@
 \subsection{Tridiagonal compact schemes}
+\label{sec:tridiagonal-compact-schemes}
 
@@
 the solution of the resulting tridiagonal system.
-The classical Thomas algorithm---also
-referred to as the Tridiagonal Matrix Algorithm---is
-widely employed for this purpose.
-The algorithm is straightforward to implement,
-and stable for diagonally dominant matrices ~\cite{numericalrecipes}.
 When $\alpha \neq 0$ and $\beta \neq  0$,
@@
 
-\subsection{Introduction}
-\subsection{Memory model}
-\subsubsection{Global memory}
-\subsubsection{Shared memory}
-\subsubsection{Registers and local memory}
-\subsection{CUDA Programming model}
+The evolution of graphics processing units (GPUs)
+has traditionally been guided by the requirements
+of graphics applications such as video gaming.
+In general, these graphics computations involve
+manipulations and calculations with pixel data.
+Because these calculations are largely independent,
+the GPU hardware is specialized
+to perform several of them \emph{in parallel}.
+Therefore, the
+GPU may be regarded as a highly multithreaded processor.
+Initially,
+programming GPUs for scientific computation tasks
+was a cumbersome process,
+as the programming interfaces were designed for graphics computations,
+and attempts to use them for other purposes were essentially ``hacks''.
+Moreover, there was little support for double precision arithmetic,
+which is critical for many applications.
+But with the introduction of the
+NVIDIA Compute Unified Device Architecture (CUDA)
+parallel computing platform
+and programming model,
+CUDA-enabled GPUs have come into the
+mainstream in high-performance computing.
+A history of the development of GPUs
+for general purpose computation,
+and of CUDA specifically is available
+in ~\cite{HwuBook}.
+We provide the essential and relevant details about CUDA here,
+and refer to the
+CUDA Programming Guide \cite{CUDAProgrammingGuide}
+for a detailed outlook.
+
+\subsection{The CUDA Programming model}
+
+CUDA is the name for the parallel computing platform
+developed by NVIDIA,
+as well as the application programming interface
+for programming their GPUs.
+CUDA allows general-purpose applications to be
+more easily programmed for the GPU.
+Two essential features of the CUDA programming modl
+are \emph{kernels} and \emph{thread organization}.
+
 \subsubsection{Kernels}
+
+The CUDA application programming interface (API)
+is availabe as an extension to
+a programming languages (C, C++, Fortran, etc.),
+allowing certain portions of code to execute on the GPU.
+The rest of the code is executed as usual on the CPU.
+In CUDA terminology,
+the CPU is referred to as the \emph{host},
+and the GPU is referred to as the \emph{device}.
+The special pieces of code
+that execute on the GPU are known as \emph{kernels}.
+Kernels have similar syntax
+as ``host code'',
+but in general,
+are more restricted in the
+features of the underlying language they can use.
+In C, for example, kernels are written as \emph{functions},
+and are called by the host code
+using (almost) the same conventions.
+Thus, a C program that uses the GPU
+looks and behaves very much like a normal C program,
+but includes special calls to these kernels.
+When the application is launched,
+a CPU thread executes the host-code as usual,
+but upon encountering a call to the kernel,
+it passes program control over to the GPU.
+After the kernel is finished executing,
+control is passed back to the CPU,
+and this process may repeat when another kernel call is encountered.
+The CPU and GPU may also operate \emph{asynchronously},
+i.e.,
+control may be passed back to the CPU \emph{before} kernel completion,
+in which case
+explicit synchronization between the host and device may be necessary.
+Apart from kernel calls, the host code can also call functions to
+allocate and deallocate device memory,
+query device information,
+perform device synchronization, etc.
+
 \subsubsection{Thread organization}
 
-\section{Tridiagonal solvers}
+A kernel is executed in parallel by
+several lightweight \emph{threads}.
+Threads are organized into groups called
+\emph{thread blocks}, or just \emph{blocks},
+and the different blocks constitute a
+\emph{grid} of blocks.
+In many cases,
+the GPU is used to process array data,
+and each thread is mapped to a single array element.
+In most applications,
+the array is logically 1-, 2- or 3-dimensional.
+Thus, for convenience, the grid and block sizes
+can be 1-, 2- or 3-dimensional
+(hence the terminology \emph{grid} and \emph{blocks}).
+This kind of thread organization is optional,
+and $n$-dimensional arrays can be processed
+using just 1-dimensional grids and blocks.
+
+\subsection{CUDA architecture and memory model}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=250pt]{img/gpu-scaling.png}
+\end{center}
+\caption{Scheduling of blocks to streaming microprocessors---GPUs
+    with more SMs are able to execute more blocks concurrently.
+(\emph{CUDA Programming Guide ~\cite{CUDAProgrammingGuide}})}
+\label{fig:gpu-scaling}
+\end{figure}
+
+From the hardware perspective,
+an NVIDIA GPU may be viewed primarily as
+a collection of so-called Streaming Microprocessors (SMs).
+When a kernel is launched with specified
+grid size (number of blocks)
+and block size (threads per block),
+each block is assigned to an SM.
+Threads within a thread block can execute concurrently on the SM,
+and an SM can execute several thread blocks concurrently.
+It is imperative that blocks are able to run
+independently, in series or in parallel.
+This feature gives CUDA programs
+their scalability \ref{fig:gpu-scaling}.
+The same program runs faster on a GPU with a larger number
+of SMs, simply because more blocks may run concurrently.
+The number of thread blocks that an
+SM can execute concurrently
+is limited by a number of factors,
+and maximizing this number is often
+key to obtaining good performance.
+Another key consideration in the implementation
+of algorithms on GPUs is the
+available \emph{memory heirarchy}.
+Threads executing a kernel can read and write
+to several different memory spaces,
+each with different capacity, latency and bandwidth.
+
+\subsubsection{Global memory}
+
+Data from the host is first read into the
+device's \emph{global memory}.
+Similarly, data from the device is read back into the host
+from global memory.
+This is done in the host-code
+using special functions provided by the CUDA API 
+(e.g., \texttt{cudaMemcpy}).
+Data transfer between the host and the device is
+extremely slow relative to data access within the device.
+This data transfer rate is limited by the bandwidth
+of the PCI-e bus between the host and device.
+All threads executing a kernel can
+read and write to locations in global memory,
+and it is the largest memory space that is
+writable by the device.
+For example, the current NVIDIA Tesla K20 accelerator
+has about 4 Gigabytes of usable global memory.
+Thread access to global memory has a long latency
+and it is especially inefficient when
+successive threads in a block access memory locations
+that are far apart.
+Data in global memory remains persistent
+throughout the execution of the program.
+
+\subsubsection{Shared memory}
+
+Threads within a block have access to a
+common, fast \emph{shared memory}.
+All threads within a block
+can read and write to the block's shared memory,
+but they may \emph{not} access
+shared memory of another block.
+Thread access to shared memory can be expected to be
+much faster than access to global memory.
+Data that may be repeatedly used in a kernel
+are good candidates for placement in shared memory
+The contents of shared memory are managed by the kernel code,
+and for this reason,
+shared memory is often viewed as explicitly managed cache.
+Data in shared memory does\emph{not} persist after
+kernel execution.
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=300pt]{img/bank-conflicts.eps}
+\end{center}
+\caption{(a) Organization of shared memory as 4 byte words
+    organized into 32 banks. Each cell (square) is a word.
+    (b) Bank conflict free access by a warp. Each warp
+    accesses a word from a different bank. (c) 2-way
+    bank conflicts arising from successive threads accessing
+alternating words.}
+\label{fig:bank-conflicts}
+\end{figure}
+
+Shared memory is organized as a collection of  4 byte ``words''.
+Each consecutive word in shared memory belongs
+to a different bank - modern GPUs have 32 banks.
+The GPU schedules
+thread execution within a block
+in groups of 32 (termed as thread \emph{warps}).
+Ideally, these 32 threads
+can access shared memory locations concurrently.
+However, when two or more threads in a warp access
+words from the same bank,
+a \emph{bank conflict} is said to occurr,
+and the access is serialized.
+Thus,
+a warp of 32 threads reading
+32 successive \texttt{float}s (4 bytes) in a shared memory array
+is perfectly parallelized.
+But a warp of 32 threads reading
+32 alternating \texttt{float}s in a shared memory array
+leads to bank conflicts (Fig. \ref{fig:bank-conflicts}).
+Because the floats are placed in consecutive banks,
+the 1st and 17th threads will acess floats from the same bank,
+as will the 2nd and 18th threads, and so on.
+This specific kind of bank conflict is termed a
+\emph{two-way} bank conflict -
+threads access two words from the same bank.
+It is observed that higher-order bank conflicts may occur,
+if threads in a warp access
+\texttt{float}s in strides of 4,
+a \emph{four-way} bank conflict can occur.
+In the worst case, successive threads may access
+\texttt{float}s in strides of 32,
+in which case all threads in the warp
+request memory from the same bank:
+a emph{32-way} bank conflict.
+
+\subsubsection{Registers and local memory}
+
+Each thread also has private \emph{local memory},
+and access to extremely fast registers.
+Unlike CPUs, the GPU has a large number of registers---a
+thread executing a kernel will
+typically attempt to store
+non-array variables defined in the kernel in registers.
+Thread access to registers has the lowest latency
+compared to all other memory spaces.
+The number of registers available to each thread is limited,
+and if exceeded,
+data is instead stored in the thread's local memory.
+Access to local memory is much slower than registers,
+so this is typically avoided.
+
+\subsubsection{Limiting shared memory and register usage}
+
+While shared memory and registers can service memory requests
+much faster than global memory,
+their overuse can lead to performance degradation.
+The amount of shared memory per SM
+and the numer of registers per SM is limited.
+For the current NVIDIA Tesla K20 accelerator,
+the amount of shared memory per SM is limited to 48 KiB,
+and the number of registers per SM is limited to 65536.
+These are also the limits
+on the resources that can be allocated for each \emph{block}.
+However,
+allocating 48 KiB shared memory or using 65536 registers 
+for each block is ill-advised,
+as this effectively restricts the number of blocks that
+each SM can run at any given time to 1.
+If each block allocates 24 KiB of shared memory,
+then an SM can run only 2 blocks concurrently.
+Thus, the resources allocated per-block affects
+the overall parallelism that can be exploited from the GPU.
+
+\subsection{Considerations to be made while programming for GPUs}
+
+As seen in the previous sections,
+several factors must be considered while
+designing and implementing algorithms for GPUs.
+Failure to include these considerations can easily
+lead to poor performance,
+and no significant speedup may be noted over CPUs.
+In fact, one may even note performance degradation.
+We list the primary considerations here:
+
+\begin{itemize}
+    \item In any application, the data transfers between
+        the host and device must be minimized.
+        Ideally,
+        data is read into the device from the host \emph{once}
+        and from the device to the host \emph{once}.
+    \item Thread access to global memory must \emph{coalesced},
+        i.e., successive threads must read successive locations
+        in global memory.
+    \item Shared memory access must be free of bank conflicts
+        as much as possible---in general, this means avoiding
+        strided shared memory access.
+    \item The amount of shared memory and registers
+        allocated for each block is kept minimum.
+\end{itemize}
+
+\section{Tridiagonal solvers for compact finite difference evaluation}
+
+As mentioned in \ref{sec:tridiagonal-compact-schemes},
+this work is concernted with
+\emph{tridiagonal} compact finite differences
+for their relative ease in evaluation.
+The resulting tridiagonal system must be solved
+repeatedly for all lines of grid points in the computational grid.
+For a regular Cartesian grid (generally employed in DNS),
+the coefficients of the tridiagonal systems
+are the same for each grid line,
+and only the right hand sides are different.
+Thus, the compact finite difference evaluation
+effectively requires the solution of a tridiagonal system
+for several right hand sides.
+In this section,
+we describe the applicability of several algorithms
+for solving this problem on the GPU.
+
 \subsection{Thomas algorithm}
+
+Traditionally,
+the Thomas algorithm is employed for solving tridiagonal systems.
+The algorithm is derived
+from the more general Gaussian (LU) algorithm
+applied to tridiagonal matrices.
+Implementations of this algorithm are relatively straightforward
+and compact ~\cite{numericalrecipes}.
+Only the non-zero diagonals of the tridiagonal system
+are stored,
+and the algorithm performs $2N$ steps
+to solve the system.
+It is also stable for diagonally dominant matrices.
+The Thomas algorithm has an algorithmic complexity of $O(N)$,
+and is the fastest serial solver for tridiagonal systems.
+It is thus well suited to solution by a single CPU thread.
+While the algorithm itself is sequential,
+in a multithreaded environment,
+the different CPU threads may be used to solve
+the different independent tridiagonal systems simultaneously.
+
+A similar parallelization strategy may be extended to GPUs,
+where each GPU thread solves an independent tridiagonal system.
+The coefficient right hand sides are assumed to be stored
+contiguously as a single array in memory.
+If each thread works entirely with global memory,
+\emph{uncoalesced} memory access is observed.
+This cost \emph{may} be amortized for a large enough
+number of systems, as is the case
+for 3-D problems \cite{sakharnykhADIconf}.
+If shared memory is used,
+then low parallelism is exhibited,
+as the amount shared memory allocated by each thread
+is relatively high ($4N$).
+An approach that uses the parallel Thomas algorithm
+effectively is described by Chang et al. \cite{chang2012scalable}.
+Here, a \emph{parallel cyclic reduction} algorithm
+is first used to reduce a
+tridiagonal system into several smaller tridiagonal system,
+and the parallel Thomas algorithm is then used to
+solve the smaller systems in parallel.
+
 \subsection{Cyclic reduction}
+
+Two other popular algorithms for solving
+tridiagonal systems are the
+cyclic reduction (CR) and parallel cyclic reduction (PCR)
+algorithms.
+
 \subsection{Other algorithms}

### 2015-11-16 thesis/introduction-and-background.tex
add notes on cyclic reduction

@@
 \subsubsection{Shared memory}
+\label{subsubsec:shared-memory}
 
@@
 
+\begin{equation} \label{eqn:general-tridiagonal-system}
+\begin{pmatrix}
+     b_1 & c_1  \\
+     a_2 & b_2  &  c_2  \\
+         & a_3  &  b_3 &  c_3  \\
+         &      &  a_4 &  b_4 &  c_4  \\
+         &      &      &      &  \ddots \\
+         &      &      &      &     &  \ddots  \\
+         &      &      &      &     &  a_n  &  b_n
+\end{pmatrix}
+\begin{bmatrix}
+    x_1 \\
+    x_2 \\
+    x_3 \\
+    x_4 \\
+    \vdots \\
+    \vdots \\
+    x_n
+ \end{bmatrix}
+=
+\begin{bmatrix}
+   d_1 \\
+   d_2 \\
+   d_3 \\
+   d_4 \\
+   \vdots \\
+   \vdots \\
+   d_{n}
+\end{bmatrix}
+\end{equation}
+
 Traditionally,
-the Thomas algorithm is employed for solving tridiagonal systems.
+the Thomas algorithm is employed for solving
+general linear systems of the form $A\bm{x} = \bm{d}$,
+where $A$ is a tridiagonal matrix with diagonals
+$\bm{a}$, $\bm{b}$ and $\bm{c}$
+[Eq. (\ref{eqn:general-tridiagonal-system})].
 The algorithm is derived
@@
 and compact ~\cite{numericalrecipes}.
-Only the non-zero diagonals of the tridiagonal system
+Only the non-zero diagonals of the tridiagonal matrix
 are stored,
@@
 tridiagonal systems are the
-cyclic reduction (CR) and parallel cyclic reduction (PCR)
+cyclic reduction (CR) and
+the related parallel cyclic reduction (PCR)
 algorithms.
 
-\subsection{Other algorithms}
+\begin{figure}
+\begin{center}
+\includegraphics[height=250pt]{img/cyclic-reduction.eps}
+\end{center}
+\caption{Cyclic reduction.}
+\label{fig:cyclic-reduction}
+\end{figure}
+
+The cyclic reduction algorithm consists of two phases:
+\emph{forward reduction} and \emph{backward substitution}
+(Fig. \ref{fig:cyclic-reduction}).
+In the forward reduction phase,
+every even-indexed equation $i$
+is expressed as a
+linear combination of equations $i$, $i-1$ and $i+1$,
+yielding a new tridiagonal system of
+$n/2$ equations in $n/2$ unknowns
+[Eqs. (\ref{eqn:forward-reduction-1}) - (\ref{eqn:forward-reduction-4})].
+The process is repeated until a system of
+2 equations in 2 unknowns is left.
+
+\begin{align} 
+& a^{\prime}_i = -a_{i-1}k_1 \
+    \label{eqn:forward-reduction-1}& \\
+& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 \
+    \label{eqn:forward-reduction-2}& \\
+& c^{\prime}_i = -c_{i+1}k_2 \
+    \label{eqn:forward-reduction-3}& \\
+& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 \
+    \label{eqn:forward-reduction-4}&
+\end{align}
+%
+where,
+
+\begin{align}
+& k_1 = \frac{a_i}{b_{i-1}} \label{eqn:k1-update}& \\
+& k_2 = \frac{c_i}{b_{i+1}} \label{eqn:k2-update}&
+\end{align}
+
+The 2-by-2 system of equations is solved trivially,
+yielding $x_n$ and $x_{n/2}$.
+In the backward substitution phase,
+every odd-indexed unknown $x_i$ is solved for by
+substituting the known values of $x_{i-1}$ and $x_{i+1}$
+[Eq. (\ref{eqn:backward-substitution})].
+
+\begin{align} \label{eqn:backward-substitution}
+x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - \
+    c^{\prime}_ix_{i+1}}{b^{\prime}_i}
+\end{align}
+%
+For the last index $i=n$,
+the forward reduction step is instead:
+\begin{align} \label{eqn:forward-reduction-last}
+    & a^{\prime}_n = -a_{n-1}k_1 & \\
+    & b^{\prime}_n = b_n - c_{n-1}k_1 & \\
+    & d^{\prime}_n = d_n - d_{n-1}k_1&
+\end{align}
+%
+And for $i=1$, the backward substitution step is instead:
+
+\begin{align} \label{eqn:backward-substitution-first}
+x_1 = \frac{d^{\prime}_1 - c^{\prime}_1x_{2}}{b^{\prime}_1}
+\end{align}
+
+Thus, in the best case ($n$ parallel processors),
+cyclic reduction requires 
+$2log_2(n) - 1$ steps.
+For even moderately large $n$,
+this is significantly smaller than
+the $2n$ steps required by the Thomas algorithm.
+This makes cyclic reduction a good fit
+for massively parallel architectures like GPUs.
+
+\subsection{Parallel cyclic reduction}
+
+The parallel cyclic reduction (PCR) algorithm ~\cite{Zhang2010FTS}
+has only the forward reduction phase.
+The first forward reduction step is applied to
+the odd and even indexed equations seperately,
+yielding two reduced systems
+of size $n/2$.
+Forward reduction applied to both of these
+systems then yields \emph{four} reduced
+systems of size $n/4$.
+The process is repeated till
+$n/2$ 2-by-2 systems are left,
+all of which can be solved trivially.
+The PCR algorithm requires half the number of steps
+required by CR ($log_2(n)$),
+but does significantly more computation per-step.
+Further, unlike CR,
+PCR can be implemented free of bank conflicts.
+
+\subsection{Cyclic reduction implementation on GPUs}
+
+The algorithm proposed in this work is based on cyclic reduction,
+so it is pertinent to discuss the
+implementation of cyclic reduction on GPUs,
+the associated issues,
+and the relevant literature.
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=200pt]{img/gpu-mapping.eps}
+\end{center}
+\caption{Mapping work to blocks and threads:
+systems are mapped to blocks and
+indices are mapped to individual threads.}
+\label{fig:gpu-mapping}
+\end{figure}
+
+In the GPU implementation of cyclic reduction,
+blocks are assigned to tridiagonal systems
+(when solving multiple systems),
+and threads within a block 
+are assigned to equations, or \emph{indices}
+(Fig. \ref{fig:gpu-mapping}).
+In this way, several grid lines are solved concurrently
+by the GPU.
+During the forward reduction phase,
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=200pt]{img/forward-reduction-step.eps}
+\end{center}
+\caption{Updating $\bm{b}$ in the first forward reduction step.}
+\label{fig:forward-reduction-step}
+\end{figure}
+
+the threads assigned to each even index $i$
+compute the coefficients
+$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$.
+In practice, the coefficients are updated \emph{in-place}.
+Figure \ref{fig:forward-reduction-step} shows the updates
+to the coefficient array $\bm{b}$
+in the first forward reduction step).
+In each step of forward reduction,
+a thread accesses values from the coefficient arrays
+$\bm{a}$, $\bm{b}$ and $\bm{c}$ in a \emph{strided} fashion.
+At every subsequent step,
+this stride is doubled, while the number of active threads is halved
+\ref{fig:cyclic-reduction}.
+In the backward substitution phase,
+the strides are halved at each step,
+while the number of active threads is doubled.
+
+This \emph{hourglass} pattern results in low GPU utilization
+towards the end of forward reduction
+and the beginning of back substitution.
+Further,
+the long-strided memory accesses towards the end
+of the forward reduction
+and the beginning of backward substitution
+is problematic for both
+global memory and shared memory implementations.
+The global memory access is \emph{uncoalesced},
+leading to severe performance penalties.
+The use of shared memory can prevent
+long-strided global memory access.
+Unfortunately,
+the power-of-two strides at each successive
+leads to bank conflicts,
+as described in Sec. \ref{subsubsec:shared-memory}.
+The bank conflicts become increasingly severe
+at each forward reduction phase,
+and decreasingly so during the back substitution phase.
+The limited amount of shared memory places restricions
+on the size of tridiagonal systems that can be solved,
+and also on the number of blocks that can run in parallel.
+
+Despite these problems,
+cyclic reduction remains an attractive algorithm
+for GPUs,
+for its low algorithmic complexity,
+and lower work per step compared to PCR.
+Much work has been done on
+optimizing cyclic reduction performance on GPUs,
+and addressing the issues discussed.
+Zhang et al. \cite{Zhang2010FTS} propose a
+\emph{hybrid} solver
+that uses both cyclic reduction and parallel cyclic reduction
+to reduce bank conflicts.
+G{\"o}dekke et al. \cite{GoSt11CR}
+use a method of separately storing
+the even and odd indexed equations
+to arrive at a bank-conflict free solver
+at the cost of additional shared memory usage.
+Davidson et al. \cite{davidson2011register}
+describe the method of
+\emph{register packing}---performing more computations
+on \emph{registers}, rather than shared memory---as
+a means to reduce shared memory usage in cyclic reduction.
+Esfahanian et. al, \cite{esfahanian2014efficient}
+avoid shared memory (and associated bank conflicts entirely)
+using a data rearrangement scheme to improve global memory access.
+Our approach takes a different route,
+and is focussed on near-Toeplitz systems---exploiting
+the relatively simple matrix structure
+to reduce the number of computations and memory accesses
+performed by each thread
+at every cyclic reduction step.

### 2015-11-17 paper-draft/paper.tex
udpate bibliography
add chapter on near-toeplitz solver

@@
 cyclic reduction requires 
-$2log(n) - 1$ steps.
+$2log_2(n) - 1$ steps.
 For even moderately large $n$,
@@
 When the matrix $A$ is near-Toeplitz tridiagonal,
-if we substitute in these equations:
+if we substitute in the Eqs.
+Eqs. (\ref{eqn:forward-reduction-1}) -
+(\ref{eqn:forward-reduction-3}):
 

### 2015-11-17 thesis/abstract.tex
add chapter on near-toeplitz solver

\chapter*{Abstract}

We study the impact of the matrix structure on
the cyclic reduction algorithm
and show that that precomputing
forward reduction coefficients can be
an especially effective strategy.
his is the abstract

### 2015-11-17 thesis/introduction-and-background.tex
add chapter on near-toeplitz solver

@@
 \begin{align}
-    f^{\prime}_1 + 2f^{\prime}_2 = \frac{-5f_1 + 4f_2 + f_3}{dx} \\
+    f^{\prime}_1 + 2f^{\prime}_2 &= \frac{-5f_1 + 4f_2 + f_3}{dx} \\
     f^{\prime}_{n} + 2f^{\prime}_{n-1}
-    =
+    &=
     \frac{5f_{n} - 4f_{n-2} -  f_{n-1}}{dx}
@@
 As mentioned in \ref{sec:tridiagonal-compact-schemes},
-this work is concernted with
+this work is concerned with
 \emph{tridiagonal} compact finite differences
@@
 In this section,
-we describe the applicability of several algorithms
+we describe the applicability of some algorithms
 for solving this problem on the GPU.
+We refer to ~\cite{chang2014guide} for
+a more comprehensive list of algorithms,
+along with details of their GPU implementation.
 
@@
 \begin{align} 
-& a^{\prime}_i = -a_{i-1}k_1 \
-    \label{eqn:forward-reduction-1}& \\
-& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 \
-    \label{eqn:forward-reduction-2}& \\
-& c^{\prime}_i = -c_{i+1}k_2 \
-    \label{eqn:forward-reduction-3}& \\
-& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 \
-    \label{eqn:forward-reduction-4}&
+    a^{\prime}_i &= -a_{i-1}k_1 \
+    \label{eqn:forward-reduction-1} \\
+    b^{\prime}_i &= b_i - c_{i-1}k_1 - a_{i+1}k_2 \
+    \label{eqn:forward-reduction-2} \\
+    c^{\prime}_i &= -c_{i+1}k_2 \
+    \label{eqn:forward-reduction-3} \\
+    d^{\prime}_i &= d_i - d_{i-1}k_1  - d_{i+1}k_2 \
+    \label{eqn:forward-reduction-4}
 \end{align}
@@
 \begin{align}
-& k_1 = \frac{a_i}{b_{i-1}} \label{eqn:k1-update}& \\
-& k_2 = \frac{c_i}{b_{i+1}} \label{eqn:k2-update}&
+    k_1 &= \frac{a_i}{b_{i-1}} \label{eqn:k1-update} \\
+    k_2 &= \frac{c_i}{b_{i+1}} \label{eqn:k2-update}
 \end{align}
@@
 
-\begin{align} \label{eqn:backward-substitution}
+\begin{equation} \label{eqn:backward-substitution}
 x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - \
     c^{\prime}_ix_{i+1}}{b^{\prime}_i}
-\end{align}
+\end{equation}
 %
@@
 \begin{align} \label{eqn:forward-reduction-last}
-    & a^{\prime}_n = -a_{n-1}k_1 & \\
-    & b^{\prime}_n = b_n - c_{n-1}k_1 & \\
-    & d^{\prime}_n = d_n - d_{n-1}k_1&
+    a^{\prime}_n &= -a_{n-1}k_1  \\
+    b^{\prime}_n &= b_n - c_{n-1}k_1  \\
+    d^{\prime}_n &= d_n - d_{n-1}k_1
 \end{align}
@@
 
-\begin{align} \label{eqn:backward-substitution-first}
+\begin{equation} \label{eqn:backward-substitution-first}
 x_1 = \frac{d^{\prime}_1 - c^{\prime}_1x_{2}}{b^{\prime}_1}
-\end{align}
+\end{equation}
+%
+In practice, the right-hand side vector can be safely
+overwritten with the solution values
+in backward substitution.
 
@@
 \subsection{Cyclic reduction implementation on GPUs}
+\label{subsec:cyclic-reduction-gpu-implementation}
 
@@
 compute the coefficients
-$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$.
-In practice, the coefficients are updated \emph{in-place}.
+and right hand side for the reduced tridiagonal system
+$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
+and $d_i^\prime$.
+In practice, the coefficients and right hand side
+are updated \emph{in-place}.
 Figure \ref{fig:forward-reduction-step} shows the updates
 to the coefficient array $\bm{b}$
-in the first forward reduction step).
+in the first forward reduction step,
+a similar pattern is seen for the arrays
+$\bm{a}$, $\bm{c}$ and $\bm{d}$.
 In each step of forward reduction,
-a thread accesses values from the coefficient arrays
-$\bm{a}$, $\bm{b}$ and $\bm{c}$ in a \emph{strided} fashion.
+a thread accesses values from the
+coefficient arrays and right hand side
+in a \emph{strided} fashion.
 At every subsequent step,
@@
 
-This \emph{hourglass} pattern results in low GPU utilization
-towards the end of forward reduction
-and the beginning of back substitution.
-Further,
-the long-strided memory accesses towards the end
-of the forward reduction
-and the beginning of backward substitution
-is problematic for both
-global memory and shared memory implementations.
-The global memory access is \emph{uncoalesced},
-leading to severe performance penalties.
-The use of shared memory can prevent
-long-strided global memory access.
+Thus, the following issues are encountered
+while implementing cyclic reduction for GPUs:
+
+\begin{enumerate}
+    \item GPU utilization is low towards the end of forward reduction,
+        and in the beginning of backward substitution.
+
+    \item Because coefficients and right hand sides
+        are updated \emph{in-place},
+        synchronization between the blocks is required at
+        the end of each step.
+
+    \item If the threads work entirely with global memory,
+        memory accesses are increasingly
+        \emph{uncoalesced} in the forward reduction phase
+        (and become increasingly coalesced in the
+        backward substitution phase).
+
+    \item The use of shared memory prevents
+        uncoalesced global memory access.
         Unfortunately,
@@
         and decreasingly so during the back substitution phase.
-The limited amount of shared memory places restricions
+
+    \item The limited amount of shared memory places restricions
         on the size of tridiagonal systems that can be solved,
-and also on the number of blocks that can run in parallel.
+        and also on the number of systems that can be solved
+        concurrently.
+\end{enumerate}
 
-Despite these problems,
+Despite these issues,
 cyclic reduction remains an attractive algorithm
@@
 that uses both cyclic reduction and parallel cyclic reduction
-to reduce bank conflicts.
+to reduce the number and severity of bank conflicts,
+and also to have better thread activity overall.
 G{\"o}dekke et al. \cite{GoSt11CR}
@@
 Our approach takes a different route,
-and is focussed on near-Toeplitz systems---exploiting
+and is focussed on \emph{near-Toeplitz} systems---exploiting
 the relatively simple matrix structure

### 2015-11-17 thesis/proposed-tridiagonal-algorithm.tex
add chapter on near-toeplitz solver

@@
 
-\section{Modified cyclic reduction}
-\subsection{Analysis of forward reduction coefficiets}
-\subsection{Precomputing coefficients}
+Here, we explore the idea of
+exploiting the relatively simple matrix structure
+of the coefficient matrix appearing in
+tridiagonal compact finite difference schemes.
+The coefficient matrix $A$ for tridiagonal schemes
+has the following general form:
+
+\begin{equation} \label{eqn:toeplitz-matrix}
+A = 
+\begin{pmatrix}
+     b_1 & c_1  \\
+     a_0 & b_0  &  c_0  \\
+         & a_0  &  b_0 &  c_0  \\
+         &      &  a_0 &  b_0 &  c    \\
+         &      &      &      &  \ddots \\
+         &      &      &      &     &  \ddots  \\
+         &      &      &      &     &  a_n  &  b_n
+\end{pmatrix}
+\end{equation}
+%
+where, in general
+\begin{align*}
+    a_0 &\neq a_n  \\
+    b_1 &\neq b_0 \neq b_n \\
+    c_1 &\neq c_0 
+\end{align*}
+%
+We refer to matrices with this specific structure as
+\emph{near-Toeplitz tridiagonal matrices},
+and the corresponding linear systems as
+\emph{near-Toeplitz tridiagonal systems}.
+These matrices appear in a wide range of applications
+~\cite{sun1995application}
+such as alternating direction implicit methods,
+line relaxation methods,
+and numerical solutions to one-dimensional differential equations.
+In this chapter,
+we will describe an efficient solver for solving such
+near-Toeplitz tridiagonal systems efficiently on GPUs.
+
+\section{Modified cyclic reduction for near-Toeplitz systems}
+
 \subsection{Forward reduction}
-\subsection{Back substitution}
+
+The forward reduction phase reduces
+a $n-by-n$ system of equations to a
+$2-by-2$ system of equations in $log_2(n)-1$ steps,
+by applying
+Eqs. (\ref{eqn:forward-reduction-1}) -
+(\ref{eqn:forward-reduction-4})
+to every even-indexed equation at each step.
+When solving the tridiagonal systems
+with the same coefficient matrix,
+but repeatedly for different right hand sides,
+we note that the results of
+Eqs. (\ref{eqn:forward-reduction-1}) -
+(\ref{eqn:forward-reduction-3})
+remain unchanged for the different right hand sides.
+These results correspond to the
+coefficients of the tridiagonal system
+produced at each forward reduction step.
+Thus, given a tridiagonal system,
+we may precompute
+the coefficients of all
+the reduced systems appearing in the
+forward reduction steps,
+and reuse them for each right-hand side.
+We note that for a general tridiagonal system,
+this requires storage for $3.(2N - 2)$ coefficients,
+in addition to the $3N$ coefficients
+for the original tridiagonal system.
+
+Let us consider the case
+when the tridiagonal system is \emph{near-Toeplitz},
+i.e., when the coefficient matrix is of the form
+$A$ in Eq. \ref{eqn:toeplitz-matrix}.
+We examine the effect of the
+first forward reduction step by making the following substitutions
+in Eqs. (\ref{eqn:forward-reduction-1}) -
+(\ref{eqn:forward-reduction-3}):
+
+\begin{align*}
+    a_2 &= a_3 = a_4 = \hdots &\equiv a_0 \\
+    b_2 &= b_3 = b_4 = \hdots &\equiv b_0 \\
+    c_2 &= c_3 = c_4 = \hdots &\equiv c_0
+\end{align*}
+%
+We observe that the resulting coefficients
+$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
+correspond to the coefficients of a tridiagonal matrix with
+exactly the near-Toeplitz structure of $A$.
+This form-preserving property of cyclic reduction
+has been reported for block Toeplitz tridiagonal systems
+by Bini et al. \cite{bini}.
+
+The fact that the reduced system at each step
+is near-Toeplitz can be exploited
+to reduce the cost of
+precomputing and storing the coefficients.
+Each near-Toeplitz matrix is completely defined
+by only a handful of coefficients:
+$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
+making its storage extremely compact
+compared to the case of general tridiagonal systems.
+In addition,
+it is advantageous to precompute and store the
+auxiliary variables $k_1$ and $k_2$,
+which can similarly be stored compactly.
+With all the forward reduction coefficient matrices
+and auxillary variables precomputed and stored,
+the $m$th forward reduction step for equation $i$
+is reduced only to the right hand side update:
+
+\begin{equation}
+d^{\prime}_i = d_i - d_{i-1}k_1^{m}  - d_{i+1}k_2^{m}
+\label{eqn:precomputed-forward-reduction-step}
+\end{equation}
+%
+where $k_1^m$ and $k_2^m$ are
+precomputed values of $k_1$ and $k_2$
+for all ``inner'' equations at the $m$th step.
+For the ``outer'' equations $i=2$ and $i=n$,
+we have instead:
+%
+\begin{align}
+    d^{\prime}_2 &= d_2 - d_{1}k_{1,1}^{m}  - d_{3}k_2^{m} \
+    \label{eqn:precomputed-forward-reduction-step-2} \\
+    d^{\prime}_n &= d_n - d_{n-1}k_{1,n}^{m} \
+    \label{eqn:precomputed-forward-reduction-step-n} 
+\end{align}
+%
+here $k_{1,2}^m$ and $k_{1,n}^m$ are
+precomputed values of $k_1$
+for the ``outer'' equations at the $m$th step.
+
+\subsection{Backward substitution}
+
+The backward substitution step for all equations $i > 1$ is
+
+\begin{equation}
+x_i = \frac{d^{\prime}_i - a^mx_{i-1} - \
+    c^{m}x_{i+1}}{b^m}
+\label{eqn:precomputed-backward-substitution-step}
+\end{equation}
+%
+where $a^m$, $b^m$ and $c^m$ are the precomputed
+coefficients for $i>1$ at the step $m$.
+For the first equation, we have instead:
+%
+\begin{equation}
+x_1 = \frac{d^{\prime}_1 - c^{m}x_{2}}{b_1^m}
+\label{eqn:precomputed-backward-substitution-step-1}
+\end{equation}
+%
+where $b_1^m$ is the
+coefficient computed for $i = 1$ at step $m$.
 
 \section{Implementation}
-\subsection{General strategy}
+
+\subsection{Precomputing forward reduction coefficients}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=200pt]{img/cyclic-reduction-precomputing.eps}
+\end{center}
+\caption{Maximum storage required for forward reduction coefficients
+at all steps $m=1, 2, ... log_2(n)-1$}
+\label{fig:cyclic-reduction-precomputing}
+\end{figure}
+
+Figure \ref{fig:cyclic-reduction-precomputing}
+shows the
+\emph{maximum} amount of storage required for
+storing each of the coefficients
+$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
+or auxiliary variables $k_1$ and $k_2$.
+A careful analysis of the
+forward reduction and backward substitution equations
+reveals that the actual amount of storage is somewhat less.
+For instance,
+the values $b_n^m$ are unused
+in Eqs (\ref{eqn:precomputed-forward-reduction-step})-
+(\ref{eqn:precomputed-backward-substitution-step-1}).
+The values $a_n^m$ and $b_n^m$ are
+required only for solving the $2-by-2$
+system at the end of the forward reduction phase,
+i.e., for $m=log_2(n)-1$,
+and they are not stored for the previous steps.
+Similarly, the values $c_1^m$ are equal to $c_0^m$,
+and do not require special storage.
+The set of precomputed coefficients required to be stored is then
+$\{a_0^m, b_0^m, c_0^m, k_1^m, k_2^m, b_1^m, k_{1,1}^m, k_{1,n}^m\}$.
+Each of these ``coefficient arrays'' are computed on the CPU
+and transferred to the GPU.
+Additionally, the two scalars $a_n^{log_2(n)-1}$ and $b_n^{log_2(n)-1}$
+are required for the 2-by-2 solve.
+%
+\begin{figure}
+\begin{center}
+\includegraphics[height=200pt]{img/block-mapping.eps}
+\end{center}
+\caption{Storing right hand sides and mapping to thread blocks.}
+\label{fig:block-mapping}
+\end{figure}
+%
+The right hand sides that the system must be solved for
+are stored in a single contiguous array in GPU memory
+\ref{fig:block-mapping}.
+The precomputed coefficient arrays and the right hand side array
+are passed as inputs to the compute kernels that implement
+the modified cyclic reduction.
+We describe two implementations of the kernels,
+the first works entirely with global memory,
+and the second leverages the GPU's \emph{shared} memory.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=400pt]{img/global-and-shared.eps}
+\end{center}
+\caption{Thread activity in shared memory (top) and global memory (bottom) implementations.}
+\label{fig:global-and-shared}
+\end{figure}
+
 \subsection{Global memory implementation}
-\subsubsection{Algorithm}
-\subsubsection{Kernels}
-\subsubsection{Grid and thread block configuration}
+
+This implementation works entirely on the GPU's global memory,
+Here, we define two kernels - one for the forward reduction step,
+and the other for the backward substitution step.
+Each kernel is called $log_2(n)-1$ times.
+An extra call to the
+forward reduction kernel performs
+the two-by-two solve.
+At each step,
+the size of the thread blocks is determined by the \emph{stride}
+between elements accessed at that step.
+For the forward reduction phase,
+we use $n/2$ threads per block for the first step,
+$n/4$ threads for the second step, and so on.
+The pattern is reversed for the backward substitution phase,
+beginning with 2 threads per block for the first step.
+Although this ensures that there are no inactive threads
+at any stage,
+the occupancy of the GPU is still very low
+during the end of forward reduction and the end of backward substitution.
+
+\begin{listing}[H]
+\begin{minted}[fontsize=\footnotesize]{python}
+stride = 1
+for i in range(n-1):
+    stride *= 2
+    forwardReductionKernel(num_blocks=nrhs, block_size=n/stride)
+
+solveTwoByTwo()
+
+for i in range(n-1):
+    stride /= 2
+    backSubstitutionKernel(num_blocks=nrhs, block=n/stride)
+\end{minted}
+\end{listing}
+
+The precomputed coefficient arrays and right hand
+are accessed by the kernels from global memory.
+The kernel suffers from
+strided memory access for the right-hand side,
+but the precomputed coefficient values are accessed
+without major coalescing problems.
+Further, by precomputing the forward reduction coefficients,
+we greatly reduce the number of computations
+(and thus the number of uncoalesced memory accesses).
 
 \subsection{Shared memory implementation}
-\subsubsection{Algorithm}
-\subsubsection{Kernel}
-\subsubsection{Grid and thread block configuration}
+
+In the shared memory approach (Fig. \ref{fig:global-and-shared}),
+we launch a single kernel to perform the entire
+cyclic reduction solve.
+The kernel is launched with $n/2$ threads per block.
+Each thread block is allocated a block of shared memory of size $n/2$.
+Each thread of a block performs the first reduction step
+[Eq. (\ref{eqn:precomputed-forward-reduction-step})]
+by accessing the required values
+$d_i$, $d_{i-1}$ and $d_{i+1}$ from global memory,
+storing the result in shared memory.
+In subsequent reduction steps,
+$d_i$, $d_{i-1}$ and $d_{i+1}$
+are accessed from shared memory,
+avoiding the uncoalesced global memory accesses
+seen in the global memory implementation.
+In each back substitution step,
+threads overwrite the existing values in shared memory
+with the values of the solution.
+In the final step,
+shared memory is filled completely
+with the even-indexed solution values.
+Each thread then computes an odd-indexed solution value,
+storing it directly in global memory
+and copies the even-indexed solution value
+from shared memory to global memory.
+The entire solution is done in a single kernel launch
+to hide the latency of
+memory transfers between global and shared memory.
+Explicit synchronization between the threads of a block
+is required within the kernel at the end of each step.
+
+The shared memory implementation suffers from two major issues:
+first, the number of active threads is halved at each forward reduction step,
+(and subsequently doubled at each backward reduction step).
+Synchronization between threads of a block is necessary at each step.
+Thus, a significant portion of the kernel execution time is
+spent by idle threads waiting for active threads to complete execution.
+Secondly, the strided access to the
+right-hand side values leads to bank-conflicts.
+However, the number of bank conflicts is significantly smaller
+than in cyclic reduction implementations for general tridiagonal systems,
+due the reduced number of computations performed.

### 2015-11-17 thesis/compact-finite-difference.tex
updates

@@
 
-\section{Compact finite difference evaluation strategy}
-    \subsection{Shared memory systems}
-    \subsection{Distributed memory systems}
-    \subsection{Heterogeneous systems}
+\section{Compact finite difference evaluation on parallel systems}
+
+Because of the high computational cost
+and memory requirements of direct numerical simulation,
+parallelism is mandatory.
+The computation of derivatives using
+compact finite differences involves two primary steps:
+
+\begin{enumerate}
+\item The evaluation of the right hand sides
+    of Eq. (\ref{eqn:general-compact})
+    at each grid point.
+\item The solution of the tridiagonal system
+    [Eq. (\ref{eqn:general-compact})]
+    for each line of grid points.
+\end{enumerate}
+
+Both of these operations are amenable to parallel operations.
+The right hand side calculation is a pointwise
+\emph{stencil} operation, i.e.,
+at every point in the domain,
+the value of the right hand side
+is computed as a combination of function values
+at that point and its neighbouring points.
+The stencil operations at individual points
+are independent,
+making the overall calculation highly parallelizable.
+The solution of the tridiagonal systems,
+is of course parallelizable
+as has been discussed in previous sections.
+
+Without loss of generality,
+we consider the parallelization of
+compact finite difference evaluations for 2-dimensional problems.
+We assume that the derivative is being calulated
+for the coordinate direction along which
+elements are stored contiguously in memory, i.e.,
+the ``fastest'' coordinate direction.
+
+\subsection{Single GPU}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=150pt]{img/compact-single-gpu.eps}
+\caption{Computational domain in 3-D.}
+\label{fig:compact-single-gpu}
+\end{center}
+\end{figure}
+
+For a single GPU \ref{fig:compact-single-gpu},
+when calculating the right hand sides,
+each point in the grid is mapped to a single thread,
+and each thread applies the required stencil operation
+to compute the right hand side at that point.
+We note that threads near the left and right boundaries
+apply a different stencil from the interior threads.
+The implementation of GPU kernels for
+stencil operations such as these is a topic of wide study.
+The most important considerations were brought out
+in the paper by Micikevicius et al. ~\ref{micikevicius20093d}.
+For solving the tridiagonal systems,
+each thread block of the GPU is mapped to a
+different \emph{grid line}
+aligned along the
+direction the derivatives are being calculated.
+The right hand sides are stored
+contiguously along these grid lines
+(as in Fig. \ref{fig:block-mapping}),
+and the modified cyclic reduction algorithm developed
+can be used to solve for the derivatives.
+
+\subsection{Multiple GPUs on a single node}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=250pt]{img/compact-shared-gpu.eps}
+\caption{Compact finite differences - multiple GPUs
+    on same node}
+\label{fig:compact-shared-gpu}
+\end{center}
+\end{figure}
+
+Here, we consider the case of multiple GPUs
+on a single shared memory node, i.e.,
+multiple GPUs attached to the same PCI-e bus.
+In this case, every GPU is visible to the host,
+and the GPUs read from and write into
+the same host memory space.
+The domain is divided into a number of
+into ``subdomains'',
+as shown in \ref{fig:compact-shared-gpu}.
+The domain decomposition is done such that
+only a single subdomain is used along the
+coordinate direction of the derivatives.
+This is the method presented by
+Sakharnykh et al. ~\cite{sakharnykhADIconf},
+and it has the advantage that
+no coordination between GPUs is required:
+each GPU is assigned an independent
+set of grid lines to solve.
+
+\subsection{Distributed GPUs - restricted in one direction}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=200pt]{img/compact-distributed-restricted.eps}
+\caption{Compact finite differences - distributed GPUs and
+    restriced in one direction}
+\label{fig:compact-distributed-restricted}
+\end{center}
+\end{figure}
+
+For larger problems,
+a distributed system may be required.
+Here, the simplest strategy is to use a domain decomposition
+as shown in \ref{fig:compact-distributed-restricted}.
+Here again, 
+no inter-GPU communication is required.
+By restricting the distribution along one coordinate direction,
+the ease of solution of the tridaigonal systems is maintained.
+
+\subsection{Distributed GPUs in all directions}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=250pt]{img/compact-restricted-transpose.eps}
+\caption{Compact finite difference evaluation in both coordinate directions}
+\label{fig:compact-restricted-transpose}
+\end{center}
+\end{figure}
+
+The above domain decomposition strategies are convenient,
+but can be impractical for some cases.
+For instance,
+let us consider the evaluation of derivatives
+in the other coordinate direction
+(Fig. \ref{fig:compact-restricted-transpose}).
+Because the grid lines aligned along this direction
+must reside on the same GPU,
+it follows that:
+%
+\begin{enumerate}
+    \item A \emph{global} transposition or rearrangement
+        of the data is required,
+        such that each GPU now contains data for
+        grid lines aligned in the direction orthogonal
+        to the previous direction.
+        For distributed systems,
+        this transposition can be an extremely expensive process.
+
+    \item For domains that are much longer along one coordinate
+        direction compared to the other(s),
+        the subdomains may become impractically \emph{slender}.
+\end{enumerate}
+%
+Such decomposition strategies are therefore, generally applicable
+when compact finite difference schemes are used only
+in a single direction.
+For the other coordinate directions,
+explicit finite difference schemes may be used,
+which has the disadvantages discussed earlier.
+To accomodate compact finite difference schemes in all the coordinate
+directions,
+the domain decomposition generally must be performed in all directions,
+as shown in \ref{fig:compact-distributed-all}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=200pt]{img/compact-distributed-all.eps}
+\caption{Compact finite differences - distributed GPUs in all
+    directions}
+\label{fig:compact-distributed-all}
+\end{center}
+\end{figure}
 
 \section{Distributed tridiagonal solver}
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=150pt]{img/computational-domain.eps}
+\caption{Computational domain in 3-D.}
+\label{fig:computational-domain}
+\end{center}
+\end{figure}
+%
+The computational domain considered
+is a 3-dimensional,
+regular, Cartesian grid with
+$nx$, $ny$ and $nz$ grid points and grid spacing of
+$dx$, $dy$ and $dz$ in the
+respective coordinate directions,
+\ref{fig:computational-domain},
+Structured grids such as these
+are widely used in DNS because they lend themselves naturally
+to finite-difference methods.
+
     \subsection{General algorithm}
+
+    Algorithm by Mattor et al.
+
     \subsection{Specialization for compact finite difference
@@
 
+    Right hand side evaluation
+
 \section{GPU application}
-    \subsection{General strategy}
+
     \subsection{GPU-GPU communication}
         \subsubsection{Naive strategy}
+
+        GPU-CPU-CPU-GPU
+
         \subsubsection{NVIDIA GPUDirect Technology}
+
+        GPU-GPU
+
     \subsection{Implementation}
         \subsubsection{Precomputing coefficients}
+
+        Recognize the coefficients of the
+        local tridiagonal system
+
         \subsubsection{Solving secondary systems}
+
+        The computation is offline,
+        solve only one system
+        Move results to GPU
+
         \subsubsection{Computing right hand sides}
+
+        Stencil code
+        Show the halo swapping strategy
+
         \subsubsection{Solving primary systems}
+
+        Each of the systems is solved using the NEATO algorithm
+
         \subsubsection{Constructing and solving reduced systems}
+
+        Interleaved tridagonal systems,
+        solved using a p-Thomas strategy (reference?)
+
         \subsubsection{Summing primary and secondary solutions}
+
+        Completely pointwise
+
         \subsubsection{Evaluating derivatives in other
             coordinate directions}
+
+        Permutation kernels

### 2015-11-17 thesis/introduction-and-background.tex
updates

@@
 \subsection{Wavenumber analysis of finite difference methods}
+\label{subsec:wavenumber-analysis}
 
@@
 may be considered.
-While they are traightforward to compute,
+While they are straightforward to compute,
 they are associated with large stencil widths.
@@
 for a given stencil width.
-Thus, they are widely used in
-CFD for evaluation of spatial derivatives.
+This makes them the superior choice for DNS,
+in which the flow quantities
+exhibit spatial variation associated with high wavenumbers.
 

### 2015-11-17 thesis/compact-finite-difference.tex
updates

@@
 
-\section{Compact finite difference evaluation on parallel systems}
+\section{Introduction}
 
+Here, we discuss the application of the
+tridiagonal solver developed in the previous section
+to the evaluation of compact finite differences---which
+are used widely in the direct numerical simulation
+of fluid flows.
+
+\begin{figure}
+\begin{center}
+\includegraphics[height=150pt]{img/computational-domain.eps}
+\caption{Computational domain in 3-D.}
+\label{fig:computational-domain}
+\end{center}
+\end{figure}
+%
+In typical DNS applications,
+the computational domain considered
+is a 3-dimensional,
+regular, Cartesian grid with
+$nx$, $ny$ and $nz$ grid points and grid spacing of
+$dx$, $dy$ and $dz$ in the
+respective coordinate directions,
+\ref{fig:computational-domain},
+Structured grids such as these
+are widely used because they lend themselves naturally
+to finite-difference methods.
+To resolve all the spatial features of the flow,
+the grid spacing is kept very small.
+Consequently,
+to simulate flows of practical sizes,
+the number of grid points must be very large.
 Because of the high computational cost
-and memory requirements of direct numerical simulation,
-parallelism is mandatory.
+and memory requirements associated with
+the large number of grid points
+parallelism becomes mandatory.
+We have spoken so far of GPUs as massively parallel systems.
+However, each GPU has a very limited amount of global memory:
+current NVIDIA Tesla accelerators have about 4 Gigabytes.
+Thus, to accomodate the
+larger problem sizes in DNS,
+the use of \emph{multiple} GPUs is necessary,
+which introduces a second level of parallelism.
+In the next section,
+we describe strategies to distribute
+the problem domain among the GPUs
+in such dual-level parallel systems.
+
+\section{Compact finite difference evaluation on parallel GPU systems}
+
 The computation of derivatives using
@@
 \includegraphics[height=150pt]{img/compact-single-gpu.eps}
-\caption{Computational domain in 3-D.}
+\caption{Compact finite differences - single GPU}
 \label{fig:compact-single-gpu}
@@
 explicit finite difference schemes may be used,
-which has the disadvantages discussed earlier.
+which have the disadvantages discussed earlier.
 To accomodate compact finite difference schemes in all the coordinate
@@
 the domain decomposition generally must be performed in all directions,
-as shown in \ref{fig:compact-distributed-all}.
+as shown in Fig. \ref{fig:compact-distributed-all}.
+In this strategy,
+the grid lines in all coordinate directions
+are interrupted by the subdomain boundaries.
+Thus,
+inter-GPU communication is required.
+For the right hand sides evaluation,
+each GPU must communicate information at the subdomain boundaries
+with neighbouring GPUs.
+For example, in Fig. \ref{fig:compact-distributed-all},
+the right-most grid points in the subdomain of
+GPU 2 require data from the
+left-most grid points in the subdomain of
+GPU 6 when evaluating derivatives in the horizontal direction.
+Similarly,
+the left-most grid points in the subdomain of
+GPU 6 require data from the
+right-most grid points in the subdomain of
+GPU 2.
+Thus, a ``swapping'' of the boundary information
+is required at each of the subdomain boundaries.
+
+The solution of the distributed tridiagonal systems
+involves much more complexity,
+and is discussed in detail in the next section.
 
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[height=150pt]{img/computational-domain.eps}
-\caption{Computational domain in 3-D.}
-\label{fig:computational-domain}
-\end{center}
-\end{figure}
+\subsection{General algorithm}
+\label{subsec:distributed-tridiagonal-algorithm}
+
+We use the algorithm
+proposed by Mattor et al. \cite{mattor1995algorithm}
+for solving the distributed tridiagonal systems.
+For a general tridiagonal system $A\bm{x}=\bm{r}$,
+the algorithm begins by partitioning
+the matrix and right-hand side among the $P$ processes.
+Each process $p$ is then associated with the
+following $\emph{subsystems}$
+
+\begin{align}
+    & A^p\bm{u}^p = \bm{r_u}^p & \label{eqn:secondary-system-1} \\ 
+    & A^p\bm{l}^p = \bm{r_l}^p & \label{eqn:secondary-system-2} \\
+    & A^p\bm{x_r}^p = \bm{r}^p & \label{eqn:primary-system} 
+\end{align}
 %
-The computational domain considered
-is a 3-dimensional,
-regular, Cartesian grid with
-$nx$, $ny$ and $nz$ grid points and grid spacing of
-$dx$, $dy$ and $dz$ in the
-respective coordinate directions,
-\ref{fig:computational-domain},
-Structured grids such as these
-are widely used in DNS because they lend themselves naturally
-to finite-difference methods.
+expanded as:
 
-    \subsection{General algorithm}
+\begin{align}
+& \begin{bmatrix}
+b_1^p & c_1^p \\
+a_2^p & b_2^p & c_2^p \\
+      & a_3^p & b_3^p & c_3^p \\
+      &       & a_4^p & b_4^p & c_4^p \\
+      &       &       &       &  \ddots & c_{m-1}^p\\
+      &       &       &       &     a_{m}^p  & b_{m}^p
+\end{bmatrix}
+\begin{bmatrix}
+x_{r,1}^p \\
+x_{r,2}^p \\
+x_{r,3}^p \\
+x_{r,4}^p \\
+\vdots \\
+x_{r,m}^p
+\end{bmatrix}
+=
+\begin{bmatrix}
+r_1^p \\
+r_2^p \\
+r_3^p \\
+r_4^p \\
+\vdots \\
+r_m^p
+\end{bmatrix} & \label{eqn:primary-system-expanded} \\
+\end{align}
 
-    Algorithm by Mattor et al.
+\begin{align}
+& \begin{bmatrix}
+b_1^p & c_1^p \\
+a_2^p & b_2^p & c_2^p \\
+      & a_3^p & b_3^p & c_3^p \\
+      &       & a_4^p & b_4^p & c_4^p \\
+      &       &       &       &  \ddots & c_{m-1}^p\\
+      &       &       &       &     a_{m}^p  & b_{m}^p
+\end{bmatrix}
+\begin{bmatrix}
+u_1^p \\
+u_2^p \\
+u_3^p \\
+u_4^p \\
+\vdots \\
+u_m^p
+\end{bmatrix}
+=
+\begin{bmatrix}
+-a_1^p \\
+0 \\
+0 \\
+0 \\
+\vdots \\
+0
+\end{bmatrix} & \label{eqn:secondary-system-1-expanded} \\
+\end{align}
+%
+\begin{align}
+& \begin{bmatrix}
+b_1^p & c_1^p \\
+a_2^p & b_2^p & c_2^p \\
+      & a_3^p & b_3^p & c_3^p \\
+      &       & a_4^p & b_4^p & c_4^p \\
+      &       &       &       &  \ddots & c_{m-1}^p\\
+      &       &       &       &     a_{m}^p  & b_{m}^p
+\end{bmatrix}
+\begin{bmatrix}
+l_1^p \\
+l_2^p \\
+l_3^p \\
+l_4^p \\
+\vdots \\
+l_m^p
+\end{bmatrix}
+=
+\begin{bmatrix}
+0 \\
+0 \\
+0 \\
+0 \\
+\vdots \\
+-c_m^p
+\end{bmatrix} & \label{eqn:secondary-system-2-expanded}
+\end{align}
+%
+We refer to the subsystem in Eq. (\ref{eqn:primary-system-expanded})
+the ``primary'' system, and the subsystems in
+Eqs. (\ref{eqn:secondary-system-1-expanded}) and
+(\ref{eqn:secondary-system-2-expanded})
+the ``secondary'' systems.
+Additionally, the following ``reduced'' system is solved:
+
+\begin{align} \label{eqn:reduced-system}
+&
+\begin{bmatrix}
+l^1_m & -1 \\
+-1    & u^2_1 & l^2_1 \\
+      & u^2_m & l^2_m & -1 \\
+      &       & -1    & u^3_1 & l^3_1 \\
+      &       &       & u^3_m & l^3_m  & -1 \\
+      &       &       &       & \ddots & \ddots & \ddots \\
+      &       &       &       &        & -1     & u^P_1
+\end{bmatrix}
+\begin{bmatrix}
+\beta^1 \\
+\alpha^2 \\
+\beta^2 \\
+\alpha^3 \\
+\beta^3 \\
+\vdots \\
+\alpha^P
+\end{bmatrix}
+=
+\begin{bmatrix}
+x_{r,m}^1 \\
+x_{r,1}^2 \\
+x_{r,m}^2 \\
+x_{r,1}^3 \\
+x_{r,m}^3 \\
+\vdots \\
+x_{r,1}^P \\
+\end{bmatrix}
+&
+\end{align}
+The local part of the solution
+is obtained as a linear combination of
+the solutions to the primary and secondary systems:
+
+\begin{equation}
+    \bm{x}^p = \bm{x}_r^p + \
+        \alpha^p \bm{u}^p + \beta^p \bm{l}^p
+    \label{eqn:sum-of-systems}
+\end{equation}
+%
+where $\alpha^p$ and $\beta^p$ are obtained
+from the solution of the reduced system.
 
@@
 
-    Right hand side evaluation
+The algorithm described in
+Sec. \ref{subsec:distributed-tridiagonal-algorithm}
+is applicable to a single tridiagonal system distributed
+accross several processors.
+When solving a distributed tridiagonal system
+for several distributed grid lines,
+we note that
+the secondary systems
+[Eq. (\ref{eqn:secondary-system-1-expanded}) and
+ Eq. (\ref{eqn:secondary-system-2-expanded})]
+are dependent only on $\bm{a}$, $\bm{b}$ and $\bm{c}$,
+the tridiagonal coefficients of the global tridiagonal system.
+These are the same for all local grid lines,
+so they are solved for a single local grid line in each subdomain.
+Only the system in Eq. (\ref{eqn:primary-system-expanded})
+is solved for all the grid lines in a subdomain.
+For a single tridiagonal system,
+the reduced system requires \emph{global} communication
+of the values $\{u^p_1, u^p_m\, l^p_1, l^p_m\, x^p_{r,1} x^p_{r,m}\}$
+from each process $p$.
+When solving for several grid lines,
+the left hand side may be assembled for a single grid line.
+The right hand sides, however, must be assembled for every grid line,
+and the system is solved for each right hand side.
 
-\section{GPU application}
+\section{GPU implementation}
 
-    \subsection{GPU-GPU communication}
-        \subsubsection{Naive strategy}
+Our algorithm is outlined in \ref{fig:compact-algorithm},
+and is described for the case of evaluating derivatives
+in the direction along which successive
+function values in a subdomain are stored contigiously in memory,
+i.e., the ``fastest'' coordinate direction.
+Each step of the algorithm must be implemented on the GPU
+to avoid data transfer to and from the CPU,
+which is prohibitively expensive for large problems.
+Thus, we have several kernels to implement the algorithm.
+For communication of data between processes,
+we use the Message Passing Interface (MPI),
+and leverage the NVIDIA GPUDirect Technology
+for GPU-GPU communication.
+MPI is interfaced via the mpi4py \cite{dalcin2005mpi}
+Python library.
+The purpose of each CUDA kernel and MPI call used
+is described in Table \ref{table:compact-algorithm}.
 
-        GPU-CPU-CPU-GPU
+\begin{figure}
+\begin{center}
+\includegraphics[trim={0 7cm 0 7cm},clip,width=500pt]{img/compact-algorithm.pdf}
+\centering
+\caption{Algorithm for evaluating compact finite differences on multiple GPUs,
+(right: CUDA kernels and MPI calls used)}
+\label{fig:compact-algorithm}
+\end{center}
+\end{figure}
 
-        \subsubsection{NVIDIA GPUDirect Technology}
+\begin{table}[]
+\centering
+\caption{Purpose of kernels and MPI calls in compact finite difference application}
+\label{table:compact-algorithm}
+\begin{tabular}{|l|p{6cm}|}
+\hline
+Memcpy3d            & Copy noncontiguous boundary information
+                      of the function values
+                      to and from contiguous halo arrays \\ \hline
+ISend, IRecv        & Perform halo swaps with i-1 and i+1 processes \\ \hline
+computeRHSKernel    & Apply pointwise stencil operator to compute
+                      RHS at each grid point in the subdomain,
+                      using the halo values near the boundaries\\ \hline
+NEATO               & Solve the primary near-Toeplitz tridiagonal systems
+                      for the computed right hand sides, giving $x_r$ \\ \hline
+copyFacesKernel     & Copy the left and right faces of $x_r$
+                      into a single contiguous array \\ \hline
+MPI\_Gather         & Gather the data required to assemble
+                      the reduced system at rank 0 \\ \hline
+reducedSolverKernel & Solve the reduced systems for parameters
+                      $\alpha^p$ and $\beta^p$
+                      for each grid line \\ \hline
+MPI\_Scatter        & Scatter the parameters 
+                      $\alpha^p$ and $\beta^p$
+                      from rank 0 to all the processes \\ \hline
+sumSolutionsKernel  & Sum the primary and secondary solutions
+                      to compute the local part of the solution \\ \hline
+\end{tabular}
+\end{table}
 
-        GPU-GPU
-
-    \subsection{Implementation}
-        \subsubsection{Precomputing coefficients}
-
-        Recognize the coefficients of the
-        local tridiagonal system
-
-        \subsubsection{Solving secondary systems}
-
-        The computation is offline,
-        solve only one system
-        Move results to GPU
-
-        \subsubsection{Computing right hand sides}
-
-        Stencil code
-        Show the halo swapping strategy
-
-        \subsubsection{Solving primary systems}
-
-        Each of the systems is solved using the NEATO algorithm
-
-        \subsubsection{Constructing and solving reduced systems}
-
-        Interleaved tridagonal systems,
-        solved using a p-Thomas strategy (reference?)
-
-        \subsubsection{Summing primary and secondary solutions}
-
-        Completely pointwise
-
-        \subsubsection{Evaluating derivatives in other
-            coordinate directions}
-
-        Permutation kernels
+\begin{figure}
+\begin{center}
+\includegraphics[width=200pt]{img/constructing-reduced-system.eps}
+\caption{Construction of the reduced system and scattering of parameters.}
+\label{fig:constructing-reduced-system}
+\end{center}
+\end{figure}
+%
+The secondary systems
+[Eq. (\ref{eqn:secondary-system-1-expanded}) and
+ Eq. (\ref{eqn:secondary-system-2-expanded})]
+are solved on the CPU,
+and the results are transferred to the GPU.
+%
+The primary system [Eq. (\ref{eqn:primary-system-expanded})] must be solved
+for each of the local grid lines in a subdomain,
+as the right hand sides are different at each of the local grid lines. 
+The evaluation of the right hand sides are
+pointwise stencil computations,
+which require communication of the
+function values at the boundaries of the subdomains.
+This is achieved using a halo-swapping technique
+with dedicated contiguous halo arrays for each subdomain,
+as communication of non-contiguous MPI
+data types is expensive on the GPU.
+The primary system is solved for all the local grid lines
+using the NEATO solver.
+%
+The set up and solution of the reduced system
+requires global communication of the boundary information
+from the solutions $\bm{u}^p$, $\bm{l}^p$ and $\bm{x_r}^p$.
+The tridiagonal coefficients of the reduced system
+are set up easily,
+by communicating the boundary elements from
+$\bm{u}^p$ and $\bm{l}^p$,
+which are the same for every local grid line in each subdomain.
+The right hand sides require significantly more communication,
+as they are assembled from the boundary elements of $\bm{x_r}^p$,
+which are different for each local grid line in each subdomain.
+Thus, the boundary ``faces'' of each subdomain need to be communicated.
+These faces are first copied into a contiguous array,
+and these arrays are gathered at rank 0
+to assemble the right-hand sides of the reduced system
+(Fig. \ref{fig:constructing-reduced-system}).
+This communication strategy has the effect of
+producing \emph{interleaved} right hand sides
+aligned along the ``slowest'' co-ordinate direction, i.e.,
+the right hand side values for neighbouring grid points
+are located far apart in memory.
+As the reduced systems are quite small relative to the primary systems,,
+we use the convenient, but rather inefficient
+p-Thomas algorithm to solve the systems.
+The solution of the reduced systems produces the parameters
+$\alpha^p$ and $\beta^p$,
+which are scattered back to the respective processes, $p$.
+%
+Finally, the summing of the solutions
+is a pointwise operation that is easily implemented
+on the GPU.
+For evaluation of compact finite differences in other
+coordinate directions,
+a local permutation of the data is performed on the input data
+(function values)
+before applying the above algorithm.
+and again on the output data
+(derivative values).

### 2015-11-17 thesis/abstract.tex
updates

@@
 
+\emph{Compact} finite difference schemes are widely used
+in the direct numerical simulation of fluid flows,
+for their ability to better resolve the small scales of turbulence.
+However, they can be expensive to evaluate and
+difficult to parallelize.
+In this work, we present an approach for the computation
+of compact finite differences
+on graphics processing units (GPUs).
+We present a variant of the \emph{cyclic reduction} 
+algorithm for solving the
+tridiagonal linear systems that arise in such numerical schemes.
 We study the impact of the matrix structure on
@@
 forward reduction coefficients can be
-an especially effective strategy.
-his is the abstract
+especially effective for obtaining good performance.
+Our tridiagonal solver is able to outperform
+the NVIDIA CUSPARSE
+and the multithreaded Intel MKL tridiagonal solvers
+on GPU and CPU respectively.
+In addition,
+we present a parallelization strategy for GPU-accelerated clusters,
+and show scalabality
+of a 3-D compact finite difference application
+for up to 64 GPUs on the Palmetto cluster.

### 2015-11-17 thesis/acknowledgments.tex
updates

@@
 
-I acknowledge everyone.
+I owe my deepest thanks
+to my advisor, Dr. Richard S. Miller,
+for hist immense guidance and support;
+and the members of my committee,
+Dr. Gang Li and Dr. Lonny Thompson
+for their valuable time.
+I would like to acknowledge
+Dr. Daniel Livescu at Los Alamos National Lab
+for introducing us to this interesting problem,
+for his helpful comments,
+and for generous funding.
+I also thank
+Dr. Melissa Smith and Karan Sapra
+for their helpful guidance,
+and for organizing an excellent course on GPU computing.
+I owe a great deal of gratitude to
+Dr. Christopher Cox at Clemson University
+for his guidance on a range of topics
+and for many interesting conversations.
+I would like to thank the
+staff of the Cyberinfrastructure Technology Integration,
+and especially
+Dr. Galen Collier and Dr. Marcin Ziolkowski
+for their valuable comments and support.
+I owe thanks to Clemson University for the generous
+allocation of compute time on the Palmetto cluster.
+I acknowledge
+Dr. K. N. Seetharamu and Dr. V. Krishna,
+my advisors at PES Institute of Technology
+for inspiring in me a passion for
+scientific research.
+I owe a special thanks to the
+Software Carpentry Foundation
+and its members,
+for helping me be a
+a more productive programmer,
+a more deliberate learner,
+a more effective teacher,
+and a better person.
+
+Finally,
+I would like to thank my family and friends
+for everything they have done for me.

### 2015-11-17 thesis/compact-finite-difference.tex
updates

@@
 of fluid flows.
-
+%
 \begin{figure}
@@
 $dx$, $dy$ and $dz$ in the
-respective coordinate directions,
-\ref{fig:computational-domain},
+respective coordinate directions
+(Fig. \ref{fig:computational-domain}).
 Structured grids such as these
@@
 However, each GPU has a very limited amount of global memory:
-current NVIDIA Tesla accelerators have about 4 Gigabytes.
+current NVIDIA Tesla K20 GPUs have about 4 Gigabytes.
 Thus, to accomodate the
@@
 making the overall calculation highly parallelizable.
-The solution of the tridiagonal systems,
-is of course parallelizable
+The solution of the tridiagonal systems
+is of course parallelizable,
 as has been discussed in previous sections.
@@
 The most important considerations were brought out
-in the paper by Micikevicius et al. ~\ref{micikevicius20093d}.
+in the paper by Micikevicius et al. ~\cite{micikevicius20093d}.
 For solving the tridiagonal systems,

### 2015-11-17 thesis/conclusions.tex
updates

\chapter{Conclusions and Future Work}

We have presented an efficient approach for evaluating
compact finite differences on GPU-accelerated clusters.

The integration of this approach to our
current DNS codes is yet to be performed.
To ameliorate the cost of host-device transfers,
it is likely that other portions of the code
will also need to be ported to GPUs.
However,
most of the computational patterns
followed in the rest of the code
have been covered in the current work,
and fewer challenges are likely to be faced
in this process.

At the core of our approach is a
fast tridiagonal solver for the resulting linear systems.
Here, we make use of the simple matrix structure
to obtain better performance.
The applicability of this strategy to the
parallel cyclic reduction (PCR) algorithm,
and to other hybrid approaches is yet to be studied.

Strategies for the evaluation of compact finite differences
on other parallel architectures is of great interest,
especially for the Intel Many Integrated Core (MIC)
architectures.

### 2015-11-17 thesis/introduction-and-background.tex
updates

@@
 
-As mentioned, DNS involves the use of extremely fine grids.
-At each of the grid points,
+When solving for flows numerically,
+at each of the grid points in the computational domain,
 various data about the flow must be stored---for
@@
 pressure, temperature and velocities at that point.
-The amount of memory required to store this data
-for even small problem sizes
+For even small problem sizes,
+the amount of memory required to store this data
 can quickly exceed the capacity of modern workstations/PCs.
@@
 In the latter case,
-each group of cores constitutesa \emph{shared memory} system,
+each group of cores constitutes a \emph{shared memory} system,
 and the overall system is referred to as a \emph{hybrid system}.
@@
 \end{enumerate}
-
+%
 The approximation of higher order derivatives generally requires
@@
 with larger stencils associated with higher accuracy~\cite{fornberg1988generation}.
+In the evaluation of spatial derivatives,
+the above schemes are referred to as \emph{explicit}
+schemes, as the derivative at each point
+can be expressed as some explicit combination of function values.
+
+\subsection{General form of compact finite difference schemes}
+\label{subsec:gen-form-compact}
+
+Compact schemes express the derivative at a point $i$
+in terms of
+function values \emph{and derivatives}
+at neighbouring points,
+i.e., the derivative is expressed \emph{implicitly}.
+For example,
+if $f_i$ represents the value of
+a uniformly sampled function evaluated at the $i$th sample point,
+the first derivative $f^{\prime}_i$ can be approximated from
+a relation of the form:
+%
+\begin{equation}
+\begin{split}
+    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
+        f^{\prime}_i
+    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
+    \hdots
+    = 
+    a\frac{f_{i+1} - f_{i-1}}{dx} + \
+    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
+    c\frac{f_{i+3} - f_{i-3}}{dx} + \
+    \hdots
+\end{split}
+\label{eqn:general-compact}
+\end{equation}
+%
+where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
+are parameters that must satisfy certain constraints ~\cite{lele1992compact}.
+We note that for equations of the form in Eq. \ref{eqn:general-compact},
+the derivative at any points $i$ cannot be computed explicitly.
+Instead, we must write similar equations for \emph{all} points $i$
+in the range.
+This results in a system of linear equations
+with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
+which may be represented by the matrix system $A\bm{x}=\bm{d}$,
+where $\bm{x}$ is the vector
+\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
+$\bm{d}$ is a vector of right hand sides [Eq.\ref{eqn:general-compact}],
+and $A$ is, in general, a \emph{banded} matrix.
 
@@
 exhibit spatial variation associated with high wavenumbers.
-
-\subsection{General form of compact finite difference schemes}
-\label{subsec:gen-form-compact}
-
-Compact schemes express the derivative at a point $i$
-in terms of
-function values \emph{and derivatives}
-at neighbouring points,
-For example,
-if $f_i$ represents the value of
-a uniformly sampled function evaluated at the $i$th sample point,
-the first derivative $f^{\prime}_i$ can be approximated from
-a relation of the form:
-%
-\begin{equation}
-\begin{split}
-    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
-        f^{\prime}_i
-    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
-    \hdots
-    = 
-    a\frac{f_{i+1} - f_{i-1}}{dx} + \
-    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
-    c\frac{f_{i+3} - f_{i-3}}{dx} + \
-    \hdots
-\end{split}
-\label{eqn:general-compact}
-\end{equation}
-%
-where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
-are parameters that must satisfy certain constraints ~\cite{lele1992compact}.
-We note that for equations of the form in Eq. \ref{eqn:general-compact},
-the derivative at any points $i$ cannot be computed explicitly.
-Instead, we must write similar equations for \emph{all} points $i$
-in the range.
-This results in a system of linear equations
-with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
-which may be represented by the matrix system $A\bm{x}=\bm{d}$,
-where $\bm{x}$ is the vector
-\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
-$\bm{d}$ is a vector of right hand sides [Eq.\ref{eqn:general-compact}],
-and $A$ is, in general, a \emph{banded} matrix.
 
@@
 
-The parallel cyclic reduction (PCR) algorithm ~\cite{Zhang2010FTS}
+The parallel cyclic reduction (PCR) algorithm
 has only the forward reduction phase.
@@
 Further, unlike CR,
-PCR can be implemented free of bank conflicts.
+PCR can be implemented free of bank conflicts
+~\cite{Zhang2010FTS}.
 

### 2015-11-17 thesis/proposed-tridiagonal-algorithm.tex
updates

@@
 \chapter{Proposed tridiagonal algorithm}
+
+\section{Modified cyclic reduction for near-Toeplitz systems}
 
@@
 and numerical solutions to one-dimensional differential equations.
-In this chapter,
-we will describe an efficient solver for solving such
+We describe an efficient solver for solving such
 near-Toeplitz tridiagonal systems efficiently on GPUs.
-
-\section{Modified cyclic reduction for near-Toeplitz systems}
-
 \subsection{Forward reduction}
@@
 
+In this section, we describe the implementation of a GPU solver
+for solving a given near-Toeplitz tridiagonal system
+for multiple right hand sides.
+Our implementation uses
+NVIDIA CUDA platform for programming the GPU,
+but is easily translated to OpenCL.
+The Python programming language is used
+to interface with CUDA, by use of the
+PyCUDA \cite{kloeckner_pycuda_2012} library.
+We develop two approaches based on the common idea
+of precomputed forward reduction coefficients---one that leverages
+the GPU's \emph{shared memory},
+and the other working directly with global memory.
+The GPU kernels for both implementations are relatively
+straightforward and compact,
+spanning no more than 100 lines of code.
+
 \subsection{Precomputing forward reduction coefficients}
@@
 during the end of forward reduction and the end of backward substitution.
-
-\begin{listing}[H]
-\begin{minted}[fontsize=\footnotesize]{python}
-stride = 1
-for i in range(n-1):
-    stride *= 2
-    forwardReductionKernel(num_blocks=nrhs, block_size=n/stride)
-
-solveTwoByTwo()
-
-for i in range(n-1):
-    stride /= 2
-    backSubstitutionKernel(num_blocks=nrhs, block=n/stride)
-\end{minted}
-\end{listing}
-
 The precomputed coefficient arrays and right hand

### 2015-11-17 thesis/results.tex
updates

@@
 
-\section{Validation}
-    \subsection{Validation of tridiagonal solver}
-    \subsection{Validation of compact finite difference application}
-        \subsubsection{Tests for individual kernels}
-        \subsubsection{Tests for communication routines}
-        \subsubsection{Test functions for derivatives}
-        \subsubsection{Order of convergence tests}
-    \subsubsection{Validation results}
+\section{Performance of GPU tridiagonal solver}
 
-\section{Performance of GPU tridiagonal solver}
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/bench-2d.eps}
+\caption{Relative solver performance for 2-D problems. Relative time defined as:}
+$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
+\label{fig:bench-2d}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/bench-3d.eps}
+\caption{Relative solver performance for 3-D problems. Relative time defined as:}
+$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
+\label{fig:bench-3d}
+\end{center}
+\end{figure}
+
+In this section we present a performance overview
+of our Near-Toeplitz tridiagonal solver (NEATO)
+against a multi-threaded Intel MKL solver and
+the CUSPARSE GPU solver
+(\texttt{dgtsv} and \texttt{dgtsvStridedBatch} respectively).
+The MKL solver is based on Gaussian Elimination with partial pivoting,
+Boundary conditions may prevent
+the matrix from being symmetric and/or diagonally dominant,
+precluding the use of more specialized tridiagonal solvers.
+and the CUSPARSE solver uses a combination of
+Cyclic Reduction and Parallel Cyclic Reduction.
+Both solvers are optimized for the underlying architectures.
+
+The CPU code is compiled with the Intel C compiler (version 15.0),
+and run (with OpenMP support) on up to
+16 independent cores of the the same shared-memory node
+(one thread per core).
+The CPU is an
+Intel Xeon Processor E5-2670 v2 (2.50 GHz, 25 MB Smart Cache).
+GPU code is compiled with the CUDA toolkit (version 6.5.14),
+and run on the
+NVIDIA Tesla K20 Accelerator.
+When measuring GPU performance (both CUSPARSE and NEATO),
+we do \emph{not} include the cost
+of data transfer between the CPU and GPU.
+\texttt{-O2} level compiler optimizations are turned on for both
+CPU and GPU code;
+no further optimization options are enabled in either case.
+Of course, we use double precision for all solvers.
+The timings reported are
+kernel \emph{execution} times, i.e.,
+the time for all kernel(s) to execute completely
+before returning the program control to the CPU.
+All timings are averaged over 100 tridiagonal solves.
+
+\subsection{NEATO: global memory v/s shared memory performance}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/global-vs-shared-2d.eps}
+\caption{Comparison of global memory and shared
+    memory implementations of NEATO (2D problems).}
+\label{fig:global-vs-shared-2d}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/global-vs-shared-3d.eps}
+\caption{Comparison of global memory and shared memory
+    implementations of NEATO (3D problems).}
+\label{fig:global-vs-shared-3d}
+\end{center}
+\end{figure}
+
+In Figs. \ref{fig:global-vs-shared-2d} and \ref{fig:global-vs-shared-3d},
+we report the performance of the two solvers for the case
+$N_{rhs} = n$ and $N_{rhs} = n^2$.
+These cases correspond to tridiagonal systems
+arising in 2-D and 3-D problems respectively.
+We note that the shared memory implementation
+offers better performance in nearly all cases.
+However, the relative speedup from using
+shared memory diminishes with increasing problem size.
+For larger problem sizes,
+the synchronization costs associated with inactive threads
+leads to poor shared memory performance.
+
+\subsection{Comparison of NEATO with Intel MKL and CUSPARSE solvers}
+
+In Fig. \ref{fig:bench-2d} and \ref{fig:bench-3d},
+we provide the relative performance of
+Intel MKL and CUSPARSE solvers and compare against
+the NEATO shared memory implementation.
+The relative performance for
+each problem size is obtained by
+normalizing the solver timings
+by the timing for the fastest solver for that problem size.
+Table \ref{table:bench} shows the timings of the various
+solvers to solve different problem sizes.
+
+\begin{sidewaystable}[]
+\scriptsize
+\centering
+\caption{Performance of Intel MKL, CUSPARSE and NEATO solvers.}
+\label{table:bench}
+\input{table/bench.tex}
+\end{sidewaystable}
+
 \section{Performance of compact finite difference application}
+
+The experiments were run on the
+Clemson University Palmetto Cluster,
+using NVIDIA Tesla K20 and K40 GPUs.
+The K40 GPUs were used for the largest problem sizes.
+Each compute node on the cluster is equipped with up to 2 GPUs,
+and nodes are connected by 56 Gbps Infiniband interconnect.
+We use Open MPI 1.8.1 configured with OpenFabrics support.
+
+The timings reported are \emph{wall clock} times
+with global synchronization between processes performed 
+before and after evaluation of the derivatives.
+All timings are averaged over 100 evaluations of the function derivatives
+in each coordinate direction.
+We make it clear that our reported
+problem sizes represent the \emph{actual size of problem data}.
+Although it may be considered sufficient to
+run tests on a single line of processes
+for measuring the compact finite difference solver performance,
+we set up and solve the problem for the entire computational domain.
+In the context of a larger simulation,
+global synchronization between the processes
+is typically performed before and after the evalaution of derivatives,
+and it is important to consider the related overhead.
+
+Figures \ref{fig:compact-profiling-1024-64}-\ref{fig:compact-profiling-2048-64}
+show the time taken by the different steps of
+the compact finite difference solver.
+We note that for the larger problem size,
+the evaluation of the primary systems (using the NEATO solver)
+constitutes a larger majority of the total runtime,
+which justifies our efforts in optimizing the tridiagonal solver.
+For evaluation of the derivatives in the y- and z- directions,
+we note that a significant portion of the runtime is dedicated
+to performing permutations of the input and output data.
+We attribute this to our naive implementation of the
+permutation kernels (no shared memory).
+Figures \ref{fig:strong-scaling-256}-\ref{fig:weak-scaling-256}
+show the strong and weak scaling of the compact finite difference solver
+for evaluating derivatives in all three coordinate directions.
+We note good scaling for both 2-D and 3-D cases.
+The strong scaling for 3-D problems is somewhat better
+as the GPU is kept busy for all problem sizes.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/profiling-1024-64.eps}
+\caption{Solving problem sized $1024^3$ on 64 GPUs}
+\label{fig:compact-profiling-1024-64}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/profiling-2048-64.eps}
+\caption{Solving problem sized $2048^3$ on 64 GPUs}
+\label{fig:compact-profiling-2048-64}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/strong-scaling-256.eps}
+\caption{Strong scaling for multi-GPU compact finite difference, problem size: $256^3$.}
+\label{fig:strong-scaling-256}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/strong-scaling-512.eps}
+\caption{Strong scaling for multi-GPU compact finite difference, problem size: $512^3$.}
+\label{fig:strong-scaling-512}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/weak-scaling-128.eps}
+\caption{Weak scaling for multi-GPU compact finite difference, problem size: $128^3$ per process.}
+\label{fig:weak-scaling-128}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/weak-scaling-256.eps}
+\caption{Weak scaling for multi-GPU compact finite difference, problem size: $256^3$ per process.}
+\label{fig:weak-scaling-256}
+\end{center}
+\end{figure}

### 2015-11-18 thesis/compact-finite-difference.tex
fix typos

@@
 current NVIDIA Tesla K20 GPUs have about 4 Gigabytes.
-Thus, to accomodate the
+Thus, to accommodate the
 larger problem sizes in DNS,
@@
 \caption{Compact finite differences - distributed GPUs and
-    restriced in one direction}
+    restricted in one direction}
 \label{fig:compact-distributed-restricted}
@@
 By restricting the distribution along one coordinate direction,
-the ease of solution of the tridaigonal systems is maintained.
+the ease of solution of the tridiagonal systems is maintained.
 
@@
 which have the disadvantages discussed earlier.
-To accomodate compact finite difference schemes in all the coordinate
+To accommodate compact finite difference schemes in all the coordinate
 directions,
@@
 is applicable to a single tridiagonal system distributed
-accross several processors.
+across several processors.
 When solving a distributed tridiagonal system
@@
 in the direction along which successive
-function values in a subdomain are stored contigiously in memory,
+function values in a subdomain are stored contiguously in memory,
 i.e., the ``fastest'' coordinate direction.

### 2015-11-18 thesis/introduction-and-background.tex
fix typos

@@
 is likely not take full advantage of the GPU's
-memory heirarchy and compute ability.
+memory hierarchy and compute ability.
 
@@
 The ability of a finite difference scheme
-to accomodate large wavenumbers
+to accommodate large wavenumbers
 is extremely relevant in
@@
 The CUDA application programming interface (API)
-is availabe as an extension to
+is available as an extension to
 a programming languages (C, C++, Fortran, etc.),
@@
 of algorithms on GPUs is the
-available \emph{memory heirarchy}.
+available \emph{memory hierarchy}.
 Threads executing a kernel can read and write
@@
 words from the same bank,
-a \emph{bank conflict} is said to occurr,
+a \emph{bank conflict} is said to occur,
 and the access is serialized.
@@
 Because the floats are placed in consecutive banks,
-the 1st and 17th threads will acess floats from the same bank,
+the 1st and 17th threads will access \texttt{float}s from the same bank,
 as will the 2nd and 18th threads, and so on.
@@
 request memory from the same bank:
-a emph{32-way} bank conflict.
+a \emph{32-way} bank conflict.
 
@@
 The amount of shared memory per SM
-and the numer of registers per SM is limited.
+and the number of registers per SM is limited.
 For the current NVIDIA Tesla K20 accelerator,
@@
 
-    \item The limited amount of shared memory places restricions
+    \item The limited amount of shared memory places restrictions
         on the size of tridiagonal systems that can be solved,

### 2015-11-18 thesis/proposed-tridiagonal-algorithm.tex
fix typos

@@
 With all the forward reduction coefficient matrices
-and auxillary variables precomputed and stored,
+and auxiliary variables precomputed and stored,
 the $m$th forward reduction step for equation $i$

### 2015-11-18 thesis/results.tex
fix typos

@@
 global synchronization between the processes
-is typically performed before and after the evalaution of derivatives,
+is typically performed before and after the evaluation of derivatives,
 and it is important to consider the related overhead.

### 2015-11-18 thesis/abstract.tex
corrections - round 1

--- 2015-11-17 thesis/abstract.tex
@@
 
+\emph{Compact} finite difference schemes are widely used
+in the direct numerical simulation of fluid flows
+for their ability to better resolve the small scales of turbulence.
+However, they can be expensive to evaluate and
+difficult to parallelize.
+In this work, we present an approach for the computation
+of compact finite differences
+and similar tridiagonal schemes
+on graphics processing units (GPUs).
+We present a variant of the \emph{cyclic reduction} 
+algorithm for solving the
+tridiagonal linear systems that arise in such numerical schemes.
 We study the impact of the matrix structure on
 the cyclic reduction algorithm
-and show that that precomputing
+and show that precomputing
 forward reduction coefficients can be
-an especially effective strategy.
-his is the abstract
+especially effective for obtaining good performance.
+Our tridiagonal solver is able to outperform
+the NVIDIA CUSPARSE
+and the multithreaded Intel MKL tridiagonal solvers
+on GPU and CPU respectively.
+In addition,
+we present a parallelization strategy for GPU-accelerated clusters,
+and show scalabality
+of a 3-D compact finite difference application
+for up to 64 GPUs on Clemson's Palmetto cluster.

### 2015-11-18 thesis/acknowledgments.tex
corrections - round 1

@@
 scientific research.
+I am indebted to Dr. Anush Krishnan
+for introducing me to the fascinating field of
+scientific computing.
 I owe a special thanks to the
@@
 and its members,
-for helping me be a
+for helping me be
 a more productive programmer,

### 2015-11-18 thesis/introduction-and-background.tex
corrections - round 1

@@
 flow equations on this grid.
-By far the most popular numerical approach for
+Among the most popular numerical approaches for
 solving the flow equations
@@
 adoption in DNS codes
-for their high order of accuracy.
+for their high order of accuracy and small stencil widths.
 The evaluation of compact finite differences
@@
 approach for evaluating
-compact finite differences
-on multiple GPUs in a distributed system.
+compact finite differences---and
+other numerical schemes leading to tridiagonal systems---on
+multiple GPUs in a distributed system.
 This is of interest because
@@
 \begin{split}
+    f_i^{\prime} + \
     \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
-        f^{\prime}_i
     \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
@@
 where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
-are parameters that must satisfy certain constraints ~\cite{lele1992compact}.
-We note that for equations of the form in Eq. \ref{eqn:general-compact},
+are parameters that must satisfy certain constraints ~\cite{kennedy1994several,lele1992compact},
+and $dx$ is the space between two grid points.
+We note that for equations of the form in Eq. (\ref{eqn:general-compact}),
 the derivative at any points $i$ cannot be computed explicitly.
@@
 \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
-$\bm{d}$ is a vector of right hand sides [Eq.\ref{eqn:general-compact}],
+$\bm{d}$ is a vector of right hand sides [Eq. (\ref{eqn:general-compact})],
 and $A$ is, in general, a \emph{banded} matrix.
@@
 \end{center}
-\caption{Modified wavenumbers for three different schemes
-    with the same stencil width.
+\caption{Modified wavenumbers for different finite difference schemes.
+    $k^\prime$ is the wavenumber of the approximated derivative,
+    while $k$ is wavenumber of the exact derivative.
+    $h$ represents the spacing between two grid points.
     The compact schemes better estimate the derivative
@@
 
-From \ref{fig:modified-wavenumbers},
+From Fig. \ref{fig:modified-wavenumbers},
 it is clear that the compact schemes are better
@@
 This is in contrast to other methods used in DNS,
-such as the \emph{spectral} methods.
+such as \emph{spectral} methods.
 We note that the Eq. (\ref{eqn:general-compact})
@@
 \begin{equation}
-    f_1^{\prime} + \alpha f_2^{\prime} = \
-        \frac{1}{h}(af_1 + bf_2 + cf_3 + df_4)
+    f_1^{\prime} + \alpha^{\prime}f_2^{\prime} = \
+        \frac{1}{dx}(a^{\prime}f_1 + b^{\prime}f_2 + c^{\prime}f_3 + d^{\prime}f_4)
 \label{eqn:boundary-compact}
@@
 When $\alpha \neq 0, \beta = \hdots$ = 0,
-$A$ is \emph{tridiagonal}.
+$A$ is tridiagonal.
 The evaluation of the derivatives requires
@@
 When $\alpha \neq 0$ and $\beta \neq  0$,
-pentadiagonal and higher-bandwidth systems arise.
+pentadiagonal systems arise.
 These can be more expensive and difficult to evaluate
@@
 $\beta = 0$, $\alpha = \frac{1}{4}$ and $a = \frac{3}{4}$
-in Eq. \ref{eqn:general-compact}
+in Eq. (\ref{eqn:general-compact})
 (all other coefficients are set to zero).
@@
  \begin{bmatrix}
-     \frac{-5f_1 + 4f_2 + f_3}{2}\\
-     \frac{3(f_{3} - f_{1})}{4}\\
-     \frac{3(f_{4} - f_{2})}{4}\\
+     \frac{-5f_1 + 4f_2 + f_3}{2dx}\\
+     \frac{3(f_{3} - f_{1})}{4dx}\\
+     \frac{3(f_{4} - f_{2})}{4dx}\\
      \vdots\\
@@
      \vdots\\
-     \frac{3(f_{n} - f_{n-2})}{4}\\
-     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2}
+     \frac{3(f_{n} - f_{n-2})}{4dx}\\
+     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2dx}
   \end{bmatrix}
@@
 more easily programmed for the GPU.
-Two essential features of the CUDA programming modl
+Two essential features of the CUDA programming model
 are \emph{kernels} and \emph{thread organization}.
@@
 This feature gives CUDA programs
-their scalability \ref{fig:gpu-scaling}.
+their scalability (Fig. \ref{fig:gpu-scaling}).
 The same program runs faster on a GPU with a larger number
@@
 shared memory is often viewed as explicitly managed cache.
-Data in shared memory does\emph{not} persist after
+Data in shared memory does \emph{not} persist after
 kernel execution.
@@
         and from the device to the host \emph{once}.
-    \item Thread access to global memory must \emph{coalesced},
+    \item Thread access to global memory must be \emph{coalesced},
         i.e., successive threads must read successive locations
@@
 \end{figure}
-
+%
 In the GPU implementation of cyclic reduction,
@@
 During the forward reduction phase,
-
-\begin{figure}
-\begin{center}
-\includegraphics[height=200pt]{img/forward-reduction-step.eps}
-\end{center}
-\caption{Updating $\bm{b}$ in the first forward reduction step.}
-\label{fig:forward-reduction-step}
-\end{figure}
-
 the threads assigned to each even index $i$
@@
 
-Thus, the following issues are encountered
-while implementing cyclic reduction for GPUs:
+\begin{figure}
+\begin{center}
+\includegraphics[height=200pt]{img/forward-reduction-step.eps}
+\end{center}
+\caption{Updating $\bm{b}$ in the first forward reduction step.}
+\label{fig:forward-reduction-step}
+\end{figure}
+
+Several issues are encountered in the GPU implementation:
 
@@
 Much work has been done on
-optimizing cyclic reduction performance on GPUs,
-and addressing the issues discussed.
+addressing these problems and
+optimizing cyclic reduction performance on GPUs.
 Zhang et al. \cite{Zhang2010FTS} propose a
@@
 a means to reduce shared memory usage in cyclic reduction.
-Esfahanian et. al, \cite{esfahanian2014efficient}
+Esfahanian et al. \cite{esfahanian2014efficient}
 avoid shared memory (and associated bank conflicts entirely)
@@
 Our approach takes a different route,
-and is focussed on \emph{near-Toeplitz} systems---exploiting
-the relatively simple matrix structure
+and is focussed on exploiting the specific matrix structure
 to reduce the number of computations and memory accesses

### 2015-11-18 thesis/proposed-tridiagonal-algorithm.tex
corrections - round 1

@@
 and numerical solutions to one-dimensional differential equations.
-We describe an efficient solver for solving such
+We describe an approach for solving such
 near-Toeplitz tridiagonal systems efficiently on GPUs.
+
+Below, we present the effect of the matrix structure
+[Eq. (\ref{eqn:toeplitz-matrix})] on the
+forward reduction and backward substitution phases of
+the cyclic reduction algorithm.
+
 \subsection{Forward reduction}
@@
 \begin{center}
-\includegraphics[height=200pt]{img/cyclic-reduction-precomputing.eps}
+\includegraphics[height=250pt]{img/cyclic-reduction-precomputing.eps}
 \end{center}

### 2015-11-18 thesis/compact-finite-difference.tex
fixing more typos

@@
 
-For a single GPU \ref{fig:compact-single-gpu},
+For a single GPU (Fig. \ref{fig:compact-single-gpu}),
 when calculating the right hand sides,
@@
 The most important considerations were brought out
-in the paper by Micikevicius et al. ~\cite{micikevicius20093d}.
+in the paper by Micikevicius et al. \cite{micikevicius20093d}.
 For solving the tridiagonal systems,

### 2015-11-18 thesis/introduction-and-background.tex
fixing more typos

@@
     = 
-    a\frac{f_{i+1} - f_{i-1}}{dx} + \
-    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
-    c\frac{f_{i+3} - f_{i-3}}{dx} + \
+    a\frac{f_{i+1} - f_{i-1}}{h} + \
+    b\frac{f_{i+2} - f_{i-2}}{h} + \\
+    c\frac{f_{i+3} - f_{i-3}}{h} + \
     \hdots
@@
 are parameters that must satisfy certain constraints ~\cite{kennedy1994several,lele1992compact},
-and $dx$ is the space between two grid points.
+and $h$ is the space between two grid points.
 We note that for equations of the form in Eq. (\ref{eqn:general-compact}),
@@
     f_1^{\prime} + \alpha^{\prime}f_2^{\prime} = \
-        \frac{1}{dx}(a^{\prime}f_1 + b^{\prime}f_2 + c^{\prime}f_3 + d^{\prime}f_4)
+        \frac{1}{h}(a^{\prime}f_1 + b^{\prime}f_2 + c^{\prime}f_3 + d^{\prime}f_4)
 \label{eqn:boundary-compact}

### 2015-11-18 thesis/proposed-tridiagonal-algorithm.tex
fixing more typos

@@
 We describe an approach for solving such
-near-Toeplitz tridiagonal systems efficiently on GPUs.
+\emph{near-Toeplitz} tridiagonal systems efficiently on GPUs.
 
@@
 We note that for a general tridiagonal system,
-this requires storage for $3.(2N - 2)$ coefficients,
-in addition to the $3N$ coefficients
+this requires storage for $3.(2n - 2)$ coefficients,
+in addition to the $3n$ coefficients
 for the original tridiagonal system.
@@
 i.e., when the coefficient matrix is of the form
-$A$ in Eq. \ref{eqn:toeplitz-matrix}.
+$A$ in Eq. (\ref{eqn:toeplitz-matrix}).
 We examine the effect of the
@@
 correspond to the coefficients of a tridiagonal matrix with
-exactly the near-Toeplitz structure of $A$.
+exactly the \emph{near-Toeplitz} structure of $A$.
 This form-preserving property of cyclic reduction
@@
 The fact that the reduced system at each step
-is near-Toeplitz can be exploited
+is \emph{near-Toeplitz} can be exploited
 to reduce the cost of
 precomputing and storing the coefficients.
-Each near-Toeplitz matrix is completely defined
+Each \emph{near-Toeplitz} matrix is completely defined
 by only a handful of coefficients:
@@
 In this section, we describe the implementation of a GPU solver
-for solving a given near-Toeplitz tridiagonal system
+for solving a given \emph{near-Toeplitz} tridiagonal system
 for multiple right hand sides.
-Our implementation uses
+Our implementation uses the
 NVIDIA CUDA platform for programming the GPU,
@@
 Similarly, the values $c_1^m$ are equal to $c_0^m$,
-and do not require special storage.
+and do not require separate storage.
 The set of precomputed coefficients required to be stored is then
@@
 than in cyclic reduction implementations for general tridiagonal systems,
-due the reduced number of computations performed.
+due to the reduced number of computations performed.

### 2015-11-18 thesis/compact-finite-difference.tex
remove space after cite

@@
 This is the method presented by
-Sakharnykh et al. ~\cite{sakharnykhADIconf},
+Sakharnykh et al. \cite{sakharnykhADIconf},
 and it has the advantage that

### 2015-11-18 thesis/introduction-and-background.tex
remove space after cite

@@
 may be derived,
-with larger stencils associated with higher accuracy~\cite{fornberg1988generation}.
+with larger stencils associated with higher accuracy \cite{fornberg1988generation}.
 In the evaluation of spatial derivatives,
@@
 where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
-are parameters that must satisfy certain constraints ~\cite{kennedy1994several,lele1992compact},
+are parameters that must satisfy certain constraints \cite{kennedy1994several,lele1992compact},
 and $h$ is the space between two grid points.
@@
 computational fluid dynamics applications
-~\cite{KravchenkoEffNumErr}.
+\cite{KravchenkoEffNumErr}.
 In DNS, the scheme
@@
 as higher order boundary stencils are unstable
-~\cite{kennedy1994several}.
+\cite{kennedy1994several}.
 In general, boundary equations for the first derivative
-are of the following form ~\cite{lele1992compact}:
+are of the following form \cite{lele1992compact}:
 
@@
 second, third and fourth derivatives
-are derived in ~\cite{lele1992compact}.
+are derived in \cite{lele1992compact}.
 
@@
 and of CUDA specifically is available
-in ~\cite{HwuBook}.
+in \cite{HwuBook}.
 We provide the essential and relevant details about CUDA here,
@@
     with more SMs are able to execute more blocks concurrently.
-(\emph{CUDA Programming Guide ~\cite{CUDAProgrammingGuide}})}
+(\emph{CUDA Programming Guide \cite{CUDAProgrammingGuide}})}
 \label{fig:gpu-scaling}
@@
 for solving this problem on the GPU.
-We refer to ~\cite{chang2014guide} for
+We refer to \cite{chang2014guide} for
 a more comprehensive list of algorithms,
@@
 Implementations of this algorithm are relatively straightforward
-and compact ~\cite{numericalrecipes}.
+and compact \cite{numericalrecipes}.
 Only the non-zero diagonals of the tridiagonal matrix
@@
 PCR can be implemented free of bank conflicts
-~\cite{Zhang2010FTS}.
+\cite{Zhang2010FTS}.
 

### 2015-11-18 thesis/proposed-tridiagonal-algorithm.tex
remove space after cite

@@
 These matrices appear in a wide range of applications
-~\cite{sun1995application}
+\cite{sun1995application}
 such as alternating direction implicit methods,

### 2015-11-18 thesis/compact-finite-difference.tex
Corrections

@@
 the same host memory space.
-The domain is divided into a number of
-into ``subdomains'',
+The domain is divided into a number of `subdomains'',
 as shown in \ref{fig:compact-shared-gpu}.
@@
 For larger problems,
-a distributed system may be required.
+a distributed system is nearly always required.
 Here, the simplest strategy is to use a domain decomposition
-as shown in \ref{fig:compact-distributed-restricted}.
+as shown in Fig. \ref{fig:compact-distributed-restricted}.
 Here again, 
@@
 is required at each of the subdomain boundaries.
-
 The solution of the distributed tridiagonal systems
@@
 0
-\end{bmatrix} & \label{eqn:secondary-system-1-expanded} \\
+\end{bmatrix} & \label{eqn:secondary-system-1-expanded} 
 \end{align}
@@
 We refer to the subsystem in Eq. (\ref{eqn:primary-system-expanded})
-the ``primary'' system, and the subsystems in
+as the ``primary'' system,
+and the subsystems in
 Eqs. (\ref{eqn:secondary-system-1-expanded}) and
 (\ref{eqn:secondary-system-2-expanded})
-the ``secondary'' systems.
-Additionally, the following ``reduced'' system is solved:
+as the ``secondary'' systems.
+Additionally, the following ``reduced'' system
+must be constructed andsolved:
 
@@
 
-Our algorithm is outlined in \ref{fig:compact-algorithm},
+Our algorithm is outlined in Fig. \ref{fig:compact-algorithm},
 and is described for the case of evaluating derivatives
@@
 \caption{Purpose of kernels and MPI calls in compact finite difference application}
+
 \label{table:compact-algorithm}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \begin{center}

### 2015-11-19 thesis/introduction-and-background.tex
updates

@@
 \end{itemize}
+
+\subsection{Software for programming NVIDIA GPUs}
+
+As described,
+the CUDA programming interface
+allows developing general-purpose GPU applications---however,
+there are other options.
+Notably,
+the OpenACC \cite{openacc} toolkit allows
+portions of code to be offloaded to GPU accelerators
+simply by use of compiler directives.
+This makes porting existing applications
+extremely easy,
+and introduces almost no extra code.
+Some libraries for scientific computation,
+such as PETSc \cite{petsc-web-page},
+also allow the \emph{transparent} use of GPUs
+through their interfaces.
+We use the CUDA interface because
+it fully exposes the
+features of the underlying architecture
+that can be exploited.
+This is especially important due to the
+complex computational and memory access patterns involved
+in our algorithms.
 

### 2015-11-19 thesis/results.tex
updates

@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=300pt]{fig/bench-2d.eps}
-\caption{Relative solver performance for 2-D problems. Relative time defined as:}
-$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
-\label{fig:bench-2d}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=300pt]{fig/bench-3d.eps}
-\caption{Relative solver performance for 3-D problems. Relative time defined as:}
-$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
-\label{fig:bench-3d}
-\end{center}
-\end{figure}
-
 In this section we present a performance overview
-of our Near-Toeplitz tridiagonal solver (NEATO)
+of our NEATO solver
 against a multi-threaded Intel MKL solver and
@@
 (\texttt{dgtsv} and \texttt{dgtsvStridedBatch} respectively).
-The MKL solver is based on Gaussian Elimination with partial pivoting,
+The MKL solver uses Gaussian elimination with partial pivoting,
+and the CUSPARSE solver uses a combination of
+Cyclic Reduction and Parallel Cyclic Reduction
+as described by Zhang et al. \cite{Zhang2010FTS}.
+These solvers represent the most straightforward way
+to compute solutions for tridiagonal systems
+and are highly optimized for performance on
+underlying architectures.
 Boundary conditions may prevent
@@
 precluding the use of more specialized tridiagonal solvers.
-and the CUSPARSE solver uses a combination of
-Cyclic Reduction and Parallel Cyclic Reduction.
-Both solvers are optimized for the underlying architectures.
 
@@
 of data transfer between the CPU and GPU.
-\texttt{-O2} level compiler optimizations are turned on for both
+This is because the tridiagonal solver is expected to be
+part of a larger application.
+This is in keeping with the timing strategies
+in related literature.
+
+The \texttt{-O2} level compiler optimizations are turned on for both
 CPU and GPU code;
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/bench-2d.eps}
+\caption{Relative solver performance for 2-D problems. Relative time defined as:}
+Time taken by solver/Time taken by NEATO solver
+\label{fig:bench-2d}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/bench-3d.eps}
+\caption{Relative solver performance for 3-D problems. Relative time defined as:}
+Time taken by solver/Time taken by NEATO solver
+\label{fig:bench-3d}
+\end{center}
+\end{figure}
+
 In Fig. \ref{fig:bench-2d} and \ref{fig:bench-3d},
@@
 normalizing the solver timings
-by the timing for the fastest solver for that problem size.
+by the timing for the NEATO solver for that problem size.
 Table \ref{table:bench} shows the timings of the various
@@
 
-The experiments were run on the
+The timings for the
+compact finite difference application
+were measured on the
 Clemson University Palmetto Cluster,
@@
 
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=300pt]{fig/profiling-1024-64.eps}
+\caption{Solving problem sized $1024^3$ on 64 GPUs}
+\label{fig:compact-profiling-1024-64}
+\end{center}
+\end{figure}
+%
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=300pt]{fig/profiling-2048-64.eps}
+\caption{Solving problem sized $2048^3$ on 64 GPUs}
+\label{fig:compact-profiling-2048-64}
+\end{center}
+\end{figure}
+%
 Figures \ref{fig:compact-profiling-1024-64}-\ref{fig:compact-profiling-2048-64}
@@
 to performing permutations of the input and output data.
-We attribute this to our naive implementation of the
+We attribute this to our na\"{\i}ve implementation of the
 permutation kernels (no shared memory).
-Figures \ref{fig:strong-scaling-256}-\ref{fig:weak-scaling-256}
-show the strong and weak scaling of the compact finite difference solver
-for evaluating derivatives in all three coordinate directions.
-We note good scaling for both 2-D and 3-D cases.
-The strong scaling for 3-D problems is somewhat better
-as the GPU is kept busy for all problem sizes.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=300pt]{fig/profiling-1024-64.eps}
-\caption{Solving problem sized $1024^3$ on 64 GPUs}
-\label{fig:compact-profiling-1024-64}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=300pt]{fig/profiling-2048-64.eps}
-\caption{Solving problem sized $2048^3$ on 64 GPUs}
-\label{fig:compact-profiling-2048-64}
-\end{center}
-\end{figure}
-
+%
 \begin{figure}
@@
 \end{figure}
-
+%
 \begin{figure}
@@
 \end{figure}
+%
+Figures \ref{fig:strong-scaling-256}-\ref{fig:weak-scaling-256}
+show the strong and weak scaling of the compact finite difference solver
+for evaluating derivatives in all three coordinate directions.
+We note good scaling for both 2-D and 3-D cases.
+The strong scaling for 3-D problems is somewhat better
+as the GPU is kept busy for all problem sizes.
+%
+\input{table/compact-refimpl-timings.tex}
+%
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/compact-refimpl-speedups.eps}
+\caption{Speedups over reference implementation}
+\label{fig:compact-refimpl-speedups}
+\end{center}
+\end{figure}
+%
+We also compare the performance of our compact finite difference
+solver with the implementation by
+Mohd-Yusof et al. \cite{mohd2010adapting}.
+This implementation uses a
+distributed tridiagonal solver based on the LU algorithm
+to solve the linear systems.
+For comparing timings,
+we use the number of CPU sockets as the basis.
+Each CPU socket uses 8 CPU cores and 1 GPU.
+Thus we maintain a ratio of 1:8 between GPUs and CPU cores
+in our comparison.
+Table \ref{table:compact-refimpl-timings}
+shows timings for computing derivatives for problem
+sized up to $2048^3$,
+and Fig. \ref{fig:compact-refimpl-speedups}
+shows the respective speedup using our implementation.

### 2015-11-19 thesis/table/compact-refimpl-timings.tex
updates

\begin{table}[]
\centering
\caption{Time (ms) to compute derivatives in $x$ direction - comparison with reference implementation \cite{mohd2010adapting}}
\label{table:compact-refimpl-timings}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multirow{2}{*}{Size} & \multicolumn{3}{c|}{Ref. impl, \#CPU cores} & \multicolumn{3}{c|}{NEATO-based, \#GPUs} \\ \cline{2-7}
         & 8         & 64        & 512      & 1       & 8       & 64      \\ \hline
$256^3$  & 79.5      & 20.8      & 11.1     & 19.9    & 5.17    & 2.79    \\ \hline
$512^3$  & 556.8     & 146.5     & 29.2     & 164.5   & 23.24   & 5.62    \\ \hline
$1024^3$ & 5188      & 1092      & 223.7    & -       & 174.9   & 24.49   \\ \hline
$2048^3$ & -         & -         & 1741     & -       & -       & 297.07  \\ \hline
\end{tabular}
\end{table}

### 2015-11-19 paper-draft/scripts/compact-refimpl/compact-refimpl.py

@@ -1,0 +1,45 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+plt.style.use('bmh')
+
+sizes = [256, 512, 1024, 2048]
+
+cpu_8 = np.array([79.5, 556.8, 5188, np.nan])
+cpu_64 = np.array([20.8, 146.5, 1092, np.nan])
+cpu_512 = np.array([11.1, 29.2, 223.7, 1741.])
+
+gpu_1 = np.array([19.9, 164.5, np.nan, np.nan])
+gpu_8 = np.array([5.17, 23.24, 174.9, np.nan])
+gpu_64 = np.array([2.79, 5.62, 24.49, 297.07])
+
+speedup_1 = cpu_8/gpu_1
+speedup_8 = cpu_64/gpu_8
+speedup_64 = cpu_512/gpu_64
+
+ind = np.arange(4)*2
+fig, ax = plt.subplots()
+width = 0.5
+
+speedup_1_bars = ax.bar(ind+0*width, speedup_1, width, color='dodgerblue')
+speedup_8_bars = ax.bar(ind+1*width, speedup_8, width, color='darkred')
+speedup_64_bars = ax.bar(ind+2*width, speedup_64, width, color='mediumpurple')
+
+ax.set_xticks(ind+2*width)
+ax.set_xticklabels(['$256^3$', '$512^3$',
+    '$1024^3$', '$2048^3$'])
+ax.set_xlabel('Problem size')
+ax.set_ylabel('Speedup over reference implementation')
+ax.grid(True)
+
+leg = ax.legend((
+            speedup_1_bars,
+            speedup_8_bars,
+            speedup_64_bars),
+            ('1 socket',
+             '8 sockets',
+             '64 sockets'),
+            loc='best',
+            fontsize='medium')
+leg.draw_frame(True)
+plt.savefig('compact-refimpl-speedups.eps')

### 2015-11-24 paper-draft/scripts/compact-refimpl/compact-refimpl.py
generate a prettier graph

@@ -21,11 +21,11 @@ ind = np.arange(4)*2
 fig, ax = plt.subplots()
 width = 0.5
 
-speedup_1_bars = ax.bar(ind+0*width, speedup_1, width, color='dodgerblue')
-speedup_8_bars = ax.bar(ind+1*width, speedup_8, width, color='darkred')
-speedup_64_bars = ax.bar(ind+2*width, speedup_64, width, color='mediumpurple')
+speedup_1_bars = ax.bar(ind+0*width, speedup_1, width, color='c')
+speedup_8_bars = ax.bar(ind+1*width, speedup_8, width, color='m')
+speedup_64_bars = ax.bar(ind+2*width, speedup_64, width, color='y')
 
-ax.set_xticks(ind+2*width)
+ax.set_xticks(ind+1.5*width)
 ax.set_xticklabels(['$256^3$', '$512^3$',
     '$1024^3$', '$2048^3$'])
 ax.set_xlabel('Problem size')

### 2015-11-24 thesis/acknowledgments.tex
some more updates to thesis

@@
 to my advisor, Dr. Richard S. Miller,
-for hist immense guidance and support;
+for his immense guidance and support;
 and the members of my committee,
@@
 I would like to thank the
-staff of the Cyberinfrastructure Technology Integration,
+staff of the Cyberinfrastructure Technology Integration group at Clemson,
 and especially
@@
 
-Finally,
-I would like to thank my family and friends
+Finally, I thank my family and friends
 for everything they have done for me.

### 2015-11-24 thesis/introduction-and-background.tex
some more updates to thesis

@@
 various data about the flow must be stored---for
-example, the geometric coordinates (x, y and z),
+example, the geometric coordinates ($x$, $y$ and $z$),
 pressure, temperature and velocities at that point.
@@
 \end{equation}
-
+%
 Similar tridiagonal schemes are available
@@
 there are other options.
-Notably,
-the OpenACC \cite{openacc} toolkit allows
+The OpenCL framework \cite{opencl}
+is used to write applications
+that can be ported across
+a variety architectures including NVIDIA GPUs.
+and introduces almost no extra code.
+Several GPU accelerated libraries \cite{gpulibraries}
+allow existing applications to be ported to GPUs
+by offering compatibility with industry standard CPU libraries.
+The OpenACC \cite{openacc} toolkit allows
 portions of code to be offloaded to GPU accelerators
-simply by use of compiler directives.
+simply by use of compiler directives,
+similar to OpenMP.
 This makes porting existing applications
-extremely easy,
-and introduces almost no extra code.
-Some libraries for scientific computation,
-such as PETSc \cite{petsc-web-page},
-also allow the \emph{transparent} use of GPUs
-through their interfaces.
+extremely easy.
+Many frameworks for scientific computing
+such at PETSc \cite{petsc-web-page}
+also allow the use of GPUs
+for most of their functionality.
+
 We use the CUDA interface because

### 2015-11-24 thesis/proposed-tridiagonal-algorithm.tex
some more updates to thesis

@@
 straightforward and compact,
-spanning no more than 100 lines of code.
+spanning no more than 100 lines of code
+(see Appendix).
 
@@
 the values $b_n^m$ are unused
-in Eqs (\ref{eqn:precomputed-forward-reduction-step})-
+in Eqs (\ref{eqn:precomputed-forward-reduction-step}) -
 (\ref{eqn:precomputed-backward-substitution-step-1}).

### 2015-11-24 thesis/results.tex
some more updates to thesis

@@
 solvers to solve different problem sizes.
+Note that data is missing for the CUSPARSE solver
+for the $512^3$ 3-D case,
+as the GPU was unable to accomodate this problem size---this
+is due to the large amount of scratch space required
+by the CUSPARSE implementation.
 
@@
 
+\subsection{Performance profiling}
+
 \begin{figure}[h!]
@@
 %
-Figures \ref{fig:compact-profiling-1024-64}-\ref{fig:compact-profiling-2048-64}
+Figures \ref{fig:compact-profiling-1024-64} - \ref{fig:compact-profiling-2048-64}
 show the time taken by the different steps of
@@
 which justifies our efforts in optimizing the tridiagonal solver.
-For evaluation of the derivatives in the y- and z- directions,
+For evaluation of the derivatives in the $y-$ and $z-$ directions,
 we note that a significant portion of the runtime is dedicated
@@
 We attribute this to our na\"{\i}ve implementation of the
-permutation kernels (no shared memory).
-%
+permutation kernels (no shared memory usage).
+
+\subsection{Strong and weak scaling}
 \begin{figure}
@@
 %
-Figures \ref{fig:strong-scaling-256}-\ref{fig:weak-scaling-256}
+Figures \ref{fig:strong-scaling-256} - \ref{fig:weak-scaling-256}
 show the strong and weak scaling of the compact finite difference solver
 for evaluating derivatives in all three coordinate directions.
+For the strong scaling measurement,
+we keep the problem size fixed
+and increase the number of GPUs used to solve the problem.
+For the weak scaling measurement,
+we keep the problem size \emph{per GPU}
+fixed, and increase the number of GPUs used.
 We note good scaling for both 2-D and 3-D cases.
@@
 as the GPU is kept busy for all problem sizes.
-%
+
+\subsection{Comparison with a CPU-only approach}
+
 \input{table/compact-refimpl-timings.tex}
@@
 \includegraphics[width=300pt]{fig/compact-refimpl-speedups.eps}
-\caption{Speedups over reference implementation}
+\caption{Speedups over reference implementation
+    for computing derivative in the fastest coordinate direction}
 \label{fig:compact-refimpl-speedups}
@@
 \end{figure}
-%
+
 We also compare the performance of our compact finite difference
-solver with the implementation by
-Mohd-Yusof et al. \cite{mohd2010adapting}.
-This implementation uses a
-distributed tridiagonal solver based on the LU algorithm
-to solve the linear systems.
+solver with the approach described by
+Mohd-Yusof et al.\ \cite{mohd2010adapting},
+implemented for CPUs.
+The approach uses a distributed tridiagonal solver based on
+the LU decomposition specialized for tridiagonal systems.
+The problem is divided among individual CPU cores,
+communicating via MPI.
 For comparing timings,
@@
 Each CPU socket uses 8 CPU cores and 1 GPU.
-Thus we maintain a ratio of 1:8 between GPUs and CPU cores
+Thus, we maintain a ratio of 1:8 between GPUs and CPU cores
 in our comparison.
 Table \ref{table:compact-refimpl-timings}
-shows timings for computing derivatives for problem
-sized up to $2048^3$,
+shows timings for computing derivatives
+in the fastest coordinate direction
+for problems sized up to $2048^3$,
 and Fig. \ref{fig:compact-refimpl-speedups}

### 2015-11-29 presentation/background-gpus.tex
add background on GPUs

\subsection{Graphics processing units}

\begin{frame}[t]
\frametitle{Introduction}
\pause
\begin{columns}[T]
\begin{column}{0.4\textwidth}
\begin{itemize}
    \item <2-> Highly multi-threaded processor
        \begin{itemize}
            \item NVIDIA Tesla K20 accelerator: 2496 cores
        \end{itemize}
    \item <3-> Spectacular performance for
        \emph{compute-intensive, data-parallel} operations
    \item <4-> Not suited for general-purpose computation
    \item <5-> Requires careful redesign of algorithms,
        substantial code changes, and tuning efforts
\end{itemize}
\end{column}
\begin{column}{0.6\textwidth}
\visible<2-> {
    \vspace{2cm}
    \includegraphics[width=180px]
    {img/device-comparison.png}
}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Programming GPUs}
\pause
Scientific applications use GPUs at various levels:
\begin{itemize}
    \item Low-level
        \begin{itemize}
            \item CUDA
            \item OpenCL
        \end{itemize}
    \item High-level
        \begin{itemize}[<+->]
            \item OpenACC compiler directives
            \item Drop-in support for scientific libraries such as Intel MKL, BLAS
        \end{itemize}
    \item Transparent
        \begin{itemize}[<+->]
            \item MATLAB: over 200 functions
            \item Software packages: Abaqus, ANSYS (Mechanical/Fluent)
        \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{CUDA programming model}
\pause
\begin{columns}
\begin{column}{0.5\textwidth}
\textbf{Kernels}
\begin{itemize}[<+->]
    \item Special pieces of code that execute on the GPU
    \item In Fortran: subroutines; in C: functions
    \item Executed concurrently by several GPU threads
\end{itemize}
\textbf{Threads}
\begin{itemize}[<+->]
    \item Threads organized into a grid of blocks
    \item Kernels launched with specified grid size
        (number of blocks) and block size (threads per block)
    \item Grid and blocks can be 2-D or 3-D for convenience
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
    \only<5-6>{\includegraphics[width=150px]{img/program-structure.png}}
    \only<7>{\includegraphics[width=150px]{img/grid-of-thread-blocks.png}}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{GPU architecture and memory model}
\pause
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}[<+->]
    \item GPU viewed as a collection of \emph{streaming microprocessors} (SMs)
    \item When kernel is launched, each block gets assigned to an SM
    \item Threads within a block execute concurrently
    \item SM may execute several blocks concurrently
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
    \visible<3->{\includegraphics[width=150px]{img/gpu-scaling.png}}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{GPU architecture and memory model}
\pause
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Global memory{
        \begin{itemize}
            \item large but slow (~5 GB for Tesla K20)
            \item all threads can access
            \item persists between kernel launches
        \end{itemize}}
    \item Shared memory{
        \begin{itemize}
            \item small but fast (~48 KiB per SM)
            \item local to threads within a block
            \item explicitly managed cache
        \end{itemize}}
    \item Registers{
        \begin{itemize}
            \item limited (65536 per SM)
            \item local to individual threads
            \item fastest
        \end{itemize}}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
    \visible<2->{\includegraphics[width=150px]{img/memory-hierarchy.png}}
\end{column}
\end{columns}
\end{frame}

### 2015-11-29 presentation/background-gpus.tex
updates

@@
-\subsection{Graphics processing units}
+\section{Background on graphics processing units (GPUs)}
 
@@
 \frametitle{Introduction}
-\pause
 \begin{columns}[T]
-\begin{column}{0.4\textwidth}
+\begin{column}{0.5\textwidth}
 \begin{itemize}
-    \item <2-> Highly multi-threaded processor
+\item Highly multi-threaded processor
     \begin{itemize}
@@
     \end{itemize}
-    \item <3-> Spectacular performance for
+\item Spectacular performance for
     \emph{compute-intensive, data-parallel} operations
-    \item <4-> Not suited for general-purpose computation
-    \item <5-> Requires careful redesign of algorithms,
+\item Not suited for general-purpose computation
+\item Requires careful redesign of algorithms,
     substantial code changes, and tuning efforts
@@
 \end{column}
-\begin{column}{0.6\textwidth}
-\visible<2-> {
-    \vspace{2cm}
-    \includegraphics[width=180px]
-    {img/device-comparison.png}
-}
+\begin{column}{0.5\textwidth}
+\begin{figure}
+\begin{center}
+\includegraphics[width=150px]{img/device-comparison.png}
+\centering
+\caption{Design philosophy of GPU (CUDA Programming Guide)}
+\end{center}
+\end{figure}
 \end{column}
@@
 \begin{frame}
-\frametitle{Programming GPUs}
-\pause
-Scientific applications use GPUs at various levels:
+\frametitle{CUDA Programming model}
 \begin{itemize}
-    \item Low-level
+\item Several options for programming GPUs
+    \begin{itemize}[<+->]
+        \item underlying models: CUDA and OpenCL
+    \end{itemize}
+\item Important to understand for good performance
+\item Primary features:
     \begin{itemize}
-            \item CUDA
-            \item OpenCL
-        \end{itemize}
-    \item High-level
-        \begin{itemize}[<+->]
-            \item OpenACC compiler directives
-            \item Drop-in support for scientific libraries such as Intel MKL, BLAS
-        \end{itemize}
-    \item Transparent
-        \begin{itemize}[<+->]
-            \item MATLAB: over 200 functions
-            \item Software packages: Abaqus, ANSYS (Mechanical/Fluent)
+        \item \emph{Kernels} and thread organization
+        \item GPU Architecture
+        \item Memory hierarchy
     \end{itemize}
@@
 \begin{frame}
-\frametitle{CUDA programming model}
-\pause
+\frametitle{Kernels and thread organization}
 \begin{columns}
@@
 \textbf{Kernels}
-\begin{itemize}[<+->]
+\begin{itemize}
 \item Special pieces of code that execute on the GPU
@@
 \textbf{Threads}
-\begin{itemize}[<+->]
+\begin{itemize}
 \item Threads organized into a grid of blocks
@@
 \begin{column}{0.5\textwidth}
-    \only<5-6>{\includegraphics[width=150px]{img/program-structure.png}}
-    \only<7>{\includegraphics[width=150px]{img/grid-of-thread-blocks.png}}
+\includegraphics[width=120px]{img/program-structure.png}
+\vspace{1cm}
+\includegraphics[width=120px]{img/grid-of-thread-blocks.png}
 \end{column}
@@
 \begin{frame}
-\frametitle{GPU architecture and memory model}
-\pause
+\frametitle{GPU Architecture}
 \begin{columns}
 \begin{column}{0.5\textwidth}
-\begin{itemize}[<+->]
+\begin{itemize}
 \item GPU viewed as a collection of \emph{streaming microprocessors} (SMs)
@@
 \begin{column}{0.5\textwidth}
-    \visible<3->{\includegraphics[width=150px]{img/gpu-scaling.png}}
+\includegraphics[width=150px]{img/gpu-scaling.png}
 \end{column}
@@
 \begin{frame}
-\frametitle{GPU architecture and memory model}
-\pause
+\frametitle{Memory hierarchy}
 \begin{columns}
@@
     \item large but slow (~5 GB for Tesla K20)
+    \item CPU can access (via PCI-e bus)
     \item all threads can access
@@
 \begin{itemize}
-            \item small but fast (~48 KiB per SM)
+    \item small but fast (~48 KiB per SM and block)
     \item local to threads within a block
+    \item threads in a block read and write into shared memory
     \item explicitly managed cache
@@
 \begin{itemize}
-            \item limited (65536 per SM)
+    \item limited (65536 per SM and block)
     \item local to individual threads
@@
 \begin{column}{0.5\textwidth}
-    \visible<2->{\includegraphics[width=150px]{img/memory-hierarchy.png}}
+    \visible{\includegraphics[width=150px]{img/memory-hierarchy.png}}
 \end{column}
@@
 \end{frame}
+
+\begin{frame}
+\frametitle{Performance considerations}
+\begin{itemize}
+\item CPU-GPU transfers
+\item Coalesced global memory access
+    \begin{itemize}
+        \item successive threads must access nearby locations
+    \end{itemize}
+\item Shared memory and bank conflicts
+    \begin{itemize}
+        \item \emph{strided} memory access is serialized
+    \end{itemize}
+\item Limiting use of shared memory and registers
+\end{itemize}
+\end{frame}

### 2015-11-29 presentation/tridiagonal-solver.tex
updates

\section{Developing a tridiagonal solver}

\begin{frame}
\frametitle{Requirements of tridiagonal solver}

Each line of grid points yields a tridiagonal system:
\small
\begin{equation*}
 \begin{bmatrix}
     1&2\\
     1/4&1&1/4\\
     &1/4&1&1/4\\
     &&1/4&1&1/4\\
     &&&1/4&1&1/4\\
     &&&&&\ddots\\
     &&&&&&\ddots\\
     &&&&&&&\ddots\\
     &&&&&&&2&1
  \end{bmatrix}
  \begin{bmatrix}
      f^{\prime}_1 \\
      f^{\prime}_2 \\
      f^{\prime}_3 \\
      \vdots \\
      \vdots \\
      \vdots \\
      \vdots \\
      f^{\prime}_{n-1} \\
      f^{\prime}_n
   \end{bmatrix}
 =
 \begin{bmatrix}
     \frac{-5f_1 + 4f_2 + f_3}{2dx}\\
     \frac{3(f_{3} - f_{1})}{4dx}\\
     \frac{3(f_{4} - f_{2})}{4dx}\\
     \vdots\\
     \vdots\\
     \vdots\\
     \vdots\\
     \frac{3(f_{n} - f_{n-2})}{4dx}\\
     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2dx}
  \end{bmatrix}
\end{equation*}
\end{frame}

\begin{frame}
\frametitle{Requirements of tridiagonal solver}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item For a 1-D grid, solution gives the derivative
    at all points simultaneously
\item For 2-D and 3-D grids, we solve a tridiagonal system
    for each grid line
\item The coefficient matrix is the same for each system
\item Only right hand sides are different
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
    \hspace{1cm}
    \includegraphics[width=100px]{img/grid-lines.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Thomas algorithm}
\begin{itemize}
\item Derived from Gaussian elimination
\item Requires $2n$ steps and $4n$ storage
\item The most efficient serial implementation
\item What about for GPUs?
    \begin{itemize}
        \item inherently sequential
        \item can use multiple threads to solve
            individual tridiagonal systems
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Issues with GPU implementation}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Uncoalesced global memory accesses
    \item What about shared memory?
        \begin{itemize}
            \item 4N storage \emph{per block}
            \item SM shared memory filled up quickly
            \item very few threads per SM
            \item under-utilizes the GPU parallelism
        \end{itemize}
    \item $2n$ steps - can do better on GPU
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
    \includegraphics[width=150px]{img/pThomas.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Tridiagonal solvers for the GPU}
\begin{itemize}
\item Zhang et al. (2010) describe the implementation
    of three algorithms for tridiagonal systems on GPUs
    \begin{itemize}
        \item Cyclic reduction
        \item Parallel cyclic reduction
        \item Recursive doubling
    \end{itemize}
\item Exhibit \emph{fine-grained} parallelism
\item Subsequent efforts based on CR and PCR primarily
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Cyclic reduction}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Buzbee and Goleb (1970)
\item \textbf{Forward reduction} ($log(n)-1$ steps):
    \begin{itemize}
    \item Eliminate odd-indexed equations at each step
    \item Two-by-two system is left
    \end{itemize}
\item \textbf{Backward substitution} ($log(n)-1$ steps):
    \begin{itemize}
    \item Solve for odd-indexed equations using even-indexed values
    \end{itemize}
\item Best case ($n$ parallel threads): requires $2log(n)-1$ steps
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=150px]{img/cyclic-reduction.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{GPU implementation}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Each tridiagonal system mapped to a thread block;
    individual threads mapped to equations
\item Several systems can be solved at concurrently
\item Forward reduction: diagonal arrays $a$, $b$, $c$ and RHS $d$
    updated \emph{in-place}
\item Backward substitution: solution $x$ computed
    from $a^\prime$, $b^\prime$, $c^\prime$ and $d^\prime$
\item Synchronization required at each step
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\only<1>{
\includegraphics[width=150px]{img/gpu-mapping.eps}
}
\only<2>{
\centering
Forward reduction
\scalebox{0.8}{
\vbox{
\begin{align*} 
    k_1 &= \frac{a_i}{b_{i-1}}, k_2 = \frac{c_i}{b_{i+1}} \\
    a^{\prime}_i &= -a_{i-1}k_1 \\
    b^{\prime}_i &= b_i - c_{i-1}k_1 - a_{i+1}k_2 \\
    c^{\prime}_i &= -c_{i+1}k_2 \\
    d^{\prime}_i &= d_i - d_{i-1}k_1  - d_{i+1}k_2 \\
\end{align*}}}

\centering
Backward substitution
\scalebox{0.8}{
\vbox{
\begin{align*}
x_i &= \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - \
    c^{\prime}_ix_{i+1}}{b^{\prime}_i}
\end{align*}}}}

\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Issues}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Thread activity
    \item Uncoalesced global memory access
    \item Bank conflicts
    \item Limited shared memory
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=150px]{img/cyclic-reduction.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Efforts for improving performance}
\begin{itemize}
    \item Zhang et al.: hybrid CR+PCR solver
    \item G{\"o}ddeke et al.: separate even and odd indexed
        coefficients
    \item Davidson et al.: register packing
    \item Esfahanian et al.: global memory with data rearrangement
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{A novel approach}
\begin{itemize}
\item Solving systems with same RHS
\item Computing $a^\prime$, $b^\prime$ and $c^\prime$
    for each system and at every time step redundant
\item Storing coefficients (similar to LU) - expensive
\item The coefficient matrix has relatively simple structure
\item Can be exploited to improve cyclic reduction performance
\end{itemize}
\end{frame}

\begin{frame}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Diagonals are nearly constant (\emph{near-Toeplitz})
    \item Symmetry may be broken by boundary conditions
    \item Near-Toeplitz matrices occur in other numerical schemes
    \begin{itemize}
        \item Alternating direct implicit methods
        \item Line relaxation methods
        \item Poisson solvers
        \item One-dimensional ODEs and PDEs
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\scalebox{0.6}{%
\vbox{
\begin{equation*}
\begin{bmatrix}
     1&2\\
     1/4&1&1/4\\
     &1/4&1&1/4\\
     &&1/4&1&1/4\\
     &&&1/4&1&1/4\\
     &&&&&\ddots\\
     &&&&&&\ddots\\
     &&&&&&&\ddots\\
     &&&&&&&2&1
  \end{bmatrix}
\end{equation*}}}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{General approach}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item {First reduction step, substitute:
    \scalebox{0.8}{
    \vbox{
    \begin{align*}
        a_2 &= a_3 = a_4 = \hdots &\equiv a_0 \\
        b_2 &= b_3 = b_4 = \hdots &\equiv b_0 \\
        c_2 &= c_3 = c_4 = \hdots &\equiv c_0
    \end{align*}}}}
\item $a^\prime$, $b^\prime$, $c^\prime$
    represent coefficients of the ``reduced'' tridiagonal system
\item \textbf{The reduced system is also near-Toeplitz}
\item Each forward reduction step produces a near-Toeplitz
    tridiagonal system
\item Coefficients can be stored compactly
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
Forward reduction
\scalebox{0.8}{
\vbox{
\begin{align*} 
    k_1 &= \frac{a_i}{b_{i-1}}, k_2 = \frac{c_i}{b_{i+1}} \\
    a^{\prime}_i &= -a_{i-1}k_1 \\
    b^{\prime}_i &= b_i - c_{i-1}k_1 - a_{i+1}k_2 \\
    c^{\prime}_i &= -c_{i+1}k_2 \\
    d^{\prime}_i &= d_i - d_{i-1}k_1  - d_{i+1}k_2 \\
\end{align*}}}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
Cyclic reduction reduced to:

\vspace{1cm}

\textbf{Forward reduction}
\begin{align*}
d^{\prime}_i = d_i - d_{i-1}k_1^{m}  - d_{i+1}k_2^{m}
\end{align*}

\textbf{Backward substitution}
\begin{align*}
x_i = \frac{d^{\prime}_i - a^mx_{i-1} - \
    c^{m}x_{i+1}}{b^m}
\end{align*}

where $a^m$, $b^m$ and $c^m$ are the precomputed
coefficients for $i>1$ at the step $m$.
\end{frame}

### 2015-11-30 presentation/compact-finite-difference-application.tex
add section on compact finite difference application and results

\section{Application to compact finite differences}

\begin{frame}
\frametitle{Application to compact finite difference schemes}
\begin{itemize}
\item Compact finite difference evaluation involves two steps:
    \begin{enumerate}
        \item Evaluation of the right hand sides
        \begin{itemize}
            \item easily performed on the GPU
        \end{itemize}
        \item Solving the tridiagonal systems for all the grid lines
    \end{enumerate}
\item Typical DNS problems too large to fit on a single GPU
\item Need a strategy for multiple GPUs
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Parallelization strategy: single GPU}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Map grid lines to blocks
    \item Solve multiple blocks simaltaneously
    \item For other coordinate directions,
        transpose the data (physically or logically)
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=150px]{img/compact-single-gpu.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Parallelization strategy: multiple GPUs on a single node}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Method used by Sakharnykh et al.
\item Decompose domain into \emph{subdomains};
    each GPU assigned a subdomain
\item Single subdomain used along derivative direction
\item No coordination between GPUs required
\item For other coordinate directions,
    transpose the data (physically or logically)
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=100px]{img/compact-shared-gpu.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Parallelization strategy: multiple GPUs on distinct nodes}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Single subdomain used along derivative direction
\item No coordination between GPUs required
\item \textbf{Physical} global transpose required
\item Subdomains can become impractically slender
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=100px]{img/compact-distributed-restricted.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Parallelization strategy: multiple GPUs on distinct nodes}
\centering
\includegraphics[width=200px]{img/compact-restricted-transpose.eps}
\end{frame}

\begin{frame}
\frametitle{Parallelization strategy: multiple GPUs on distinct nodes}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Need a distributed tridiagonal solver
\item Can ease communication costs
\item Remove restrictions on domain sizes
\item Greatly increases overall complexity
\item Especially difficult for GPUs
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=100px]{img/compact-distributed-all.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Distributed tridiagonal solver}

Solving a tridiagonal system of size
$mP$, split among $P$ processes:

\footnotesize
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{enumerate}
\item Each process solves the local systems:
    \begin{align*}
        A^px_r^{p} &= r_p \\
        A^pu^p &= \{-a_1^p, 0, 0, 0, \hdots\} \\
        A^pl^p &= \{0, 0, 0 \hdots, -c_m^p\}
    \end{align*}
\item Construct and assemble a ``reduced'' system
    of the form:
    \scalebox{0.6}{
        \vbox{
    \begin{align*}
     \begin{bmatrix}
l^1_m & -1 \\
-1    & u^2_1 & l^2_1 \\
  & u^2_m & l^2_m & -1 \\
  &       & -1    & u^3_1 & l^3_1 \\
  &       &       & u^3_m & l^3_m  & -1 \\
  &       &       &       & \ddots & \ddots & \ddots \\
  &       &       &       &        & -1     & u^P_1
\end{bmatrix}
\begin{bmatrix}
\beta^1 \\
\alpha^2 \\
\beta^2 \\
\alpha^3 \\
\beta^3 \\
\vdots \\
\alpha^P
\end{bmatrix}
=
\begin{bmatrix}
x_{r,m}^1 \\
x_{r,1}^2 \\
x_{r,m}^2 \\
x_{r,1}^3 \\
x_{r,m}^3 \\
\vdots \\
x_{r,1}^P \\
\end{bmatrix}  
        \end{align*}}}
\item Local part of solution given by:
\begin{align*}
x^p = x_r^p + \
    \alpha^p u^p + \beta^p l^p
\end{align*}
\end{enumerate}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=120px]{img/dividing-tridiagonal.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\begin{itemize}
\item 3 local tridiagonal systems ($m$ equations)
\item 1 ``global'' tridiagonal system ($2P-2$ equations)
\item Must be done for each grid line
\item Sufficient to consider a single line of processes
\end{itemize}
\begin{center}
\includegraphics[height=100px]{img/single-line-of-processes.eps}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Right hand side evaluation}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Stencil pattern: easily performed on GPU
\item Each process communicates with its neighbours
    (halo-swapping)
\item NVIDIA GPUDirect technology for efficient
    GPU-GPU communication
\item Developed a tool to manage structured grid data
    and transparently perform halo swaps
    for multiple GPUs
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=150px]{img/rhs-halos.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Solving local systems}
\begin{itemize}
\item Local tridiagonal systems solved using
    algorithm described in previous sections
\item For 2-D problems,
    $N_{rhs} = m$
    for 3-D problems
    $N_{rhs} = m^2$
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Constructing reduced system}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Involves global communication
\item Produces tridiagonal system with
    several right-hand sides
\item Systems are \emph{interleaved}
\item Solved by thread-parallel Thomas
    (pThomas) algorithm
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\includegraphics[width=100px]{img/constructing-reduced-system.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Summing the solutions}
\begin{itemize}
\item Pointwise-parallel: perfect fit for GPU
\item No communication involved
\end{itemize}
\end{frame}

### 2015-11-30 presentation/results.tex
add section on compact finite difference application and results

\section{Results}

\begin{frame}
\frametitle{Overview of results}
\begin{itemize}
\item Results for tridiagonal solver
\begin{itemize}
    \item Global memory v/s shared memory implementations
    \item Compared against library solvers
        (multicore CPU and GPU)
\end{itemize}
\item Results for compact finite difference application
\begin{itemize}
    \item Profiling results
    \item Scaling results
    \item Comparison with CPU-only approach
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Global memory v/s shared memory implementations}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item For 2-D problems, shared memory clearly better
\item For 3-D problems, shared memory speedup
    is smaller
\item Global memory makes better use of
    GPU parallelism in 3-D case
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=140px]{img/global-vs-shared-2d.eps}

\includegraphics[width=140px]{img/global-vs-shared-3d.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Comparison with library solvers - timings}
\begin{table}
\resizebox{\textwidth}{!}{%
\input{table/bench.tex}
}
\end{table}
\end{frame}

\begin{frame}
\frametitle{Comparison with library solvers - speedup}
\centering
\includegraphics[width=180px]{img/bench-2d.eps}
\includegraphics[width=180px]{img/bench-3d.eps}
\begin{table}
\end{table}
\end{frame}

\begin{frame}
\frametitle{Note on speedups}
\begin{itemize}
\item Comparing against 8 CPU cores, speedups in the range 1-2x
\item Solving the tridiagonal systems
    is the \emph{least amenable to efficient parallel solution}
\item Objective is to keep the problem on the GPU
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Compact finite difference evaluation - profiling}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
    \item Problem sizes: $1024^3$ and $2048^3$ on 64 GPUs
    \item For larger problems, more time spent
        on tridiagonal systems
    \item Significant portion of time spent on
        data permutation
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\includegraphics[width=170px]{img/profiling-1024-64.eps}

\includegraphics[width=170px]{img/profiling-2048-64.eps}
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Compact finite difference evaluation - strong scaling}
\centering
\includegraphics[width=180px]{img/strong-scaling-256.eps}
\includegraphics[width=180px]{img/strong-scaling-512.eps}

Problem sizes: $256^3$ and $512^3$
\end{frame}

\begin{frame}
\frametitle{Compact finite difference evaluation - weak scaling}
\centering
\includegraphics[width=180px]{img/weak-scaling-128.eps}
\includegraphics[width=180px]{img/weak-scaling-256.eps}

Problem sizes: $128^3$ and $256^3$ \emph{per process}
\end{frame}

\begin{frame}
\frametitle{Comparison with CPU-only approach}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Reference implementation: approach in CFDNS code
    (Los Alamos National Lab)
\item Maintain a ratio of 8 CPU cores : 1 GPU in comparison
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\centering
\begin{table}
\resizebox{0.75\textwidth}{!}{%
\input{table/compact-refimpl-timings.tex}
}
\end{table}

\includegraphics[width=140px]{img/compact-refimpl-speedups.eps}
\end{column}
\end{columns}
\end{frame}

### 2015-12-01 presentation/background-gpus.tex
updates to presentation

@@
 \begin{frame}[t]
-\frametitle{Introduction}
+\frametitle{Graphics processing units}
 \begin{columns}[T]
@@
 \begin{itemize}
-\item Highly multi-threaded processor
+\item Highly parallel and multi-threaded accelerator
     \begin{itemize}
         \item NVIDIA Tesla K20 accelerator: 2496 cores
+        \item Focus on \emph{throughput}, not execution speed
     \end{itemize}
-\item Spectacular performance for
-    \emph{compute-intensive, data-parallel} operations
-\item Not suited for general-purpose computation
-\item Requires careful redesign of algorithms,
-    substantial code changes, and tuning efforts
+\item Spectacular performance for workloads that
+    are compute-intensive, data-parallel
+    (\emph{fine grained parallelism})
+\item Effective use requires careful redesign
+    of algorithms, code modification,
+    and substantial tuning efforts
+\item Vendors: NVIDIA, AMD
 \end{itemize}
@@
 \centering
-\caption{Design philosophy of GPU (CUDA Programming Guide)}
 \end{center}
@@
 \begin{frame}
-\frametitle{CUDA Programming model}
+\frametitle{Performance considerations}
 \begin{itemize}
-\item Several options for programming GPUs
-    \begin{itemize}[<+->]
-        \item underlying models: CUDA and OpenCL
+\item CPU-GPU transfers expensive -
+    hide communication cost by keeping GPU busy
+\item Expose as much parallelism as possible
+    to utilize GPU cores effectively
+\item Avoid thread divergence and excessive
+    synchronization
+\item Memory access:
+    \begin{itemize}
+        \item Threads must access memory in \emph{coalesced}
+            fashion
+        \item Threads must do enough work to hide
+            memory access latency
     \end{itemize}
-\item Important to understand for good performance
-\item Primary features:
-    \begin{itemize}
-        \item \emph{Kernels} and thread organization
-        \item GPU Architecture
-        \item Memory hierarchy
-    \end{itemize}
+\item Strided or uncoalesced memory access patterns:
+    use explicitly managed cache called (\emph{shared} memory)
 \end{itemize}
 \end{frame}
-
-\begin{frame}
-\frametitle{Kernels and thread organization}
-\begin{columns}
-\begin{column}{0.5\textwidth}
-\textbf{Kernels}
-\begin{itemize}
-\item Special pieces of code that execute on the GPU
-\item In Fortran: subroutines; in C: functions
-\item Executed concurrently by several GPU threads
-\end{itemize}
-\textbf{Threads}
-\begin{itemize}
-\item Threads organized into a grid of blocks
-\item Kernels launched with specified grid size
-    (number of blocks) and block size (threads per block)
-\item Grid and blocks can be 2-D or 3-D for convenience
-\end{itemize}
-\end{column}
-\begin{column}{0.5\textwidth}
-\includegraphics[width=120px]{img/program-structure.png}
-\vspace{1cm}
-\includegraphics[width=120px]{img/grid-of-thread-blocks.png}
-\end{column}
-\end{columns}
-\end{frame}
-
-\begin{frame}
-\frametitle{GPU Architecture}
-\begin{columns}
-\begin{column}{0.5\textwidth}
-\begin{itemize}
-\item GPU viewed as a collection of \emph{streaming microprocessors} (SMs)
-\item When kernel is launched, each block gets assigned to an SM
-\item Threads within a block execute concurrently
-\item SM may execute several blocks concurrently
-\end{itemize}
-\end{column}
-\begin{column}{0.5\textwidth}
-\includegraphics[width=150px]{img/gpu-scaling.png}
-\end{column}
-\end{columns}
-\end{frame}
-
-\begin{frame}
-\frametitle{Memory hierarchy}
-\begin{columns}
-\begin{column}{0.5\textwidth}
-\begin{itemize}
-\item Global memory{
-\begin{itemize}
-    \item large but slow (~5 GB for Tesla K20)
-    \item CPU can access (via PCI-e bus)
-    \item all threads can access
-    \item persists between kernel launches
-\end{itemize}}
-\item Shared memory{
-\begin{itemize}
-    \item small but fast (~48 KiB per SM and block)
-    \item local to threads within a block
-    \item threads in a block read and write into shared memory
-    \item explicitly managed cache
-\end{itemize}}
-\item Registers{
-\begin{itemize}
-    \item limited (65536 per SM and block)
-    \item local to individual threads
-    \item fastest
-\end{itemize}}
-\end{itemize}
-\end{column}
-\begin{column}{0.5\textwidth}
-    \visible{\includegraphics[width=150px]{img/memory-hierarchy.png}}
-\end{column}
-\end{columns}
-\end{frame}
-
-\begin{frame}
-\frametitle{Performance considerations}
-\begin{itemize}
-\item CPU-GPU transfers
-\item Coalesced global memory access
-    \begin{itemize}
-        \item successive threads must access nearby locations
-    \end{itemize}
-\item Shared memory and bank conflicts
-    \begin{itemize}
-        \item \emph{strided} memory access is serialized
-    \end{itemize}
-\item Limiting use of shared memory and registers
-\end{itemize}
-\end{frame}

### 2015-12-01 presentation/compact-finite-difference-application.tex
updates to presentation

@@
 \end{enumerate}
+\item Current efforts:
+\begin{itemize}
+    \item Tutkun et al. (2012)
+    \item Esfahanian et al. (2014)
+\end{itemize}
 \item Typical DNS problems too large to fit on a single GPU
@@
 \end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Parallelization strategy: single GPU}
-\begin{columns}
-\begin{column}{0.5\textwidth}
-\begin{itemize}
-    \item Map grid lines to blocks
-    \item Solve multiple blocks simaltaneously
-    \item For other coordinate directions,
-        transpose the data (physically or logically)
-\end{itemize}
-\end{column}
-\begin{column}{0.5\textwidth}
-\includegraphics[width=150px]{img/compact-single-gpu.eps}
-\end{column}
-\end{columns}
 \end{frame}
@@
 \item Subdomains can become impractically slender
+\item Scalable approach needs a distributed tridiagonal solver
 \end{itemize}
@@
 \begin{column}{0.5\textwidth}
-\includegraphics[width=100px]{img/compact-distributed-restricted.eps}
-\end{column}
-\end{columns}
-\end{frame}
-
-\begin{frame}
-\frametitle{Parallelization strategy: multiple GPUs on distinct nodes}
-\centering
-\includegraphics[width=200px]{img/compact-restricted-transpose.eps}
-\end{frame}
-
-\begin{frame}
-\frametitle{Parallelization strategy: multiple GPUs on distinct nodes}
-\begin{columns}
-\begin{column}{0.5\textwidth}
-\begin{itemize}
-\item Need a distributed tridiagonal solver
-\item Can ease communication costs
-\item Remove restrictions on domain sizes
-\item Greatly increases overall complexity
-\item Especially difficult for GPUs
-\end{itemize}
-\end{column}
-\begin{column}{0.5\textwidth}
-\centering
-\includegraphics[width=100px]{img/compact-distributed-all.eps}
+    \only<1>{\includegraphics[width=100px]{img/compact-distributed-restricted.eps}}
+    \only<2>{\includegraphics[width=150px]{img/compact-restricted-transpose.eps}}
+    \only<3>{\includegraphics[width=120px]{img/compact-distributed-all.eps}}
 \end{column}
@@
 \begin{frame}
-\begin{itemize}
-\item 3 local tridiagonal systems ($m$ equations)
-\item 1 ``global'' tridiagonal system ($2P-2$ equations)
-\item Must be done for each grid line
-\item Sufficient to consider a single line of processes
-\end{itemize}
-\begin{center}
-\includegraphics[height=100px]{img/single-line-of-processes.eps}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Right hand side evaluation}
+\frametitle{Compact finite difference on multi-GPU system}
+\footnotesize
 \begin{columns}
 \begin{column}{0.5\textwidth}
+\begin{enumerate}
+\item Computing RHS for all local grid points - \textbf{requires communication}
 \begin{itemize}
-\item Stencil pattern: easily performed on GPU
-\item Each process communicates with its neighbours
-    (halo-swapping)
-\item NVIDIA GPUDirect technology for efficient
-    GPU-GPU communication
-\item Developed a tool to manage structured grid data
-    and transparently perform halo swaps
-    for multiple GPUs
+    \footnotesize
+    \item halo swaps with neighbouring GPUs
+    \item pointwise stencil
 \end{itemize}
+\item Solving local tridiagonal systems - no communication
+\item Assembling and solving reduced tridiagonal systems - \textbf{requires communication}
+\begin{itemize}
+    \footnotesize
+    \item global communication along line of GPUs
+    \item produces interleaved tridiagonal sysems - pThomas algorithm
+\end{itemize}
+\item Summing solutions from steps 2 and 3 - no communication
+\begin{itemize}
+    \footnotesize
+    \item pointwise sum
+\end{itemize}
+\item For other coordinate directions: \emph{local} permutation of data
+\end{enumerate}
 \end{column}
 \begin{column}{0.5\textwidth}
-\includegraphics[width=150px]{img/rhs-halos.eps}
+\only<1>{\includegraphics[width=150px]{img/rhs-halos.eps}}
+\only<2>{\includegraphics[width=150px]{img/constructing-reduced-system.eps}}
 \end{column}
@@
 \end{frame}
-
-\begin{frame}
-\frametitle{Solving local systems}
-\begin{itemize}
-\item Local tridiagonal systems solved using
-    algorithm described in previous sections
-\item For 2-D problems,
-    $N_{rhs} = m$
-    for 3-D problems
-    $N_{rhs} = m^2$
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Constructing reduced system}
-\begin{columns}
-\begin{column}{0.5\textwidth}
-\begin{itemize}
-\item Involves global communication
-\item Produces tridiagonal system with
-    several right-hand sides
-\item Systems are \emph{interleaved}
-\item Solved by thread-parallel Thomas
-    (pThomas) algorithm
-\end{itemize}
-\end{column}
-\begin{column}{0.5\textwidth}
-\includegraphics[width=100px]{img/constructing-reduced-system.eps}
-\end{column}
-\end{columns}
-\end{frame}
-
-\begin{frame}
-\frametitle{Summing the solutions}
-\begin{itemize}
-\item Pointwise-parallel: perfect fit for GPU
-\item No communication involved
-\end{itemize}
-\end{frame}

### 2015-12-01 presentation/results.tex
updates to presentation

@@
 \begin{itemize}
-\item Results for tridiagonal solver
+\item Results for tridiagonal solver (NEATO)
 \begin{itemize}
@@
 \frametitle{Global memory v/s shared memory implementations}
+\footnotesize
 \begin{columns}
 \begin{column}{0.5\textwidth}
+
+\begin{block}{Benchmark details}
+\begin{itemize}
+\item Solving multiple tridiagonal systems on single GPU
+\item Upto $2048$ equations in $2048$ variables (2-D)
+    and $512^2$ equations in $512$ variables (3-D)
+\item GPU: NVIDIA Tesla K20
+\end{itemize}
+\end{block}
+
 \begin{itemize}
@@
     is smaller
-\item Global memory makes better use of
-    GPU parallelism in 3-D case
+\item Thread inactivity in shared memory leads
+    to reduced performance
 \end{itemize}
@@
 
+\begin{frame}{Comparison with library solvers}
+\footnotesize
+\begin{block}{Benchmark details}
+\begin{itemize}
+\item Solving multiple tridiagonal systems on single GPU and multicore CPU
+\item Upto $2048$ equations in $2048$ variables (2-D)
+    and $512^2$ equations in $512$ variables (3-D)
+\item GPU: NVIDIA Tesla K20
+\item CPU: Intel Xeon E5 v2
+\item Do \emph{not} include cost of CPU-GPU transfer
+\item Library solver details:
+\begin{itemize}
+    \footnotesize
+    \item Intel MKL \texttt{dgtsv} (Gaussian elimination with partial pivoting)
+    \item CUSPARSE \texttt{dgtsv} (Hybrid CR+PCR)
+\end{itemize}
+\item \texttt{-02} level compiler optimizations
+\item Double precision
+\end{itemize}
+\end{block}
+\end{frame}
+
+\begin{frame}
+\frametitle{Comparison with library solvers - speedup}
+Speedups over library implementations:
+\begin{center}
+\includegraphics[width=180px]{img/bench-2d.eps}
+\includegraphics[width=180px]{img/bench-3d.eps}
+\end{center}
+\end{frame}
+
 \begin{frame}
 \frametitle{Comparison with library solvers - timings}
+Timings for various solvers:
 \begin{table}
@@
 \begin{frame}
-\frametitle{Comparison with library solvers - speedup}
-\centering
-\includegraphics[width=180px]{img/bench-2d.eps}
-\includegraphics[width=180px]{img/bench-3d.eps}
-\begin{table}
-\end{table}
+\frametitle{Note on speedups}
+\begin{itemize}
+\item Faster than CUSPARSE in almost all cases
+\item Comparing against 8 CPU cores,
+    speedups over MKL in the range 1-2x
+\item Solving the tridiagonal systems
+    \emph{least amenable to efficient parallel solution}
+\item Objective is to keep the problem on the GPU
+\item Multicore CPU performance flattens out
+\end{itemize}
 \end{frame}
 
-\begin{frame}
-\frametitle{Note on speedups}
+\begin{frame}{Compact finite difference evaluation}
+\footnotesize
+\begin{block}{Benchmark details}
 \begin{itemize}
-\item Comparing against 8 CPU cores, speedups in the range 1-2x
-\item Solving the tridiagonal systems
-    is the \emph{least amenable to efficient parallel solution}
-\item Objective is to keep the problem on the GPU
+\item Compact finite difference evaluation in 3-D for a function $f(x, y, z)$
+\item Problem size up to $2048^3$ on up to 64 GPUs and 512 CPU cores
+\item Tests performed on Palmetto cluster (Infiniband interconnect)
 \end{itemize}
+\end{block}
 \end{frame}
@@
 \item Problem sizes: $1024^3$ and $2048^3$ on 64 GPUs
-    \item For larger problems, more time spent
-        on tridiagonal systems
+\item Most time spent on solving tridiagonal systems
+    \begin{itemize}
+        \item Justifies optimization effort
+    \end{itemize}
 \item Significant portion of time spent on
     data permutation
+    \begin{itemize}
+        \item na\"{\i}ve implementation
+    \end{itemize}
 \end{itemize}
@@
 \begin{itemize}
-\item Reference implementation: approach in CFDNS code
+\item Reference implementation: CFDNS
     (Los Alamos National Lab)
-\item Maintain a ratio of 8 CPU cores : 1 GPU in comparison
+\item Uses LU algorithm
+\item Parallelized among individual cores
+\item Maintain a ratio of 8 CPU cores : 1 GPU for comparison
 \end{itemize}
@@
 \end{frame}
+
+\begin{frame}
+\frametitle{Conclusions and future work}
+\begin{itemize}
+    \item Tridiagonal solver performance
+        is key in compact finite difference evaluation
+    \item Simple matrix structure can be exploited
+        in the cyclic reduction algorithm
+    \item Substantial speedups can be achieved
+        over CPU
+    \item Applicability to other algorithms: PCR, CR+PCR
+    \item Integration into current DNS code
+    \item Strategies for MIC architecures
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+    Thank you    
+\end{frame}

### 2015-12-01 presentation/tridiagonal-solver.tex
add section on compact finite difference application and results
updates to presentation

@@
 \section{Developing a tridiagonal solver}
-
-\begin{frame}
-\frametitle{Requirements of tridiagonal solver}
-
-Each line of grid points yields a tridiagonal system:
-\small
-\begin{equation*}
- \begin{bmatrix}
-     1&2\\
-     1/4&1&1/4\\
-     &1/4&1&1/4\\
-     &&1/4&1&1/4\\
-     &&&1/4&1&1/4\\
-     &&&&&\ddots\\
-     &&&&&&\ddots\\
-     &&&&&&&\ddots\\
-     &&&&&&&2&1
-  \end{bmatrix}
-  \begin{bmatrix}
-      f^{\prime}_1 \\
-      f^{\prime}_2 \\
-      f^{\prime}_3 \\
-      \vdots \\
-      \vdots \\
-      \vdots \\
-      \vdots \\
-      f^{\prime}_{n-1} \\
-      f^{\prime}_n
-   \end{bmatrix}
- =
- \begin{bmatrix}
-     \frac{-5f_1 + 4f_2 + f_3}{2dx}\\
-     \frac{3(f_{3} - f_{1})}{4dx}\\
-     \frac{3(f_{4} - f_{2})}{4dx}\\
-     \vdots\\
-     \vdots\\
-     \vdots\\
-     \vdots\\
-     \frac{3(f_{n} - f_{n-2})}{4dx}\\
-     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2dx}
-  \end{bmatrix}
-\end{equation*}
-\end{frame}
 
@@
     at all points simultaneously
-\item For 2-D and 3-D grids, we solve a tridiagonal system
-    for each grid line
-\item The coefficient matrix is the same for each system
-\item Only right hand sides are different
+\item For 2-D and 3-D grids,
+    must solve many independent tridiagonal systems
+\item Systems have the same tridiagonal coefficients,
+    different right hand sides
+\item Need a tridiagonal solver that solves
+    a given system for several right hand sides
+\item For 2-D problems $n \sim N_{rhs}$;
+    for 3-D problems $n << N_{rhs}$
 \end{itemize}
@@
     \hspace{1cm}
-    \includegraphics[width=100px]{img/grid-lines.eps}
+    \includegraphics[width=120px]{img/grid-lines.eps}
 \end{column}
@@
 \item Requires $2n$ steps and $4n$ storage
-\item The most efficient serial implementation
+\item The most efficient sequential algorithm
 \item What about for GPUs?
@@
         \item inherently sequential
-        \item can use multiple threads to solve
+        \item use multiple threads to solve
             individual tridiagonal systems
+            (Sakharnykh et al.)
     \end{itemize}
@@
 \begin{itemize}
-    \item Uncoalesced global memory accesses
-    \item What about shared memory?
-        \begin{itemize}
-            \item 4N storage \emph{per block}
-            \item SM shared memory filled up quickly
-            \item very few threads per SM
-            \item under-utilizes the GPU parallelism
-        \end{itemize}
+    \item Uncoalesced memory accesses
+    \item Needs data rearrangement to coalesce accesses
+    \item Does not expose enough parallelism
     \item $2n$ steps - can do better on GPU
+    \item A case in which \emph{algorithm choice
+        significantly impacts GPU performance}
 \end{itemize}
@@
 \begin{column}{0.5\textwidth}
-    \includegraphics[width=150px]{img/pThomas.eps}
+    \includegraphics[width=120px]{img/pThomas.eps}
 \end{column}
@@
     \end{itemize}
-\item Exhibit \emph{fine-grained} parallelism
+\item \textbf{Step efficient}: trade more work per step for \emph{fewer steps}
+\item Exhibit \emph{fine-grained} parallelism - more suited to GPU
 \item Subsequent efforts based on CR and PCR primarily
@@
 \begin{frame}
-\frametitle{GPU implementation}
+\frametitle{Cyclic reduction}
 \begin{columns}
@@
 \begin{itemize}
-\item Each tridiagonal system mapped to a thread block;
-    individual threads mapped to equations
-\item Several systems can be solved at concurrently
-\item Forward reduction: diagonal arrays $a$, $b$, $c$ and RHS $d$
-    updated \emph{in-place}
-\item Backward substitution: solution $x$ computed
-    from $a^\prime$, $b^\prime$, $c^\prime$ and $d^\prime$
-\item Synchronization required at each step
+\item $17$ operations per equation,
+    compared to $8$ in Thomas algorithm
+\item But exhibits more parallelism:
+\begin{itemize}
+    \item Individual systems can be solved independently
+    \item Individual equations can be computed independently
+\end{itemize}
+\item Equations updated in place: $4n$ storage (same as Thomas algorithm)
+\item Synchronization between threads required at each step
 \end{itemize}
@@
 \begin{column}{0.5\textwidth}
-\only<1>{
-\includegraphics[width=150px]{img/gpu-mapping.eps}
-}
-\only<2>{
 \centering
@@
     c^{\prime}_ix_{i+1}}{b^{\prime}_i}
-\end{align*}}}}
-
+\end{align*}}}
 \end{column}
@@
 \frametitle{Issues}
+\footnotesize
 \begin{columns}
@@
 \begin{itemize}
-    \item Thread activity
-    \item Uncoalesced global memory access
-    \item Bank conflicts
-    \item Limited shared memory
+    \item Strided memory access:
+        stride increases by a factor of 2 at each step
+        (decreases during backward substitution)
+    \item \textbf{Solution}: use \emph{shared memory} (cache)
+    \item Thread activity decreases during forward
+        reduction and increases during backward substitution
+    \item \textbf{Solution}: schedule threads for each step
+        \begin{itemize}
+            \footnotesize
+            \item conflicts with shared memory usage
+        \end{itemize}
+    \item Shared memory \emph{bank conflicts}
+        due to power-of-2 strides; reduced GPU occupancy
 \end{itemize}
@@
 \frametitle{A novel approach}
+\begin{columns}
+\begin{column}{0.5\textwidth}
 \begin{itemize}
-\item Solving systems with same RHS
+\item Solving system with same matrix and different RHS
 \item Computing $a^\prime$, $b^\prime$ and $c^\prime$
-    for each system and at every time step redundant
-\item Storing coefficients (similar to LU) - expensive
-\item The coefficient matrix has relatively simple structure
-\item Can be exploited to improve cyclic reduction performance
+    for each system and at every time step - \textbf{redundant}
+\item Storing coefficients (similar to LU) - \textbf{expensive}
+\item Approach: exploit the simple matrix structure
 \end{itemize}
+\end{column}
+\begin{column}{0.5\textwidth}
+\centering
+Forward reduction
+\scalebox{0.8}{
+\vbox{
+\begin{align*} 
+    k_1 &= \frac{a_i}{b_{i-1}}, k_2 = \frac{c_i}{b_{i+1}} \\
+    a^{\prime}_i &= -a_{i-1}k_1 \\
+    b^{\prime}_i &= b_i - c_{i-1}k_1 - a_{i+1}k_2 \\
+    c^{\prime}_i &= -c_{i+1}k_2 \\
+    d^{\prime}_i &= d_i - d_{i-1}k_1  - d_{i+1}k_2 \\
+\end{align*}}}
+\end{column}
+\end{columns}
 \end{frame}
@@
 \begin{itemize}
-    \item Diagonals are nearly constant (\emph{near-Toeplitz})
-    \item Symmetry may be broken by boundary conditions
-    \item Near-Toeplitz matrices occur in other numerical schemes
+\item Diagonals are nearly constant (\emph{near-Toeplitz tridiagonal})
+\item Near-Toeplitz trdiagonal matrices occur in other numerical schemes
 \begin{itemize}
@@
 \end{itemize}
+\item Symmetric: but may be broken by boundary conditions
 \end{itemize}
@@
 \begin{itemize}
-\item {First reduction step, substitute:
+\item {First reduction step: substitute:
     \scalebox{0.8}{
@@
     \end{align*}}}}
-\item $a^\prime$, $b^\prime$, $c^\prime$
-    represent coefficients of the ``reduced'' tridiagonal system
-\item \textbf{The reduced system is also near-Toeplitz}
-\item Each forward reduction step produces a near-Toeplitz
+\item \textbf{The resulting reduced system is also near-Toeplitz}
+\item Each subsequent forward reduction step produces a near-Toeplitz
     tridiagonal system
-\item Coefficients can be stored compactly
+\begin{itemize}
+    \item can be stored compactly
+    \item no need to compute for each RHS/time step
+\end{itemize}
 \end{itemize}

### 2015-12-02 presentation/compact-finite-difference-application.tex
updates

@@
 Solving a tridiagonal system of size
-$mP$, split among $P$ processes:
+$mP$, split among $P$ processes (Mattor et al., 1995):
 
@@
 \begin{column}{0.5\textwidth}
-\only<1>{\includegraphics[width=150px]{img/rhs-halos.eps}}
-\only<2>{\includegraphics[width=150px]{img/constructing-reduced-system.eps}}
+\only<1>{\includegraphics[width=175px]{img/rhs-halos.eps}}
+\only<2>{\includegraphics[width=175px]{img/constructing-reduced-system.eps}}
 \end{column}

### 2015-12-04 thesis/introduction-and-background.tex
updates
typos in thesis

@@
 This is done in the host-code
-using special functions provided by the CUDA API 
-(e.g., \texttt{cudaMemcpy}).
+using special functions provided by the CUDA API.
 Data transfer between the host and the device is
@@
 and also to have better thread activity overall.
-G{\"o}dekke et al. \cite{GoSt11CR}
+G{\"o}ddeke et al. \cite{GoSt11CR}
 use a method of separately storing
@@
 Our approach takes a different route,
-and is focussed on exploiting the specific matrix structure
+and is focused on exploiting the specific matrix structure
 to reduce the number of computations and memory accesses

### 2015-12-04 thesis/results.tex
thesis errata
typos in thesis

@@
 underlying architectures.
-Boundary conditions may prevent
+boundary conditions may prevent
 the matrix from being symmetric and/or diagonally dominant,
@@
 fixed, and increase the number of GPUs used.
-We note good scaling for both 2-D and 3-D cases.
-The strong scaling for 3-D problems is somewhat better
-as the GPU is kept busy for all problem sizes.
+The strong scaling for larger problems is somewhat better
+as the GPU is kept more busy.
 

### 2015-12-14 presentation/background-gpus.tex
updates
final presentation updates

@@
 \includegraphics[width=150px]{img/device-comparison.png}
+\caption{Design philosophies of CPU and GPU.\newline
+Credit: NVIDIA CUDA programming guide}
 \centering
@@
     \begin{itemize}
-        \item Threads must access memory in \emph{coalesced}
+        \item Threads must access GPU memory (\emph{global} memory) in \emph{coalesced}
             fashion
@@
 \item Strided or uncoalesced memory access patterns:
-    use explicitly managed cache called (\emph{shared} memory)
+    use explicitly managed cache (\emph{shared} memory)
 \end{itemize}

### 2015-12-14 presentation/results.tex
corrections to presentation
final presentation updates

@@
 \frametitle{Overview of results}
-\begin{itemize}
+\begin{enumerate}
 \item Results for tridiagonal solver (NEATO)
@@
     \item Global memory v/s shared memory implementations
-    \item Compared against library solvers
+    \item Compared with library solvers
         (multicore CPU and GPU)
@@
 \end{itemize}
-\end{itemize}
+\end{enumerate}
 \end{frame}
@@
 \item Solving multiple tridiagonal systems on single GPU
-\item Upto $2048$ equations in $2048$ variables (2-D)
-    and $512^2$ equations in $512$ variables (3-D)
+\item Up to $2048$ systems with $2048$ equations (2-D)
+    and $512^2$ systems with $512$ equations (3-D)
 \item GPU: NVIDIA Tesla K20
@@
     is smaller
-\item Thread inactivity in shared memory leads
+\item \textbf{Thread inactivity} in shared memory leads
     to reduced performance
@@
 \item Solving multiple tridiagonal systems on single GPU and multicore CPU
-\item Upto $2048$ equations in $2048$ variables (2-D)
+\item Up to $2048$ equations in $2048$ variables (2-D)
     and $512^2$ equations in $512$ variables (3-D)
@@
 \item Double precision
+\item Comparison with up to 16 CPU cores
 \end{itemize}
@@
 Speedups over library implementations:
-\begin{center}
-\includegraphics[width=180px]{img/bench-2d.eps}
-\includegraphics[width=180px]{img/bench-3d.eps}
-\end{center}
+
+\begin{figure}
+\begin{minipage}{.45\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{img/bench-2d.eps}
+    \caption{2-D problems}
+\end{minipage}
+\begin{minipage}{.45\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{img/bench-3d.eps}
+    \caption{3-D problems}
+\end{minipage}
+\end{figure}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Comparison with library solvers - timings}
-Timings for various solvers:
-\begin{table}
-\resizebox{\textwidth}{!}{%
-\input{table/bench.tex}
-}
-\end{table}
-\end{frame}
+\frametitle{Notes on speedups}
+\begin{itemize}
 
-\begin{frame}
-\frametitle{Note on speedups}
+\item Faster than CUSPARSE in almost all cases
 \begin{itemize}
-\item Faster than CUSPARSE in almost all cases
-\item Comparing against 8 CPU cores,
-    speedups over MKL in the range 1-2x
+    \item Can accomodate larger problems than CUSPARSE
+\end{itemize}
+\item Multicore CPU performance flattens out
+\item Compared with 8 CPU cores, speedups over MKL in the range 1-2x
 \item Solving the tridiagonal systems
     \emph{least amenable to efficient parallel solution}
-\item Objective is to keep the problem on the GPU
-\item Multicore CPU performance flattens out
+\item Primary objective is to keep the problem on the GPU
 \end{itemize}
@@
 \begin{frame}{Compact finite difference evaluation}
-\footnotesize
 \begin{block}{Benchmark details}
@@
 \begin{itemize}
-\item Problem sizes: $1024^3$ and $2048^3$ on 64 GPUs
 \item Most time spent on solving tridiagonal systems
@@
     \end{itemize}
+\item Compute/communicate ratio increases with
+    problem size
 \item Significant portion of time spent on
@@
 \centering
-\includegraphics[width=170px]{img/profiling-1024-64.eps}
+\begin{figure}
+\centering
+\includegraphics[width=150px]{img/profiling-1024-64.eps}
 
-\includegraphics[width=170px]{img/profiling-2048-64.eps}
+\includegraphics[width=150px]{img/profiling-2048-64.eps}
+\caption{Problem sizes: $1024^3$ (above)
+    and $2048^3$ (below) - 64 GPUs}
+\end{figure}
 \end{column}
@@
 \frametitle{Compact finite difference evaluation - strong scaling}
+\begin{figure}
+\begin{minipage}{.45\textwidth}
 \centering
-\includegraphics[width=180px]{img/strong-scaling-256.eps}
-\includegraphics[width=180px]{img/strong-scaling-512.eps}
-
-Problem sizes: $256^3$ and $512^3$
+\includegraphics[width=\linewidth]{img/strong-scaling-256.eps}
+\caption{Problem size: $256^3$}
+\end{minipage}
+\begin{minipage}{.45\textwidth}
+\centering
+\includegraphics[width=\linewidth]{img/strong-scaling-512.eps}
+\caption{Problem size: $512^3$}
+\end{minipage}
+\end{figure}
+\centering
 \end{frame}
@@
 \begin{frame}
-\frametitle{Compact finite difference evaluation - weak scaling}
+\frametitle{Compact finite difference evaluation - strong scaling}
+\begin{figure}
+\begin{minipage}{.45\textwidth}
 \centering
-\includegraphics[width=180px]{img/weak-scaling-128.eps}
-\includegraphics[width=180px]{img/weak-scaling-256.eps}
-
-Problem sizes: $128^3$ and $256^3$ \emph{per process}
+\includegraphics[width=\linewidth]{img/weak-scaling-128.eps}
+\caption{Problem size \emph{per GPU}: $128^3$}
+\end{minipage}
+\begin{minipage}{.45\textwidth}
+\centering
+\includegraphics[width=\linewidth]{img/weak-scaling-256.eps}
+\caption{Problem size \emph{per GPU}: $256^3$}
+\end{minipage}
+\end{figure}
+\centering
 \end{frame}
@@
 \item Maintain a ratio of 8 CPU cores : 1 GPU for comparison
+\item Speedups  in the range 3-9x
+\item NEATO has strong potential for GPU-based clusters
 \end{itemize}
@@
 
-\includegraphics[width=140px]{img/compact-refimpl-speedups.eps}
+\includegraphics[width=160px]{img/compact-refimpl-speedups.eps}
 \end{column}

### 2015-12-14 presentation/tridiagonal-solver.tex
updates
corrections to presentation
final presentation updates

@@
-\section{Developing a tridiagonal solver}
+\section{Developing a GPU tridiagonal solver}
 
@@
 \item The most efficient sequential algorithm
-\item What about for GPUs?
+\item What about parallelization/GPUs?
     \begin{itemize}
@@
 \begin{frame}
-\frametitle{Issues with GPU implementation}
+\frametitle{Issues in GPU implementation}
 \begin{columns}
@@
 \item Buzbee and Goleb (1970)
-\item \textbf{Forward reduction} ($log(n)-1$ steps):
+\item \textbf{Forward reduction} ($log_2(n)-1$ steps):
     \begin{itemize}
@@
     \end{itemize}
-\item \textbf{Backward substitution} ($log(n)-1$ steps):
+\item \textbf{Backward substitution} ($log_2(n)-1$ steps):
     \begin{itemize}
@@
     \end{itemize}
-\item Best case ($n$ parallel threads): requires $2log(n)-1$ steps
+\item Best case ($n$ parallel threads): requires $2log_2(n)-1$ steps
 \end{itemize}
@@
 \begin{itemize}
-\item $17$ operations per equation,
-    compared to $8$ in Thomas algorithm
+\item $17n$ operations,
+    compared to $8n$ in Thomas algorithm
 \item But exhibits more parallelism:
@@
 Forward reduction
+
+(diagonals $a$, $b$, $c$ and RHS $d$)
 \scalebox{0.8}{
@@
 \begin{frame}
-\frametitle{Issues}
+\frametitle{Issues in GPU implementation}
 \footnotesize
@@
         due to power-of-2 strides; reduced GPU occupancy
+    \item We provide both implementations:
+        global memory \emph{and} shared memory
 \end{itemize}
@@
 \begin{frame}
+\frametitle{A novel approach}
 \begin{columns}
@@
 \item Diagonals are nearly constant (\emph{near-Toeplitz tridiagonal})
+\item Symmetry broken by boundary conditions
 \item Near-Toeplitz trdiagonal matrices occur in other numerical schemes
@@
 \end{itemize}
-\item Symmetric: but may be broken by boundary conditions
+\item Can be stored compactly compared to general tridiagonal systems
 \end{itemize}
@@
 \begin{itemize}
+\item Each forward reduction step reduces
+    a matrix of size $n$ to $n/2$
 \item {First reduction step: substitute:
@@
 \textbf{Forward reduction}
+
+\hspace{3.85cm} \sout{$k_1 = \frac{a_i}{b_{i-1}}, k_2 = \frac{c_i}{b_{i+1}}$}
+
+\hspace{3.85cm} \sout{$a^{\prime}_i = -a_{i-1}k_1$}
+
+\hspace{3.85cm} \sout{$b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2$}
+
+\hspace{3.85cm} \sout{$c^{\prime}_i = -c_{i+1}k_2$} 
 \begin{align*}
@@
 
-where $a^m$, $b^m$ and $c^m$ are the precomputed
+where $k_1^m$, $k_2^m$, $a^m$, $b^m$ and $c^m$ are the precomputed
 coefficients for $i>1$ at the step $m$.

### 2016-01-07 paper-draft/paper.tex
updates
final updates to paper

@@
 \begin{frontmatter}
-\author{Author names}
-\title{A fast GPU solver for near-Toeplitz tridiagonal systems}
-\maketitle
+\author{Ashwin Srinath \fnref{fn1}}
+\author{Daniel Livescu \fnref{fn2}}
+\author{Richard S. Miller \fnref{fn1}\corref{cor1}}
+
+\fntext[fn1]{Department of Mechanical Engineering, Clemson University, Clemson, SC 29634-0921}
+\fntext[fn2]{Los Alamos National Laboratory, Los Alamos, NM 87544, USA}
+\cortext[cor1]{Corresponding author. Email address: rm@clemson.edu}
+
+\title{Solution of near-Toeplitz tridiagonal systems on
+GPU-accelerated clusters}
 
@@
 (on up to 16 cores).
-We also describe the implementation of a multi-GPU solver
-using MPI and CUDA,
-and show scalability for up to 64 GPUs
-for a 3-D compact finite difference application.
+Further, we present a parallelization strategy
+for GPU-accelerated clusters,
+and show scalability of a compact finite difference applications
+for up to 64 GPUs on the Clemson University Palmetto cluster.
+
 \end{abstract}
@@
 \end{frontmatter}
-
   %---------------------------------------------------------------------%
@@
 First, we describe the implementation of a GPU solver
-for the specific case of near-Toeplitz systems
+for the specific case of \emph{near-Toeplitz} tridiagonal systems
 that uses the known structure to reduce
 computations and memory accesses.
-Secondly, we describe the implementation for
-multi-GPU clusters using MPI and CUDA,
+Second, we describe the implementation for
+multi-GPU clusters, i.e., a distributed memory environment,
 and apply it to the evaluation of compact finite differences.
@@
 Section \ref{sec:proposed-algorithm}
-describes our proposed solver for near-Toeplitz tridiagonal systems.
+describes our proposed solver for \emph{near-Toeplitz} tridiagonal systems.
 Section \ref{sec:gpu-implementation}
@@
 Cyclic reduction is applicable to the solution of
-general linear systems of the form $A\bm{x} = \bm{d}$,
+linear systems of the form $A\bm{x} = \bm{d}$,
 where $A$ is a tridiagonal matrix with diagonals
@@
 \begin{center}
-\includegraphics[height=250pt]{img/cyclic-reduction.eps}
+\includegraphics[height=200pt]{img/cyclic-reduction.eps}
 \end{center}
@@
 built on the ``Kepler'' architecture,
-an overview of which is available at \cite{Keplerwhitepaper}.
+an overview of which is available in \cite{Keplerwhitepaper}.
 
@@
 Much work has been done on optimizing cyclic reduction performance
-ong GPUs.
+on GPUs.
 Zhang et al. \cite{Zhang2010FTS} propose a
@@
 to reduce bank conflicts.
-G{\"o}dekke et al. \cite{GoSt11CR}
+G{\"o}ddeke et al. \cite{GoSt11CR}
 use a method of separately storing
@@
 a means to reduce shared memory usage in cyclic reduction.
-Esfahanian et. al, \cite{esfahanian2014efficient}
+Esfahanian et al., \cite{esfahanian2014efficient}
 avoid shared memory (and associated bank conflicts entirely)
@@
 Our approach takes a different route,
-and is focussed on near-Toeplitz systems---exploiting
+and is focused on \emph{near-Toeplitz} systems---exploiting
 the relatively simple matrix structure
@@
 \begin{center}
-\includegraphics[height=200pt]{img/forward-reduction-step.eps}
+\includegraphics[height=150pt]{img/forward-reduction-step.eps}
 \end{center}
@@
 While this can be done for general tridiagonal systems,
-the applicability to near-Toeplitz systems is
+the applicability to \emph{near-Toeplitz} tridiagonal systems is
 of particular interest due to the lower storage costs.
@@
 
-When the matrix $A$ is near-Toeplitz tridiagonal,
+When the matrix $A$ is \emph{near-Toeplitz} tridiagonal,
 if we substitute in the Eqs.
@@
 $k_{1,i}$ and $k_{2,i}$
-for a near-Toeplitz matrix
+for a \emph{near-Toeplitz} tridiagonal matrix
 is, at most, $log_2n-2 + 2(log_2n-1)$.
@@
 \begin{center}
-\includegraphics[height=250pt]{img/cyclic-reduction-precomputing.eps}
+\includegraphics[height=200pt]{img/cyclic-reduction-precomputing.eps}
 \end{center}
@@
 In this section, we describe the implementation of a GPU solver
-for solving a given near-Toeplitz tridiagonal system
+for solving a given \emph{near-Toeplitz} tridiagonal system
 for $N_{rhs}$ right hand sides.
@@
 \begin{center}
-\includegraphics[width=300pt]{fig/bench-2d.eps}
-\caption{Relative solver performance for 2-D problems. Relative time defined as:}
-$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
-\label{fig:bench-2d}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=300pt]{fig/bench-3d.eps}
-\caption{Relative solver performance for 3-D problems. Relative time defined as:}
-$\frac{\text{Time taken by solver}}{\text{Time taken by NEATO solver}}$
-\label{fig:bench-3d}
+\includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
+\caption{Comparison of global memory and shared
+    memory implementations of NEATO (2D problems).}
+\label{fig:global-vs-shared-2d}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=1.0\linewidth]{fig/global-vs-shared-3d.eps}
+\caption{Comparison of global memory and shared memory
+    implementations of NEATO (3D problems).}
+\label{fig:global-vs-shared-3d}
 \end{center}
@@
 In this section we present a performance overview
-of our Near-Toeplitz tridiagonal solver (NEATO)
+of our \emph{near-Toeplitz} tridiagonal solver (NEATO):
 against a multi-threaded Intel MKL solver and
@@
 of data transfer between the CPU and GPU.
-\texttt{-O2} level compiler optimizations are turned on for both
+\texttt{-O2} level compiler optimizations are enabled for both
 CPU and GPU code;
@@
 \begin{center}
-\includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
-\caption{Comparison of global memory and shared
-    memory implementations of NEATO (2D problems).}
-\label{fig:global-vs-shared-2d}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=1.0\linewidth]{fig/global-vs-shared-3d.eps}
-\caption{Comparison of global memory and shared memory
-    implementations of NEATO (3D problems).}
-\label{fig:global-vs-shared-3d}
+\includegraphics[width=300pt]{fig/bench-2d.eps}
+\caption{Relative solver performance for 2-D problems. Relative time defined as:}
+\text{Time taken by solver}/\text{Time taken by NEATO solver}
+\label{fig:bench-2d}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=300pt]{fig/bench-3d.eps}
+\caption{Relative solver performance for 3-D problems. Relative time defined as:}
+$\text{Time taken by solver}/\text{Time taken by NEATO solver}$
+\label{fig:bench-3d}
 \end{center}
@@
 \end{figure}
+
+In this section, we present performance details
+for evaluating derivatives of a function $f(x, y, z)$
+using a compact finite difference scheme.
+The numerical algorithm and computational domain are
+discussed below:
 
@@
 This boundary treatment maintains the tridiagonal identity
-of the system to be solved, which then has the near-Toeplitz form:
+of the system to be solved, which then has the form:
 
@@
 Some work has been done on the solution of these
-tridiagonal systems using GPUs---Tutkun et. al
+tridiagonal systems using GPUs---Tutkun et al.
 \cite{tutkun2012gpu} compute the inverse of the coefficient matrix,
@@
 the 2-D ADI problem.
-More recently, Esfahanian et. al \cite{esfahanian2014efficient}
+More recently, Esfahanian et al. \cite{esfahanian2014efficient}
 use a variant of cyclic reduction with
@@
 to solve for the right hand sides at each grid line,
-which takes advantage of the known near-Toepltiz structure
+which takes advantage of the known structure
 of the coefficient matrix arising
@@
       &       & -1    & u^3_1 & l^3_1 \\
-      &       &       & u^3_m & l^3_n  & -1 \\
+      &       &       & u^3_m & l^3_m  & -1 \\
       &       &       &       & \ddots & \ddots & \ddots \\
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[trim={0cm 8cm 0cm 8cm},clip,width=500pt]{img/compact-algorithm.pdf}
+\centering
+\caption{Algorithm for evaluating compact finite differences on multiple GPUs,
+(right: CUDA kernels and MPI calls used)}
+\label{fig:compact-algorithm}
+\end{center}
+\end{figure}
+
 We present a specialization of the above algorithm
-applied to the distributed near-Toeplitz systems that
+applied to the distributed tridiagonal systems that
 appear in compact finite difference evaluations,
 and adapted for GPUs.
-Our algorithm is outlined in \ref{fig:compact-algorithm},
+Our algorithm is outlined in Fig. \ref{fig:compact-algorithm},
 and is described for the case of evaluating derivatives
 in the direction along which successive
-function values in a subdomain are stored contigiously in memory,
+function values in a subdomain are stored contiguously in memory,
 i.e., the ``fastest'' coordinate direction.
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[trim={0 7cm 0 7cm},clip,width=500pt]{img/compact-algorithm.pdf}
-\centering
-\caption{Algorithm for evaluating compact finite differences on multiple GPUs,
-(right: CUDA kernels and MPI calls used)}
-\label{fig:compact-algorithm}
-\end{center}
-\end{figure}
-
 \begin{table}[]
@@
                       using the halo values near the boundaries\\ \hline
-NEATO               & Solve the primary near-Toeplitz tridiagonal systems
+NEATO               & Solve the primary tridiagonal systems
                       for the computed right hand sides, giving $x_r$ \\ \hline
@@
 
-First, we note that the secondary systems
+First, we note that
+the secondary systems
 [Eq. (\ref{eqn:secondary-system-1-expanded}) and
@@
 ``slowest'' co-ordinate direction, i.e.,
-the right hand side values for neighbouring grid points
+the right hand side values for neighboring grid points
 are located far apart in memory.
@@
 (function values)
-before applying the above algorithm.
+before applying the above algorithm,
 and again on the output data
@@
 \subsection{Results}
+
+Here, we provide performance results for
+evaluating derivatives using the above approach
+for large, distributed problems.
 
@@
 
-We present a GPU solver for near-Toeplitz tridiagonal systems
+We present a GPU solver for \emph{near-Toeplitz} tridiagonal systems
 based on the cyclic reduction algorithm,

### 2016-01-09 conference-paper/template/paper.tex
add conference paper

\documentclass[letterpaper,10pt]{article}
\usepackage{ictam}
\usepackage{amsmath,bm}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{caption}

\begin{document}

\title{A novel strategy for compact finite difference evaluation
on GPU-accelerated clusters}

\author[1]{\underline{Ashwin Srinath}}
\author[2]{Daniel Livescu}
\author[3]{Richard S. Miller \footnote{
Corresponding author. Email: rm@clemson.edu}}
\affil[1,3]{
Department of Mechanical Engineering, Clemson University, Clemson, SC 29634-0921}
\affil[2]{Los Alamos National Laboratory, Los Alamos, NM 87544, USA}
\maketitle

\begin{abstract} 
In this work,
we present a novel strategy to solve the
tridiagonal systems arising in such numerical schemes
as compact finite differences
and alternating direction implicit methods
on graphics processing units (GPUs).
We demonstrate the impact of the simple matrix structure
on the \emph{cyclic reduction} algorithm,
and show that precomputation of coefficients appearing in the
algorithm becomes feasible and efficient.
We demonstrate that an implementation using our approach
is able to outperform
the NVIDIA CUSPARSE and multithreaded Intel MKL solvers
on GPU and CPU respectively.
We apply this tridiagonal solver
to the solution of
compact finite differences on multiple GPUs
distributed in a cluster,
and show scaling for up to 64 GPUs.

\end{abstract}

\section{Introduction}

Compact finite difference schemes are widely used
in computational fluid dynamics (CFD)
for their ability to resolve the high-wavenumber fluctuations
associated with turbulent flows.
In CFD codes that a finite difference approach,
the evaluation of the derivatives using compact schemes
generally constitutes a significant portion of the runtime.
The high computational cost associated with
compact finite difference schemes arises from the fact that they require
the solution of tridiagonal linear systems.
Classically, the Thomas algorithm has been used to solve such linear systems.
But with multi-core CPUs, graphics processing units (GPUs),
and more recently,
architectures such as the
Intel Many Integrated Core (MICs)
expected to be mainstays in scientific computing,
such sequential algorithms need to be
replaced by algorithms more suited to these architectures.
In this study,
we develop an algorithm to solve the tridiagonal systems
arising in compact finite difference evaluation
and similar numerical schemes.
Our approach is based on \emph{cyclic reduction},
an extremely successfull algorithm for solving multiple
tridiagonal systems on GPUs.
Our work fills two existing gaps in the literature.
First, we develop an approach for \emph{multiple} GPUs
in a distributed system (GPU-accelerated clusters).
Second, we specialize the cyclic reduction for the case
of \emph{near-Toeplitz} tridiagonal systems,
a special class of systems arising in numerical schemes
such as compact finite differences,
alternating direction implicit methods,
line relaxation methods, and others.

\section{Approach}

If $f_i$ represents the value of
a uniformly sampled function evaluated at the $i^{th}$ sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:

%
\begin{equation}
\begin{split}
    f_i^{\prime} + \
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{h} + \
    b\frac{f_{i+2} - f_{i-2}}{h} + \\
    c\frac{f_{i+3} - f_{i-3}}{h} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%

The derivative is defined implicitly,
requiring the solution of a linear system to solve for
the derivative at all points $i$ simultaneously.
Of particular interest are schemes that yield
\emph{tridiagonal} systems ($\beta=0$).
An example of such a scheme is the fourth-order accurate
Pad\'{e} scheme, which yields the following system:

\begin{equation}
 \label{eqn:compact-tridiagonal-system}
 \begin{bmatrix}
     1&2\\
     1/4&1&1/4\\
     &1/4&1&1/4\\
     &&1/4&1&1/4\\
     &&&1/4&1&1/4\\
     &&&&&\ddots\\
     &&&&&&\ddots\\
     &&&&&&&\ddots\\
     &&&&&&&2&1
  \end{bmatrix}
  \begin{bmatrix}
      f^{\prime}_1 \\

      f^{\prime}_2 \\
      f^{\prime}_3 \\
      \vdots \\
      \vdots \\
      \vdots \\
      \vdots \\
      f^{\prime}_{n-1} \\
      f^{\prime}_n
   \end{bmatrix}
 =
 \begin{bmatrix}
     \frac{-5f_1 + 4f_2 + f_3}{2dx}\\
     \frac{3(f_{3} - f_{1})}{4dx}\\
     \frac{3(f_{4} - f_{2})}{4dx}\\
     \vdots\\
     \vdots\\
     \vdots\\
     \vdots\\
     \frac{3(f_{n} - f_{n-2})}{4dx}\\
     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2dx}
  \end{bmatrix}
\end{equation}

We note the \emph{near-Toeplitz} nature of the tridiagonal system,
i.e., constant diagonal coefficients except near the boundaries.
The cyclic reduction algorithm consists of two phases,
each containing $log_2(N)-1$ steps.
At each step of the \emph{forward reduction} phase,
the odd-indexed equations are eliminated, yielding eventually
a 2-by-2 system.
At each step of the \emph{backward substitution} phase,
the odd-indexed equations are solved from the known solutions
of the even-indexed equations.
Cyclic reduction is well-suited to highly threaded processors
such as GPUs, as the individual equations can be processed in parallel.

When cyclic reduction is applied to
\emph{near-Toeplitz} tridiagonal systems,
we make the important observation that
each forward reduction step preserves the near-Toeplitz
nature of the linear system, i.e.,
the elimination of the odd-indexed equations at each step
yields a new linear system that has
exactly the same near-Toeplitz structure.
This makes it feasible to precompute and store
the coefficients of each linear system,
significantly reducing the amount of computation required.
As it turns out,
this approach also ameliorates several
GPU architecture-specific issues relating to
memory access patterns.
To extend the above approach to multiple GPUs,
we adapt the algorithm described by Mattor et. al
to GPUs.
Our implementation ensures that
each step of this algorithm is performed on the GPU,
avoiding memory transfers to and from the CPU entirely.

\section{Results}

\begin{figure}
\centering
\begin{minipage}[b]{0.45\textwidth}
\begin{center}
\includegraphics[width=220pt]{fig/bench-2d.eps}
\end{center}
\end{minipage}
\begin{minipage}[b]{0.45\textwidth}
\begin{center}
\includegraphics[width=220pt]{fig/bench-3d.eps}
\end{center}
\end{minipage}
\caption{Relative solver performance
for 2-D (left) and 3-D (right) problems.
Relative solver time defined as:}
Time taken by solver/Time taken by NEATO solver
\label{fig:bench}
\end{figure}

\begin{figure}
\centering
\begin{minipage}[b]{0.45\textwidth}
\begin{center}
\includegraphics[width=220pt]{fig/strong-scaling-512.eps}
\end{center}
\end{minipage}
%
\begin{minipage}[b]{0.45\textwidth}
\begin{center}
\includegraphics[width=220pt]{fig/weak-scaling-256.eps}
\end{center}
\end{minipage}
\caption{Strong scaling (left) and weak scaling (right)
    for compact finite difference application.
    Problem sizes $512^3$ and $2048^3$ respectively.}
\label{fig:scaling}
\end{figure}

In Figure \ref{fig:bench},
we compare the performance of our GPU tridiagonal solver
for near-Toeplitz systems (NEATO),
with the equivalent library solvers from
Intel MKL (CPU) and NVIDIA CUSPARSE (GPU).
Here we solve the tridiagonal system in
Equation \ref{eqn:compact-tridiagonal-system}
for sizes $N$ and multiple right hand sides $N_{rhs}$.
We present results for two cases, $N_{rhs}=N$,
(representative of 2-D problems)
and $N_{rhs}=N^2$ (3-D problems).
In Figure \ref{fig:scaling},
we present the strong and weak scaling for a
multi-GPU compact finite difference application.
Here, we solve for the derivatives of a 3-D function
in a domain distributed among multiple GPUs in a cluster.
Derivatives in the non-contiguous directions are solved
using a \emph{local} data rearrangement approach.

\section{Conclusions}

We have present an algorithm
for efficiently solving a common class of
tridiagonal linear systems
based on the cyclic reduction algorithm for GPUs,
Our solver gives better performance than
the current multicore Intel MKL and CUSPARSE solvers
Additionally, we describe an implementation of our solver
for GPU clusters,
and provide performance results for a
compact finite difference application,
run on up to 64 GPUs.lusters.

\end{document}

### 2016-03-16 conference-paper/template/paper.tex
add final changes to ocnference paper

@@
 \usepackage{caption}
+\usepackage{cite}
 
@@
 
-\title{A novel strategy for compact finite difference evaluation
-on GPU-accelerated clusters}
+\title{A NOVEL STRATEGY FOR COMPACT FINITE DIFFERENCE EVALUATION
+ON GPU-ACCELERATED CLUSTERS}
 
@@
 the NVIDIA CUSPARSE and multithreaded Intel MKL solvers
-on GPU and CPU respectively.
+on GPU and CPU, respectively.
 We apply this tridiagonal solver
@@
 
-\section{Introduction}
-
-Compact finite difference schemes are widely used
+Compact finite difference schemes
+\cite{lele1992compact,kennedy1994several}
+are widely used
 in computational fluid dynamics (CFD)
@@
 associated with turbulent flows.
-In CFD codes that a finite difference approach,
-the evaluation of the derivatives using compact schemes
-generally constitutes a significant portion of the runtime.
+In CFD codes using compact schemes,
+this generally constitutes a significant portion of the runtime.
 The high computational cost associated with
@@
 Our approach is based on \emph{cyclic reduction},
-an extremely successfull algorithm for solving multiple
+an extremely successful algorithm for solving multiple
 tridiagonal systems on GPUs.
 Our work fills two existing gaps in the literature.
-First, we develop an approach for \emph{multiple} GPUs
+First,
+whereas current efforts to develop tridiagonal solvers
+are focused on single GPUs (e.g., 
+\cite{GoSt11CR,
+davidson2011register,
+esfahanian2014efficient,
+tutkun2012gpu,
+zhao2015efficiently})
+we develop an approach for \emph{multiple} GPUs
 in a distributed system (GPU-accelerated clusters).
@@
 
-\section{Approach}
-
 If $f_i$ represents the value of
-a uniformly sampled function evaluated at the $i^{th}$ sample point,
+a uniformly sampled function with spacing $dx$,
+evaluated at the $i^{th}$ sample point,
 the first derivative $f^{\prime}_i$ can be approximated from
 a relation of the form:
-
 %
@@
     f_i^{\prime} + \
-    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
-    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
-    \hdots
+    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1})
     = 
-    a\frac{f_{i+1} - f_{i-1}}{h} + \
-    b\frac{f_{i+2} - f_{i-2}}{h} + \\
-    c\frac{f_{i+3} - f_{i-3}}{h} + \
+    a\frac{f_{i+1} - f_{i-1}}{dx} + \
+    b\frac{f_{i+2} - f_{i-2}}{dx} + \
+    c\frac{f_{i+3} - f_{i-3}}{dx} + \
     \hdots
@@
 %
-
 The derivative is defined implicitly,
@@
 the derivative at all points $i$ simultaneously.
-Of particular interest are schemes that yield
-\emph{tridiagonal} systems ($\beta=0$).
 An example of such a scheme is the fourth-order accurate
 Pad\'{e} scheme, which yields the following system:
-
+%
 \begin{equation}
@@
      1/4&1&1/4\\
-     &1/4&1&1/4\\
-     &&1/4&1&1/4\\
-     &&&1/4&1&1/4\\
-     &&&&&\ddots\\
-     &&&&&&\ddots\\
-     &&&&&&&\ddots\\
-     &&&&&&&2&1
+     &&\ddots\\
+     &&&\ddots\\
+     &&&&\ddots\\
+     &&&&2&1
   \end{bmatrix}
@@
       f^{\prime}_1 \\
-
       f^{\prime}_2 \\
-      f^{\prime}_3 \\
-      \vdots \\
-      \vdots \\
       \vdots \\
@@
      \frac{3(f_{3} - f_{1})}{4dx}\\
-     \frac{3(f_{4} - f_{2})}{4dx}\\
-     \vdots\\
-     \vdots\\
      \vdots\\
@@
 \end{equation}
-
+%
 We note the \emph{near-Toeplitz} nature of the tridiagonal system,
@@
 Cyclic reduction is well-suited to highly threaded processors
-such as GPUs, as the individual equations can be processed in parallel.
+such as GPUs \cite{Zhang2010FTS},
+as the individual equations can be processed in parallel.
+\pagebreak
 
@@
 we adapt the algorithm described by Mattor et. al
+\cite{mattor1995algorithm}
 to GPUs.
@@
 avoiding memory transfers to and from the CPU entirely.
-
-\section{Results}
-
+%
 \begin{figure}
@@
 \end{figure}
-
+%
+\iffalse
 \begin{figure}
@@
 \end{minipage}
-%
 \begin{minipage}[b]{0.45\textwidth}
@@
 \end{figure}
-
-In Figure \ref{fig:bench},
+\fi
+%
+In Fig. \ref{fig:bench},
 we compare the performance of our GPU tridiagonal solver
@@
 Here we solve the tridiagonal system in
-Equation \ref{eqn:compact-tridiagonal-system}
+Eq. (\ref{eqn:compact-tridiagonal-system})
 for sizes $N$ and multiple right hand sides $N_{rhs}$.
@@
 and $N_{rhs}=N^2$ (3-D problems).
-In Figure \ref{fig:scaling},
-we present the strong and weak scaling for a
+On the GPU, we are consistently able to
+obtain better performance than the CUSPARSE solver.
+Our GPU solver also outperforms the MKL solver
+running on up to 16 CPU cores for large 3-D problems.
+We have also investigated the strong and weak scaling for a
 multi-GPU compact finite difference application.
@@
 in a domain distributed among multiple GPUs in a cluster.
-Derivatives in the non-contiguous directions are solved
-using a \emph{local} data rearrangement approach.
+We are able to show both strong and weak scaling for up to 64 GPUs
+(the largest problem investigated).
 
-\section{Conclusions}
-
-We have present an algorithm
-for efficiently solving a common class of
-tridiagonal linear systems
-based on the cyclic reduction algorithm for GPUs,
-Our solver gives better performance than
-the current multicore Intel MKL and CUSPARSE solvers
-Additionally, we describe an implementation of our solver
-for GPU clusters,
-and provide performance results for a
-compact finite difference application,
-run on up to 64 GPUs.lusters.
+\bibliographystyle{unsrt}
+\bibliography{bibliography}
 

### 2016-03-16 paper-draft/paper.tex
changes due to Dr. Livescu

@@
 \usepackage{multirow}
+\usepackage{color,soul}
 \usepackage{rotating}
@@
 
-Thus, in the best case ($n$ parallel processors),
-cyclic reduction requires 
-$2log_2(n) - 1$ steps.
-For even moderately large $n$,
-this is significantly smaller than
-the $2n$ steps required by the classical Thomas algorithm.
-This makes cyclic reduction a good fit
-for massively parallel architectures like GPUs.
+\hl{
+Thus, cyclic reduction requires \emph{significantly} fewer steps
+($2log_2(n) - 1$) compared to Thomas algorithm ($2n$).
+Each step in cyclic reduction consists of operations on several indices
+that can each be done in parallel,
+making it a good fit for massively parallel architectures such as GPUs.}
 
@@
 r_m^p
-\end{bmatrix} & \label{eqn:primary-system-expanded} \\
+\end{bmatrix} & \label{eqn:primary-system-expanded} 
 \end{align}
@@
 the tridiagonal coefficients of the global tridiagonal system.
-These are the same for all local grid lines,
-so they are solved for a single local grid line in each subdomain.
+\hl{These are the same for all local grid lines,
+so they are \emph{precomputed} -
+solved \emph{once} for a single local grid line in each subdomain.
+This is done on the CPU.}
 

### 2016-03-17 paper-draft/paper.tex
Address Mattor et al. argument about CR/Thomas algorithm

@@
 Each step in cyclic reduction consists of operations on several indices
-that can each be done in parallel,
-making it a good fit for massively parallel architectures such as GPUs.}
+that can each be done independently,
+suggesting a parallel (multi-threaded) approach.
+When solving linear systems with the same matrix but several right hand sides,
+a thread-parallel Thomas algorithm in which
+each thread solves an independent system may be considered.
+Mattor et al. suggest that this may be a more efficient use
+of parallelization compared to cyclic reduction,
+due to the smaller amount of work required per thread.
+However, using cyclic reduction for solving the independent systems
+exposes a larger amount of parallelism -
+which is important for good performance on GPUs.
+When the number of systems to be solved is small
+(as is the case in 2-D problems and small 3-D problems),
+the thread-parallel Thomas algorithm under-utilizes the GPU.
+For larger problems,
+successive threads access far-apart memory locations,
+leading to poor GPU performance.
+Thus, we argue that cyclic reduction is a better fit for
+massively parallel architectures such as GPUs.
+}
 

### 2016-03-24 paper-draft/paper.tex
Changes to section justifying the use of cyclic reduction

--- 2016-03-16 paper-draft/paper.tex
@@
 \usepackage{rotating}
+\soulregister\cite{7}
 
@@
 \hl{
-Thus, cyclic reduction requires \emph{significantly} fewer steps
-($2log_2(n) - 1$) compared to Thomas algorithm ($2n$).
-Each step in cyclic reduction consists of operations on several indices
-that can each be done in parallel,
-making it a good fit for massively parallel architectures such as GPUs.}
+Although the total number of operations involved in cyclic reduction ($17n$)
+is larger compared to the Thomas algorithm ($8n$),
+cyclic reduction requires \emph{significantly} fewer steps
+($2log_2(n) - 1$) compared to Thomas algorithm ($2n$) \cite{Zhang2010FTS}.
+At each step, the operations at each index in cyclic reduction
+can be done independently,
+making it amenable to parallel solution.
+Further, when solving multiple tridiagonal systems with different right hand sides,
+cyclic reduction exposes a larger degree of parallelism
+than solving the systems using a thread-parallel Thomas algorithm.
+This fact, along with a more favourable memory access pattern,
+has made cyclic reduction---and the
+closely related \emph{parallel cyclic reduction} algorithm---the
+basis for most approaches to solving tridiagonal systems on GPUs \cite{chang2014guide}.
+}
 
@@
 so they are \emph{precomputed} -
-solved \emph{once} for a single local grid line in each subdomain.
-This is done on the CPU.}
+solved \emph{once} for a single local grid line in each subdomain.}
 

### 2016-03-24 paper-draft/paper.tex
add some more notes to the conclusions

@@
 solvers to solve different problem sizes.
+\hl{
+We note that NEATO performs
+at least as well as the Intel MKL on 8-16 cores
+for most cases,
+and outperforms CUSPARSE in almost all cases.
+This is primarily because NEATO is able to take advantage
+of the known matrix coefficients,
+while the CUSPARSE implementation is for the more
+general tridigaonal case.
+}
 
@@
 \hl{These are the same for all local grid lines,
-so they are \emph{precomputed} -
-solved \emph{once} for a single local grid line in each subdomain.}
+so they are solved \emph{once} for a single local grid line in each subdomain.
+The results are stored on the GPU and reused for other grid lines.}
 
@@
 in computational fluid dynamics.
-Our solver gives better performance than
-the current multicore Intel MKL and CUSPARSE solvers
+\hl{Our solver is able to take advantage of the
+special structure of these systems
+to reduce the number of computations performed and
+improve memory access patterns.
+Thus, we are able to show better performance
+compared to the Intel MKL and NVIDIA CUSPARSE solvers
+on multi-core CPUs and GPUs respectively.}
 Additionally, we describe an implementation of our solver
-for GPU clusters,
+for multi-GPU clusters,
 and provide performance results for a
